### 2020-07-22 MSc_Dissertation_B156476.tex
First commit, builds but very incomplete!

% CEG - Computational Evolutionary Biology
% Asgard Report

\documentclass[tikz, 12pt,a4paper,oneside,fleqn]{article}
\usepackage[margin=2cm]{geometry}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{fancyvrb}
\usepackage{setspace}
\usepackage{enumitem}
% \setlist[description]{leftmargin=\parindent,labelindent=\parindent}
% \usepackage{marginnote}

\usepackage{caption}
\captionsetup[figure]{labelfont=bf,format=hang,font={stretch=1.2, small}}
\captionsetup[table]{labelfont=bf,format=hang,font={stretch=1.2, small}}

\usepackage[linewidth=1pt]{mdframed}

\usepackage{graphicx}

\graphicspath{{/home/ipoole/Dropbox/MSc_SSB/Disertation_Ovarian_Cancer/images/}}

\newcommand{\plotspath}[1]{{/home/ipoole/Documents/gitrepos/HgsocTromics/Plots/#1}}

\usepackage{hyperref}
\usepackage{subfig}
\usepackage{booktabs}  % Needed for pandas to_latex() output
\usetikzlibrary{positioning}

\newcommand{\etal}{{\em et al\/}}
\newcommand{\comment}[1]{\marginpar{{\tiny\singlespacing #1 \par}}}

\linespread{1.5}
% This sets Arial font (bit ugly!)
\usepackage{helvet}
\renewcommand{\familydefault}{\sfdefault}

%\usepackage{titlesec}
%\titleformat{\subsubsection}
%   {\itshape\normalsize}{\thesubsubsection}{1em}{}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% SUBMISSION PAGE %%%%%%%%%%%%%%%%%%%%%%%%%%%%

\thispagestyle{empty}

\title{Phylogenetic study of Asgard Archaea}
\author{}

\begin{document}

%%%%%% Assesment Title Page
\begin{center}
\includegraphics[scale=0.3]{images/UoE_SBO_logo.png}
\end{center}

\vspace{0.3in}

\begin{mdframed}
\begin{center}
\huge
\vspace{0.3in}
\bf
Disentangling patterns of gene expression in high grade serous ovarian cancer
\vspace{0.2in}
\end{center}
\vspace{0.2in}
\end{mdframed}

\vspace{0.3in}

\begin{mdframed}
\begin{center}
\large
\vspace{0.2in}
Student Exam Number: \bf{B156476}
\vspace{0.2in}
\end{center}
\end{mdframed}

\vspace{0.3in}

\begin{mdframed}
\begin{center}
\large
\vspace{0.2in}
In partial fulfilment of the requirement for the Degree of
Master of Science in Systems and Synthetic Biology at the
University of Edinburgh,
2019 / 2020
\vspace{0.2in}
\end{center}
\end{mdframed}

\vspace{0.3in}

\begin{mdframed}
\begin{center}
\large
\vspace{0.2in}

Disssertation Supervisor:  Dr. Ailith Ewing
\vspace{0.2in}
\end{center}
\end{mdframed}

\newpage
\setcounter{page}{1}
%\pagestyle{headings}

%%%%%%%%%%%%%%%%%%%%%%%%%%% Start of report for real %%%%%%%%%%%%%%%%%%%

\section{Introduction}

Research questions:  
\begin{enumerate}
\item What are the known patterns of gene expression in high-grade serous ovarian cancer?
\item How can blind-source separation techniques such as independent component analysis and non-negative matrix factorization be applied to transcrptomics studies?
\item How can biological databases be used to link genes into oncogenesis pathways?
\end{enumerate}

\subsection{Matrix factorization methods}
\label{sec-matrix-factorization-intro}

A good introduction is found in ref \cite{Stein-OBrien2018} matrix factorization (MF) methods are discussed as a means of extraction the low dimensional structure from a high dimensional ``-omics" datasets, with gene expression analysis being the prime example.  
Conventionally, transcriptomics expression arrays are oriented with genes (or transcripts) as rows, and samples (e.g. patients) in columns.   
A typical expression array might have in the order of tens of thousands of rows and hundreds of columns.  
MF methods reduces this large $M \times N$ (genes $\times$ samples) matrix into two two smaller matrices.
In the terminology of Stein-O'Brien \etal\ these are the $M \times K$ \emph{pattern matrix} and the $K \times N$ \emph{amplitude matrix}.  $K$ is typically small of the order 10, and refers to the number of extracted \emph{factors} (or components).

\begin{figure}[ht]
\def\M{2.2cm}
\def\N{1.3cm}
\def\K{0.3cm}
\def\fudge{-0.8mm}
\centering

\subfloat[Optimising independence of Meta\underline{genes}]{
\begin{tikzpicture}[node distance = 0.25cm]
	\node (X) [draw, fill=green, minimum height=\M, minimum width=\N] {\large X};
	\node [right=of X] (eq) {\large $\approx$};
	\node [right=of eq] (S) [draw, fill=yellow, minimum width=\K, minimum height=\M] {\large S};
	\node [right=of S] (times) {\LARGE$\times$};
	\node [right=of times] (A) [draw, fill=cyan, minimum width=\N, minimum height=\K] {\large A};

	\node [below=\fudge of X] {\tiny{$N$ Samples}};
	\node [left=\fudge of X] {\rotatebox{90}{\tiny{$M$ Genes}}};
	\node [below=of X, text width=2cm, align=center] {\small Expression matrix};

	\node [below=\fudge of S] {\tiny{$K$ Metagenes}};
	\node [left=\fudge of S] {\rotatebox{90}{\tiny{$M$ Genes}}};
	\node [below=of S, text width=2cm, align=center] {\small Metagenes};

	\node [below=\fudge of A] {\tiny{$N$ Samples}};
	\node [left=\fudge of A] {\tiny{$K$}};
	\node [below=of A, text width=2cm, align=center] {\small Metasamples};
\end{tikzpicture}}

\subfloat[Optimising independence of Meta\underline{samples}]{
\begin{tikzpicture}[node distance = 0.25cm]
	\node (X) [draw, fill=green, minimum height=\N, minimum width=\M] {\large X};
	\node [right=of X] (eq) {\large $\approx$};
	\node [right=of eq] (S) [draw, fill=cyan, minimum width=\K, minimum height=\N]{\large S};
	\node [right=of S] (times) {\LARGE$\times$};
	\node [right=of times] (A) [draw, fill=yellow, minimum width=\M, minimum height=\K]{\large A};

	\node [below=\fudge of X] {\tiny{$M$ Genes}};
	\node [left=\fudge of X] {\rotatebox{90}{\tiny{$N$ Samples}}};
	\node [below=of X, text width=2cm, align=center] {\small Expression matrix};

	\node [below=\fudge of S] {\tiny{$K$ Metasamples}};
	\node [left=\fudge of S] {\rotatebox{90}{\tiny{$N$ Samples}}};
	\node [below=of S, text width=2cm, align=center] {\small Metasamples};

	\node [below=\fudge of A] {\tiny{$M$ Genes}};
	\node [left=\fudge of A] {\tiny{$K$}};
	\node [below=of A, text width=2cm, align=center] {\small Metagenes};
\end{tikzpicture}}
\caption{Two ways of configuring ICA.  The $S$ matrix is optimised for independence of components, thus configuration (a) optimises for metagenes, while (b) optimises for metasamples. The $X \approx S \times A$ notation is conventional for ICA.}
\label{fig:matrix-factorization}
\end{figure}

There are three matrix factorization methods of potential interest to this project:
\begin{enumerate}
\item Independent component analysis (ICA)
\item Non-negative matrix factorization (NMF)
\item Principal component analysis (PCA)
\end{enumerate}

\subsubsection{Independent Component Analysis (ICA)}

The most commonly used matrix notation for ICA, (independent of any -omics applications) is:

\begin{equation}
X \approx S A
\end{equation}
Adding subscripts to indicate the row,column orientation of the matrices:
\begin{equation}
X_{M,N} \approx S_{M,K} A_{K,N}
\end{equation}

$X$ is assumed to have been pre-whitened.  $S$ is referred to variously as the pattern matrix, latent variable or independent component matrix.  $A$ is referred to as the mixing or amplitude matrix.   If $X$ is sized ($M$, $N$) then $S$ will be ($M$, $K$) and $A$ ($K$, $N$) by the rules of matrices.  Crucially, it is the $S$ matrix which captures the independent components.

Authors use different conventions in the nomencalture and orientation of the expression array.  This is partly a matter of convention but there are ramifications of substance.   The terms ``metagenes'' is used commonly and consistently to refer to the $K$ components of $M$ long gene expression profiles.  Similarly, ``metasamples'' refers to the $K$ components of $N$ long sample profiles.   However, the packing of these entities into matrices varies in orientation, so that metagenes are sometimes $M \times K$, sometimes $K \times M$, and metasamples are sometimes $N \times K$, sometimes $K \times N$, depending on the orientation of the input expression array.

The point of substance (as opposed to notational convention) is whether ICA is used to maximise independence of metagenes or of metasamples.  This relates to the choice of orientation of the input expression matrix.  Given the standard definition of ICA above \cite{ICA_definition}, the following rule applies to the (row, column) orientation of the expression matrix, as passed to ICA:
\begin{description}
\item[to optimise metagenes] the expression matrix is oriented as ($M$ genes, $N$ samples) resulting in a ($M$, $K$) metagenes matrix. 
\item[to optimise metasamples] the expression matrix is oriented as ($N$ samples, $M$ genes), resulting in a ($N$, $K$) metasamples matrix.
\end{description}

See figure \ref{fig:matrix-factorization}.

There is substantial confusion an lack of clarity in the way that ICA is applied in transcriptomics research.  ``Surprisingly, both ways of applying ICA to omics data are wide-spread, and sometimes it makes an effort to figure out in which way ICA was applied." \cite{Sompairac2019a}, and ``Different protocols to apply ICA to transcriptomic data exist and currently no single standard approach has been defined. The main difference in the existing approaches consists in what is considered as source signal matrix in the decomposition" \cite{Cantini}.   According to Cantini \etal\, references \cite{Au-Yeung2014}, \cite{Kairov2017}, \cite{Kong2008} and \cite{Lee2003} optimise metagenes, while references \cite{Meng2016}

The most commonly used matrix notation for ICA, (independent of any -omics applications) is:
\begin{equation}
X \approx S A
\end{equation}

$X$ is assumed to have been pre-whitened.  $S$ is referred to variously as the pattern matrix, latent variable or independent component matrix.  $A$ is referred to as the mixing or amplitude matrix.   If $X$ is sized ($M$, $N$) then $S$ will be ($M$, $K$) and $A$ ($K$, $N$) by the rules of matrices.  Crucially, it is the $S$ matrix which captures the independent components.

\subsubsection{Determining the optimum number of factors}

The number of factors, $K$, to extract is a key decision in any matrix factorization approach.  In this regard, PCA differs from ICA and NMF.  In PCA it is reasonable to extract all factors, setting $K=N$, the factors being ranked by the associated eigenvalue which also ranks the proportion of variance explained.   That approach is not reasonable for ICA and NMF, since differing sets of factors will be obtained for different $K$, and for a given choice of $K$ all factors are equally important -- they do not rank \cite{Stein-OBrien2018}.

The problem of determining the optimal $K$ for a given expression matrix is addressed  by Kairov \etal\ \cite{Kairov2017}, based on optimizing the \emph{stability} of the components over multiple runs.

\subsubsection{Non-negative Matrix Factorization}

When the NMF is used, the pattern and amplitude matrices are typically notated as $W$ and $H$ respectively, while the original expression matrix is notated as $V$, or $A$ and oriented as ($M$ genes, $N$ samples).   The facorization is thus written $V \approx W H$, Thus $W$ represents the ($M$, $K$) metagenes matrix and $H$ represents the ($K$, $N$) metasamples.

\subsubsection{Principal component analysis}

\subsection{Ovarian cancer sub-type prediction}

A recent pre-publication develops a pipeline for sub-type prediction which the potential to be used clinically \cite{Talhouk2020}.
The NanoString gene expression platform is used. Approach is to focus on small set of 513 genes (vs 20,000 protein coding genes) known have relevance to subtyping, as identified from literature.

Two approaches were followed by different teams.  ``All Array": used expression array data from 1650 patients across 14 studies evaluating 9 supervised learning algorithms by bootstrap, selecting an AdaBoost -like method. ``TCGA" used 434 patients from TCGA evaluating 5 algorithms by cross-validation selecting a random forest.

"All Array" had the advantage of more data but needed attention to batch effects. A final classifier took a consensus of the two methods based on a minimal gene set (order 40 genes), validated by a leave-one-out (patient level) approach.

Survival analysis was caried out stratified by prediced subtype.

\begin{figure}
\begin{center}
\includegraphics[scale=0.6]{talhouk_figure_1_part.png}
\end{center}
\caption{HGSOC subtype prediction by expression analysis of a focussed gene set and concensus across two independent pipelines.  From \cite{Talhouk2020}}
\label{fig-talhouk_figure_1_part.png}
\end{figure}

\subsection{Gene expression in HGSOC}

This section focusses on biological and clinical conclusions relating to gene expression in HGSOC and mentions analysis techniques only in passing.

Obesity is known to negatively impact prognosis in ovarian cancer \cite{Cuello2018}, \cite{Au-Yeung2014}.  Cuello \etal\ \cite{Cuello2018} used NMF based clustering on mRNA microarrays and reverse-phase protein arrays to demonstrate an association between cancer driver genes and obesity related genes.

Wang \etal\ \cite{Wang2017c} also use NMF clustering to identify five subtypes of HGSOC informative of outcomes: 1:mesenchymal, 2:immunoreactive, 3:proliferative, 4:differentiated and 5:anti-mesenchymal.  Subtypes 2, 5 are found to be associated with longer survival.

Mairinger \etal\ screened 770 immune related genes to identify 11 differentially expressed genes associated with response to platinum treatment.  They find that expression of HS11B1, DNBT1, CKLF, NUP107. CCL18, LY96, ATG7, SLAMF7, CXCL9 is associated with long survival, while IKBKG and SDHA with short survival.

BRCA1 inactivation is known to cause chromosomal instability in many cancers.  Pradhan \etal\ \cite{Pradhan2010} investigated the role of BRCA1 in HGSOC by copy number and expression analysis, finding surprisingly that inactivation has no relationship with gross genomic alteration.  They leave open the question whether DNA repair by PARP plays a role.   The relationship between BRCA1/2 and DNA repair -- specifically homologous repair deficiency (HRD) -- is picked up by Ewing \etal \cite{Ewing2020}.  They study the complex interplay of single nucleotide variants and large structural variants as they impact BRCA1/2 disruption and thence HRD.  This work suggests that when BRCA1/2 loss is detected in HGSOC patients, there may be a clinical role for PARP inhibitors, since this would prevent error-prone non-homologous repair by PARP, and so selectively kill BRCA1/2 disrupted cells.

Note that several of the studies reviewed here use data from the Australian Ovarian Cancer Study (AOCS): \cite{Winterhoff2016,Patch2015,Ewing2020,Zhang2018,Cuello2018,Au-Yeung2014}.

Stratification of HGSOC by immunohistochemistry is addresses Kamble \etal. \cite{Kamble2019}.  They identify a panel of six biomarkers -- TCF21, E-cadherin, PARP1, Slug and AnnexinA2.   These markers allow stratification into categories including Mesenchymal-to-epithelial transition, Homologous recombination repair and Epithelial-to-mesenchymal transition.  This, they argue, will allow selection of class-specific inhibitor drugs such as Olaparib and Rucaparib.

The platinum based drug Cisplatin is a key treatment in ovarian cancer, yet tumours often develop resistance, possibly related to the host's immune response.  Understanding and predicting such resistance is therefore of clinical importance and is addressed by Maringer \etal\ \cite{Mairinger2019} through expression analysis using the NanoString platform with a panel of 770 immune related genes.   They find the following to be significantly related to platinum resistance: KLRC1, TCF7, CD274, HSD11B1, COLEC12, PDGFC, FCF1, BMI1, TNFRSF9, ATG10, EWSR1.

\subsection{Research Questions}
\begin{enumerate}
\item
Which are the influential genes in HGSOC according
to expression data? Do these confirm published
results?
\item
Which of PCA, ICA and NMF is best suited to this
analysis?
\item
For identified genes, what is the underlying biology of
their action?
\item
Can we discriminate subtypes of HGSOC from
expression data?
\item
Is there value in incorporating somatic and germline
genomic data?
\end{enumerate}

%%%%%%%%%%%%%%% METHODOLOGY %%%%%%%%%%%%%%%%%

\section{Methodology}

\subsection{Outline}

\subsubsection{Datasets}
\begin{enumerate}
\item TCGA, n=274, g=19,601
\item AOCS, n=80, g=19,730
\end{enumerate}

\subsubsection{Method outline}
In outline our method is follows:
\begin{enumerate}
\item
Unsupervised factor extraction by Matrix
factorization methods : NMF, ICA and PCA.
\item
Identification of a consistent gene list between the TCGA and AOCS datasets.
\item
Factor selection for robustness by
k-means clustering and silhouette score
\item
Investigation of biological significance of determined factors by gene enrichment analysis against the Gene Ontology (GO) and Kyoto Encyclopedia of Genes and Genomes (KEGG).
\item
Survival analysis on TCGA to identify key predictive factors
\item
Transfer of identified factors to the AOCS dataset by matrix psuedo-inverse to validate efficacy of identified features.
\item
Revisit enrichment analysis for most significant factors to study biological significance.
\end{enumerate}

\subsubsection{Tools}
\begin{itemize}
\item Python 3.6
\item Numpy for high performance matrix manipulation
\item Pandas for data frame handing
\item Matplotlib for plotting
\item Scikit-learn for matrix factorization and k-means clustering
\item GOATOOLS package for gene enrichment analysis against GO
\item Lifelines package for survival analysis.
\end{itemize}

A full list of software and libraries used, with versions, is given in appendix \ref{sec-software-versions}

\subsection{Data normalization and quality control}

Expression data for the AOCS and TCGA datasets was received for this project in a normalised matrix format.  The full pre-processing pipeline is described in \ref{Ewing2020}, in brief using the DESeq2 package with normalization by variance stabilizing transformation.

\subsection{Gene set intersection}

In order to allow factorizations found in the TCGA dataset dataset to be applied to the AOCS dataset it is necessary (or at least convenient) to synchronize the set of genes over which the expression matrices are defined.   As provided to this project, the TCGA and AOCS datasets cover 19,610 and 19,730 genes respectively, 19,566 or which common to both (according to ENSG encodings).  Thus, both datasets were pruned to the 19,566 intersection set and ordered consistently.

\subsection{Matrix factorization computation}
One of the aims of this work is to compare the efficacy of three dimensionality reduction methods -- NMF, ICA and PCA -- as explained in the introduction.   The three methods have different properties and notational conventions, but to simplify the discussion here we adopt $X \approx W H$ for all three methods, where $X$ is the (genes, patients) expression matrix  $W$ is the (genes, factors) metagenes matrix and $H$ is the (factors, patients) metasamples matrix.   Please refer back to section \ref{sec-matrix-factorization-intro} for further explanation.

Some of the key algorithm hyper parameters of each method were investigated and tuned with respect to reconstruction accuracy, specifically the root-mean square (RMS) difference between $X$ and $WH$. The following parameters were explored in each case:
\begin{description}
\item[NMF:] Parameters {\tt max\_iter} (algorithm iterations) and {\tt tol} (tolerance of convergence) were optimised for good accuracy and acceptable execution time.  Other parameters of interest are {\tt alpha} (multipluer for regulation term) and {\tt l1\_ratio} (multiplier for L1 regularization, when {\tt alpha} $> 0$).  L1 regularization  favours elements being precisely zero, whereas L2 regularization will encourage them to be small.  These parameters have not been explored in this project, and the default {\tt alpha = 0} (no regularization) was used.
\item[ICA:]  Parameters {\tt max\_iter} and {\tt tol} we investigated as for NMF above.  Additionally, options for the entropy functional which forms the basis of the optimiziation were investigated.
\item[PCA:] This is in principle a deterministic algorithm based on eigenvector decomposition.  However, {\tt sklearn.decomposition.PCA} uses a more efficient 'randomized' algorithm when the given matrix is larger than 500 in both dimensions.  Thus in our use case PCA is seen to have stochastic behaviour.

\end{description}

\subsection{Factor selection by cluster coherence}

As outlined in the discussion, deciding the number of components (metagenes) to extract is key.  Taking more components results in more accurate representation of the observed expression matrix and provides more avenues to explore the underlying biology.  However, it is important that the metagenes are \emph{stable}, that is likely to have reliable meaning when transfered to other datasets.   There are two sources of variation or instability to consider.

NMF and ICA are inherently stochastic algorithms, sensitive to their starting state, so repeated runs give different results.

A second and more fundamental source of variation of relevance to all three factorization methods is \emph{sampling error}.   Our factorizations are based on a small (N=80 or N=374) sample of patients drawn from the population of HGSOC patients; our particular datasets are just two examples of many different `draws' which could have been made.

\emph{Bootstrap sampling} (also know as \emph{Monte Carlo} simulation) is a common method of empirically propagating the consequence of sampling error when the distribution or processing operations are difficult to model mathematically.   This is implemented by performing factorizations multiple times, at each iterations choosing N samples from the N available \emph{with replacement}.   For this work 50 repeats were performed, being a compromise between achieving an adequate simulation without overly burdensome computation time.  The overall process is summarised as pseudocode in figure \ref{fig-clustering-psuedocode}.

Each of the three factorizer methods is evaluated for rank $k$ between 2 and 10.  In each case, 50 iterations of factorization are performed on bootstrap samples, generating $50 k$ 19,566 dimensioned metagenes.   t-SNE plots \ref{tsne-reference} visualise these high-dimensioned metagenes in a 2-D view.   If sampling and algorithm initialisation error are modest then we expect to see $k$ clusters of points.   To avoid lengthy computation in the t-SNE clustering, the metagenes are first reduced to $r=20$ dimensions by PCA; brief experiments showed that this reduction had negligible effect on the t-SNE visualisation for $r \geq 10$.

In order to obtain median estimates of the $k$ metagenes from the $50 k$ which were generated, k-means clustering was performed in the PCA reduced (20 dimension) space, delivering $k$ sets of points.  These were referenced back to the original 19,566 dimensioned metagenes and the per-dimension median calculated.  These $k$ median metagenes were saved a file named for the specific factorizer and rank $k$.

The k-means clustering further allowed for a quantitative assessment of cluster coherence for the particular choice of factorizer and rank, via the \emph{silhouette score}.  In brief, this is a measure of how close (by Euclidean distance) each point in a cluster is to other points in the cluster versus points \emph{not} in the same cluster.  See 
% \href{https://en.wikipedia.org/wiki/Silhouette_(clustering)}{Wikipedia: Silouette (clustering)}.  
The score is in the range -1 to +1, with 0 implying random scatter and 1 implying all points of a cluster overlay.

The t-SNE plots and silhouette scores were assessed visually (see figures \ref{fig-first-cluster-result} to \ref{fig-last-custer-result} to decide, for each of the three factorizers, the highest rank with good cluster coherence, those median metagenes to take forward to gene enrichment and survival analysis.

Initial investigation on the N=80 AOCS dataset showed very poor coherence, even for k=2.   Thus it was decided compute and select metagenes only from the N=374 TCGA dataset.   We refer to those selected ranks as $K_{\mbox{\tiny NMF}}$ etc., thus yielding $K_{\mbox{\tiny NMF}} + K_{\mbox{\tiny ICA}} + K_{\mbox{\tiny PCA}}$ metagenes in total.

\begin{figure}
\begin{center}
\begin{Verbatim}[baselinestretch=1, frame=single, rulecolor=\color{blue}, label=Metagene Stability Assessment, fontfamily=courier, fontsize=\small]

 for factorizer in NMF, ICA, PCA:
    for k in 2..11:
       Ws = []
       for j in 1..50:
          X = bootstrap sample N from N
          W, H = factorizer(X, rank=k, seed=j)
          append W to Ws

          # Ws is a list of 50*k metagenes, each of length 19,566
          # Pragmatically reduce metagenes to 20 dimensions by PCA

       Ws_reduced = PCA(Ws, rank=20)
       plot t-SNE(Ws_reduced)
       clustering = k-means clustering(Ws_reduced, n_clusters=k)
       score[factorizer, k] = silhouette score(clustering)	  
       median_metagenes = calculate medians for k clusters of metagenes
       save median_metagenes to file by factorizer and k
       for each factorizer 
          plot score vs k

\end{Verbatim}
\end{center}
\caption{Pseudocode for generating median metagenes and assessing there stability to random algorithm initialization and sampling error.}
\label{fig-clustering-psuedocode}
\end{figure}

\subsection{Gene enrichment analysis}

\subsection{Survival analysis}

Survival analysis was performed to investigate whether the selected metagene factors correlated with patient survival.  Overall  survival (OS) data is available for both the TCGA and AOCS datsets.  Additionally, progression free survival (PFS) data is available for AOCS.   As discussed in section \ref{sec-clustering-method}, metagenes were obtained by factorization on the TCGA dataset only.  Application to TCGA (for OS) thus represents an in-sample test, while application to AOCS (for OS and PFS) is a more exacting out-of-sample test.

Analysis was performed using the Python {\tt Lifelines} package \cite{Davidson-Pilon2020}.  Metasamples ($H$ matrices) were calculated from the metagenes and expression matrix as explained in the following sub-section.  Metasample values were binariased to 0, 1 by thresholding at the median value.   Kaplan-Meier plots for made in respect of derived metasample for each application -- making $11 \times 3$ plots each with two survival curves (with 95\% confidence intervals).  A hazard ratio (HR)
\comment{Do I need to explain the definition and meaning of HR?} 
was calculated for each plot by fitting Cox's proportional hazards model with p-value relating to the hypothesis that HR is significantly different to 1.0.

\subsection{Transfer of learned factors to novel a dataset}

\newcommand{\trainset}{\mathcal{T}}
\newcommand{\validset}{\mathcal{V}}

Rank $k$ matrix factorization on the training dataset $\trainset$ (specifically the TCGA dataset) results in:
\begin{equation}
 X_\trainset \approx W_\trainset H_\trainset
\end{equation}
where $X_\trainset$ is the expression matrix of shape $(g_\trainset, n_\trainset)$, with $g_\trainset$ the number genes and $n_\trainset$ the number of patients. $W_\trainset$ is the metagene matrix of shape $(g_\trainset, k)$ and $H_t$ is the metasamples matrix of shape $(k, n_\trainset)$.

We wish to apply the factorization learned on $\trainset$ to a novel dataset $\validset$ (specifically the AOCS dataset) of shape $(g_\validset, n_\validset)$. Importantly, $n_\validset = 1$ reflects the application of these methods to a single patient in a clinical setting.  To apply the learned factorization we need to find $H_\validset$ as in the factorization
\begin{equation}
	X_\validset \approx W_\validset H_\validset
\end{equation}
In both the experimental and clinical situation we are given $X_\validset$ but not $W_\validset$ and require to find $H_\validset$.  We \emph{cannot} simply perform the matrix factorization on dataset $\validset$ since $n_\validset$ may be small, or even a single patient.   However, if patients in datasets $\trainset$ and $\validset$ are drawn from the same population (ovarian cancer patients), then $X_\trainset$ and $X_\validset$ can be expected to have similar distributions w.r.t to their columns, and thus $W_\trainset$ and $W_\validset$ can be expected to be equivalent within sampling error.  This only makes sense if the two expression matrices $X_\trainset$ and $X_\validset$ are defined over the \emph{same list of genes}, so that $g_\trainset = g_\validset = g$, in which case $W_\trainset$ and $W_\validset$ have the same shape of $(g, k)$.

We thus need to solve for $H_\validset$ in
\begin{equation}
	X_\validset  \approx  W_\trainset H_\validset \label{eq_Xv_WtHv}
\end{equation}

This can be solved by the method of least squares.  In the case that the original factorization was by NMF, we use non-negative least square regression (NNLS):
\begin{equation}
	H_\validset = \mbox{argmin}_{H\geq 0}{\lvert W_\trainset H - X_\validset \rvert}_2
	\label{eq_Hv_WdtXv}
\end{equation}
where ${\lvert \mathbf{\cdot} \rvert}_2$ indicates Euclidean distance or L2-norm.
The Python library function {\tt scipy.optimize.nnls} is used.

For ICA and PCA we use ordinary least square regression, formulated above by dropping the $H\geq 0$ constraint, using {\tt sklearn.linear\_model.LinearRegression}.

An internal check on the validity of the statistical assumptions can be made my substituting  $H_\validset$ from (\ref{eq_Hv_WdtXv}) into (\ref{eq_Xv_WtHv}) and assessing the accuracy of the approximation by RMS difference.

The end result of the analysis the $H_\validset$ matrix of shape $(k, n_\validset)$, thus delivering for dataset in our $\validset$ a feature vector of length $k$ -- a small number typically less than a dozen.

Since in this work we wish to evaluate the efficacy of the three factorization methods -- NMF, ICA and PCA -- we in fact take forward three different $H$ matrices, having not necessarily the same rank $k$.

\subsection{Survival analysis on novel dataset}

\subsection{Study of biological significance of predictive factors}

\subsection{Codebase}

The described method was implemented in Python, except for the KEGG gene enrichment analysis which was implemented in R.  All code is availailable a github repository: \url{https://github.com/ipoole/HgsocTromics}.  Good software engineering practice has been followed, with an object-oriented design, frequent commits and unit testing.  Unit tests are based on

%%%%%%%%%%% RESULTS %%%%%%%%%%%%%

\section{Results}

\subsection{Investigation of optimal parameters}

\subsection{Cluster coherence analysis results}

Figure \ref{fig-AOCS-ica-fixed-vs-bootstrap} demonstrates that resampling error on a small N=80 dataset makes metagene extraction highly unstable, effectively un-usable.  These figures show results for ICA only in the interest of space, but a very similar effect is observable for PCA and NMF.   For this reason it was decided to perform all metagene extraction on the larger N=374 TCGA dataset, in which the impact of sampling error is not so severe -- see figure \ref{fig-TCGA-ica-fixed-vs-bootstrap}.

The critical decision of the choosing the facorization rank for each method -- $K_{\mbox{\tiny NMF}}$, $K_{\mbox{\tiny ICA}}$ and $K_{\mbox{\tiny PCA}}$ -- is based on figure \ref{fig-TCGA-all-bootstrap}.   On this bases, considering both the t-SNE plots and silhouette scores, for following choices were made:
$K_{\mbox{\tiny NMF}}  =  3,
K_{\mbox{\tiny ICA}}  =  5,
K_{\mbox{\tiny PCA}}  =  3$.

The rationalle for these choices is that, firstly $k>2$ is desirable to have sufficient information to work with.  NMF cluster coherence seems to deteriorate after $k=3$.   Looking at the graph of silhouette scores, ICA appears to have a sweet spot at $k=5$.  For PCA, $k = 3$ or $4$ both seem reasonable, $k=3$ is safe.

\def\S{0.4}
\begin{figure}
\begin{center}
\includegraphics[scale=\S]{\plotspath{AOCS_Protein/FactorClustering/multiple_single_factors_scatter_AOCS_ICA_2_3_5_10_fixed.pdf}} \\
\includegraphics[scale=\S]{\plotspath{AOCS_Protein/FactorClustering/multiple_single_factors_scatter_AOCS_ICA_2_3_5_10_bootstrap.pdf}}\\
\caption{Clustering of metagenes from ICA facorizations on the N=80 AOCS dataset, comparing \emph{fixed} sampling (top row) with \emph{bootstrap} sampling (bottom row) for a selection of factorization ranks.  With sampling error excluded, clusters appear reasonably coherent, but when sampling error is modelled by bootstrap sampling then we see the factorizations become much less stable.}
\label{fig-AOCS-ica-fixed-vs-bootstrap}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[scale=\S]{\plotspath{TCGA_OV_VST/FactorClustering/multiple_single_factors_scatter_TCGA_ICA_2_3_5_10_fixed.pdf}} 
\includegraphics[scale=\S]{\plotspath{TCGA_OV_VST/FactorClustering/multiple_single_factors_scatter_TCGA_ICA_2_3_5_10_bootstrap.pdf}}
\caption{Clustering of metagenes from ICA facorizations on the N=374 TCGA dataset, again comparing fixed and bootstrap sampling.  In this larger dataset, sampling error (bottom row) is not so severe; compare with figure \ref{fig-AOCS-nmf-fixed-vs-bootstrap}}
\label{fig-TCGA-ica-fixed-vs-bootstrap}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[scale=\S]{\plotspath{TCGA_OV_VST/FactorClustering/multiple_single_factors_scatter_TCGA_NMF_2_3_5_10_bootstrap.pdf}} \\
\includegraphics[scale=\S]{\plotspath{TCGA_OV_VST/FactorClustering/multiple_single_factors_scatter_TCGA_ICA_2_3_5_10_bootstrap.pdf}} \\
\includegraphics[scale=\S]{\plotspath{TCGA_OV_VST/FactorClustering/multiple_single_factors_scatter_TCGA_PCA_2_3_5_10_bootstrap.pdf}} \\
\includegraphics[scale=0.45]{\plotspath{/TCGA_OV_VST/FactorClustering/silhouette_plots_TCGA_2_3_4_5_6_7_8_9_10_bootstrap.pdf}} 
\caption{Metagene clustering for all three methods applied to the N=374 TCGA dataset with bootstrap sampling, over a range of factorization ranks.  The silhoutte scores are also plotted (bottom).  It is on the basis of this figure that the factorization ranks for each method were selected.}
\label{fig-TCGA-all-bootstrap}
\end{center}
\end{figure}

\subsection{Survival analysis results}

\begin{table}
\begin{center}
\small
% \renewcommand{\arraystretch}{0.75}
\input{/home/ipoole/Documents/gitrepos/HgsocTromics/Plots/SurvivalAnalysis/survival_analysis_table_os_TCGA_TCGA.tex}
\end{center}
\caption{Survival analysis (overall survival) for TCGA stratified at the median of each of the selected components}
\label{tab-survival-hr}
\end{table}

\begin{table}
\begin{center}
\small
% \renewcommand{\arraystretch}{0.75}
\input{/home/ipoole/Documents/gitrepos/HgsocTromics/Plots/SurvivalAnalysis/survival_analysis_table_os_TCGA_AOCS.tex}
\end{center}
\caption{Survival analysis (overall survival) for OACS stratified at the median of each of the selected components}
\label{tab-survival-hr}
\end{table}

\begin{table}
\begin{center}
\small
% \renewcommand{\arraystretch}{0.75}
\input{/home/ipoole/Documents/gitrepos/HgsocTromics/Plots/SurvivalAnalysis/survival_analysis_table_os_TCGA_AOCS.tex}
\end{center}
\caption{Survival analysis (progression free survival) for OACS stratified at the median of each of the selected components}
\label{tab-survival-hr}
\end{table}

\section{Discussion}
\begin{itemize}
\item Importance of taking account of sampling error.
\item Little correlation with survival time - not surprising.
\end{itemize}

\section{Conclusions}

\bibliography{OvarianCancer}
\bibliographystyle{ieeetr}

\section{Appendices}

\subsection{Software libraries and versions}
\label{sec-software-versions}

\end{document}

### 2020-07-22 MSc_Dissertation_B156476.tex
Survival analysis basic method and Kaplan-meier plots in

@@
-% CEG - Computational Evolutionary Biology
-% Asgard Report
+% MSc Dissertation on pattern expression in Ovarian Cancer
 
@@
 \usepackage{enumitem}
-% \setlist[description]{leftmargin=\parindent,labelindent=\parindent}
-% \usepackage{marginnote}
+\setlist[description]{leftmargin=\parindent,labelindent=\parindent}
 
@@
 \begin{center}
-\includegraphics[scale=0.6]{talhouk_figure_1_part.png}
+\includegraphics[scale=0.35]{talhouk_figure_1_part.png}
 \end{center}
-\caption{HGSOC subtype prediction by expression analysis of a focussed gene set and concensus across two independent pipelines.  From \cite{Talhouk2020}}
+\caption{HGSOC subtype prediction by expression analysis of a focussed gene set and concensus across two independent pipelines.  From \cite{Talhouk2020}.}
 \label{fig-talhouk_figure_1_part.png}
@@
 Application to TCGA (for OS) thus represents an in-sample test, while application to AOCS (for OS and PFS) is a more exacting out-of-sample test.
+There are thus three analyses to consider: 1) TCGA$\rightarrow$TCGA (OS), 2) TCGA$\rightarrow$OACS (OS) and 3) TCGA$\rightarrow$OACS (PFS).
 
 Analysis was performed using the Python {\tt Lifelines} package \cite{Davidson-Pilon2020}.
-Metasamples ($H$ matrices) were calculated from the metagenes and expression matrix as explained in the following sub-section.
+Metasamples ($H$ matrices) were calculated from the metagenes and expression matrix as explained in section \ref{sec-transfer-to-novel} below.
 Metasample values were binariased to 0, 1 by thresholding at the median value.
-Kaplan-Meier plots for made in respect of derived metasample for each application -- making $11 \times 3$ plots each with two survival curves (with 95\% confidence intervals).
+Kaplan-Meier plots were made in respect of derived metasample for each of the three analysis -- making $11 \times 3$ plots each with two survival curves (with 95\% confidence intervals).
 A hazard ratio (HR)
@@
 \subsection{Transfer of learned factors to novel a dataset}
+\label{sec-transfer-to-novel}
 
@@
 
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.5]{\plotspath{SurvivalAnalysis/multiple_kaplan_meier_os_TCGA_TCGA.pdf}} 
+\caption{Kaplan-Meier plots for each metasample component, stratified at the median value, for the TCGA $\rightarrow$ TCGA for overall survival (OS) case.
+Hazard ratio and p-value is shown for each case.
+The final plot (bottom right) is unstratified overall survival}
+\label{fig-kaplan-meier-os-TCGA-TCGA}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.5]{\plotspath{SurvivalAnalysis/multiple_kaplan_meier_os_TCGA_AOCS.pdf}} 
+\caption{Kaplan-Meier for the TCGA $\rightarrow$ TCGA for overall survival (OS) case.}
+\label{fig-kaplan-meier-os-TCGA-OACS}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.5]{\plotspath{SurvivalAnalysis/multiple_kaplan_meier_pfs_TCGA_AOCS.pdf}} 
+\caption{Kaplan-Meier for the TCGA $\rightarrow$ TCGA for progression free survival (PFS) case.  }
+\label{fig-kaplan-meier-pfs-TCGA-OACS}
+\end{center}
+\end{figure}
+
 \begin{table}
@@
 \caption{Survival analysis (overall survival) for TCGA stratified at the median of each of the selected components}
-\label{tab-survival-hr}
+\label{tab-survival-os_TCGA_TCGA}
 \end{table}
@@
 \caption{Survival analysis (overall survival) for OACS stratified at the median of each of the selected components}
-\label{tab-survival-hr}
+\label{tab-survival-os_TCGA_AOCS}
 \end{table}
@@
 % \renewcommand{\arraystretch}{0.75}
-\input{/home/ipoole/Documents/gitrepos/HgsocTromics/Plots/SurvivalAnalysis/survival_analysis_table_os_TCGA_AOCS.tex}
+\input{/home/ipoole/Documents/gitrepos/HgsocTromics/Plots/SurvivalAnalysis/survival_analysis_table_pfs_TCGA_AOCS.tex}
 \end{center}
 \caption{Survival analysis (progression free survival) for OACS stratified at the median of each of the selected components}
-\label{tab-survival-hr}
+\label{tab-survival-hr_pfs_TCGA_AOCS}
 \end{table}

### 2020-07-22 MSc_Dissertation_B156476.tex
Results incorporate survival analysis summary bar plots

@@
 
-\graphicspath{{/home/ipoole/Dropbox/MSc_SSB/Disertation_Ovarian_Cancer/images/}}
+\graphicspath{{/home/ipoole/Documents/gitrepos/MScDissertation/images/}}  % trailin '/' is required!
 
@@
 \includegraphics[scale=0.5]{\plotspath{SurvivalAnalysis/multiple_kaplan_meier_os_TCGA_TCGA.pdf}} 
-\caption{Kaplan-Meier plots for each metasample component, stratified at the median value, for the TCGA $\rightarrow$ TCGA for overall survival (OS) case.
+\caption{Kaplan-Meier plots for each metasample component, stratified at the median value, for TCGA $\rightarrow$ TCGA for overall survival (OS) case.
 Hazard ratio and p-value is shown for each case.
@@
 \includegraphics[scale=0.5]{\plotspath{SurvivalAnalysis/multiple_kaplan_meier_os_TCGA_AOCS.pdf}} 
-\caption{Kaplan-Meier for the TCGA $\rightarrow$ TCGA for overall survival (OS) case.}
+\caption{Kaplan-Meier plots for TCGA $\rightarrow$ TCGA for overall survival (OS) case.}
 \label{fig-kaplan-meier-os-TCGA-OACS}
@@
 \includegraphics[scale=0.5]{\plotspath{SurvivalAnalysis/multiple_kaplan_meier_pfs_TCGA_AOCS.pdf}} 
-\caption{Kaplan-Meier for the TCGA $\rightarrow$ TCGA for progression free survival (PFS) case.  }
+\caption{Kaplan-Meier for TCGA $\rightarrow$ TCGA for progression free survival (PFS) case.  }
 \label{fig-kaplan-meier-pfs-TCGA-OACS}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.5]{\plotspath{/SurvivalAnalysis/survival_summary_barplots_NMF.pdf}} \\
+\includegraphics[scale=0.5]{\plotspath{/SurvivalAnalysis/survival_summary_barplots_ICA.pdf}} \\
+\includegraphics[scale=0.5]{\plotspath{/SurvivalAnalysis/survival_summary_barplots_PCA.pdf}} \\
+\caption{Visual summary of survival analysis as applied to TCGA(OS), AOCS(OS) and AOCS(PFS).
+Plots are divided by factorization method.
+Bar heights show $\log_2(HR)$ with p-value also shown.
+For a component to have robust correlation with survival we expect all three of the analysis bars to pass a significance threshold {\em and} for the $\log{HR}$ to to be {\em in the same sense\/}.
+None achieve this.}
+\label{fig-summary-survival-plots}
 \end{center}

### 2020-07-24 MSc_Dissertation_B156476.tex
Minor changes mostly in Results

@@
 The described method was implemented in Python, except for the KEGG gene enrichment analysis which was implemented in R.
-All code is availailable a github repository: \url{https://github.com/ipoole/HgsocTromics}.
+All code is available in a github repository: \url{https://github.com/ipoole/HgsocTromics}.
 Good software engineering practice has been followed, with an object-oriented design, frequent commits and unit testing.
-Unit tests are based on
+Unit tests are based on tiny expression matrices, just 100 genes by 10 patients, so that the whole test suite executes in around 10 seconds.
+These tests are particularly valuable when re-factoring code by providing a rapid means of detecting errors.
+All plots are generated in vector graphics pdf format to ensure smooth scaling to any resolution.
 
@@
 \section{Results}
+
+Results of the above described methodology are explained and presented here.
+Deeper interpretation and discussion is deferred to the following section.
+Throughout, consistent colours of \textcolor{blue}{blue}, \textcolor{orange}{orange} and \textcolor{green}{green} are used for results relating to NMF, ICA and PCA respectively.
 
@@
 Figure \ref{fig-AOCS-ica-fixed-vs-bootstrap} demonstrates that resampling error on a small N=80 dataset makes metagene extraction highly unstable, effectively un-usable.
-These figures show results for ICA only in the interest of space, but a very similar effect is observable for PCA and NMF.
+For brevity only results for ICA are shown, but a very similar effect is observable for PCA and NMF.
 For this reason it was decided to perform all metagene extraction on the larger N=374 TCGA dataset, in which the impact of sampling error is not so severe -- see figure \ref{fig-TCGA-ica-fixed-vs-bootstrap}.
@@
 
-\begin{table}
-\begin{center}
-\small
-% \renewcommand{\arraystretch}{0.75}
-\input{/home/ipoole/Documents/gitrepos/HgsocTromics/Plots/SurvivalAnalysis/survival_analysis_table_os_TCGA_TCGA.tex}
-\end{center}
-\caption{Survival analysis (overall survival) for TCGA stratified at the median of each of the selected components}
-\label{tab-survival-os_TCGA_TCGA}
-\end{table}
-
-\begin{table}
-\begin{center}
-\small
-% \renewcommand{\arraystretch}{0.75}
-\input{/home/ipoole/Documents/gitrepos/HgsocTromics/Plots/SurvivalAnalysis/survival_analysis_table_os_TCGA_AOCS.tex}
-\end{center}
-\caption{Survival analysis (overall survival) for OACS stratified at the median of each of the selected components}
-\label{tab-survival-os_TCGA_AOCS}
-\end{table}
-
-\begin{table}
-\begin{center}
-\small
-% \renewcommand{\arraystretch}{0.75}
-\input{/home/ipoole/Documents/gitrepos/HgsocTromics/Plots/SurvivalAnalysis/survival_analysis_table_pfs_TCGA_AOCS.tex}
-\end{center}
-\caption{Survival analysis (progression free survival) for OACS stratified at the median of each of the selected components}
-\label{tab-survival-hr_pfs_TCGA_AOCS}
-\end{table}
+%\begin{table}
+%\begin{center}
+%\small
+%% \renewcommand{\arraystretch}{0.75}
+%\input{/home/ipoole/Documents/gitrepos/HgsocTromics/Plots/SurvivalAnalysis/survival_analysis_table_os_TCGA_TCGA.tex}
+%\end{center}
+%\caption{Survival analysis (overall survival) for TCGA stratified at the median of each of the selected components}
+%\label{tab-survival-os_TCGA_TCGA}
+%\end{table}
+%
+%
+%\begin{table}
+%\begin{center}
+%\small
+%% \renewcommand{\arraystretch}{0.75}
+%\input{/home/ipoole/Documents/gitrepos/HgsocTromics/Plots/SurvivalAnalysis/survival_analysis_table_os_TCGA_AOCS.tex}
+%\end{center}
+%\caption{Survival analysis (overall survival) for OACS stratified at the median of each of the selected components}
+%\label{tab-survival-os_TCGA_AOCS}
+%\end{table}
+%
+%
+%\begin{table}
+%\begin{center}
+%\small
+%% \renewcommand{\arraystretch}{0.75}
+%\input{/home/ipoole/Documents/gitrepos/HgsocTromics/Plots/SurvivalAnalysis/survival_analysis_table_pfs_TCGA_AOCS.tex}
+%\end{center}
+%\caption{Survival analysis (progression free survival) for OACS stratified at the median of each of the selected components}
+%\label{tab-survival-hr_pfs_TCGA_AOCS}
+%\end{table}
 

### 2020-07-24 MSc_Dissertation_B156476.tex
Gene enrichment analysis additions * GEA method mostly done * Includes example plot of the GO * GEA results included in GO plot format, however some at too small to read * Figures include list of genes input directly from the Plots folder * Set draft mode for speedier writing / biulding.

@@
 
-\documentclass[tikz, 12pt,a4paper,oneside,fleqn]{article}
+\documentclass[draft, tikz, 12pt,a4paper,oneside,fleqn]{article}
 \usepackage[margin=2cm]{geometry}
@@
 \captionsetup[table]{labelfont=bf,format=hang,font={stretch=1.2, small}}
+\usepackage{graphicx}
+\usepackage{rotating}
+\usepackage{setspace}
 
@@
 
-\usepackage{graphicx}
-
 \graphicspath{{/home/ipoole/Documents/gitrepos/MScDissertation/images/}}  % trailin '/' is required!
-
 \newcommand{\plotspath}[1]{{/home/ipoole/Documents/gitrepos/HgsocTromics/Plots/#1}}
@@
 \subsection{Gene set intersection}
+\label{sec:gene-set-intersection}
 
@@
 \subsection{Gene enrichment analysis}
+
+The metagenes extracted by the above described factorization process provide valuable information into which genes varying in expression in the study samples; thus in our case, which genes are influential in HGSOC.
+
+In order to gain insights into what biological processes are involved, gene enrichment analysis against the Gene Ontology (GO) was carried out for each of the $K_{\mbox{\tiny NMF}} + K_{\mbox{\tiny ICA}} + K_{\mbox{\tiny PCA}}$ metagenes individually, as follows.
+
+The essence of gene enrichment analysis is to compare a candidate set of genes with many function gene sets, asking the question "are there significantly more (or less, i.e.
+depletion) intersecting genes than would be expected for the same number of genes being drawn at random (without replacement) from the total population of genes in the study.
+
+For each metagene, the candidate gene set was determined as those genes having weights outwith three standard deviations of the mean.
+This set was analysed using the Python GOATOOLS package \cite{GOATOOLS}.
+
+The gene ontology was downloaded from \url{http://purl.obolibrary.org/obo/go/go-basic.obo}.
+({\tt Purl.org} is a resource for managing permanent URLs; {\tt obo} refers to the Open Biological and Biomedical Ontology (OBO).
+Annotations linking human genes to GO concepts was downloaded from the GO website, specifically \url{http://geneontology.org/gene-associations/goa_human.gaf.gaf}.   
+The gene population was defined as the common 19,566 protein coding genes (see section \ref{sec:gene-set-intersection}) against which the metagenes were computed.
+
+Uncorrected p-value threshold was set to 0.01.
+Multiple hypothesis significance testing used the false discovery rate (FDR) method of Bengamini and Hochberg, the threshold being set to 0.01.
+This is more stringent than the 0.05 value which is typically used, and was chosen since multiple metagenes are being analysed, implying multiple hypothesis testing over and above that which is accounted for by FDR filtering within a single metagene run.
+
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[scale=0.13]{GO_0042773_lineage.png}
+\caption{A small section of the Gene Ontology (GO), focussed GO:0042773 (in purple), showing parents and children of the term.   Generated by GOATOOLS.}
+\label{fig:eg_go_graphic}
+\end{center}
+\end{figure}
+
+The result of this analysis, per metagene, is a list of enriched (or depleted) GO terms, each with an associated list of involved genes and a FDR significance level.
+This list can be inspected directly for insights, but that misses the point of the GO, which is to organise the terms hierarchically by 'is a' relationships.
+Thus, the enriched terms were rendered graphically to show them in the context of their parent terms.
+For some metagenes, additional filtering was required to reduce to plot to a manageable size; this is explained in the results section.
 
@@
 
+\subsection{Gene Enrichment Results}
+
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.14]{\plotspath{TCGA_OV_VST/GeneEnrichment/go_lineage_1_of_NMF_3.png}} \\
+\begin{spacing}{0.6}
+{\small Genes: \tiny \input{/home/ipoole/Documents/gitrepos/HgsocTromics/Plots/TCGA_OV_VST/GeneEnrichment/go_gene_set_1_of_NMF_3.txt}}
+\end{spacing}
+\vspace{0.5cm}
+{\small Component NMF 1 of 3} \\
+\vspace{1cm}
+\includegraphics[scale=0.14]{\plotspath{TCGA_OV_VST/GeneEnrichment/go_lineage_2_of_NMF_3.png}} \\
+\begin{spacing}{0.6}
+{\small Genes: \tiny \input{/home/ipoole/Documents/gitrepos/HgsocTromics/Plots/TCGA_OV_VST/GeneEnrichment/go_gene_set_2_of_NMF_3.txt}}
+\end{spacing}
+\vspace{0.5cm}
+{\small Component NMF 2 of 3} \\
+\caption{Lineage maps of enriched Gene Ongology (GO) terms for components 1 and 2 from the rank=3 NMF factorization.  (Component 3 produced no significant enrichment results)}
+\label{fig-NMF-go-lineage}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.1]{\plotspath{TCGA_OV_VST/GeneEnrichment/go_lineage_3_of_ICA_5.png}} \\
+\begin{spacing}{0.6}
+{\small Genes: \tiny \input{/home/ipoole/Documents/gitrepos/HgsocTromics/Plots/TCGA_OV_VST/GeneEnrichment/go_gene_set_3_of_ICA_5.txt}}
+\end{spacing}
+\vspace{0.5cm}
+{\small Component ICA 3 of 5} \\
+\vspace{1cm}
+\includegraphics[scale=0.08]{\plotspath{TCGA_OV_VST/GeneEnrichment/go_lineage_5_of_ICA_5.png}} \\
+\begin{spacing}{0.6}
+{\small Genes: \tiny \input{/home/ipoole/Documents/gitrepos/HgsocTromics/Plots/TCGA_OV_VST/GeneEnrichment/go_gene_set_5_of_ICA_5.txt}}
+\end{spacing}
+\vspace{0.5cm}
+{\small Component ICA 5 of 5} \\
+\caption{Lineage maps of enriched Gene Ongology (GO) terms for components 3 and 5 of the rank=5 ICA factorization.  }
+\label{fig-ICA-3-5-go-lineage}
+\end{center}
+\end{figure}
+
+\begin{sidewaysfigure}[ht]
+\begin{center}
+\includegraphics[scale=0.075]{\plotspath{TCGA_OV_VST/GeneEnrichment/go_lineage_4_of_ICA_5.png}} \\
+\begin{spacing}{0.6}
+{\small Genes: \tiny \input{/home/ipoole/Documents/gitrepos/HgsocTromics/Plots/TCGA_OV_VST/GeneEnrichment/go_gene_set_4_of_ICA_5.txt}}
+\end{spacing}
+\vspace{0.5cm}
+\caption{Lineage maps of enriched Gene Ongology (GO) terms for components 4 from the rank=5 ICA factorization. }
+\label{fig-ICA-4-go-lineage}
+\end{center}
+\end{sidewaysfigure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.1]{\plotspath{TCGA_OV_VST/GeneEnrichment/go_lineage_1_of_PCA_3.png}} \\
+
+\begin{spacing}{0.6}
+{\small Genes: \tiny \input{/home/ipoole/Documents/gitrepos/HgsocTromics/Plots/TCGA_OV_VST/GeneEnrichment/go_gene_set_1_of_PCA_3.txt}}
+\end{spacing}
+\vspace{0.5cm}
+{\small Component PCA 1 of 3} \\
+\vspace{1cm}
+\includegraphics[scale=0.09]{\plotspath{TCGA_OV_VST/GeneEnrichment/go_lineage_3_of_PCA_3.png}} \\
+\begin{spacing}{0.6}
+{\small Genes: \tiny \input{/home/ipoole/Documents/gitrepos/HgsocTromics/Plots/TCGA_OV_VST/GeneEnrichment/go_gene_set_3_of_PCA_3.txt}}
+\end{spacing}
+\vspace{0.5cm}
+{\small Component PCA 3 of 3} \\
+\caption{Lineage maps of enriched Gene Ongology (GO) terms for components 1 and 3 from the rank=5 PCA factorization. (Component 2 produced no significant enrichment results)  }
+\label{fig-PCA-1-3-go-lineage}
+\end{center}
+\end{figure}
+
 \subsection{Survival analysis results}

### 2020-07-27 MSc_Dissertation_B156476.tex
Various substantial additions * Heatmaps plots * Heatmap method and results text * Metasample / metadata reconcilliation text * Tweaked GO lineage plot sizes * Tweaked hyperref colours

@@
 
-\documentclass[draft, tikz, 12pt,a4paper,oneside,fleqn]{article}
-\usepackage[margin=2cm]{geometry}
+\documentclass[tikz, 12pt,a4paper,oneside,fleqn]{article}
+\usepackage[margin=1.5cm]{geometry}
 \usepackage[utf8]{inputenc}
@@
 \usepackage{setspace}
+\usepackage{afterpage}
 
@@
 
+\definecolor{darkblue}{rgb}{0.0, 0.0, 0.55}
 \usepackage{hyperref}
+\hypersetup{
+    colorlinks=true,
+    linkcolor=darkblue,
+    filecolor=blue,      
+    urlcolor=darkblue,
+}
 \usepackage{subfig}
@@
 \renewcommand{\familydefault}{\sfdefault}
+\setlength{\parindent}{2em}
+\setlength{\parskip}{1em}
+\renewcommand{\baselinestretch}{1.5}
 
@@
 
-The most commonly used matrix notation for ICA, (independent of any -omics applications) is:
-
+The most commonly used matrix notation for ICA, (independent of
+ -omics applications) is:
 \begin{equation}
@@
 
-Survival analysis was caried out stratified by prediced subtype.
+Survival analysis was carried out stratified by predicted subtype.
 
@@
 
-\subsection{Factor selection by cluster coherence}
-
+\subsection{Metagene selection by cluster coherence}
+\label{sec-metagene-selection}
 As outlined in the discussion, deciding the number of components (metagenes) to extract is key.
@@
 
-\subsection{Survival analysis}
-
-Survival analysis was performed to investigate whether the selected metagene factors correlated with patient survival.
-Overall  survival (OS) data is available for both the TCGA and AOCS datsets.
-Additionally, progression free survival (PFS) data is available for AOCS.
-As discussed in section \ref{sec-clustering-method}, metagenes were obtained by factorization on the TCGA dataset only.
-Application to TCGA (for OS) thus represents an in-sample test, while application to AOCS (for OS and PFS) is a more exacting out-of-sample test.
-There are thus three analyses to consider: 1) TCGA$\rightarrow$TCGA (OS), 2) TCGA$\rightarrow$OACS (OS) and 3) TCGA$\rightarrow$OACS (PFS).
-
-Analysis was performed using the Python {\tt Lifelines} package \cite{Davidson-Pilon2020}.
-Metasamples ($H$ matrices) were calculated from the metagenes and expression matrix as explained in section \ref{sec-transfer-to-novel} below.
-Metasample values were binariased to 0, 1 by thresholding at the median value.
-Kaplan-Meier plots were made in respect of derived metasample for each of the three analysis -- making $11 \times 3$ plots each with two survival curves (with 95\% confidence intervals).
-A hazard ratio (HR)
-\comment{Do I need to explain the definition and meaning of HR?} 
-was calculated for each plot by fitting Cox's proportional hazards model with p-value relating to the hypothesis that HR is significantly different to 1.0.
-
-\subsection{Transfer of learned factors to novel a dataset}
+\subsection{Transfer of learned metagenes to novel a dataset}
 \label{sec-transfer-to-novel}
@@
 
-\subsection{Survival analysis on novel dataset}
+\subsection{Reconciling computed metasamples and metadata}
+\label{sec-reconciling-metasamples-metadata}
 
-\subsection{Study of biological significance of predictive factors}
+To carry out the patient level analysis -- survival analysis, heatmaps and boxplot -- as described in the following sections, it was necessary to compile tables (Panda DataFrames) which bring together per-patient metasamples relating to each of the chosen metagenes, with associated metadata -- such as celluarity and survival information.
+The metagenes always derive from the TCGA dataset (because of its larger size as explained in section \ref{sec-metagene-selection}), but we wish to apply these metagenes to the expression data and associated metadata of either the TCGA or AOCS datasets.
+The mathematics of transferring metagenes across datasets has been described above.
+Care was required to ensure that expression matrices aligned correctly with metagenes with respect to their genes (rows), and with the metadata with respect to patient identifiers in the columns.
+
+% \subsection{Study of biological significance of predictive factors}
+
+\subsection{Survival analysis}
+
+Survival analysis was performed to investigate whether the selected metagene factors correlated with patient survival.
+Overall  survival (OS) data is available for both the TCGA and AOCS datsets.
+Additionally, progression free survival (PFS) data is available for AOCS.
+As discussed in section \ref{sec-clustering-method}, metagenes were obtained by factorization on the TCGA dataset only.
+Application to TCGA (for OS) thus represents an in-sample test, while application to AOCS (for OS and PFS) is a more exacting out-of-sample test.
+There are thus three analyses to consider: 1) TCGA$\rightarrow$TCGA (OS), 2) TCGA$\rightarrow$OACS (OS) and 3) TCGA$\rightarrow$OACS (PFS).
+
+Analysis was performed using the Python {\tt Lifelines} package \cite{Davidson-Pilon2020}.
+Metasamples ($H$ matrices) were calculated from the metagenes and expression matrix as explained in section \ref{sec-transfer-to-novel} below.
+Metasample values were binariased to 0, 1 by thresholding at the median value.
+Kaplan-Meier plots were made in respect of derived metasample for each of the three analysis -- making $11 \times 3$ plots each with two survival curves (with 95\% confidence intervals).
+A hazard ratio (HR)
+\comment{Do I need to explain the definition and meaning of HR?} 
+was calculated for each plot by fitting Cox's proportional hazards model with p-value relating to the hypothesis that HR is significantly different to 1.0.
+
+\subsection{Metasample heatmap analysis}
+
+To further investigate the relationship between selected metasamples and to available metadata of patients, clustered heatmaps were generated using the \href{https://seaborn.pydata.org/generated/seaborn.clustermap.html}{Seaborn {\tt clustermap()}} function using the reconciled metasamples and metadata (section \ref{sec-reconciling-metasamples-metadata}).
+Heatmaps were generated for TCGA and AOCS datasets (\emph{Should say something about stratifying the metadata, when I figure it out}).
 
@@
 %%%%%%%%%%% RESULTS %%%%%%%%%%%%%
+% \afterpage{\clearpage}
 
@@
 
-\subsection{Investigation of optimal parameters}
+% \subsection{Investigation of optimal parameters}
 
@@
 
-The critical decision of the choosing the facorization rank for each method -- $K_{\mbox{\tiny NMF}}$, $K_{\mbox{\tiny ICA}}$ and $K_{\mbox{\tiny PCA}}$ -- is based on figure \ref{fig-TCGA-all-bootstrap}.
+The critical decision of the choosing the factorization rank for each method -- $K_{\mbox{\tiny NMF}}$, $K_{\mbox{\tiny ICA}}$ and $K_{\mbox{\tiny PCA}}$ -- is based on figure \ref{fig-TCGA-all-bootstrap}.
 On this bases, considering both the t-SNE plots and silhouette scores, for following choices were made:
@@
 
-The rationalle for these choices is that, firstly $k>2$ is desirable to have sufficient information to work with.
+The rationale for these choices is that, firstly $k>2$ is desirable to have sufficient information to work with.
 NMF cluster coherence seems to deteriorate after $k=3$.
-Looking at the graph of silhouette scores, ICA appears to have a sweet spot at $k=5$.
-For PCA, $k = 3$ or $4$ both seem reasonable, $k=3$ is safe.
+Looking at the graph of silhouette scores (figure \ref{fig-TCGA-all-bootstrap}, bottom), ICA appears to have a sweet spot at $k=5$.
+For PCA, $k = 3$ or $4$ both seem reasonable; $k=3$ was selected.
 
@@
 \includegraphics[scale=\S]{\plotspath{TCGA_OV_VST/FactorClustering/multiple_single_factors_scatter_TCGA_PCA_2_3_5_10_bootstrap.pdf}} \\
-\includegraphics[scale=0.45]{\plotspath{/TCGA_OV_VST/FactorClustering/silhouette_plots_TCGA_2_3_4_5_6_7_8_9_10_bootstrap.pdf}} 
+\includegraphics[scale=0.55]{\plotspath{/TCGA_OV_VST/FactorClustering/silhouette_plots_TCGA_2_3_4_5_6_7_8_9_10_bootstrap.pdf}} 
 \caption{Metagene clustering for all three methods applied to the N=374 TCGA dataset with bootstrap sampling, over a range of factorization ranks.
@@
 
-\subsection{Gene Enrichment Results}
+\afterpage{\clearpage}
 
-\begin{figure}
+\subsection{Gene Enrichment (against GO) Results}
+
+Gene enrichment analysis against the GO results for, each of the 11 metagenes, in a table of GO terms with a list of the candidate genes which have an annotated association with that term.   
+To take full advantage of the hierarchical nature of the GO, results are presented as lineage maps in figures \ref{fig-NMF-go-lineage} to \ref{fig-PCA-1-3-go-lineage}.  
+The contributing candidate genes for all significantly enriched terms are shown beneath each figure.  
+(Ideally, the specific genes would be listed alongside each term, but this is difficult to achieve graphically).   
+High-level terms at depth less than 3 were removed.   
+One component -- ICA 4 of 5 -- was problematic in that it 49 GO terms (with depth $\geq 3$) were identified as significant.  
+For this component it was necessary to limit the graphic to the 12 terms having the largest number of associated candidate genes.   
+The full table of terms for this component is available in appendix \ref{tab-ica-4-of-5-enrichment-table}.
+
+The four components NMF-3-of-3, ICA-1-of-5, ICA-2-of-5 and PCA-2-of-3 yielded no significant enriched terms (for FDR $\leq$ 0.01).
+
+\begin{figure}[hb]
 \begin{center}
@@
 \vspace{1cm}
-\includegraphics[scale=0.08]{\plotspath{TCGA_OV_VST/GeneEnrichment/go_lineage_5_of_ICA_5.png}} \\
+\includegraphics[scale=0.07]{\plotspath{TCGA_OV_VST/GeneEnrichment/go_lineage_5_of_ICA_5.png}} \\
 \begin{spacing}{0.6}
@@
 \begin{center}
-\includegraphics[scale=0.075]{\plotspath{TCGA_OV_VST/GeneEnrichment/go_lineage_4_of_ICA_5.png}} \\
+\includegraphics[width=\textwidth, draft=false]{\plotspath{TCGA_OV_VST/GeneEnrichment/go_lineage_4_of_ICA_5.png}} \\
 \begin{spacing}{0.6}
@@
 \vspace{0.5cm}
-\caption{Lineage maps of enriched Gene Ongology (GO) terms for components 4 from the rank=5 ICA factorization. }
+\caption{Lineage maps of enriched Gene Ongology (GO) terms for components 4 from the rank=5 ICA factorization.
+For this component it was necessary to additionally filter the enriched terms -- see text.}
 \label{fig-ICA-4-go-lineage}
@@
 \begin{center}
-\includegraphics[scale=0.1]{\plotspath{TCGA_OV_VST/GeneEnrichment/go_lineage_1_of_PCA_3.png}} \\
-
+\includegraphics[width=\textwidth]{\plotspath{TCGA_OV_VST/GeneEnrichment/go_lineage_1_of_PCA_3.png}} \\
 \begin{spacing}{0.6}
@@
 \vspace{1cm}
-\includegraphics[scale=0.09]{\plotspath{TCGA_OV_VST/GeneEnrichment/go_lineage_3_of_PCA_3.png}} \\
+\includegraphics[width=\textwidth]{\plotspath{TCGA_OV_VST/GeneEnrichment/go_lineage_3_of_PCA_3.png}} \\
 \begin{spacing}{0.6}
@@
 
+\afterpage{\clearpage}
+
 \subsection{Survival analysis results}
+
+Kaplan-Meier overall survival (OS) plots stratified by each metasample component are shown for the TCGA dataset  in figure \ref{fig-kaplan-meier-os-TCGA-TCGA} and for the AOCS dataset in figure \ref{fig-kaplan-meier-os-TCGA-OACS}.
+Plots for progression free survival (PFS) on OACS are shown in figure \ref{fig-kaplan-meier-pfs-TCGA-OACS}.
+All plots show 95\% confidence intervals, hazard ratio (HR) with associated p-value.
+
+All these results are summarised in figure \ref{fig-summary-survival-plots}, which brings together the three sets of results -- TCGA (OS), OACS OS) and OACS (PFS) with respect to the 11 metasample components.
+$\log_2{HR}$ is used in order that the sense of the survival impact can be readily appreciated.
+If a metasample component had a robust correlation with survival, then we would expect the p-value for all three sets of results to show significance \emph{and} for the effects to be in the same direction.
+None of the 11 components pass this test, thus we have a negative result.
 
@@
 \includegraphics[scale=0.5]{\plotspath{SurvivalAnalysis/multiple_kaplan_meier_pfs_TCGA_AOCS.pdf}} 
-\caption{Kaplan-Meier for TCGA $\rightarrow$ TCGA for progression free survival (PFS) case.  }
+\caption{Kaplan-Meier plots for TCGA $\rightarrow$ TCGA for progression free survival (PFS) case.  }
 \label{fig-kaplan-meier-pfs-TCGA-OACS}
@@
 \includegraphics[scale=0.5]{\plotspath{/SurvivalAnalysis/survival_summary_barplots_ICA.pdf}} \\
-\includegraphics[scale=0.5]{\plotspath{/SurvivalAnalysis/survival_summary_barplots_PCA.pdf}} \\
+\includegraphics[scale=0.5]{\plotspath{/SurvivalAnalysis/survival_summary_barplots_PCA.pdf}}
 \caption{Visual summary of survival analysis as applied to TCGA(OS), AOCS(OS) and AOCS(PFS).
@@
 
-%\begin{table}
-%\begin{center}
-%\small
-%% \renewcommand{\arraystretch}{0.75}
-%\input{/home/ipoole/Documents/gitrepos/HgsocTromics/Plots/SurvivalAnalysis/survival_analysis_table_os_TCGA_TCGA.tex}
-%\end{center}
-%\caption{Survival analysis (overall survival) for TCGA stratified at the median of each of the selected components}
-%\label{tab-survival-os_TCGA_TCGA}
-%\end{table}
-%
-%
-%\begin{table}
-%\begin{center}
-%\small
-%% \renewcommand{\arraystretch}{0.75}
-%\input{/home/ipoole/Documents/gitrepos/HgsocTromics/Plots/SurvivalAnalysis/survival_analysis_table_os_TCGA_AOCS.tex}
-%\end{center}
-%\caption{Survival analysis (overall survival) for OACS stratified at the median of each of the selected components}
-%\label{tab-survival-os_TCGA_AOCS}
-%\end{table}
-%
-%
-%\begin{table}
-%\begin{center}
-%\small
-%% \renewcommand{\arraystretch}{0.75}
-%\input{/home/ipoole/Documents/gitrepos/HgsocTromics/Plots/SurvivalAnalysis/survival_analysis_table_pfs_TCGA_AOCS.tex}
-%\end{center}
-%\caption{Survival analysis (progression free survival) for OACS stratified at the median of each of the selected components}
-%\label{tab-survival-hr_pfs_TCGA_AOCS}
-%\end{table}
+\afterpage{\clearpage}
+
+\subsection{Metasample heatmap clustering results}
+
+As with survival analysis, metasamples were derived from the metagenes factorized from the TCGA dataset, then applied to both TCGA and AOCS datasets, as shown in figures \ref{fig-heatmap-TCGA-TCGA} and \ref{fig-heatmap-TCGA-OACS} respectively.
+Metadata was available for the AOCS dataset, and this is presented in colour coded columns on the left of figure \ref{fig-heatmap-TCGA-OACS}.
+The meanings of each column are as follows:
+\begin{description}
+\item[WGD]:
+\item[Cellularity]: proportion of cells belonging to the tumour (as opposed to surrounding normal tissue).
+\item[HR Detect]: score from homologous repair (HR) analysis, deriving from Ewing \etal's work \cite{Ewing2020}
+\item[Mutational Load]:
+\item[CNV Load]: Copy-number variation ...
+\item[Donor survival time]: (should probably remove).
+\end{description}
+
+The dendrogram clustering of metasamples (columns, top of figures) gives insights as to the similarity of each metasample component, while the clustering of patients (rows, left of figures) offers insights into sub-populations of patients and how (in the case of AOCS only) how these relate to metadata variables such as cellularity and mutational load.
+These topics will be taken up int he discussion.
+
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.7, draft=false]{\plotspath{HeatMaps/clustered_heatmap_TCGA_TCGA.pdf}} 
+\caption{Heatmap of the metasamples matrix for the TCGA dataset, computed from metagenes factorized also TCGA.
+Clustering dendrograms (based on a correlation distance) are shown between the metasamples (columns) and patients (rows).}
+\label{fig-heatmap-TCGA-TCGA}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.7, draft=false]{\plotspath{HeatMaps/clustered_heatmap_TCGA_AOCS.pdf}} 
+\caption{Heatmap of the metasamples matrix for the AOCS dataset, computed from metagenes factorized from the TCGA dataset.
+Clustering dendrograms (based on a correlation distance) are shown between the metasamples (columns) and patients (rows).
+Several patient patient metadata variables are shown in columns to the left.
+(ToDo: key for metadata colouring).}
+\label{fig-heatmap-TCGA-OACS}
+\end{center}
+\end{figure}
+
+\clearpage
 
@@
 
-\bibliography{OvarianCancer}
-\bibliographystyle{ieeetr}
+%\bibliography{OvarianCancer}
+%\bibliographystyle{ieeetr}
 

### 2020-07-27 MSc_Dissertation_B156476.tex
Read through method and resutls section, fixing and adding

@@
 
-\documentclass[tikz, 12pt,a4paper,oneside,fleqn]{article}
+\documentclass[draft, tikz, 12pt,a4paper,oneside,fleqn]{article}
 \usepackage[margin=1.5cm]{geometry}
@@
 \subsubsection{Datasets}
+Two gene expression datasets were assailable for this work (n refers to number of patients, g refers to number of genes).
 \begin{enumerate}
-\item TCGA, n=274, g=19,601
-\item AOCS, n=80, g=19,730
+\item TCGA, n=274, g=19,601, with metadata on survival
+\item AOCS, n=80, g=19,730, with metadata on survival, cellularity and additional genomic features
 \end{enumerate}
@@
 \subsubsection{Method outline}
+An overview of the methodology adopted in this work can be seen in figure \ref{fig-method-overview}.
 In outline our method is follows:
@@
 \item
-Unsupervised factor extraction by Matrix
-factorization methods : NMF, ICA and PCA.
-\item
 Identification of a consistent gene list between the TCGA and AOCS datasets.
 \item
-Factor selection for robustness by
-k-means clustering and silhouette score
+Unsupervised metagene extraction by matrix factorization methods : NMF, ICA and PCA.
 \item
-Investigation of biological significance of determined factors by gene enrichment analysis against the Gene Ontology (GO) and Kyoto Encyclopedia of Genes and Genomes (KEGG).
+Metagene selection for robustness by k-means clustering of bootstrap resampled factorizations evaluated by silhouette score, based on the TCGA dataset.
 \item
-Survival analysis on TCGA to identify key predictive factors
+Investigation of biological significance of determined factors by gene enrichment analysis against the Gene Ontology (GO).
 \item
-Transfer of identified factors to the AOCS dataset by matrix psuedo-inverse to validate efficacy of identified features.
+Computation by least squares optimisation of metasamples associated with with each TCGA derived metagene against TCGA and AOCS datasets.
 \item
-Revisit enrichment analysis for most significant factors to study biological significance.
+Alignment of available metadata with computed metasamples, for TCGA and AOCS.
+\item 
+Survival analysis on TCGA and AOCS to investigate relationship between metasamples and patient survival.
+\item
+Clustered heatmaps to visualise relationships between metasamples, patients and metadata.
 \end{enumerate}
@@
 \item Python 3.6
+\item PyCharm IDE
 \item Numpy for high performance matrix manipulation
 \item Pandas for data frame handing
-\item Matplotlib for plotting
+\item Matplotlib for general plotting
+\item Seaborn for clustered heatmaps
 \item Scikit-learn for matrix factorization and k-means clustering
@@
 
-Expression data for the AOCS and TCGA datasets was received for this project in a normalised matrix format.
-The full pre-processing pipeline is described in \ref{Ewing2020}, in brief using the DESeq2 package with normalization by variance stabilizing transformation.
+Expression data for the AOCS and TCGA datasets was received for this project in a spreadsheet format, having been derived from the RNA-Seq data as described in \ref{Ewing2020}, including  normalization by variance stabilizing transformation.
 
@@
 
-In order to allow factorizations found in the TCGA dataset dataset to be applied to the AOCS dataset it is necessary (or at least convenient) to synchronize the set of genes over which the expression matrices are defined.
-As provided to this project, the TCGA and AOCS datasets cover 19,610 and 19,730 genes respectively, 19,566 or which common to both (according to ENSG encodings).
+In order to allow factorizations found in the TCGA dataset to be applied to the AOCS dataset it is necessary (or at least convenient) to synchronize the set of genes over which the expression matrices are defined.
+As provided to this project, the TCGA and AOCS datasets cover 19,610 and 19,730 protein coding genes respectively, 19,566 or which common to both (according to ENSG encodings).
 Thus, both datasets were pruned to the 19,566 intersection set and ordered consistently.
@@
 \subsection{Matrix factorization computation}
-One of the aims of this work is to compare the efficacy of three dimensionality reduction methods -- NMF, ICA and PCA -- as explained in the introduction.
-The three methods have different properties and notational conventions, but to simplify the discussion here we adopt $X \approx W H$ for all three methods, where $X$ is the (genes, patients) expression matrix  $W$ is the (genes, factors) metagenes matrix and $H$ is the (factors, patients) metasamples matrix.
-Please refer back to section \ref{sec-matrix-factorization-intro} for further explanation.
+One of the aims of this work is to compare the efficacy of three methods of dimensionality reduction -- NMF, ICA and PCA -- as explained in the introduction.
+The three methods have different properties and notational conventions, but to simplify the discussion here we adopt the notation $X \approx W H$ for all three methods, where $X$ is the (genes, patients) expression matrix  $W$ is the (genes, factors) metagenes matrix and $H$ is the (factors, patients) metasamples matrix.
 
-Some of the key algorithm hyper parameters of each method were investigated and tuned with respect to reconstruction accuracy, specifically the root-mean square (RMS) difference between $X$ and $WH$.
+Some of the key algorithm hyper parameters of each method were investigated and tuned with respect to reconstruction accuracy, specifically the root-mean square (RMS) difference between $X$ and $W H$.
 The following parameters were explored in each case:
@@
 These parameters have not been explored in this project, and the default {\tt alpha = 0} (no regularization) was used.
-\item[ICA:]  Parameters {\tt max\_iter} and {\tt tol} we investigated as for NMF above.  Additionally, options for the entropy functional which forms the basis of the optimiziation were investigated.
+\item[ICA:]  Parameters {\tt max\_iter} and {\tt tol} were investigated as for NMF above.
+Additionally, options for the entropy functional which forms the basis of the optimiziation were investigated.
 \item[PCA:] This is in principle a deterministic algorithm based on eigenvector decomposition.
@@
 \label{sec-metagene-selection}
-As outlined in the discussion, deciding the number of components (metagenes) to extract is key.
+As outlined in the discussion, deciding on the number of components (metagenes) to extract -- that is the factorization rank -- is key.
 Taking more components results in more accurate representation of the observed expression matrix and provides more avenues to explore the underlying biology.
-However, it is important that the metagenes are \emph{stable}, that is likely to have reliable meaning when transfered to other datasets.
+However, it is important that the metagenes are \emph{stable}, that is that they have reliable meaning when transfered to other datasets.
 There are two sources of variation or instability to consider.
 
-NMF and ICA are inherently stochastic algorithms, sensitive to their starting state, so repeated runs give different results.
+Firstly, NMF and ICA are inherently stochastic algorithms, sensitive to their starting state, so repeated runs give different results.
 
 A second and more fundamental source of variation of relevance to all three factorization methods is \emph{sampling error}.
-Our factorizations are based on a small (N=80 or N=374) sample of patients drawn from the population of HGSOC patients; our particular datasets are just two examples of many different `draws' which could have been made.
+Our factorizations are based on a small (N=80 or N=374) sample of patients drawn from the population of HGSOC patients; our particular datasets are just two examples of many different `draws' which could have been made from that population.
 
@@
 
-Each of the three factorizer methods is evaluated for rank $k$ between 2 and 10.
-In each case, 50 iterations of factorization are performed on bootstrap samples, generating $50 k$ 19,566 dimensioned metagenes.
-t-SNE plots \ref{tsne-reference} visualise these high-dimensioned metagenes in a 2-D view.
+Each of the three factorizer methods was evaluated for rank $k$ between 2 and 10.
+In each case, 50 iterations of factorization are performed on bootstrap samples, generating $50 k$ instances of 19,566 dimensioned metagenes.
+Two dimensional t-SNE plots \ref{tsne-reference} were generated for visualization.
 If sampling and algorithm initialisation error are modest then we expect to see $k$ clusters of points.
-To avoid lengthy computation in the t-SNE clustering, the metagenes are first reduced to $r=20$ dimensions by PCA; brief experiments showed that this reduction had negligible effect on the t-SNE visualisation for $r \geq 10$.
+To avoid lengthy computation in the t-SNE clustering, the metagenes were first reduced to $r=20$ dimensions by PCA; brief experiments showed that this reduction had negligible effect on the t-SNE visualisation for $r \geq 10$.
 
@@
 These were referenced back to the original 19,566 dimensioned metagenes and the per-dimension median calculated.
-These $k$ median metagenes were saved a file named for the specific factorizer and rank $k$.
+These $k$ median metagenes were saved to a file named for the specific factorizer and rank $k$.
 
@@
 See
-% \href{https://en.wikipedia.org/wiki/Silhouette_(clustering)}{Wikipedia: Silouette (clustering)}.  
-The score is in the range -1 to +1, with 0 implying random scatter and 1 implying all points of a cluster overlay.
+\href{https://en.wikipedia.org/wiki/Silhouette_(clustering)}{Wikipedia: Silouette (clustering)}.
+The score is in the range -1 to +1, with 0 implying random scatter and 1 implying all points of a cluster perfectly overlay.
 
 The t-SNE plots and silhouette scores were assessed visually (see figures \ref{fig-first-cluster-result} to \ref{fig-last-custer-result} to decide, for each of the three factorizers, the highest rank with good cluster coherence, those median metagenes to take forward to gene enrichment and survival analysis.
+Note that the t-SNE plots are used only for visualisation; they are not involved in the computation of median metagenes or silhouette score.
 
-Initial investigation on the N=80 AOCS dataset showed very poor coherence, even for k=2.
-Thus it was decided compute and select metagenes only from the N=374 TCGA dataset.
+Initial investigation on the N=80 AOCS dataset showed very poor cluster coherence, even for k=2.
+Thus it was decided to compute and select metagenes only from the N=374 TCGA dataset.
 We refer to those selected ranks as $K_{\mbox{\tiny NMF}}$ etc., thus yielding $K_{\mbox{\tiny NMF}} + K_{\mbox{\tiny ICA}} + K_{\mbox{\tiny PCA}}$ metagenes in total.
+\footnote{Would it be better to state the selected ranks here?  As it is, this is left open until the results section}
 
@@
  for factorizer in NMF, ICA, PCA:
-    for k in 2..11:
+    for k in 2..10:
        Ws = []
@@
 
-The essence of gene enrichment analysis is to compare a candidate set of genes with many function gene sets, asking the question "are there significantly more (or less, i.e.
+The essence of gene enrichment analysis is to compare a candidate set of genes with many functional gene sets, asking the question "are there significantly more (or less, i.e.
 depletion) intersecting genes than would be expected for the same number of genes being drawn at random (without replacement) from the total population of genes in the study.
@@
 The gene ontology was downloaded from \url{http://purl.obolibrary.org/obo/go/go-basic.obo}.
-({\tt Purl.org} is a resource for managing permanent URLs; {\tt obo} refers to the Open Biological and Biomedical Ontology (OBO).
+({\tt Purl.org} is a resource for managing permanent URLs; {\tt obo} refers to the Open Biological and Biomedical Ontology (OBO)).
 Annotations linking human genes to GO concepts was downloaded from the GO website, specifically \url{http://geneontology.org/gene-associations/goa_human.gaf.gaf}.   
@@
 Uncorrected p-value threshold was set to 0.01.
-Multiple hypothesis significance testing used the false discovery rate (FDR) method of Bengamini and Hochberg, the threshold being set to 0.01.
-This is more stringent than the 0.05 value which is typically used, and was chosen since multiple metagenes are being analysed, implying multiple hypothesis testing over and above that which is accounted for by FDR filtering within a single metagene run.
+Multiple hypothesis significance testing used the false discovery rate (FDR) method of Bengamini and Hochberg, the FDR threshold being set to 0.01.
+This is more stringent than the 0.05 value which is typically used, and was chosen since multiple metagenes are being analysed, implying multiple hypothesis testing over and above that which is accounted for by FDR filtering within a single gene enrichment analysis run.
 
@@
 \includegraphics[scale=0.13]{GO_0042773_lineage.png}
-\caption{A small section of the Gene Ontology (GO), focussed GO:0042773 (in purple), showing parents and children of the term.   Generated by GOATOOLS.}
+\caption{Example of a small section of the Gene Ontology (GO), focussed GO:0042773 (in purple), showing parents and children of the term.   Generated by GOATOOLS.}
 \label{fig:eg_go_graphic}
@@
 Thus, the enriched terms were rendered graphically to show them in the context of their parent terms.
-For some metagenes, additional filtering was required to reduce to plot to a manageable size; this is explained in the results section.
 
@@
 
-Rank $k$ matrix factorization on the training dataset $\trainset$ (specifically the TCGA dataset) results in:
+It is fundamental to our approach that metagenes determined on the basis of one cohort of data can be used to generate metasamples for a different cohort.
+In the exposition below we refer to these cohorts as the \emph{training} and \emph{validation} cohorts respectively.
+
+Rank $k$ matrix factorization on the training dataset $\trainset$ (e.g. TCGA) results in:
 \begin{equation}
@@
 
-For ICA and PCA we use ordinary least square regression, formulated above by dropping the $H\geq 0$ constraint, using {\tt sklearn.linear\_model.LinearRegression}.
+For ICA and PCA we use ordinary least square regression, formulated above by dropping the $H\geq 0$ constraint, using {\tt scipy.linalg.lstsq}.
 
@@
 
-The end result of the analysis the $H_\validset$ matrix of shape $(k, n_\validset)$, thus delivering for dataset in our $\validset$ a feature vector of length $k$ -- a small number typically less than a dozen.
+The end result of the analysis is the $H_\validset$ matrix of shape $(k, n_\validset)$, thus delivering for dataset in our $\validset$ a feature vector of length $k$ -- a small number typically less than a dozen.
 
-Since in this work we wish to evaluate the efficacy of the three factorization methods -- NMF, ICA and PCA -- we in fact take forward three different $H$ matrices, having not necessarily the same rank $k$.
+Since in this work we wish to evaluate the efficacy of the three factorization methods -- NMF, ICA and PCA -- we in fact take forward three different $H$ matrices, each having ranks as for the associated metagenes.
 
@@
 
-To carry out the patient level analysis -- survival analysis, heatmaps and boxplot -- as described in the following sections, it was necessary to compile tables (Panda DataFrames) which bring together per-patient metasamples relating to each of the chosen metagenes, with associated metadata -- such as celluarity and survival information.
+To carry out patient level analysis -- survival analysis, heatmaps and boxplot -- as described in the following sections, it is necessary to compile tables (Panda DataFrames) which bring together per-patient metasamples relating to each of the chosen metagenes, with associated metadata -- such as cellularity and survival information.
 The metagenes always derive from the TCGA dataset (because of its larger size as explained in section \ref{sec-metagene-selection}), but we wish to apply these metagenes to the expression data and associated metadata of either the TCGA or AOCS datasets.
 The mathematics of transferring metagenes across datasets has been described above.
-Care was required to ensure that expression matrices aligned correctly with metagenes with respect to their genes (rows), and with the metadata with respect to patient identifiers in the columns.
-
-% \subsection{Study of biological significance of predictive factors}
+Care is required to ensure that expression matrices aligned correctly with metagenes with respect to their genes (rows), and with the metadata with respect to patient identifiers in the columns.
+Note that the same transferring approach is taken when metasamples are required for TCGA, even though we could in that case obtain the $H$ matrix directly from the factorization.
 
@@
 
-Survival analysis was performed to investigate whether the selected metagene factors correlated with patient survival.
+Survival analysis was performed to investigate whether the selected metagene factors correlate with patient survival.
 Overall  survival (OS) data is available for both the TCGA and AOCS datsets.
@@
 Analysis was performed using the Python {\tt Lifelines} package \cite{Davidson-Pilon2020}.
-Metasamples ($H$ matrices) were calculated from the metagenes and expression matrix as explained in section \ref{sec-transfer-to-novel} below.
+%Metasamples ($H$ matrices) were calculated from the metagenes and expression matrix as explained in section \ref{sec-transfer-to-novel} below.
 Metasample values were binariased to 0, 1 by thresholding at the median value.
@@
 A hazard ratio (HR)
-\comment{Do I need to explain the definition and meaning of HR?} 
+\footnote{Do I need to explain the definition and meaning of HR?}
 was calculated for each plot by fitting Cox's proportional hazards model with p-value relating to the hypothesis that HR is significantly different to 1.0.
@@
 
-To further investigate the relationship between selected metasamples and to available metadata of patients, clustered heatmaps were generated using the \href{https://seaborn.pydata.org/generated/seaborn.clustermap.html}{Seaborn {\tt clustermap()}} function using the reconciled metasamples and metadata (section \ref{sec-reconciling-metasamples-metadata}).
+To further investigate the relationship between metasamples and to available patient metadata, clustered heatmaps were generated using the \href{https://seaborn.pydata.org/generated/seaborn.clustermap.html}{Seaborn {\tt clustermap()}} function from the reconciled metasamples and metadata (section \ref{sec-reconciling-metasamples-metadata}).
 Heatmaps were generated for TCGA and AOCS datasets (\emph{Should say something about stratifying the metadata, when I figure it out}).
+Heatmap clustering by correlation distance was applied.
 
@@
 
-The described method was implemented in Python, except for the KEGG gene enrichment analysis which was implemented in R.
+The described method was implemented in Python.
+%except for the KEGG gene enrichment analysis which was implemented in R.
 All code is available in a github repository: \url{https://github.com/ipoole/HgsocTromics}.
 Good software engineering practice has been followed, with an object-oriented design, frequent commits and unit testing.
-Unit tests are based on tiny expression matrices, just 100 genes by 10 patients, so that the whole test suite executes in around 10 seconds.
+Unit tests are based on tiny expression matrices, just 100 genes by 10 patients.
+The whole test suite of over 60 tests executes in around 30 seconds.
 These tests are particularly valuable when re-factoring code by providing a rapid means of detecting errors.
@@
 %%%%%%%%%%% RESULTS %%%%%%%%%%%%%
-% \afterpage{\clearpage}
+\afterpage{\clearpage}
 
@@
 
-Results of the above described methodology are explained and presented here.
+Results of applying the above described methodology are explained and presented here.
 Deeper interpretation and discussion is deferred to the following section.
 Throughout, consistent colours of \textcolor{blue}{blue}, \textcolor{orange}{orange} and \textcolor{green}{green} are used for results relating to NMF, ICA and PCA respectively.
+Specific metagenes are referred to by, for example, ``NMF-2-of-3" -- meaning the 2nd component of the rank $k=3$ NMF factorization.
 
@@
 
-Figure \ref{fig-AOCS-ica-fixed-vs-bootstrap} demonstrates that resampling error on a small N=80 dataset makes metagene extraction highly unstable, effectively un-usable.
+Figure \ref{fig-AOCS-ica-fixed-vs-bootstrap} demonstrates that \emph{sampling} error is far greater then errors due to algorithm initialisation when the dataset is small (N=80), making metagene extraction highly unstable.
 For brevity only results for ICA are shown, but a very similar effect is observable for PCA and NMF.
 For this reason it was decided to perform all metagene extraction on the larger N=374 TCGA dataset, in which the impact of sampling error is not so severe -- see figure \ref{fig-TCGA-ica-fixed-vs-bootstrap}.
-
-The critical decision of the choosing the factorization rank for each method -- $K_{\mbox{\tiny NMF}}$, $K_{\mbox{\tiny ICA}}$ and $K_{\mbox{\tiny PCA}}$ -- is based on figure \ref{fig-TCGA-all-bootstrap}.
-On this bases, considering both the t-SNE plots and silhouette scores, for following choices were made:
-$K_{\mbox{\tiny NMF}}  =  3,
-K_{\mbox{\tiny ICA}}  =  5,
-K_{\mbox{\tiny PCA}}  =  3$.
-
-The rationale for these choices is that, firstly $k>2$ is desirable to have sufficient information to work with.
-NMF cluster coherence seems to deteriorate after $k=3$.
-Looking at the graph of silhouette scores (figure \ref{fig-TCGA-all-bootstrap}, bottom), ICA appears to have a sweet spot at $k=5$.
-For PCA, $k = 3$ or $4$ both seem reasonable; $k=3$ was selected.
 
@@
 
+The critical decision of the choosing the factorization rank for each method -- $K_{\mbox{\tiny NMF}}$, $K_{\mbox{\tiny ICA}}$ and $K_{\mbox{\tiny PCA}}$ -- is based on figure \ref{fig-TCGA-all-bootstrap}.
+On this bases, considering both the t-SNE plots and silhouette scores, the following choices were made:
+$K_{\mbox{\tiny NMF}}  =  3,
+K_{\mbox{\tiny ICA}}  =  5,
+K_{\mbox{\tiny PCA}}  =  3$.
+
+The rationale for these choices is that, firstly $k>2$ is desirable to have sufficient information to work with.
+NMF cluster coherence seems to deteriorate after $k=3$.
+Looking at the graph of silhouette scores (figure \ref{fig-TCGA-all-bootstrap}, bottom), ICA appears to have a sweet spot at $k=5$.
+For PCA, $k = 3$ or $4$ both seem reasonable; $k=3$ was selected.
+
 \begin{figure}
@@
 \caption{Metagene clustering for all three methods applied to the N=374 TCGA dataset with bootstrap sampling, over a range of factorization ranks.
-The silhoutte scores are also plotted (bottom).
+The silhouette scores are also plotted (bottom).
 It is on the basis of this figure that the factorization ranks for each method were selected.}
@@
 
-Gene enrichment analysis against the GO results for, each of the 11 metagenes, in a table of GO terms with a list of the candidate genes which have an annotated association with that term.   
+Gene enrichment analysis against the GO results in, each of the 11 metagenes, a table of GO terms with a list of the candidate genes which have an annotated association to that term.   
 To take full advantage of the hierarchical nature of the GO, results are presented as lineage maps in figures \ref{fig-NMF-go-lineage} to \ref{fig-PCA-1-3-go-lineage}.  
@@
 (Ideally, the specific genes would be listed alongside each term, but this is difficult to achieve graphically).   
-High-level terms at depth less than 3 were removed.   
-One component -- ICA 4 of 5 -- was problematic in that it 49 GO terms (with depth $\geq 3$) were identified as significant.  
+High-level terms at depth less than 3 were removed, since these are uninteresting.   
+One component -- ICA-4-of-5 -- was problematic in that 49 GO terms (with depth $\geq 3$) were identified as significant.  
 For this component it was necessary to limit the graphic to the 12 terms having the largest number of associated candidate genes.   
@@
 \vspace{0.5cm}
-{\small Component NMF 1 of 3} \\
+{\small Component NMF-1 of-3} \\
 \vspace{1cm}
@@
 \vspace{0.5cm}
-{\small Component NMF 2 of 3} \\
-\caption{Lineage maps of enriched Gene Ongology (GO) terms for components 1 and 2 from the rank=3 NMF factorization.  (Component 3 produced no significant enrichment results)}
+{\small Component NMF 2-of-3} \\
+\caption{Lineage maps of enriched Gene Ongology (GO) terms for components NMF-1-of-3 and NMF-2-of-3.  (NMF-3-of-3 produced no significant enrichment results)}
 \label{fig-NMF-go-lineage}
@@
 \vspace{0.5cm}
-{\small Component ICA 3 of 5} \\
+{\small Component ICA-3-of-5} \\
 \vspace{1cm}
@@
 \vspace{0.5cm}
-{\small Component ICA 5 of 5} \\
-\caption{Lineage maps of enriched Gene Ongology (GO) terms for components 3 and 5 of the rank=5 ICA factorization.  }
+{\small Component ICA-5-of-5} \\
+\caption{Lineage maps of enriched Gene Ongology (GO) terms for components ICA-3-of-5 and ICA-5-of-5}
 \label{fig-ICA-3-5-go-lineage}
@@
 \vspace{0.5cm}
-\caption{Lineage maps of enriched Gene Ongology (GO) terms for components 4 from the rank=5 ICA factorization.
-For this component it was necessary to additionally filter the enriched terms -- see text.}
+\caption{Lineage maps of enriched Gene Ongology (GO) terms for component ICA-4-of-5. For this component it was necessary to additionally filter the enriched terms -- see text.}
 \label{fig-ICA-4-go-lineage}
@@
 \vspace{0.5cm}
-{\small Component PCA 1 of 3} \\
+{\small Component PCA-1-of-3} \\
 \vspace{1cm}
@@
 \vspace{0.5cm}
-{\small Component PCA 3 of 3} \\
-\caption{Lineage maps of enriched Gene Ongology (GO) terms for components 1 and 3 from the rank=5 PCA factorization. (Component 2 produced no significant enrichment results)  }
+{\small Component PCA-3-of-3} \\
+\caption{Lineage maps of enriched Gene Ongology (GO) terms for components 1 and 3 from the rank=5 PCA factorization. (Component PCA-2-of-3 produced no significant enrichment results)  }
 \label{fig-PCA-1-3-go-lineage}
@@
 Plots for progression free survival (PFS) on OACS are shown in figure \ref{fig-kaplan-meier-pfs-TCGA-OACS}.
-All plots show 95\% confidence intervals, hazard ratio (HR) with associated p-value.
+All plots show 95\% confidence intervals and hazard ratio (HR) with associated p-value.
 
-All these results are summarised in figure \ref{fig-summary-survival-plots}, which brings together the three sets of results -- TCGA (OS), OACS OS) and OACS (PFS) with respect to the 11 metasample components.
+All of these results are summarised in figure \ref{fig-summary-survival-plots}, which brings together the three sets of results -- TCGA (OS), OACS OS) and OACS (PFS) with respect to the 11 metasample components.
 $\log_2{HR}$ is used in order that the sense of the survival impact can be readily appreciated.
-If a metasample component had a robust correlation with survival, then we would expect the p-value for all three sets of results to show significance \emph{and} for the effects to be in the same direction.
-None of the 11 components pass this test, thus we have a negative result.
+If a metasample component has a robust correlation with survival, then we expect the p-value for all three sets of results to show significance \emph{and} for the effects to have the same sense -- be in the same direction.
+None of the 11 components pass this test, thus we have failed to demonstrate a robust correlation between survival and any of the 11 metasamples.
 
@@
 \includegraphics[scale=0.5]{\plotspath{SurvivalAnalysis/multiple_kaplan_meier_os_TCGA_AOCS.pdf}} 
-\caption{Kaplan-Meier plots for TCGA $\rightarrow$ TCGA for overall survival (OS) case.}
+\caption{Kaplan-Meier plots for TCGA $\rightarrow$ TCGA for overall survival (OS).}
 \label{fig-kaplan-meier-os-TCGA-OACS}
@@
 \includegraphics[scale=0.5]{\plotspath{SurvivalAnalysis/multiple_kaplan_meier_pfs_TCGA_AOCS.pdf}} 
-\caption{Kaplan-Meier plots for TCGA $\rightarrow$ TCGA for progression free survival (PFS) case.  }
+\caption{Kaplan-Meier plots for TCGA $\rightarrow$ TCGA for progression free survival (PFS).  }
 \label{fig-kaplan-meier-pfs-TCGA-OACS}
@@
 Plots are divided by factorization method.
-Bar heights show $\log_2(HR)$ with p-value also shown.
-For a component to have robust correlation with survival we expect all three of the analysis bars to pass a significance threshold {\em and} for the $\log{HR}$ to to be {\em in the same sense\/}.
+Bar heights show $\log_2(\mbox{HR})$ with p-value also shown.
+For a component to have robust correlation with survival we expect all three of the analysis bars to pass a significance threshold {\em and} for the $\log{\mbox{HR}}$ to to be {\em in the same sense\/}.
 None achieve this.}
@@
 
-As with survival analysis, metasamples were derived from the metagenes factorized from the TCGA dataset, then applied to both TCGA and AOCS datasets, as shown in figures \ref{fig-heatmap-TCGA-TCGA} and \ref{fig-heatmap-TCGA-OACS} respectively.
+As with survival analysis, metasamples were derived from the metagenes factorized from the TCGA dataset, then applied to both TCGA and AOCS datasets.
+The resulting heatmaps are shown in figures \ref{fig-heatmap-TCGA-TCGA} and \ref{fig-heatmap-TCGA-OACS} respectively.
 Metadata was available for the AOCS dataset, and this is presented in colour coded columns on the left of figure \ref{fig-heatmap-TCGA-OACS}.
@@
 
-The dendrogram clustering of metasamples (columns, top of figures) gives insights as to the similarity of each metasample component, while the clustering of patients (rows, left of figures) offers insights into sub-populations of patients and how (in the case of AOCS only) how these relate to metadata variables such as cellularity and mutational load.
-These topics will be taken up int he discussion.
+The dendrogram clustering of metasamples (columns, top of figures) gives insights into the similarity of each metasample component, while the clustering of patients (rows, left of figures) offers insights into sub-populations of patients and how (in the case of AOCS only) how these relate to metadata variables such as cellularity and mutational load.
+These topics will be taken up in the following discussion section.
 
@@
 \caption{Heatmap of the metasamples matrix for the AOCS dataset, computed from metagenes factorized from the TCGA dataset.
-Clustering dendrograms (based on a correlation distance) are shown between the metasamples (columns) and patients (rows).
 Several patient patient metadata variables are shown in columns to the left.

### 2020-07-28 MSc_Dissertation_B156476.tex
Added cool methodology_overview; some WIP on introduction

@@
 \section{Introduction}
-
-Research questions:  
-\begin{enumerate}
-\item What are the known patterns of gene expression in high-grade serous ovarian cancer?
-\item How can blind-source separation techniques such as independent component analysis and non-negative matrix factorization be applied to transcrptomics studies?
-\item How can biological databases be used to link genes into oncogenesis pathways?
-\end{enumerate}
+\emph{Say something about ovarian cancer!   An image perhaps.}
 
@@
 
-A good introduction is found in ref \cite{Stein-OBrien2018} matrix factorization (MF) methods are discussed as a means of extraction the low dimensional structure from a high dimensional ``-omics" datasets, with gene expression analysis being the prime example.
+A good introduction to field is ref \cite{Stein-OBrien2018}, in which matrix factorization (MF) methods are discussed as a means of extraction the low dimensional structure from a high dimensional ``-omics" datasets, with gene expression analysis being the prime example.
+
+There are three matrix factorization methods of interest to this project:
+\begin{enumerate}
+\item Independent component analysis (ICA)
+\item Non-negative matrix factorization (NMF)
+\item Principal component analysis (PCA)
+\end{enumerate}
+
 Conventionally, transcriptomics expression arrays are oriented with genes (or transcripts) as rows, and samples (e.g. patients) in columns.   
@@
 MF methods reduces this large $M \times N$ (genes $\times$ samples) matrix into two two smaller matrices.
-In the terminology of Stein-O'Brien \etal\ these are the $M \times K$ \emph{pattern matrix} and the $K \times N$ \emph{amplitude matrix}.
+In the general terminology of Stein-O'Brien \etal\ these are the $M \times K$ \emph{pattern matrix} and the $K \times N$ \emph{amplitude matrix}.
 $K$ is typically small of the order 10, and refers to the number of extracted \emph{factors} (or components).
+Since we are focussing on transriptomics, rather than pattern and amplitude we will use the terms \emph{metagenes} and \emph{metasamples} respectively, which are in common useage.
+The symbol conventions for these matrices varies depending whether NMF, ICA or PCA is being discussed.
+NMF generally uses $W$ and $H$ respectively while ICA generally uses $S$ and $A$.
+PCA is tends to be differently formulated, but $W$ and $T$ sometimes used.
+The original matrix to be factorized is variously named $X$ or $V$.
+
+In this work we wish to treat and discuss the three factorization methods in a unified way.  Thus, for the remainder of this dissertation we will adopt the following conventions:
+\begin{description}
+\item[Expression matrix:] $X$ $(M \times N)$
+\item[Metagene matrix:] $W$ $(M \times K)$
+\item[Metasample matrix:] $H$ $(K \times N)$
+\end{description}
+where $N$ is the number of patients (or samples), $M$ is the number of genes and $K$ is the factorization rank, i.e. the number of components or factors.  Thus the factorization is written
+\begin{equation}
+X \approx W H
+\end{equation}
+Adding subscripts to indicate the row,column orientation of the matrices:
+\begin{equation}
+X_{M,N} \approx W_{M,K} H_{K,N}
+\end{equation}
+as illustrated in figure \ref{fig-matrix-factorization}
 
@@
 \centering
-
 \subfloat[Optimising independence of Meta\underline{genes}]{
@@
 	\node [right=of X] (eq) {\large $\approx$};
-	\node [right=of eq] (S) [draw, fill=yellow, minimum width=\K, minimum height=\M] {\large S};
+	\node [right=of eq] (S) [draw, fill=yellow, minimum width=\K, minimum height=\M] {\large W};
 	\node [right=of S] (times) {\LARGE$\times$};
-	\node [right=of times] (A) [draw, fill=cyan, minimum width=\N, minimum height=\K] {\large A};
+	\node [right=of times] (A) [draw, fill=cyan, minimum width=\N, minimum height=\K] {\large H};
 
@@
 	\node [left=\fudge of A] {\tiny{$K$}};
-	\node [below=of A, text width=2cm, align=center] {\small Metasamples};
+	\node [below=of A, text width=2.5cm, align=center] {\small Metasamples};
 \end{tikzpicture}}
@@
 	\node [right=of X] (eq) {\large $\approx$};
-	\node [right=of eq] (S) [draw, fill=cyan, minimum width=\K, minimum height=\N]{\large S};
+	\node [right=of eq] (S) [draw, fill=cyan, minimum width=\K, minimum height=\N]{\large W};
 	\node [right=of S] (times) {\LARGE$\times$};
-	\node [right=of times] (A) [draw, fill=yellow, minimum width=\M, minimum height=\K]{\large A};
+	\node [right=of times] (A) [draw, fill=yellow, minimum width=\M, minimum height=\K]{\large H};
 
@@
 	\node [left=\fudge of S] {\rotatebox{90}{\tiny{$N$ Samples}}};
-	\node [below=of S, text width=2cm, align=center] {\small Metasamples};
+	\node [below=of S, text width=2.5cm, align=center] {\small Metasamples};
 
@@
 The $X \approx S \times A$ notation is conventional for ICA.}
-\label{fig:matrix-factorization}
+\label{fig-matrix-factorization}
 \end{figure}
-
-There are three matrix factorization methods of potential interest to this project:
-\begin{enumerate}
-\item Independent component analysis (ICA)
-\item Non-negative matrix factorization (NMF)
-\item Principal component analysis (PCA)
-\end{enumerate}
 
@@
 
-See figure \ref{fig:matrix-factorization}.
+See figure \ref{fig-matrix-factorization}.
 
@@
 
-\subsection{Research Questions}
+\subsection{Research questions}
+The following research goals were set out early in project. 
 \begin{enumerate}
-\item
-Which are the influential genes in HGSOC according
-to expression data? Do these confirm published
-results?
-\item
-Which of PCA, ICA and NMF is best suited to this
-analysis?
-\item
-For identified genes, what is the underlying biology of
-their action?
-\item
-Can we discriminate subtypes of HGSOC from
-expression data?
-\item
-Is there value in incorporating somatic and germline
-genomic data?
+\item Which are the influential genes in HGSOC according to gene expression data?   Do these confirm published results?
+\item Which of PCA, ICA and NMF is best suited to this analysis?
+\item For identified genes, what is the underlying biology?
+\item Can we discriminate subtypes of HGSOC from expression data?
+\item Is there value in incorporating somatic and germline genomic data?
 \end{enumerate}
@@
 An overview of the methodology adopted in this work can be seen in figure \ref{fig-method-overview}.
+
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.7, draft=false]{methodlology_overview.pdf}
+\end{center}
+\caption{Overview of methodology as information flow}
+\label{fig-method-overview}
+\end{figure}
 In outline our method is follows:
@@
 %%%%%%%%%%% RESULTS %%%%%%%%%%%%%
-\afterpage{\clearpage}
+% \afterpage{\clearpage}
 

### 2020-07-28 MSc_Dissertation_B156476.tex
Introduction improved; some references fixed

@@
 \end{equation}
-as illustrated in figure \ref{fig-matrix-factorization}
+as illustrated in figure \ref{fig-matrix-factorization} (a).
 
@@
 \end{tikzpicture}}
-\caption{Two ways of configuring ICA.
-The $S$ matrix is optimised for independence of components, thus configuration (a) optimises for metagenes, while (b) optimises for metasamples.
-The $X \approx S \times A$ notation is conventional for ICA.}
+\caption{Two ways of configuring matrix factorization in the context of gene expression analysis.
+In configuration (a) NMF and ICA will optimise the \emph{metagenes}, while in (b) the \emph{metasamples} are optimised. }
 \label{fig-matrix-factorization}
@@
 
-\subsubsection{Independent Component Analysis (ICA)}
+However, in the case of NMF and ICA, the $W$ and $H$ matrices cannot be trivially exchanged and transposed.
+This is because the optimisations (sparsity and independence respectively) which define these algorithms are focussed on the $W$ matrix.
+It is perfectly possible to apply these algorithms to gene expression analysis with exchanged and transposed meanings of $W$ and $H$, and this is illustrated in figure \ref{fig-matrix-factorization} (b).
+In this case it is the properties of the \emph{metasamples} which will be optimised.
+Thus, the two forms are different in substance, not simply in notational convention.
 
-The most commonly used matrix notation for ICA, (independent of
- -omics applications) is:
-\begin{equation}
-X \approx S A
-\end{equation}
-Adding subscripts to indicate the row,column orientation of the matrices:
-\begin{equation}
-X_{M,N} \approx S_{M,K} A_{K,N}
-\end{equation}
-
-$X$ is assumed to have been pre-whitened.
-$S$ is referred to variously as the pattern matrix, latent variable or independent component matrix.
-$A$ is referred to as the mixing or amplitude matrix.
-If $X$ is sized ($M$, $N$) then $S$ will be ($M$, $K$) and $A$ ($K$, $N$) by the rules of matrices.
-Crucially, it is the $S$ matrix which captures the independent components.
-
-Authors use different conventions in the nomencalture and orientation of the expression array.
-This is partly a matter of convention but there are ramifications of substance.
-The terms ``metagenes'' is used commonly and consistently to refer to the $K$ components of $M$ long gene expression profiles.
-Similarly, ``metasamples'' refers to the $K$ components of $N$ long sample profiles.
-However, the packing of these entities into matrices varies in orientation, so that metagenes are sometimes $M \times K$, sometimes $K \times M$, and metasamples are sometimes $N \times K$, sometimes $K \times N$, depending on the orientation of the input expression array.
-
-The point of substance (as opposed to notational convention) is whether ICA is used to maximise independence of metagenes or of metasamples.
-This relates to the choice of orientation of the input expression matrix.
-Given the standard definition of ICA above \cite{ICA_definition}, the following rule applies to the (row, column) orientation of the expression matrix, as passed to ICA:
-\begin{description}
-\item[to optimise metagenes] the expression matrix is oriented as ($M$ genes, $N$ samples) resulting in a ($M$, $K$) metagenes matrix. 
-\item[to optimise metasamples] the expression matrix is oriented as ($N$ samples, $M$ genes), resulting in a ($N$, $K$) metasamples matrix.
-\end{description}
-
-See figure \ref{fig-matrix-factorization}.
-
-There is substantial confusion an lack of clarity in the way that ICA is applied in transcriptomics research.
+There seems to be substantial confusion an lack of clarity in the way that matrix factorization, particularly ICA, is applied in transcriptomics research.  
 ``Surprisingly, both ways of applying ICA to omics data are wide-spread, and sometimes it makes an effort to figure out in which way ICA was applied." \cite{Sompairac2019a}, and ``Different protocols to apply ICA to transcriptomic data exist and currently no single standard approach has been defined.
-The main difference in the existing approaches consists in what is considered as source signal matrix in the decomposition" \cite{Cantini}.
-According to Cantini \etal\, references \cite{Au-Yeung2014}, \cite{Kairov2017}, \cite{Kong2008} and \cite{Lee2003} optimise metagenes, while references \cite{Meng2016}
-
-The most commonly used matrix notation for ICA, (independent of any -omics applications) is:
-\begin{equation}
-X \approx S A
-\end{equation}
-
-$X$ is assumed to have been pre-whitened.
-$S$ is referred to variously as the pattern matrix, latent variable or independent component matrix.
-$A$ is referred to as the mixing or amplitude matrix.
-If $X$ is sized ($M$, $N$) then $S$ will be ($M$, $K$) and $A$ ($K$, $N$) by the rules of matrices.
-Crucially, it is the $S$ matrix which captures the independent components.
+The main difference in the existing approaches consists in what is considered as source signal matrix in the decomposition" \cite{Cantini2019}.
+According to Cantini \etal\, references \cite{Au-Yeung2014}, \cite{Kairov2017}, \cite{Kong2008} and \cite{Lee2003} optimise metagenes, while references \cite{Meng2016} and \cite{Barillot2013} optimises metasamples.
 
@@
 The problem of determining the optimal $K$ for a given expression matrix is addressed  by Kairov \etal\ \cite{Kairov2017}, based on optimizing the \emph{stability} of the components over multiple runs.
-
-\subsubsection{Non-negative Matrix Factorization}
-
-When the NMF is used, the pattern and amplitude matrices are typically notated as $W$ and $H$ respectively, while the original expression matrix is notated as $V$, or $A$ and oriented as ($M$ genes, $N$ samples).
-The facorization is thus written $V \approx W H$, Thus $W$ represents the ($M$, $K$) metagenes matrix and $H$ represents the ($K$, $N$) metasamples.
-
-\subsubsection{Principal component analysis}
 
@@
 
-Expression data for the AOCS and TCGA datasets was received for this project in a spreadsheet format, having been derived from the RNA-Seq data as described in \ref{Ewing2020}, including  normalization by variance stabilizing transformation.
+Expression data for the AOCS and TCGA datasets was received for this project in a spreadsheet format, having been derived from the RNA-Seq data as described in \cite{Ewing2020}, including  normalization by variance stabilizing transformation.
 
@@
 For each metagene, the candidate gene set was determined as those genes having weights outwith three standard deviations of the mean.
-This set was analysed using the Python GOATOOLS package \cite{GOATOOLS}.
+This set was analysed using the Python GOATOOLS package \cite{Klopfenstein2018}.
 
@@
 \label{fig-AOCS-ica-fixed-vs-bootstrap}
+\label{fig-first-cluster-result}
 \end{center}
@@
 \label{fig-TCGA-all-bootstrap}
+\label{fig-last-cluster-result}
 \end{center}
@@
 \begin{center}
-\includegraphics[scale=0.7, draft=false]{\plotspath{HeatMaps/clustered_heatmap_TCGA_TCGA.pdf}} 
+\includegraphics[scale=0.7]{\plotspath{HeatMaps/clustered_heatmap_TCGA_TCGA.pdf}} 
 \caption{Heatmap of the metasamples matrix for the TCGA dataset, computed from metagenes factorized also TCGA.
@@
 \begin{center}
-\includegraphics[scale=0.7, draft=false]{\plotspath{HeatMaps/clustered_heatmap_TCGA_AOCS.pdf}} 
+\includegraphics[scale=0.7]{\plotspath{HeatMaps/clustered_heatmap_TCGA_AOCS.pdf}} 
 \caption{Heatmap of the metasamples matrix for the AOCS dataset, computed from metagenes factorized from the TCGA dataset.
@@
 
-%\bibliography{OvarianCancer}
-%\bibliographystyle{ieeetr}
+\bibliography{OvarianCancer}
+\bibliographystyle{ieeetr}
 

### 2020-07-29 MSc_Dissertation_B156476.tex
Substantial additions * More explanations / observations in results * Added Jaccard heatmap of candidate gene set similarities * Good start at Discussion, analysing each component.

@@
 
+Our overall approach is to use \emph{unsupervised} machine learning methods to represent gene expression data in a small number of features, then to study whether these features correlate with clinical and biological variables.
+This is appropriate in our case, since of the two datasets available (described below), one (TCGA) has relatively large n but little available metadata, while the other (AOCS) is much smaller but has useful metadata.
+This motivates an approach of unsupervised learning on the TCGA dataset.
+
 \subsubsection{Datasets}
@@
 
-Initial investigation on the N=80 AOCS dataset showed very poor cluster coherence, even for k=2.
-Thus it was decided to compute and select metagenes only from the N=374 TCGA dataset.
-We refer to those selected ranks as $K_{\mbox{\tiny NMF}}$ etc., thus yielding $K_{\mbox{\tiny NMF}} + K_{\mbox{\tiny ICA}} + K_{\mbox{\tiny PCA}}$ metagenes in total.
-\footnote{Would it be better to state the selected ranks here?  As it is, this is left open until the results section}
+Initial investigation on the N=80 AOCS dataset showed very poor cluster coherence, even for k=2.   Thus it was decided to compute and select metagenes only from the N=374 TCGA dataset.   
+The selection rationale is set out in the results section, but is convenient to state here that the following ranks were selected:
+$K_{\mbox{\tiny NMF}} = 3, K_{\mbox{\tiny ICA}} = 5, K_{\mbox{\tiny PCA}} =3$.  Thus, there were $3 + 5 + 3 = 11$ metagenes taken forward for follow on analysis.
 
@@
 
-In order to gain insights into what biological processes are involved, gene enrichment analysis against the Gene Ontology (GO) was carried out for each of the $K_{\mbox{\tiny NMF}} + K_{\mbox{\tiny ICA}} + K_{\mbox{\tiny PCA}}$ metagenes individually, as follows.
+In order to gain insights into what biological processes are involved, gene enrichment analysis against the Gene Ontology (GO) was carried out for each of the 11 metagenes individually, as follows.
 
@@
 
-To carry out patient level analysis -- survival analysis, heatmaps and boxplot -- as described in the following sections, it is necessary to compile tables (Panda DataFrames) which bring together per-patient metasamples relating to each of the chosen metagenes, with associated metadata -- such as cellularity and survival information.
+To carry out patient level analysis -- survival analysis, heatmaps and boxplot -- as described in the following sections, it is necessary to compile tables (Panda DataFrames) which bring together per-patient metasamples relating to each of the selected metagenes, with associated metadata -- such as cellularity and survival information.
 The metagenes always derive from the TCGA dataset (because of its larger size as explained in section \ref{sec-metagene-selection}), but we wish to apply these metagenes to the expression data and associated metadata of either the TCGA or AOCS datasets.
@@
 
-Figure \ref{fig-AOCS-ica-fixed-vs-bootstrap} demonstrates that \emph{sampling} error is far greater then errors due to algorithm initialisation when the dataset is small (N=80), making metagene extraction highly unstable.
+From figure \ref{fig-AOCS-ica-fixed-vs-bootstrap} it can be seen that demonstrates that with sampling error excluded, clusters appear reasonably coherent, but when sampling error is modelled by bootstrap sampling then we see the factorizations become much less stable.
+This demonstrates that \emph{sampling} error is far greater then errors due to algorithm initialisation when the dataset is small (N=80), making metagene extraction highly unstable.
+Figure  \ref{fig-TCGA-ica-fixed-vs-bootstrap} makes the same comparison for the lager N=374 TCGA dataset, in which it can be seen that the impact of sampling error is not so severe.
 For brevity only results for ICA are shown, but a very similar effect is observable for PCA and NMF.
-For this reason it was decided to perform all metagene extraction on the larger N=374 TCGA dataset, in which the impact of sampling error is not so severe -- see figure \ref{fig-TCGA-ica-fixed-vs-bootstrap}.
+
+For this reason, and in line with our overall strategy, it is appropriate to perform all metagene extraction on the larger TCGA dataset, in the expectation that the obtained metagenes will be more robust an likely to generalise better to other datasets.
 
@@
 \caption{Clustering of metagenes from ICA facorizations on the N=80 AOCS dataset, comparing \emph{fixed} sampling (top row) with \emph{bootstrap} sampling (bottom row) for a selection of factorization ranks.
-With sampling error excluded, clusters appear reasonably coherent, but when sampling error is modelled by bootstrap sampling then we see the factorizations become much less stable.}
+}
 \label{fig-AOCS-ica-fixed-vs-bootstrap}
@@
 \includegraphics[scale=\S]{\plotspath{TCGA_OV_VST/FactorClustering/multiple_single_factors_scatter_TCGA_ICA_2_3_5_10_bootstrap.pdf}}
-\caption{Clustering of metagenes from ICA facorizations on the N=374 TCGA dataset, again comparing fixed and bootstrap sampling.
-In this larger dataset, sampling error (bottom row) is not so severe; compare with figure \ref{fig-AOCS-nmf-fixed-vs-bootstrap}}
+\caption{Clustering of metagenes from ICA facorizations on the N=374 TCGA dataset, again comparing fixed and bootstrap sampling.}
 \label{fig-TCGA-ica-fixed-vs-bootstrap}
@@
 For PCA, $k = 3$ or $4$ both seem reasonable; $k=3$ was selected.
+
+There is a curious artefact visible in the NMF factorization clusterings of figure \ref{fig-TCGA-all-bootstrap}, top row.
+For $k=2, 3 and 5$ several of the clusters show a bi-modal character.
+This is not observed in the fixed sampling case (not shown).
+The artefact is difficult to explain.
 
@@
 (Ideally, the specific genes would be listed alongside each term, but this is difficult to achieve graphically).   
-High-level terms at depth less than 3 were removed, since these are uninteresting.   
+High-level terms at depth less than 3 were removed, since these are generic (e.g. "regulation of biological process") and so uninteresting.   
 One component -- ICA-4-of-5 -- was problematic in that 49 GO terms (with depth $\geq 3$) were identified as significant.  
@@
 The four components NMF-3-of-3, ICA-1-of-5, ICA-2-of-5 and PCA-2-of-3 yielded no significant enriched terms (for FDR $\leq$ 0.01).
+
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.5, draft=false]{\plotspath{TCGA_OV_VST/GeneEnrichment/component_jaccard_heatmap.pdf}}
+\end{center}
+\caption{Heatmap of Jaccard similarities between the candidate genes identified by the 11 components.}
+\label{fig-component-jaccard-similarities}
+\end{figure}
+
+To what extent do the 11 components overlap in the candidate genes they propose (independent of enrichment analysis)?
+Figure \ref{fig-component-jaccard-similarities} shows this by a heatmap of Jaccard similarities.
+Jaccard similarity (or index) is defined on a pair of sets, in this case sets of genes, as:
+\[
+   \mbox{Jaccard}(A, B) = \frac{A \cap B}{A \cup B}
+\]
+lying between 0 (no similarity) and 1 (identical).
+
+From this heatmap we can see that within a factorization method the components have almost no similarity -- they represent largely independent projections, as one would expect.
+Between methods, however, there is considerable overlap in some cases.
+In particular PCA 2\_of\_3 vs ICA 1\_of\_5 (J=0.61), PCA 3\_of\_3 vs ICA 5\_of\_5 (J=0.49) and NMF 1\_of\_3 vs ICA 3\_of\_5 (J=0.39).
 
@@
 \vspace{0.5cm}
-{\small Component NMF-1 of-3} \\
+{\small Component NMF-1-of-3} \\
 \vspace{1cm}
@@
 {\small Component NMF 2-of-3} \\
-\caption{Lineage maps of enriched Gene Ongology (GO) terms for components NMF-1-of-3 and NMF-2-of-3.  (NMF-3-of-3 produced no significant enrichment results)}
+\caption{Lineage maps of enriched Gene Ongology (GO) terms for components NMF-1-of-3 and NMF-2-of-3.
+In these diagrams, enriched terms are coloured purple, while there ancestors in the ontology are yellow.
+(NMF-3-of-3 produced no significant enrichment results)}
 \label{fig-NMF-go-lineage}
@@
 If a metasample component has a robust correlation with survival, then we expect the p-value for all three sets of results to show significance \emph{and} for the effects to have the same sense -- be in the same direction.
-None of the 11 components pass this test, thus we have failed to demonstrate a robust correlation between survival and any of the 11 metasamples.
+None of the 11 components pass this test.
+
+The component closest to correlation with survival would seem to be PCA 1-of-3, showing a reasonably consistent hazard ratio of around 1.3 ($2^{0.4}$) across the three experiments.
+This has $p < 0.05$ in the larger TCGA dataset, but not in the smaller AOCS dataset.
 
@@
 \begin{description}
-\item[WGD]:
+\item[WGD]: Whole genome doubling -- a marker of genome instability.
 \item[Cellularity]: proportion of cells belonging to the tumour (as opposed to surrounding normal tissue).
-\item[HR Detect]: score from homologous repair (HR) analysis, deriving from Ewing \etal's work \cite{Ewing2020}
-\item[Mutational Load]:
-\item[CNV Load]: Copy-number variation ...
-\item[Donor survival time]: (should probably remove).
+\item[HR Detect]: A predictor of homologous repair deficiency based on established mutational features. \cite{Ewing2020}
+\item[Mutational Load]: A measure of the total number of mutations present in the tumour genome.
+\item[CNV Load]: Copy-number variation, i.e. deviation the the normal diploid cell compliment.
+% \item[Donor survival time]: (should probably remove).
 \end{description}
@@
 \section{Discussion}
+
+One valuable lesson from this work is the importance of accounting for sampling error, by bootstrapping,  when considering the stability of factorizations and thence the selection of factorization rank.
+Many authors comment on the variability of NMF and ICA factorization due to algorithm initialisation, but neglect sampling error which we have shown to be far more significant.
+These considerations lead to the selection of much lower ranks (fewer metagenes) than would otherwise be the case.
+
+Considering the biological significance of each of the 11 extracted metagenes:
+\begin{description}
+%
+\item[NMF 1-of-3]: this is related to the cellular structures and processes of the extra-cellular matrix (ECM), that is the proteins such as collagens which mediate the three-dimensional organisation of cells in a tissues.
+ECM molecular composition will vary substantially between tissue types, but is also known to play a part in many disease processes \cite{Theocharis2019}.
+Thus, this metagene may simply reflect heterogeneity of tissues in the biopsied sample, or may have some deeper disease related significance.
+%
+\item[NMF 2-of-3]: here we see enrichment of genes relating to the ribosomal subunit and the processes of RNA binding, implying perhaps a link with assembly of the ribosomal RNA-protein complex.
+Ribosomes are known to have a role in carcinogenesis, by dysregulation the RNA $\rightarrow$ translation or mutations in ribosomal subunits impacting on cellular metabolism \cite{Sulima2017}.
+%
+\item[NMF 3-of-3]: no significant biological enrichment found.
+%
+\item[ICA 1-of-5 and ICA 2-of-5]: no significant biological enrichment found.
+%
+\item[ICA 3-of-5]: this component relates to processes of multicellular / extracellular organisation, and the ECM.  It is therefore similar to NMF 1-of-3 above.
+\item[ICA 4-of-5]: this component is seen to relate to the ribosomal subunit (as NMF 2-of-3), additionally membrane proteins in the respiratory complex, mitochondrial and the NADH dehydrogenase complex.
+Processes of organonitrogen biosynthesis are also highlighted.
+%
+\item[ICA 5-of-5]: processes relating to regulation of immune response are enriched in this component, featuring genes from the major histocompatibility complex group, in particular the HLA- genes, allowing the immune system to recognise self from non-self.
+GO terms relating to the endoplasmic reticulum (ER) are also highlighted.
+It may be that this component is mainly sensitive to the immunohistochemical signature of patients, and therefore not of clinical interest.
+%
+\item[PCA 1-of-3]: there is mention of ECM related terms, as for ICA 3-of-5 and NMF 1-of-3.
+However, there are also terms relating to the regulation of angiogenesis.
+It is known that tumours have a need for increased blood supply, and that expression level of factors proting angigenesis are associated with aggresiveness of tumour growth \cite{Nishida2006}.
+Two distinct subtypes of ovarian cancer have been identified, defined on the expression of angiogenesis related genes.
+Further, these subtypes have been found to inform clinical outcome \cite{Glass2015}.
+Thus, this component might contain useful prognostic value.
+%
+\item[PCA 2-of-3]: no significant biological enrichment found.
+%
+\item[PCA 3-of-3]: this component has some similarity with ICA 5-of-5, in that refers to regulation of immune processes (HLA- genes) and ER membrane.
+However, chemotaxis (cell movement) is also highlighted.
+\end{description}
+
+We have noted above that both NMF 1-of-3 and ICA 3-of-5 may be related to tissue type heterogeneity.
+Interestingly, we have also earlier noted that heatmap analysis shows these two metagenes to result in closely correlated metasamples in both the TCGA and OACS datasets.
+Further, we also observed that these components correlate negatively with cellularity.
+This is as one would expect: low cellularity implies a mixture of tumour and non-tumour cell types, resulting in variation of ECM composition.
+That these two components highlight similar terms should be no surprise -- they have a high Jaccard similarity of 0.39.
+
+Interestingly, the above summary of the gene enrichment results tentatively suggests component PCA 1-of-3 to be most likely to have clinical prognostic value, based on a cursory look at the literature.
+This perhaps supports an earlier observation that, while failing to show consistent significance, component PCA-1-of-3 comes closest to demonstrating correlation with patient survival (section \ref{sec-survival-analysis}).
+
+How should we interpret those four components which demonstrate no significant enrichment with biological meaning?
+
 \begin{itemize}
+\item Unsupervised approach
 \item Importance of taking account of sampling error.
 \item Little correlation with survival time - not surprising.
+\item Each method has at least one component with no significantly enriched genes for biological impact - could these be technical dimensions?
+\item Lookup most extreme two genes for each component.
 \end{itemize}

### 2020-07-30 MSc_Dissertation_B156476.tex
Added explanation of NMF, ICA and PCA; diagram from Sompairac

@@
 \section{Introduction}
-\emph{Say something about ovarian cancer!   An image perhaps.}
+High grade serous ovarian cancer (HGSOC) effects the epithelium of the ovaries and fallopian tube.
+``Serous" refers to the epithelial membrane which secretes serous (serum) fluid.
+It is the 7th most common cancer in women worldwide .
+Epithelial ovarian cancers are classified into two broad subtypes:  Type I are low grade with generally good prognosis; type II are more aggressive and typically carry P53 mutations with defects in mechanisms of DNA repair.
+Type II is more-or-less synonomous with HGSOC.
+\cite{Lisio2019}.
 
-\subsection{Matrix factorization methods}
+This work is concerned with analysing the patters gene expression in HGSOC dimensionality reduction techniques of matrix factorization, and relating these patterns to clinical and genomic features.
+
+\subsection{Matrix factorization}
 \label{sec-matrix-factorization-intro}
@@
 
-There are three matrix factorization methods of interest to this project:
-\begin{enumerate}
-\item Independent component analysis (ICA)
-\item Non-negative matrix factorization (NMF)
-\item Principal component analysis (PCA)
-\end{enumerate}
+The three matrix factorization methods of interest to this project are listed below.
+All are forms of dimensionality reduction, but have differing criteria they aim to optimise.
+The number of target dimensions is referred to as the rank, $k$.
+\begin{description}
+\item [Non-negative matrix factorization (NMF).]  
+As the name implies, this is only applicable to matrices with +ve or zero elements, and finds components which themselves +ve or zero.
+This property make the resulting factorization more interpretable since the components are strictly additive.
+It also allows for input samples to be directly assigned to one of $k$ clusters.
+A good description of NMF and its applications in bioinformatics can be found in \cite{Brunet2004}.
+%
+\item [Independent component analysis (ICA).] The goal of ICA is to represent the given matrix with components which are statistically independent of each other.
+It was originally proposed to solve the blind source separation problem \cite{Comon1994}.
+If the input matrix of samples were drawn from a multivariate Gaussian distribution, then the result would be no different to that of PCA.
+Where the data is non-Gaussian however, ICA results in components which separate out independent sources of variation.
+See \cite{Sompairac2019} for a full explanation, from which figure \ref{fig-pca-ica-nmp-explanation} is taken.
+\item [Principal component analysis (PCA).]  PCA is a well known technique for dimensionality reduction, based on eigenvalue decomposition.
+It is the ideal, efficient solution for multivariate Gaussian distributed data.
+\end{description}
+
+The contrasting results of the three methods is illustrated in figure.
+
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.4, draft=false]{Sompairac_ICA_NMF_PCA.png}
+\end{center}
+\caption{Illustrating the differing results of component extraction by PCA, ICA and NMF.  From \cite{Sompairac2019}}
+\label{fig-pca-ica-nmp-explanation}
+\end{figure}
 
@@
 
-The number of factors, $K$, to extract is a key decision in any matrix factorization approach.
+The number of factors, or rank $K$, to extract is a key decision in any matrix factorization approach.
 In this regard, PCA differs from ICA and NMF.
 In PCA it is reasonable to extract all factors, setting $K=N$, the factors being ranked by the associated eigenvalue which also ranks the proportion of variance explained.
-That approach is not reasonable for ICA and NMF, since differing sets of factors will be obtained for different $K$, and for a given choice of $K$ all factors are equally important -- they do not rank \cite{Stein-OBrien2018}.
+However, that approach is not valid for ICA and NMF, since differing sets of factors will be obtained for different $K$, and for a given choice of $K$ all factors are equally important -- they do not rank \cite{Stein-OBrien2018}.
 
 The problem of determining the optimal $K$ for a given expression matrix is addressed  by Kairov \etal\ \cite{Kairov2017}, based on optimizing the \emph{stability} of the components over multiple runs.
-
-\subsection{Ovarian cancer sub-type prediction}
-
-A recent pre-publication develops a pipeline for sub-type prediction which the potential to be used clinically \cite{Talhouk2020}.
-The NanoString gene expression platform is used. Approach is to focus on small set of 513 genes (vs 20,000 protein coding genes) known have relevance to subtyping, as identified from literature.
-
-Two approaches were followed by different teams.
-``All Array": used expression array data from 1650 patients across 14 studies evaluating 9 supervised learning algorithms by bootstrap, selecting an AdaBoost -like method.
-``TCGA" used 434 patients from TCGA evaluating 5 algorithms by cross-validation selecting a random forest.
-
-"All Array" had the advantage of more data but needed attention to batch effects.
-A final classifier took a consensus of the two methods based on a minimal gene set (order 40 genes), validated by a leave-one-out (patient level) approach.
-
-Survival analysis was carried out stratified by predicted subtype.
-
-\begin{figure}
-\begin{center}
-\includegraphics[scale=0.35]{talhouk_figure_1_part.png}
-\end{center}
-\caption{HGSOC subtype prediction by expression analysis of a focussed gene set and concensus across two independent pipelines.  From \cite{Talhouk2020}.}
-\label{fig-talhouk_figure_1_part.png}
-\end{figure}
 
@@
 They find the following to be significantly related to platinum resistance: KLRC1, TCF7, CD274, HSD11B1, COLEC12, PDGFC, FCF1, BMI1, TNFRSF9, ATG10, EWSR1.
+
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.35]{talhouk_figure_1_part.png}
+\end{center}
+\caption{HGSOC subtype prediction by expression analysis of a focussed gene set and concensus across two independent pipelines.  From \cite{Talhouk2020}.}
+\label{fig-talhouk_figure_1_part.png}
+\end{figure}
+
+A recent pre-publication develops a pipeline for sub-type prediction which the potential to be used clinically \cite{Talhouk2020}.
+The NanoString gene expression platform is used. Approach is to focus on small set of 513 genes (vs 20,000 protein coding genes) known have relevance to subtyping, as identified from literature.
+
+Two approaches were followed by different teams.
+``All Array": used expression array data from 1650 patients across 14 studies evaluating 9 supervised learning algorithms by bootstrap, selecting an AdaBoost -like method.
+``TCGA" used 434 patients from TCGA evaluating 5 algorithms by cross-validation selecting a random forest.
+
+"All Array" had the advantage of more data but needed attention to batch effects.
+A final classifier took a consensus of the two methods based on a minimal gene set (order 40 genes), validated by a leave-one-out (patient level) approach.
+
+Survival analysis was carried out stratified by predicted subtype.
 
@@
 Thus in our use case PCA is seen to have stochastic behaviour.
-
 \end{description}
@@
 \label{sec-metagene-selection}
-As outlined in the discussion, deciding on the number of components (metagenes) to extract -- that is the factorization rank -- is key.
+As outlined in the discussion, deciding on the number of components (factors, metagenes) to extract -- that is the factorization rank -- is key.
 Taking more components results in more accurate representation of the observed expression matrix and provides more avenues to explore the underlying biology.
@@
 The overall process is summarised as pseudocode in figure \ref{fig-clustering-psuedocode}.
+
+A complication arises in the computation of ICA and PCA facorizations, in that essentially the same factor can arise as $w$ or $-w$ vectors.
+These would appear as separate cluster in repeated sampling, and confound attempts to collect and aggregate.
+The solution adopted here is to normalise each factor by requiring that the most extreme value -- i.e.
+having the greatest absolute value -- is positive, the whole factor being negated if this is not the case.
+This is arguably over simplistic, but seems to be effective in practice.
 

### 2020-07-30 MSc_Dissertation_B156476.tex
Further work section, edits throughout, proof reading

@@
 
-\documentclass[draft, tikz, 12pt,a4paper,oneside,fleqn]{article}
+\documentclass[tikz, 12pt,a4paper,oneside,fleqn]{article}
 \usepackage[margin=1.5cm]{geometry}
@@
 
-\title{Phylogenetic study of Asgard Archaea}
+\title{Disentangling patterns of gene expression in high grade serous ovarian cancer}
 \author{}
@@
 
+\tableofcontents
+
+\newpage
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%% Start of report for real %%%%%%%%%%%%%%%%%%%
@@
 \section{Introduction}
-High grade serous ovarian cancer (HGSOC) effects the epithelium of the ovaries and fallopian tube.
-``Serous" refers to the epithelial membrane which secretes serous (serum) fluid.
+High grade serous ovarian cancer (HGSOC) effects the epithelium of the ovaries and fallopian tubes.
+``Serous" refers to the epithelial membrane which secretes serous fluid (serum).
 It is the 7th most common cancer in women worldwide .
@@
 
-This work is concerned with analysing the patters gene expression in HGSOC dimensionality reduction techniques of matrix factorization, and relating these patterns to clinical and genomic features.
+This work is concerned with analysing the patters gene expression in HGSOC by dimensionality reduction techniques of matrix factorization, and relating these patterns to clinical and genomic features.
 
@@
 
-The three matrix factorization methods of interest to this project are listed below.
+The three matrix factorization methods of interest to this work are listed below.
 All are forms of dimensionality reduction, but have differing criteria they aim to optimise.
-The number of target dimensions is referred to as the rank, $k$.
+The number of target dimensions (or components, factors, metagenes) is referred to as the rank, $k$.
 \begin{description}
 \item [Non-negative matrix factorization (NMF).]  
-As the name implies, this is only applicable to matrices with +ve or zero elements, and finds components which themselves +ve or zero.
+As the name implies, this is only applicable to matrices with +ve or zero elements, and finds components which are themselves +ve or zero.
 This property make the resulting factorization more interpretable since the components are strictly additive.
@@
 \item [Independent component analysis (ICA).] The goal of ICA is to represent the given matrix with components which are statistically independent of each other.
-It was originally proposed to solve the blind source separation problem \cite{Comon1994}.
+It was originally proposed to solve the blind source separation problems \cite{Comon1994}.
 If the input matrix of samples were drawn from a multivariate Gaussian distribution, then the result would be no different to that of PCA.
@@
 
-The contrasting results of the three methods is illustrated in figure.
+The contrasting results of the three methods is illustrated in figure \ref{fig-pca-ica-nmp-explanation}.
 
@@
 NMF generally uses $W$ and $H$ respectively while ICA generally uses $S$ and $A$.
-PCA is tends to be differently formulated, but $W$ and $T$ sometimes used.
+PCA tends to be differently formulated, but $W$ and $T$ sometimes used.
 The original matrix to be factorized is variously named $X$ or $V$.
 
-In this work we wish to treat and discuss the three factorization methods in a unified way.  Thus, for the remainder of this dissertation we will adopt the following conventions:
+In this work we wish to treat and discuss the three factorization methods in a unified way.  Thus, for the remainder of this dissertation we adopt the following conventions:
 \begin{description}
@@
 
-However, in the case of NMF and ICA, the $W$ and $H$ matrices cannot be trivially exchanged and transposed.
+In the case of NMF and ICA, the $W$ and $H$ matrices cannot be trivially exchanged and transposed.
 This is because the optimisations (sparsity and independence respectively) which define these algorithms are focussed on the $W$ matrix.
@@
 There seems to be substantial confusion an lack of clarity in the way that matrix factorization, particularly ICA, is applied in transcriptomics research.  
-``Surprisingly, both ways of applying ICA to omics data are wide-spread, and sometimes it makes an effort to figure out in which way ICA was applied." \cite{Sompairac2019a}, and ``Different protocols to apply ICA to transcriptomic data exist and currently no single standard approach has been defined.
+``Surprisingly, both ways of applying ICA to omics data are wide-spread, and sometimes it makes an effort to figure out in which way ICA was applied." \cite{Sompairac2019}, and ``Different protocols to apply ICA to transcriptomic data exist and currently no single standard approach has been defined.
 The main difference in the existing approaches consists in what is considered as source signal matrix in the decomposition" \cite{Cantini2019}.
@@
 
-The problem of determining the optimal $K$ for a given expression matrix is addressed  by Kairov \etal\ \cite{Kairov2017}, based on optimizing the \emph{stability} of the components over multiple runs.
+The problem of determining the optimal $K$ for a given expression matrix is addressed  by Kairov \etal\ \cite{Kairov2017}, based on optimizing the \emph{stability} of the components over multiple algorithm initializations.
 
@@
 They find that expression of HS11B1, DNBT1, CKLF, NUP107.
-CCL18, LY96, ATG7, SLAMF7, CXCL9 is associated with long survival, while IKBKG and SDHA with short survival.
+CCL18, LY96, ATG7, SLAMF7, CXCL9 is associated with better survival, while IKBKG and SDHA associate with with poor survival.
 
@@
 These markers allow stratification into categories including Mesenchymal-to-epithelial transition, Homologous recombination repair and Epithelial-to-mesenchymal transition.
-This, they argue, will allow selection of class-specific inhibitor drugs such as Olaparib and Rucaparib.
+This, they argue, allows selection of class-specific inhibitor drugs such as Olaparib and Rucaparib.
 
@@
 
-A recent pre-publication develops a pipeline for sub-type prediction which the potential to be used clinically \cite{Talhouk2020}.
-The NanoString gene expression platform is used. Approach is to focus on small set of 513 genes (vs 20,000 protein coding genes) known have relevance to subtyping, as identified from literature.
-
+A recent pre-publication develops a pipeline for sub-type prediction with the potential to be used clinically \cite{Talhouk2020}.
+The NanoString gene expression platform is used.
+See figure \ref{fig-talhouk_figure_1_part.png}, their strategy was to focus on small set of 513 genes (vs $\sim$20,000 protein coding genes) known from the literature to have relevance to subtyping.
 Two approaches were followed by different teams.
@@
 ``TCGA" used 434 patients from TCGA evaluating 5 algorithms by cross-validation selecting a random forest.
-
 "All Array" had the advantage of more data but needed attention to batch effects.
 A final classifier took a consensus of the two methods based on a minimal gene set (order 40 genes), validated by a leave-one-out (patient level) approach.
-
 Survival analysis was carried out stratified by predicted subtype.
@@
 Our overall approach is to use \emph{unsupervised} machine learning methods to represent gene expression data in a small number of features, then to study whether these features correlate with clinical and biological variables.
-This is appropriate in our case, since of the two datasets available (described below), one (TCGA) has relatively large n but little available metadata, while the other (AOCS) is much smaller but has useful metadata.
+This is appropriate in our case, since of the two datasets available (described below), one (TCGA) has relatively large $n$ but little available metadata, while the other (AOCS) is much smaller but has useful metadata.
 This motivates an approach of unsupervised learning on the TCGA dataset.
@@
 \begin{enumerate}
-\item TCGA, n=274, g=19,601, with metadata on survival
-\item AOCS, n=80, g=19,730, with metadata on survival, cellularity and additional genomic features
+\item The Cancer Genome Atlas (TCGA) derived, n=274, g=19,601, with metadata on survival
+\item Austrailian Ovarian Cancer Study (AOCS) \cite{Patch2015}, n=80, g=19,730, with metadata on survival, cellularity and additional genomic features.
 \end{enumerate}
+TODO: \emph{more explanation of the provenance of these datasets}.
 
 \subsubsection{Method outline}
-An overview of the methodology adopted in this work can be seen in figure \ref{fig-method-overview}.
+An overview of the methodology adopted in this work is shown in figure \ref{fig-method-overview}.
 
@@
 \end{center}
-\caption{Overview of methodology as information flow}
+\caption{Overview of methodology as information flow.
+The diagramatic convention here is that at each stage, both datasets -- TCGA and AOCS -- are processes separately by the given algorithm.
+The exception is Gene Set Intersection, which involves interaction between the datasets.}
 \label{fig-method-overview}
@@
 \item
-Identification of a consistent gene list between the TCGA and AOCS datasets.
+Identification of a consistent gene set between the TCGA and AOCS datasets.
 \item
@@
 \item Matplotlib for general plotting
-\item Seaborn for clustered heatmaps
+\item Seaborn for heatmaps
 \item Scikit-learn for matrix factorization and k-means clustering
@@
 
-A full list of software and libraries used, with versions, is given in appendix \ref{sec-software-versions}
+A full list of software and libraries used, with versions, is given in appendix \ref{sec-software-versions}.
 
@@
 In order to allow factorizations found in the TCGA dataset to be applied to the AOCS dataset it is necessary (or at least convenient) to synchronize the set of genes over which the expression matrices are defined.
-As provided to this project, the TCGA and AOCS datasets cover 19,610 and 19,730 protein coding genes respectively, 19,566 or which common to both (according to ENSG encodings).
+As provided to this project, the TCGA and AOCS datasets cover 19,610 and 19,730 protein coding genes respectively, 19,566 of which are common to both (according to ENSG encodings).
 Thus, both datasets were pruned to the 19,566 intersection set and ordered consistently.
@@
 \label{sec-metagene-selection}
-As outlined in the discussion, deciding on the number of components (factors, metagenes) to extract -- that is the factorization rank -- is key.
+Deciding on the number of components (factors, metagenes) to extract -- that is the factorization rank -- is key.
 Taking more components results in more accurate representation of the observed expression matrix and provides more avenues to explore the underlying biology.
@@
 A second and more fundamental source of variation of relevance to all three factorization methods is \emph{sampling error}.
-Our factorizations are based on a small (N=80 or N=374) sample of patients drawn from the population of HGSOC patients; our particular datasets are just two examples of many different `draws' which could have been made from that population.
+Our factorizations are based on a small ($N=80$ or $N=374$) sample of patients drawn from the population of HGSOC patients; our particular datasets are just two examples of many different `draws' which could have been made from that population.
 
@@
 
-A complication arises in the computation of ICA and PCA facorizations, in that essentially the same factor can arise as $w$ or $-w$ vectors.
+A complication arises in the computation of ICA and PCA facorizations, in that essentially the same factor can arise as $w$ or $-w$ vectors -- i.e.
+$180^\circ$ rotated.
 These would appear as separate cluster in repeated sampling, and confound attempts to collect and aggregate.
-The solution adopted here is to normalise each factor by requiring that the most extreme value -- i.e.
+The solution adopted here is to normalise each factor by requiring that the most extreme element -- i.e.
 having the greatest absolute value -- is positive, the whole factor being negated if this is not the case.
@@
 In each case, 50 iterations of factorization are performed on bootstrap samples, generating $50 k$ instances of 19,566 dimensioned metagenes.
-Two dimensional t-SNE plots \ref{tsne-reference} were generated for visualization.
-If sampling and algorithm initialisation error are modest then we expect to see $k$ clusters of points.
+Two dimensional t-SNE plots were generated for visualization purposes.
+If sampling and algorithm initialisation error are modest then we expect to see $k$ tight clusters of points.
 To avoid lengthy computation in the t-SNE clustering, the metagenes were first reduced to $r=20$ dimensions by PCA; brief experiments showed that this reduction had negligible effect on the t-SNE visualisation for $r \geq 10$.
@@
 
-The t-SNE plots and silhouette scores were assessed visually (see figures \ref{fig-first-cluster-result} to \ref{fig-last-custer-result} to decide, for each of the three factorizers, the highest rank with good cluster coherence, those median metagenes to take forward to gene enrichment and survival analysis.
+The t-SNE plots and silhouette scores were assessed visually (see figures \ref{fig-first-cluster-result} to \ref{fig-last-cluster-result}) to decide, for each of the three factorizers, the highest rank with good cluster coherence, those median metagenes to take forward to gene enrichment and survival analysis.
 Note that the t-SNE plots are used only for visualisation; they are not involved in the computation of median metagenes or silhouette score.
@@
 The selection rationale is set out in the results section, but is convenient to state here that the following ranks were selected:
-$K_{\mbox{\tiny NMF}} = 3, K_{\mbox{\tiny ICA}} = 5, K_{\mbox{\tiny PCA}} =3$.  Thus, there were $3 + 5 + 3 = 11$ metagenes taken forward for follow on analysis.
+$K_{\mbox{\tiny NMF}} = 3, K_{\mbox{\tiny ICA}} = 5, K_{\mbox{\tiny PCA}} =3$.  Thus, there were $3 + 5 + 3 = 11$ metagenes taken forward for follow-on analysis.
 
@@
 
-The metagenes extracted by the above described factorization process provide valuable information into which genes varying in expression in the study samples; thus in our case, which genes are influential in HGSOC.
+The metagenes extracted by the above described factorization and clustering process provide valuable information into which genes vary in expression in the study samples; thus in our case, which genes are influential in HGSOC.
+In order to gain insights into what biological processes are involved, gene enrichment analysis against the Gene Ontology (GO) was carried out for each of the 11 metagenes individually.
 
-In order to gain insights into what biological processes are involved, gene enrichment analysis against the Gene Ontology (GO) was carried out for each of the 11 metagenes individually, as follows.
-
-The essence of gene enrichment analysis is to compare a candidate set of genes with many functional gene sets, asking the question "are there significantly more (or less, i.e.
+The essence of gene enrichment analysis is to compare a candidate set of genes with many functional gene sets, asking the question ``are there significantly more (or less, i.e.
 depletion) intersecting genes than would be expected for the same number of genes being drawn at random (without replacement) from the total population of genes in the study.
@@
 
-The result of this analysis, per metagene, is a list of enriched (or depleted) GO terms, each with an associated list of involved genes and a FDR significance level.
-This list can be inspected directly for insights, but that misses the point of the GO, which is to organise the terms hierarchically by 'is a' relationships.
+The result of this analysis, per metagene, is a list of enriched (or depleted, however in our case no depleted terms were found) GO terms, each with an associated list of involved genes and a FDR significance level.
+This list can be inspected directly for insights, but that misses the point of the GO, which is to organise the terms hierarchically by `is a' relationships.
 Thus, the enriched terms were rendered graphically to show them in the context of their parent terms.
@@
 It is fundamental to our approach that metagenes determined on the basis of one cohort of data can be used to generate metasamples for a different cohort.
-In the exposition below we refer to these cohorts as the \emph{training} and \emph{validation} cohorts respectively.
+In the exposition below we refer to these as the \emph{training} and \emph{validation} cohorts respectively.
 
@@
 However, if patients in datasets $\trainset$ and $\validset$ are drawn from the same population (ovarian cancer patients), then $X_\trainset$ and $X_\validset$ can be expected to have similar distributions w.r.t to their columns, and thus $W_\trainset$ and $W_\validset$ can be expected to be equivalent within sampling error.
-This only makes sense if the two expression matrices $X_\trainset$ and $X_\validset$ are defined over the \emph{same list of genes}, so that $g_\trainset = g_\validset = g$, in which case $W_\trainset$ and $W_\validset$ have the same shape of $(g, k)$.
+This only makes sense if the two expression matrices $X_\trainset$ and $X_\validset$ are defined over the \emph{same set of genes}, so that $g_\trainset = g_\validset = g$, in which case $W_\trainset$ and $W_\validset$ have the same shape of $(g, k)$.
 
@@
 
-For ICA and PCA we use ordinary least square regression, formulated above by dropping the $H\geq 0$ constraint, using {\tt scipy.linalg.lstsq}.
+For ICA and PCA we use ordinary least square regression, formulated a above but dropping the $H\geq 0$ constraint, using {\tt scipy.linalg.lstsq}.
 
-An internal check on the validity of the statistical assumptions can be made my substituting  $H_\validset$ from (\ref{eq_Hv_WdtXv}) into (\ref{eq_Xv_WtHv}) and assessing the accuracy of the approximation by RMS difference.
+%An internal check on the validity of the statistical assumptions can be made my substituting  $H_\validset$ from (\ref{eq_Hv_WdtXv}) into (\ref{eq_Xv_WtHv}) and assessing the accuracy of the approximation by RMS difference.
 
-The end result of the analysis is the $H_\validset$ matrix of shape $(k, n_\validset)$, thus delivering for dataset in our $\validset$ a feature vector of length $k$ -- a small number typically less than a dozen.
+The end result of the analysis is the $H_\validset$ matrix of shape $(k, n_\validset)$, thus delivering for each dataset in our $\validset$ a feature vector -- or metasample -- of length $k$.
 
@@
 \subsection{Survival analysis}
+\label{sec-survival-analysis}
 
-Survival analysis was performed to investigate whether the selected metagene factors correlate with patient survival.
+Survival analysis was performed to investigate whether the metasamples derived from the selected metagenes correlate with patient survival.
 Overall  survival (OS) data is available for both the TCGA and AOCS datsets.
 Additionally, progression free survival (PFS) data is available for AOCS.
-As discussed in section \ref{sec-clustering-method}, metagenes were obtained by factorization on the TCGA dataset only.
+As described above,  metagenes were obtained by factorization on the TCGA dataset only.
 Application to TCGA (for OS) thus represents an in-sample test, while application to AOCS (for OS and PFS) is a more exacting out-of-sample test.
@@
 Metasample values were binariased to 0, 1 by thresholding at the median value.
-Kaplan-Meier plots were made in respect of derived metasample for each of the three analysis -- making $11 \times 3$ plots each with two survival curves (with 95\% confidence intervals).
+Kaplan-Meier plots were made in respect of derived metasample for each of the three analyses -- making $11 \times 3$ plots each with two survival curves and showing with 95\% confidence intervals.
 A hazard ratio (HR)
@@
 Good software engineering practice has been followed, with an object-oriented design, frequent commits and unit testing.
-Unit tests are based on tiny expression matrices, just 100 genes by 10 patients.
-The whole test suite of over 60 tests executes in around 30 seconds.
-These tests are particularly valuable when re-factoring code by providing a rapid means of detecting errors.
+Unit tests are based on tiny expression matrices, just 100 genes by 10 patients, thus the whole test suite of over 60 tests executes in around 30 seconds.
+These tests are particularly valuable when re-factoring code by providing a rapid means of detecting coding errors.
+The total codebase is approximately 2,800 lines of Python.
 All plots are generated in vector graphics pdf format to ensure smooth scaling to any resolution.
@@
 
-From figure \ref{fig-AOCS-ica-fixed-vs-bootstrap} it can be seen that demonstrates that with sampling error excluded, clusters appear reasonably coherent, but when sampling error is modelled by bootstrap sampling then we see the factorizations become much less stable.
-This demonstrates that \emph{sampling} error is far greater then errors due to algorithm initialisation when the dataset is small (N=80), making metagene extraction highly unstable.
+From figure \ref{fig-AOCS-ica-fixed-vs-bootstrap} it can be seen that with sampling error excluded, clusters appear reasonably coherent, but when sampling error is modelled by bootstrap sampling then  the factorizations become much less stable.
+This demonstrates that \emph{sampling} error is far greater than errors due to algorithm initialisation when the dataset is small (N=80), making metagene extraction unreliable.
 Figure  \ref{fig-TCGA-ica-fixed-vs-bootstrap} makes the same comparison for the lager N=374 TCGA dataset, in which it can be seen that the impact of sampling error is not so severe.
@@
 
-For this reason, and in line with our overall strategy, it is appropriate to perform all metagene extraction on the larger TCGA dataset, in the expectation that the obtained metagenes will be more robust an likely to generalise better to other datasets.
+For this reason, and in line with our overall strategy, it is appropriate to perform all metagene extraction on the larger TCGA dataset, in the expectation that the obtained metagenes will be more robust and likely to better generalise to other datasets.
 
@@
 There is a curious artefact visible in the NMF factorization clusterings of figure \ref{fig-TCGA-all-bootstrap}, top row.
-For $k=2, 3 and 5$ several of the clusters show a bi-modal character.
+For $k=2, 3$ and $5$ several of the clusters show a bi-modal character.
 This is not observed in the fixed sampling case (not shown).
 The artefact is difficult to explain.
+It cannot be the $180^\circ$ rotation issue discussed earlier, since this does not apply to the all +ve components.
 
@@
 
-Gene enrichment analysis against the GO results in, each of the 11 metagenes, a table of GO terms with a list of the candidate genes which have an annotated association to that term.   
+Gene enrichment analysis against the GO results in, for each of the 11 metagenes, a table of GO terms with a list of the candidate genes which have an annotated association to that term.   
 To take full advantage of the hierarchical nature of the GO, results are presented as lineage maps in figures \ref{fig-NMF-go-lineage} to \ref{fig-PCA-1-3-go-lineage}.  
@@
 \begin{center}
-\includegraphics[width=\textwidth]{\plotspath{TCGA_OV_VST/GeneEnrichment/go_lineage_1_of_PCA_3.png}} \\
+\includegraphics[width=0.85 \textwidth]{\plotspath{TCGA_OV_VST/GeneEnrichment/go_lineage_1_of_PCA_3.png}} \\
 \begin{spacing}{0.6}
@@
 \vspace{1cm}
-\includegraphics[width=\textwidth]{\plotspath{TCGA_OV_VST/GeneEnrichment/go_lineage_3_of_PCA_3.png}} \\
+\includegraphics[width=0.85 \textwidth]{\plotspath{TCGA_OV_VST/GeneEnrichment/go_lineage_3_of_PCA_3.png}} \\
 \begin{spacing}{0.6}
@@
 
-All of these results are summarised in figure \ref{fig-summary-survival-plots}, which brings together the three sets of results -- TCGA (OS), OACS OS) and OACS (PFS) with respect to the 11 metasample components.
-$\log_2{HR}$ is used in order that the sense of the survival impact can be readily appreciated.
+All of these results are summarised in figure \ref{fig-summary-survival-plots}, which brings together the three sets of results -- TCGA (OS), OACS (OS) and OACS (PFS) --  with respect to the 11 metasample components.
+$\log_2{\mbox{HR}}$ is used in order that the \emph{sense} of the survival impact can be readily appreciated.
 If a metasample component has a robust correlation with survival, then we expect the p-value for all three sets of results to show significance \emph{and} for the effects to have the same sense -- be in the same direction.
@@
 
-As with survival analysis, metasamples were derived from the metagenes factorized from the TCGA dataset, then applied to both TCGA and AOCS datasets.
+As with survival analysis, metasamples were derived from the metagenes factorized from the TCGA dataset, then applied to both TCGA and AOCS.
 The resulting heatmaps are shown in figures \ref{fig-heatmap-TCGA-TCGA} and \ref{fig-heatmap-TCGA-OACS} respectively.
@@
 \item[Cellularity]: proportion of cells belonging to the tumour (as opposed to surrounding normal tissue).
-\item[HR Detect]: A predictor of homologous repair deficiency based on established mutational features. \cite{Ewing2020}
+\item[HR Detect]: A predictor of homologous repair deficiency based on established mutational features \cite{Ewing2020}.
 \item[Mutational Load]: A measure of the total number of mutations present in the tumour genome.
-\item[CNV Load]: Copy-number variation, i.e. deviation the the normal diploid cell compliment.
+\item[CNV Load]: Copy-number variation, i.e. deviation from the normal diploid cell compliment.
 % \item[Donor survival time]: (should probably remove).
@@
 
-The dendrogram clustering of metasamples (columns, top of figures) gives insights into the similarity of each metasample component, while the clustering of patients (rows, left of figures) offers insights into sub-populations of patients and how (in the case of AOCS only) how these relate to metadata variables such as cellularity and mutational load.
+The dendrogram clustering of metasamples (columns, top of figures) gives insights into the similarity of each metasample component, while the clustering of patients (rows, left of figures) offers insights into sub-populations of patients and how (in the case of AOCS only) these relate to metadata variables such as cellularity and mutational load.
 These topics will be taken up in the following discussion section.
@@
 One valuable lesson from this work is the importance of accounting for sampling error, by bootstrapping,  when considering the stability of factorizations and thence the selection of factorization rank.
-Many authors comment on the variability of NMF and ICA factorization due to algorithm initialisation, but neglect sampling error which we have shown to be far more significant.
+Sevaral authors comment on the variability of NMF and ICA factorization due to algorithm initialisation \cite{Kairov2017,Sompairac2019,WayGregory2019}, but most neglect sampling error which we have shown to be far more significant.
+Cantini \etal\ \cite{Cantini2019a} mention bootstrapping, in reference to the BIODICA package (for ICA factorization), but it is unclear whether this is applied to the input data to model sampling error.
+In the initial stages of the current work, BIODICA was used, and was found to recommend ranks of 14, 19 or 30.
+Yet we have seen that such high ranks on the small n=80 dataset leads to highly unstable factorizations when bootstrap modelling of sampling error is included.
+
 These considerations lead to the selection of much lower ranks (fewer metagenes) than would otherwise be the case.
@@
 \item[NMF 2-of-3]: here we see enrichment of genes relating to the ribosomal subunit and the processes of RNA binding, implying perhaps a link with assembly of the ribosomal RNA-protein complex.
-Ribosomes are known to have a role in carcinogenesis, by dysregulation the RNA $\rightarrow$ translation or mutations in ribosomal subunits impacting on cellular metabolism \cite{Sulima2017}.
+Ribosomes are known to have a role in carcinogenesis, by dysregulation the RNA $\rightarrow$ protein translation, or mutations in ribosomal subunits impacting on cellular metabolism \cite{Sulima2017}.
 %
@@
 However, there are also terms relating to the regulation of angiogenesis.
-It is known that tumours have a need for increased blood supply, and that expression level of factors proting angigenesis are associated with aggresiveness of tumour growth \cite{Nishida2006}.
+It is known that tumours have a need for increased blood supply, and that expression level of factors promoting angigenesis are associated with aggresiveness of tumour growth \cite{Nishida2006}.
 Two distinct subtypes of ovarian cancer have been identified, defined on the expression of angiogenesis related genes.
@@
 %
-\item[PCA 3-of-3]: this component has some similarity with ICA 5-of-5, in that refers to regulation of immune processes (HLA- genes) and ER membrane.
+\item[PCA 3-of-3]: this component has some similarity with ICA 5-of-5, in that it refers to regulation of immune processes (HLA- genes) and ER membrane.
 However, chemotaxis (cell movement) is also highlighted.
@@
 We have noted above that both NMF 1-of-3 and ICA 3-of-5 may be related to tissue type heterogeneity.
-Interestingly, we have also earlier noted that heatmap analysis shows these two metagenes to result in closely correlated metasamples in both the TCGA and OACS datasets.
-Further, we also observed that these components correlate negatively with cellularity.
+We have also seen from metasample heatmap analysis that these two metagenes are associated with closely correlated metasamples in both the TCGA and OACS datasets, and also to have a moderately high Jaccard similarity of 0.39.
+Further, we observed that these components correlate negatively with cellularity.
 This is as one would expect: low cellularity implies a mixture of tumour and non-tumour cell types, resulting in variation of ECM composition.
-That these two components highlight similar terms should be no surprise -- they have a high Jaccard similarity of 0.39.
 
-Interestingly, the above summary of the gene enrichment results tentatively suggests component PCA 1-of-3 to be most likely to have clinical prognostic value, based on a cursory look at the literature.
-This perhaps supports an earlier observation that, while failing to show consistent significance, component PCA-1-of-3 comes closest to demonstrating correlation with patient survival (section \ref{sec-survival-analysis}).
+Interestingly, the above summary of the gene enrichment results tentatively suggests component PCA 1-of-3 is most likely to have clinical prognostic value, based on a cursory look at the literature.
+This perhaps supports an earlier observation that, while not conclusive, component PCA-1-of-3 comes closest to demonstrating correlation with patient survival (section \ref{sec-survival-analysis}, figure \ref{fig-summary-survival-plots}).
 
 How should we interpret those four components which demonstrate no significant enrichment with biological meaning?
+It is notable that all three methods have at least one such component.
+Among these four, PCA 1-of-3 and ICA 1-of-5 have high (the highest) Jaccard similarity of 0.61.
+For all other pairs among these four Jaccard similarity is negligible.
+Perhaps these two components are picking up on similar technical variation between the samples?
 
-\begin{itemize}
-\item Unsupervised approach
-\item Importance of taking account of sampling error.
-\item Little correlation with survival time - not surprising.
-\item Each method has at least one component with no significantly enriched genes for biological impact - could these be technical dimensions?
-\item Lookup most extreme two genes for each component.
-\end{itemize}
+\subsection{Research questions revisited}
+
+TODO (sorry!).
+
+\subsection{Further work}
+
+This work has thrown up several possible lines of further study:
+\begin{enumerate}
+\item
+{\bf Algorithm hyper-parameters}.
+NMF and ICA algorithms have a number of hyper-parameters, only a few of which were explored in this work.
+For example, NMF has parameters to encourage sparsity through L1 regularization.
+This should result in many (most?) metagene elements reducing to zero, and might offer a more robust means of selecting candidate genes to feed into gene enrichment analysis -- replacing the current arbitrary 3 SD from the mean rule.
+\item
+{\bf Discovering batch effects}.
+It has been claimed (e.g.
+\cite{Stein-OBrien2018}) that matrix factorization methods are an effective means of identifying and removing batch effects.
+This could be explored by horizontally concatenating the TCGA and AOCS datasets (after gene set intersection), producing an $n=80+274=354$ dataset with a substantial batch artefact.
+Factorizations of this combined dataset should be more robust because of the larger $n$.
+Does each of NMF, ICA and PCA naturally produce a component which correlates with batch?
+\item
+{\bf Selecting components from several factorization ranks}.
+In the current work, a single rank was selected for each method -- 3, 5 an 3 for NMF, ICA and PCA respectively.
+Unlike PCA, components extracted by NMF and ICA  do not `nest' with rank; that is adding rank in general yields a new set of components.
+Thus one might, for NMF and ICA, perform factorizations at $k=2,3,4,5$, say, yielding $2 \times 15 = 30$ potential components in total.
+Jaccard similarity could be used to identify that subset of components which had the least overlap in detected genes.
+In this way, a larger number of components could be obtained without use of high factorization ranks which we have seen to be unstable.
+\item
+{\bf Cross-dataset factorization stability}.
+We went to considerable effort -- through bootstrap sampling and cluster analysis -- to select factorization ranks which we hoped would be robust and generalise well to other datasets.
+A way to confirm this robustness would be to perform the factorization / clustering pipeline on \emph{both} datasets separately, yielding two sets of metagenes.
+In a perfect world, these metagenes would pair up identically (in-so-far as the two datasets had identical technical characteristics, drawn from identical population of patients).
+The Jaccard similarity heatmap could be used to verify this, and reject components which showed low cross-dataset similarity.
+\item
+{\bf Contrasting gene expression patters between cancers}.
+The current work set out to find patters of gene expression in HGSOC.
+Yet it is hard to say the highlighted patterns are specific to HGSOC, or generic to all cancers.
+One way of teasing out cancer specific patterns might be to take a similar approach to the previous item, but instead of looking for metagenes which are consistent between datasets, find those which are distinct.
+That said, there will likely be many more sophisticated approached in the literature, given the clinical importance of the topic.
+\item
+{\bf Systematic comparison with published gene expression patterns}.
+In analysing the meaning of each metagene above, some tentative links were made with the research literature.
+But this was anecdotal and frankly not particularly scientific; searching on PubMed for GO terms associated with HGSOC and looking at one or two hits!
+A more systematic and ideally automated approach is required.
+The \href{https://www.ncbi.nlm.nih.gov/geoprofiles}{Geo Profiles database} at NCBI, or the \href{https://www.ebi.ac.uk/gxa/home}{Expression Atlas} at EMBL-BI might be possible starting points.
+
+\end{enumerate}
 
 \section{Conclusions}
+TODO
 
-%\bibliography{OvarianCancer}
-%\bibliographystyle{ieeetr}
+\bibliography{OvarianCancer}
+\bibliographystyle{ieeetr}
 
 \section{Appendices}
-
+\subsection{Additional figures and plots}
+TODO - I'll likely move the AOCS survivial plots here, and perhaps some of the GO lineages.
+\subsection{Gene enrichment raw results}
+\label{tab-ica-4-of-5-enrichment-table}
+TODO
 \subsection{Software libraries and versions}
+TODO
 \label{sec-software-versions}

### 2020-07-30 MSc_Dissertation_B156476.tex
Minor tweaks, added .pdf.  As sent to Ailith

@@
 
-In this work we wish to treat and discuss the three factorization methods in a unified way.  Thus, for the remainder of this dissertation we adopt the following conventions:
+In this work we wish to treat and discuss the three factorization methods in a unified way.
+Thus, for the remainder of this dissertation we adopt the following conventions:\footnote{I should probably use bold font for X, W and H matrices}.
 \begin{description}
@@
 \caption{Overview of methodology as information flow.
-The diagramatic convention here is that at each stage, both datasets -- TCGA and AOCS -- are processes separately by the given algorithm.
+The diagramatic convention here is that at each stage, both datasets -- TCGA and AOCS -- are processed separately by the given algorithm.
 The exception is Gene Set Intersection, which involves interaction between the datasets.}
@@
 \begin{center}
-\includegraphics[width=0.85 \textwidth]{\plotspath{TCGA_OV_VST/GeneEnrichment/go_lineage_1_of_PCA_3.png}} \\
+\includegraphics[width=0.5 \textwidth]{\plotspath{TCGA_OV_VST/GeneEnrichment/go_lineage_1_of_PCA_3.png}} \\
 \begin{spacing}{0.6}
@@
 \vspace{1cm}
-\includegraphics[width=0.85 \textwidth]{\plotspath{TCGA_OV_VST/GeneEnrichment/go_lineage_3_of_PCA_3.png}} \\
+\includegraphics[width= \textwidth]{\plotspath{TCGA_OV_VST/GeneEnrichment/go_lineage_3_of_PCA_3.png}} \\
 \begin{spacing}{0.6}

### 2020-08-08 MSc_Dissertation_B156476.tex
Substantial work on 'Epithelial Ovarian Cancer' and 'Patterns of ... in HGSOC'

@@
 
-\documentclass[tikz, 12pt,a4paper,oneside,fleqn]{article}
+\documentclass[draft, tikz, 12pt,a4paper,oneside,fleqn]{article}
+
 \usepackage[margin=1.5cm]{geometry}
@@
 
-\title{Disentangling patterns of gene expression in high grade serous ovarian cancer}
+\title{Disentangling patterns of gene expression in high grade serous ovarian cancer by matrix factorization}
 \author{}
@@
 %%%%%% Assesment Title Page
+\begin{center}
 \begin{center}
@@
 \end{mdframed}
+\end{center}
 
@@
 
+\begin{abstract}
+Here is the abstract
+\end{abstract}
+\newpage
+
 \section{Introduction}
-High grade serous ovarian cancer (HGSOC) effects the epithelium of the ovaries and fallopian tubes.
-``Serous" refers to the epithelial membrane which secretes serous fluid (serum).
-It is the 7th most common cancer in women worldwide .
-Epithelial ovarian cancers are classified into two broad subtypes:  Type I are low grade with generally good prognosis; type II are more aggressive and typically carry P53 mutations with defects in mechanisms of DNA repair.
-Type II is more-or-less synonomous with HGSOC.
-\cite{Lisio2019}.
-
-This work is concerned with analysing the patters gene expression in HGSOC by dimensionality reduction techniques of matrix factorization, and relating these patterns to clinical and genomic features.
+
+\subsection{Epithelial Ovarian Cancer}
+
+Ovarian cancer (OC) is a heterogeneous disease, presenting with a wide range of pathologies and molecular characteristics .
+The World Health Organisation (WHO) sets out five subtypes of epithelial ovarian carcinomas (EOC) grouped into two main types\cite{Kossai2018, Lisio2019}:
+
+\begin{description}
+\item[Type I] are low grade with generally good prognosis, with two subtypes:
+\begin{description}
+\item[Low grade serous] affecting the epithelial membrane which secretes serous fluid), $< 5$\% of cases
+\item[Mucosal] affecting membrane rich mucous glands, 2 - 3\% 
+\item[Clear cell] relating to the presence of clear cells in histology, 5 - 10\%.
+\end{description}
+\item[Type II] are more aggressive (rapidly growing), typically carrying P53 mutations with defects in mechanisms of DNA repair. Three subtypes:
+\begin{description}
+\item[\emph{High grade serous OC (HGSOC)}] affects the epithelium of the ovaries and fallopian tubes.
+It is the most common, accounting for 70\% of cases, and most serious of the OC subtypes and the focus of this dissertation.
+\item[Endometrial] affecting the the inner lining of the uterus, 10 \% of cases.
+\end{description}
+\end{description}
+
+HGSOC typically occur in older patients and are diagnosed at a late stage.   Histology of HGSOC is similar to the low grade subtype, but develops along distinct molecular pathways.
+
+\emph{Treatment options for HGSOC...}
+
+\subsection{Patterns of gene expression in HGSOC}
+
+Identifying the patterns -- or ``signatures'' -- of gene expression has been an active research area, for the purposes of understanding the carcinogenesis pathway, subtype classification, predicting survival prognosis, predicting treatment response and understanding the causative mechanism of risk factors; these are clearly related, however.
+
+\subsubsection{Carcinogenesis pathways including homologous repair deficiency
+}
+
+BRCA1 inactivation is known to cause chromosomal instability in many cancers.
+Pradhan \etal\ \cite{Pradhan2010} investigated the role of BRCA1 in HGSOC by copy number and expression analysis, finding surprisingly that inactivation has no relationship with gross genomic alteration.
+They leave open the question whether DNA repair by PARP plays a role.
+The relationship between BRCA1/2 and DNA repair -- specifically homologous repair deficiency (HRD) -- is picked up by Ewing \etal \cite{Ewing2020}.
+They study the complex interplay of single nucleotide variants and large structural variants as they impact BRCA1/2 disruption and thence HRD.
+This work suggests that when BRCA1/2 loss is detected in HGSOC patients, there may be a clinical role for PARP inhibitors, since this would prevent error-prone non-homologous repair by PARP, and so selectively kill BRCA1/2 disrupted cells.
+
+Studying the combined landscape of chromosomal rearrangements, driver mutations (particularly TP53, BRCA1/2), methylation and gene expression profiles has lead to a better understanding of the molecular events involved in the progression HGSOC and the development of chemoresistance \cite{Patch2015}.
+The sequence of events is complex, however, and not easily summarised here.
+A similar integrative / multiplatform approach is taken in \cite{He2017a} and \cite{Hoadley2014}.
+
+\subsubsection{Subtype classification}
+
+A relatively early (2003) expression microarray based work demonstrated clustering into for example lymphocyte related genes (IGH3, IGKC, IGLJ3), extracellular matrix/stromal related (COL11A1, Col3A1, MMP2, SPARC, RBBP1) and six others \cite{Schaner2003}.
+
+Wang \etal\ \cite{Wang2017c} also use NMF clustering to identify five subtypes of HGSOC informative of outcomes: 1:mesenchymal, 2:immunoreactive, 3:proliferative, 4:differentiated and 5:anti-mesenchymal.
+Subtypes 2, 5 are found to be associated with longer survival.
+
+Disruption of the PI3K-AKT signalling pathway is known to be significant in several cancers, due to its role in regulating apoptosis.  
+Espinosa \etal\ \cite{Espinosa2011} used unsupervised clustering techniques to investigate expression of 22 genes of the pathway.  
+They found that HGSOC cases formed two separate clusters; the cluster with high expression of CASP3, XIAP, NFKB1, FAS and GSK3B was linked with better outcomes.
+
+A recent pre-publication develops a pipeline for sub-type prediction with the potential to be used clinically \cite{Talhouk2020}.
+The NanoString gene expression platform is used.
+See figure \ref{fig-talhouk_figure_1_part.png}, their strategy was to focus on small set of 513 genes (vs $\sim$20,000 protein coding genes) known from the literature to have relevance to subtyping.
+Two approaches were followed by different teams.
+``All Array": used expression array data from 1650 patients across 14 studies evaluating 9 supervised learning algorithms by bootstrap, selecting an AdaBoost -like method.
+``TCGA" used 434 patients from TCGA evaluating 5 algorithms by cross-validation selecting a random forest.
+"All Array" had the advantage of more data but needed attention to batch effects.
+A final classifier took a consensus of the two methods based on a minimal gene set (order 40 genes), validated by a leave-one-out (patient level) approach.
+Survival analysis was carried out stratified by predicted subtype.
+
+\subsubsection{Treatment response and survival}
+
+As we have seen in section \ref{sec-epithelial-ovarian-cancer}, EOC subtypes are closely linked with survival, however much research is directed at finding direct links between expression and survival.
+
+The prognostic value of expression signatures is demonstrated by the ``Classification of Ovarian Cancer" (CLOVAR) system \cite{Verhaak2013}.  
+Single-sample gene set enrichment analysis (ssGSEA) to obtain scores for previously reported expression signatures of differentiated, immunoreative, mesenchymal and proliferative.  
+A survival outcome prediction model based on these signatures gave good survival stratification (likelihood ratio 10.8 on a validation dataset) which improved to 15.7 when augmented with BRCA1/2 mutation status and other clinical post-operative details.
+
+An explicit supervised learning approach can be taken, as in Berchuck \etal, where a hybrid decision tree plus linear discriminant model demonstrated survival stratification in both cross-validation and external validation experiments, based on microarray expression data.
+They found expression of CSFT3, ABCD3, MAL and APMCF1 to be the most influential; in fact the decision tree was almost entirely driven by the first two of these
+
+Mairinger \etal\ screened 770 immune related genes to identify 11 differentially expressed genes associated with response to platinum treatment.  
+They find that expression of HS11B1, DNBT1, CKLF, NUP107. CCL18, LY96, ATG7, SLAMF7, CXCL9 is associated with better survival, while IKBKG and SDHA associate with with poor survival.
+
+The platinum based drug Cisplatin is a key treatment in ovarian cancer, yet tumours often develop resistance, possibly related to the host's immune response.  
+Understanding and predicting such resistance is therefore of clinical importance and is addressed by Maringer \etal\ \cite{Mairinger2019} through expression analysis using the NanoString platform with a panel of 770 immune related genes.
+They find the following to be significantly related to platinum resistance: KLRC1, TCF7, CD274, HSD11B1, COLEC12, PDGFC, FCF1, BMI1, TNFRSF9, ATG10, EWSR1.
+
+Finding a clear correlation between gene expression patterns and survival is not straight forward.  
+One substantial meta-analysis applied 16 previously published gene sets, to two studies from GEO (199 samples) which included survival information, and found no significant predictive power \cite{Fekete2012a}.
+Although, within the same study, the authors did identify predictive genes, in particular SNGC, MAPT, ESR2 and PGR.
+
+Expression of a single gene -- CD38, which codes for a transmembrane glycoprotein -- has recently been shown to have survival prognostic value on its own \cite{Zhu2020}.
+CD38 has a role in the breakdown of nicotinamide adenine dinucleotide ($\mbox{NAD}^+$), and is also a cell marker of lymphocytes.
+CD38 is over expressed in myloma cells, thence its importance as a prognostic marker and a potential drug treatment target.  
+It is worth mentioning that the analysis in this paper made use of the Gene Expression Profiling Interactive Analysis (GEPIA) database, which conveniently supports survival analysis with respect of any given gene, against a range of TCGA gene expression databases.
+
+\subsubsection{Association with other risk factors}
+
+Obesity has been shown to negatively impact prognosis in ovarian cancer, possibly through difficulties in matching chemotherapy dose with patient BMI.
+\cite{Au-Yeung2014}  A more direct molecular link is proposed by Cuello \etal\ \cite{Cuello2018}.
+They used NMF based clustering (see section \ref{sec-nmf}) to demonstrate that expression of genes linked with obesity and lipid metabolism impart poorer progression free survival, independent of known HGSOC driver gene expression.
+The link is complex however, there been many confounding non-molecular based reasons for a link between high BMI and HGSOC survival, including later diagnosis and compromised operability.
+Some studies (e.g.
+\cite{Hew2013}) conclude no overall link at all.
+
+Note that several of the studies reviewed here use data from the Australian Ovarian Cancer Study (AOCS): \cite{Patch2015,Ewing2020,Cuello2018,Au-Yeung2014}.
+
+%Stratification of HGSOC by immunohistochemistry is addresses Kamble \etal.
+\cite{Kamble2019}.
+They identify a panel of six biomarkers -- TCF21, E-cadherin, PARP1, Slug and AnnexinA2.
+These markers allow stratification into categories including Mesenchymal-to-epithelial transition, Homologous recombination repair and Epithelial-to-mesenchymal transition.
+This, they argue, allows selection of class-specific inhibitor drugs such as Olaparib and Rucaparib.
+
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.35]{talhouk_figure_1_part.png}
+\end{center}
+\caption{HGSOC subtype prediction by expression analysis of a focussed gene set and concensus across two independent pipelines.  From \cite{Talhouk2020}.}
+\label{fig-talhouk_figure_1_part.png}
+\end{figure}
 
@@
 \begin{center}
-\includegraphics[scale=0.4, draft=false]{Sompairac_ICA_NMF_PCA.png}
+\includegraphics[scale=0.4]{Sompairac_ICA_NMF_PCA.png}
 \end{center}
@@
 
-In this work we wish to treat and discuss the three factorization methods in a unified way.
-Thus, for the remainder of this dissertation we adopt the following conventions:\footnote{I should probably use bold font for X, W and H matrices}.
+In this work I wish to treat and discuss the three factorization methods in a unified way.  Thus, for the remainder of this dissertation the following notational conventions are adopted:
 \begin{description}
@@
 
-\subsection{Gene expression in HGSOC}
-
-This section focusses on biological and clinical conclusions relating to gene expression in HGSOC and mentions analysis techniques only in passing.
-
-Obesity is known to negatively impact prognosis in ovarian cancer \cite{Cuello2018}, \cite{Au-Yeung2014}.
-Cuello \etal\ \cite{Cuello2018} used NMF based clustering on mRNA microarrays and reverse-phase protein arrays to demonstrate an association between cancer driver genes and obesity related genes.
-
-Wang \etal\ \cite{Wang2017c} also use NMF clustering to identify five subtypes of HGSOC informative of outcomes: 1:mesenchymal, 2:immunoreactive, 3:proliferative, 4:differentiated and 5:anti-mesenchymal.
-Subtypes 2, 5 are found to be associated with longer survival.
-
-Mairinger \etal\ screened 770 immune related genes to identify 11 differentially expressed genes associated with response to platinum treatment.
-They find that expression of HS11B1, DNBT1, CKLF, NUP107.
-CCL18, LY96, ATG7, SLAMF7, CXCL9 is associated with better survival, while IKBKG and SDHA associate with with poor survival.
-
-BRCA1 inactivation is known to cause chromosomal instability in many cancers.
-Pradhan \etal\ \cite{Pradhan2010} investigated the role of BRCA1 in HGSOC by copy number and expression analysis, finding surprisingly that inactivation has no relationship with gross genomic alteration.
-They leave open the question whether DNA repair by PARP plays a role.
-The relationship between BRCA1/2 and DNA repair -- specifically homologous repair deficiency (HRD) -- is picked up by Ewing \etal \cite{Ewing2020}.
-They study the complex interplay of single nucleotide variants and large structural variants as they impact BRCA1/2 disruption and thence HRD.
-This work suggests that when BRCA1/2 loss is detected in HGSOC patients, there may be a clinical role for PARP inhibitors, since this would prevent error-prone non-homologous repair by PARP, and so selectively kill BRCA1/2 disrupted cells.
-
-Note that several of the studies reviewed here use data from the Australian Ovarian Cancer Study (AOCS): \cite{Winterhoff2016,Patch2015,Ewing2020,Zhang2018,Cuello2018,Au-Yeung2014}.
-
-Stratification of HGSOC by immunohistochemistry is addresses Kamble \etal.
-\cite{Kamble2019}.
-They identify a panel of six biomarkers -- TCF21, E-cadherin, PARP1, Slug and AnnexinA2.
-These markers allow stratification into categories including Mesenchymal-to-epithelial transition, Homologous recombination repair and Epithelial-to-mesenchymal transition.
-This, they argue, allows selection of class-specific inhibitor drugs such as Olaparib and Rucaparib.
-
-The platinum based drug Cisplatin is a key treatment in ovarian cancer, yet tumours often develop resistance, possibly related to the host's immune response.
-Understanding and predicting such resistance is therefore of clinical importance and is addressed by Maringer \etal\ \cite{Mairinger2019} through expression analysis using the NanoString platform with a panel of 770 immune related genes.
-They find the following to be significantly related to platinum resistance: KLRC1, TCF7, CD274, HSD11B1, COLEC12, PDGFC, FCF1, BMI1, TNFRSF9, ATG10, EWSR1.
-
-\begin{figure}
-\begin{center}
-\includegraphics[scale=0.35]{talhouk_figure_1_part.png}
-\end{center}
-\caption{HGSOC subtype prediction by expression analysis of a focussed gene set and concensus across two independent pipelines.  From \cite{Talhouk2020}.}
-\label{fig-talhouk_figure_1_part.png}
-\end{figure}
-
-A recent pre-publication develops a pipeline for sub-type prediction with the potential to be used clinically \cite{Talhouk2020}.
-The NanoString gene expression platform is used.
-See figure \ref{fig-talhouk_figure_1_part.png}, their strategy was to focus on small set of 513 genes (vs $\sim$20,000 protein coding genes) known from the literature to have relevance to subtyping.
-Two approaches were followed by different teams.
-``All Array": used expression array data from 1650 patients across 14 studies evaluating 9 supervised learning algorithms by bootstrap, selecting an AdaBoost -like method.
-``TCGA" used 434 patients from TCGA evaluating 5 algorithms by cross-validation selecting a random forest.
-"All Array" had the advantage of more data but needed attention to batch effects.
-A final classifier took a consensus of the two methods based on a minimal gene set (order 40 genes), validated by a leave-one-out (patient level) approach.
-Survival analysis was carried out stratified by predicted subtype.
-
-\subsection{Research questions}
-The following research goals were set out early in project. 
+\subsection{Research goals}
+This work is concerned with analysing the patters gene expression in HGSOC by dimensionality reduction techniques of matrix factorization, and relating these patterns to clinical and genomic features.
+
+The following research questions were set out early in the project. 
 \begin{enumerate}
@@
 Our overall approach is to use \emph{unsupervised} machine learning methods to represent gene expression data in a small number of features, then to study whether these features correlate with clinical and biological variables.
-This is appropriate in our case, since of the two datasets available (described below), one (TCGA) has relatively large $n$ but little available metadata, while the other (AOCS) is much smaller but has useful metadata.
+This is appropriate in our case, since of the two datasets available (described below), one (TCGA) has relatively large $n$ but little available metadata, while the other (AOCS) is much smaller but has more useful metadata of genomic features.
 This motivates an approach of unsupervised learning on the TCGA dataset.
@@
 \begin{center}
-\includegraphics[scale=0.7, draft=false]{methodlology_overview.pdf}
+\includegraphics[scale=0.7]{methodlology_overview.pdf}
 \end{center}
@@
        save median_metagenes to file by factorizer and k
-       for each factorizer 
+       for each factorizer:
           plot score vs k
@@
 
-It is fundamental to our approach that metagenes determined on the basis of one cohort of data can be used to generate metasamples for a different cohort.
+It is fundamental to our approach that \emph{metagenes} determined on the basis of one cohort of data can be used to generate \emph{metasamples} for a different cohort.
 In the exposition below we refer to these as the \emph{training} and \emph{validation} cohorts respectively.
@@
 % \subsection{Investigation of optimal parameters}
-
 \subsection{Cluster coherence analysis results}
@@
 Gene enrichment analysis against the GO results in, for each of the 11 metagenes, a table of GO terms with a list of the candidate genes which have an annotated association to that term.   
-To take full advantage of the hierarchical nature of the GO, results are presented as lineage maps in figures \ref{fig-NMF-go-lineage} to \ref{fig-PCA-1-3-go-lineage}.  
+To take full advantage of the hierarchical nature of the GO, results are presented as lineage maps in figures \ref{fig-NMF-go-lineage} to \ref{fig-PCA-1-3-genesiso-lineage}.  
 The contributing candidate genes for all significantly enriched terms are shown beneath each figure.  
@@
 \begin{center}
-\includegraphics[scale=0.5, draft=false]{\plotspath{TCGA_OV_VST/GeneEnrichment/component_jaccard_heatmap.pdf}}
+\includegraphics[scale=0.5]{\plotspath{TCGA_OV_VST/GeneEnrichment/component_jaccard_heatmap.pdf}}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=\textwidth, draft=false]{\plotspath{TCGA_OV_VST/GeneEnrichment/go_lineage_4_of_ICA_5.png}} \\
+\includegraphics[width=\textwidth]{\plotspath{TCGA_OV_VST/GeneEnrichment/go_lineage_4_of_ICA_5.png}} \\
 \begin{spacing}{0.6}
@@
 \subsection{Metasample heatmap clustering results}
-
 As with survival analysis, metasamples were derived from the metagenes factorized from the TCGA dataset, then applied to both TCGA and AOCS.

### 2020-08-09 MSc_Dissertation_B156476.tex
Treatment options; improved HGSOC expression section; removed copied figure

@@
 \subsection{Epithelial Ovarian Cancer}
+\label{{sec-epithelial-ovarian-cancer}
 
 Ovarian cancer (OC) is a heterogeneous disease, presenting with a wide range of pathologies and molecular characteristics .
-The World Health Organisation (WHO) sets out five subtypes of epithelial ovarian carcinomas (EOC) grouped into two main types\cite{Kossai2018, Lisio2019}:
+The World Health Organisation sets out five subtypes of epithelial ovarian carcinomas (EOC) grouped into two main types\cite{Kossai2018, Lisio2019}:
 
@@
 \begin{description}
+\item[Endometrial] affecting the the inner lining of the uterus, 10 \% of cases.
 \item[\emph{High grade serous OC (HGSOC)}] affects the epithelium of the ovaries and fallopian tubes.
-It is the most common, accounting for 70\% of cases, and most serious of the OC subtypes and the focus of this dissertation.
-\item[Endometrial] affecting the the inner lining of the uterus, 10 \% of cases.
+It is the most common type of OC, accounting for 70\% of cases, and most serious of the OC subtypes and the focus of this dissertation.
 \end{description}
@@
 
-HGSOC typically occur in older patients and are diagnosed at a late stage.   Histology of HGSOC is similar to the low grade subtype, but develops along distinct molecular pathways.
-
-\emph{Treatment options for HGSOC...}
+HGSOC typically occurs in older patients and is diagnosed at a late stage.   Histology of HGSOC is similar to the low grade subtype, but the cancer develops along distinct molecular pathways.
+
+Treatment options for HGSOC depend on tumour stage and include surgery (tumour debulking) or platinum-based chemotherapy (e.g.
+Cisplatin).
+Acquired resistance to platinum is a major cause of disease recurrance\cite{Patch2015} and so motivates research into the genomic events which lead to such resistance.
 
@@
 
-Identifying the patterns -- or ``signatures'' -- of gene expression has been an active research area, for the purposes of understanding the carcinogenesis pathway, subtype classification, predicting survival prognosis, predicting treatment response and understanding the causative mechanism of risk factors; these are clearly related, however.
+Identifying the patterns -- or ``signatures'' -- of gene expression has been an active research area, for the purposes of understanding the carcinogenesis pathway, subtype classification, predicting survival prognosis, predicting treatment response and understanding the causative mechanism of risk factors; there is much cross-over between these areas, however.
 
@@
 
-A relatively early (2003) expression microarray based work demonstrated clustering into for example lymphocyte related genes (IGH3, IGKC, IGLJ3), extracellular matrix/stromal related (COL11A1, Col3A1, MMP2, SPARC, RBBP1) and six others \cite{Schaner2003}.
-
-Wang \etal\ \cite{Wang2017c} also use NMF clustering to identify five subtypes of HGSOC informative of outcomes: 1:mesenchymal, 2:immunoreactive, 3:proliferative, 4:differentiated and 5:anti-mesenchymal.
-Subtypes 2, 5 are found to be associated with longer survival.
+A relatively early (2003) expression microarray based work demonstrated clustering into lymphocyte related genes (IGH3, IGKC, IGLJ3), extracellular matrix/stromal related (COL11A1, COL3A1, MMP2, SPARC, RBBP1) and six other clusters \cite{Schaner2003}.
+
+Wang \etal\ \cite{Wang2017c} use NMF clustering (see section \ref{sec-NMF}) to identify five subtypes of HGSOC informative of outcomes: 1:mesenchymal, 2:immunoreactive, 3:proliferative, 4:differentiated and 5:anti-mesenchymal.
+Subtypes 2 and 5 are found to be associated with better survival.
 
 Disruption of the PI3K-AKT signalling pathway is known to be significant in several cancers, due to its role in regulating apoptosis.  
-Espinosa \etal\ \cite{Espinosa2011} used unsupervised clustering techniques to investigate expression of 22 genes of the pathway.  
+Espinosa \etal\ \cite{Espinosa2011} used unsupervised clustering techniques to investigate expression of 22 genes of this pathway.  
 They found that HGSOC cases formed two separate clusters; the cluster with high expression of CASP3, XIAP, NFKB1, FAS and GSK3B was linked with better outcomes.
@@
 A recent pre-publication develops a pipeline for sub-type prediction with the potential to be used clinically \cite{Talhouk2020}.
-The NanoString gene expression platform is used.
-See figure \ref{fig-talhouk_figure_1_part.png}, their strategy was to focus on small set of 513 genes (vs $\sim$20,000 protein coding genes) known from the literature to have relevance to subtyping.
+The NanoString gene expression platform is used.  Their strategy was to focus on small set of 513 genes (vs $\sim$21,000 protein coding genes) known from the literature to have relevance to subtyping.
 Two approaches were followed by different teams.
 ``All Array": used expression array data from 1650 patients across 14 studies evaluating 9 supervised learning algorithms by bootstrap, selecting an AdaBoost -like method.
-``TCGA" used 434 patients from TCGA evaluating 5 algorithms by cross-validation selecting a random forest.
+The ``TCGA" team used 434 patients from The Cancer Genome Atlas (TCGA\footnote{I'm always amused by the geneticist's pun in the naming of this resource!}) evaluating 5 algorithms by cross-validation selecting a random forest.
 "All Array" had the advantage of more data but needed attention to batch effects.
-A final classifier took a consensus of the two methods based on a minimal gene set (order 40 genes), validated by a leave-one-out (patient level) approach.
+A final classifier took a consensus of the two methods based on a minimal gene set (order 40 of genes), validated by a leave-one-out (patient level) approach.
 Survival analysis was carried out stratified by predicted subtype.
@@
 
-As we have seen in section \ref{sec-epithelial-ovarian-cancer}, EOC subtypes are closely linked with survival, however much research is directed at finding direct links between expression and survival.
+As we have seen in section \ref{sec-epithelial-ovarian-cancer}, EOC subtypes are closely linked with survival, never-the-less much research is directed at finding direct links with expression.
 
 The prognostic value of expression signatures is demonstrated by the ``Classification of Ovarian Cancer" (CLOVAR) system \cite{Verhaak2013}.  
-Single-sample gene set enrichment analysis (ssGSEA) to obtain scores for previously reported expression signatures of differentiated, immunoreative, mesenchymal and proliferative.  
-A survival outcome prediction model based on these signatures gave good survival stratification (likelihood ratio 10.8 on a validation dataset) which improved to 15.7 when augmented with BRCA1/2 mutation status and other clinical post-operative details.
+Single-sample gene set enrichment analysis (ssGSEA) is used to obtain scores for previously reported expression signatures: differentiated, immunoreative, mesenchymal and proliferative.  
+An outcome prediction model based on these signatures gave good survival stratification (likelihood ratio 10.8 between above/below median groups) on a validation dataset) which improved to 15.7 when augmented with BRCA1/2 mutation status and other post-operative clinical factors.
 
@@
 Understanding and predicting such resistance is therefore of clinical importance and is addressed by Maringer \etal\ \cite{Mairinger2019} through expression analysis using the NanoString platform with a panel of 770 immune related genes.
-They find the following to be significantly related to platinum resistance: KLRC1, TCF7, CD274, HSD11B1, COLEC12, PDGFC, FCF1, BMI1, TNFRSF9, ATG10, EWSR1.
-
-Finding a clear correlation between gene expression patterns and survival is not straight forward.  
-One substantial meta-analysis applied 16 previously published gene sets, to two studies from GEO (199 samples) which included survival information, and found no significant predictive power \cite{Fekete2012a}.
+They find the following genes to be significantly related to platinum resistance: KLRC1, TCF7, CD274, HSD11B1, COLEC12, PDGFC, FCF1, BMI1, TNFRSF9, ATG10, EWSR1.
+
+Finding a clear, robust correlation between gene expression patterns and survival is not straight forward.  
+One substantial meta-analysis applied 16 previously published gene sets to two studies from GEO (199 samples) which included survival information, and found no significant predictive power \cite{Fekete2012a}.
 Although, within the same study, the authors did identify predictive genes, in particular SNGC, MAPT, ESR2 and PGR.
@@
 \cite{Au-Yeung2014}  A more direct molecular link is proposed by Cuello \etal\ \cite{Cuello2018}.
-They used NMF based clustering (see section \ref{sec-nmf}) to demonstrate that expression of genes linked with obesity and lipid metabolism impart poorer progression free survival, independent of known HGSOC driver gene expression.
+They used NMF based clustering to demonstrate that expression of genes linked with obesity and lipid metabolism impart poorer progression free survival, independent of known HGSOC driver gene expression.
 The link is complex however, there been many confounding non-molecular based reasons for a link between high BMI and HGSOC survival, including later diagnosis and compromised operability.
@@
 \cite{Hew2013}) conclude no overall link at all.
-
-Note that several of the studies reviewed here use data from the Australian Ovarian Cancer Study (AOCS): \cite{Patch2015,Ewing2020,Cuello2018,Au-Yeung2014}.
 
@@
 This, they argue, allows selection of class-specific inhibitor drugs such as Olaparib and Rucaparib.
-
-\begin{figure}
-\begin{center}
-\includegraphics[scale=0.35]{talhouk_figure_1_part.png}
-\end{center}
-\caption{HGSOC subtype prediction by expression analysis of a focussed gene set and concensus across two independent pipelines.  From \cite{Talhouk2020}.}
-\label{fig-talhouk_figure_1_part.png}
-\end{figure}
 
@@
 \item Austrailian Ovarian Cancer Study (AOCS) \cite{Patch2015}, n=80, g=19,730, with metadata on survival, cellularity and additional genomic features.
+Several of the studies reviewed earlier also use data from the Australian Ovarian Cancer Study (AOCS): \cite{Patch2015,Ewing2020,Cuello2018,Au-Yeung2014}.
 \end{enumerate}

### 2020-08-09 MSc_Dissertation_B156476.tex
Treatment options; improved HGSOC expression section; removed copied figure
Figure[ht], WGD boxplots, grid of scatter plots * Added [ht] to all begin{figure} * Incorporated boxplot for WGD (may remove) and grid of scatter plots,   with caption.

@@
 \subsection{Epithelial Ovarian Cancer}
-\label{{sec-epithelial-ovarian-cancer}
+\label{sec-epithelial-ovarian-cancer}
 
@@
 
-\begin{figure}
+\begin{figure}[ht]
 \begin{center}
@@
 \end{enumerate}
+
+\afterpage{\clearpage}
 
@@
 \def\S{0.4}
-\begin{figure}
+\begin{figure}[ht]
 \begin{center}
@@
 
-\begin{figure}[hb]
+\begin{figure}[ht]
 \begin{center}
@@
 \caption{Heatmap of the metasamples matrix for the AOCS dataset, computed from metagenes factorized from the TCGA dataset.
-Several patient patient metadata variables are shown in columns to the left.
+Several patient metadata variables are shown in columns to the left.
 (ToDo: key for metadata colouring).}
@@
 
+\subsection{Metasample correlation with genomic features}
+
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[scale=0.6, draft=false]{\plotspath{BoxAndScatters/wgd_boxplot_NMF_TCGA_AOCS.pdf}}
+\includegraphics[scale=0.6, draft=false]{\plotspath{BoxAndScatters/wgd_boxplot_ICA_TCGA_AOCS.pdf}}
+\includegraphics[scale=0.6, draft=false]{\plotspath{BoxAndScatters/wgd_boxplot_PCA_TCGA_AOCS.pdf}}
+\caption{Boxplots showing relationship between metasamples (derived from TCGA metagenes, applied to AOCS data) and whole genome doubling (WGD) status.
+Metasamples from NMF, ICA and PCA are shown separately.
+To allow convenient visualisation, each metasample is normalised to zero mean unit variance, hence between metasample variation is removed.}
+\end{center}
+\label{fig-wgd-boxplots}
+\end{figure}
+
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[scale=0.54, draft=false]{\plotspath{BoxAndScatters/genomic_feature_scatters_TCGA_AOCS.pdf}} 
+\caption{Grid of scatter plots to visualise the correlation between metasamples and genomic features.
+In the case of WGD which has binary values (0, 1), jitter is applied for visual effect only.
+Above each plot is shown a correlation coefficient $r$ and associated p-value, highlighted with '***' where $p < 0.01$.
+In the case of WGD , the Point-Biserial correlation is used; for all others Pearson's r is used.}
+\label{fig-genomic-scatters}
+\end{center}
+\end{figure}
+
+\clearpage
 \section{Discussion}

### 2020-08-10 MSc_Dissertation_B156476.tex
Much improvement, rearrangements, corrections and additions.. Sending this version to Athina.

@@
 
-\documentclass[draft, tikz, 12pt,a4paper,oneside,fleqn]{article}
+\documentclass[tikz, 12pt,a4paper,oneside,fleqn]{article}
 
@@
 
-\subsubsection{Carcinogenesis pathways including homologous repair deficiency
-}
+\subsubsection{Carcinogenesis pathways including homologous repair deficiency}
 
@@
 \subsubsection{Subtype classification}
+\label{sec-subtype-classification}
 
@@
 
-Wang \etal\ \cite{Wang2017c} use NMF clustering (see section \ref{sec-NMF}) to identify five subtypes of HGSOC informative of outcomes: 1:mesenchymal, 2:immunoreactive, 3:proliferative, 4:differentiated and 5:anti-mesenchymal.
+A relatively early (2003) expression microarray based work demonstrated clustering into lymphocyte related genes (IGH3, IGKC, IGLJ3), extracellular matrix/stromal related (COL11A1, COL3A1, MMP2, SPARC, RBBP1) and six other clusters \cite{Schaner2003}.
+
+Wang \etal\ \cite{Wang2017c} use NMF clustering (see section \ref{sec-nmf}) to identify five subtypes of HGSOC informative of outcomes: 1:mesenchymal, 2:immunoreactive, 3:proliferative, 4:differentiated and 5:anti-mesenchymal.
 Subtypes 2 and 5 are found to be associated with better survival.
@@
 They found that HGSOC cases formed two separate clusters; the cluster with high expression of CASP3, XIAP, NFKB1, FAS and GSK3B was linked with better outcomes.
+
+An integrative network approach has identified two distinct subtypes of ovarian cancer, distinguished by differences in expression of genes related to transcription factors which influence angiogenesis \cite{Glass2015}.
+The authors suggest that the regulatory networks highlighted by their analysis can lead to targetting with specific drugs.
 
@@
 ``All Array": used expression array data from 1650 patients across 14 studies evaluating 9 supervised learning algorithms by bootstrap, selecting an AdaBoost -like method.
-The ``TCGA" team used 434 patients from The Cancer Genome Atlas (TCGA\footnote{I'm always amused by the geneticist's pun in the naming of this resource!}) evaluating 5 algorithms by cross-validation selecting a random forest.
+The ``TCGA" team used 434 patients from The Cancer Genome Atlas (TCGA\footnote{I'm always amused by the genetics pun in the naming of this resource!}) evaluating 5 algorithms by cross-validation selecting a random forest.
 "All Array" had the advantage of more data but needed attention to batch effects.
@@
 Finding a clear, robust correlation between gene expression patterns and survival is not straight forward.  
-One substantial meta-analysis applied 16 previously published gene sets to two studies from GEO (199 samples) which included survival information, and found no significant predictive power \cite{Fekete2012a}.
+One substantial meta-analysis applied 16 previously published gene sets to two studies from GEO (199 samples) which included survival information, and found no significant predictive power \cite{Fekete2012}.
 Although, within the same study, the authors did identify predictive genes, in particular SNGC, MAPT, ESR2 and PGR.
@@
 
-Obesity has been shown to negatively impact prognosis in ovarian cancer, possibly through difficulties in matching chemotherapy dose with patient BMI.
-\cite{Au-Yeung2014}  A more direct molecular link is proposed by Cuello \etal\ \cite{Cuello2018}.
+Obesity has been shown to negatively impact prognosis in ovarian cancer, possibly through difficulties in matching chemotherapy dose with patient BMI  \cite{Au-Yeung2014}.
+A more direct molecular link is proposed by Cuello \etal\ \cite{Cuello2018}.
 They used NMF based clustering to demonstrate that expression of genes linked with obesity and lipid metabolism impart poorer progression free survival, independent of known HGSOC driver gene expression.
@@
 
-\subsection{Matrix factorization}
+\subsection{Matrix factorization for dimensionality reduction}
 \label{sec-matrix-factorization-intro}
 
-A good introduction to field is ref \cite{Stein-OBrien2018}, in which matrix factorization (MF) methods are discussed as a means of extraction the low dimensional structure from a high dimensional ``-omics" datasets, with gene expression analysis being the prime example.
-
-The three matrix factorization methods of interest to this work are listed below.
+Dimensionality reduction concerns representing high dimensional data in far fewer dimensions, with minimal loss of information.
+This is possible in many cases because dimensions -- or equivalently elements of a feature vector -- are highly correlated.
+In the case of gene expression analysis, the dimensionality is of the order 20,000, i.e.
+the number of transcripts (mRMA transcribed genes) being considered.
+However, it is known that sets of genes act in concert, their expression thus being correlated, so reducing the true degrees of freedom in the data.
+The challenge is to uncover these true degrees of freedom, which constitute our ``patterns of gene expression''.
+
+The three matrix factorization methods considered in this work are described below.
 All are forms of dimensionality reduction, but have differing criteria they aim to optimise.
-The number of target dimensions (or components, factors, metagenes) is referred to as the rank, $k$.
-\begin{description}
-\item [Non-negative matrix factorization (NMF).]  
+The number of target dimensions (or components, factors, metagenes -- these terms are used somewhat synonymously) is referred to as the rank, $k$.
+\subsubsection{Non-negative matrix factorization (NMF)}
+\label{sec-nmf}
 As the name implies, this is only applicable to matrices with +ve or zero elements, and finds components which are themselves +ve or zero.
 This property make the resulting factorization more interpretable since the components are strictly additive.
-It also allows for input samples to be directly assigned to one of $k$ clusters.
-A good description of NMF and its applications in bioinformatics can be found in \cite{Brunet2004}.
+NMF has been applied widely in -omics research, originally to gene expression microarrays  and latterly to RNA-seq datasets.  
+A particular value of NMF is that it allows for input samples to be directly assigned to one of $k$ clusters.   
+The optimization function on which NMF is based can be tuned to achieve a \emph{sparse} factorization result, that is one in which only a small number of elements in the achieved factorization are non-zero, a property which further simplifies interpretation.
+\cite{Brunet2004}.
 %
-\item [Independent component analysis (ICA).] The goal of ICA is to represent the given matrix with components which are statistically independent of each other.
+\subsubsection{Independent component analysis (ICA)} 
+The goal of ICA is to represent the given matrix with components which are statistically independent of each other.
 It was originally proposed to solve the blind source separation problems \cite{Comon1994}.
@@
 See \cite{Sompairac2019} for a full explanation, from which figure \ref{fig-pca-ica-nmp-explanation} is taken.
-\item [Principal component analysis (PCA).]  PCA is a well known technique for dimensionality reduction, based on eigenvalue decomposition.
-It is the ideal, efficient solution for multivariate Gaussian distributed data.
-\end{description}
-
-The contrasting results of the three methods is illustrated in figure \ref{fig-pca-ica-nmp-explanation}.
+\subsubsection{Principal component analysis (PCA)}
+PCA is a well known technique for dimensionality reduction, based on eigenvalue decomposition.
+It is arguably the ideal, efficient solution for data which is multivariate Gaussian distributed -- or can assumed to be so.
+PCA results in components which are ordered by the proportion of total variance they explain, and are mutually orthogonal.
+
+The contrasting results of the three methods is illustrated in figure \ref{fig-pca-ica-nmp-explanation}. \emph{TODO: add explanation of this diagram}
 
@@
 A typical expression array might have in the order of tens of thousands of rows and hundreds of columns.  
-MF methods reduces this large $M \times N$ (genes $\times$ samples) matrix into two two smaller matrices.
+MF methods reduces this large $M \times N$ (genes $\times$ samples) matrix into two smaller matrices.
 In the general terminology of Stein-O'Brien \etal\ these are the $M \times K$ \emph{pattern matrix} and the $K \times N$ \emph{amplitude matrix}.
-$K$ is typically small of the order 10, and refers to the number of extracted \emph{factors} (or components).
+$K$ (or $k$ as we will use) is typically small of the order 10, and refers to the number of extracted \emph{factors} (or components); it is the rank of the factorization.
 Since we are focussing on transriptomics, rather than pattern and amplitude we will use the terms \emph{metagenes} and \emph{metasamples} respectively, which are in common useage.
@@
 
-There seems to be substantial confusion an lack of clarity in the way that matrix factorization, particularly ICA, is applied in transcriptomics research.  
-``Surprisingly, both ways of applying ICA to omics data are wide-spread, and sometimes it makes an effort to figure out in which way ICA was applied." \cite{Sompairac2019}, and ``Different protocols to apply ICA to transcriptomic data exist and currently no single standard approach has been defined.
+There is substantial confusion and lack of clarity in the way that matrix factorization, particularly ICA, is applied in transcriptomics research.  
+``Surprisingly, both ways of applying ICA to omics data are wide-spread, and sometimes it takes an effort to figure out in which way ICA was applied" \cite{Sompairac2019}, and ``Different protocols to apply ICA to transcriptomic data exist and currently no single standard approach has been defined.
 The main difference in the existing approaches consists in what is considered as source signal matrix in the decomposition" \cite{Cantini2019}.
@@
 
-The number of factors, or rank $K$, to extract is a key decision in any matrix factorization approach.
+The number of factors, or rank $k$, to extract is a key decision in any matrix factorization approach.
 In this regard, PCA differs from ICA and NMF.
-In PCA it is reasonable to extract all factors, setting $K=N$, the factors being ranked by the associated eigenvalue which also ranks the proportion of variance explained.
-However, that approach is not valid for ICA and NMF, since differing sets of factors will be obtained for different $K$, and for a given choice of $K$ all factors are equally important -- they do not rank \cite{Stein-OBrien2018}.
-
-The problem of determining the optimal $K$ for a given expression matrix is addressed  by Kairov \etal\ \cite{Kairov2017}, based on optimizing the \emph{stability} of the components over multiple algorithm initializations.
+In PCA it is reasonable to extract all factors, setting $k=N$, the factors being ranked by the associated eigenvalue which also ranks the proportion of variance explained.
+However, that approach is not valid for ICA and NMF, since differing sets of factors will be obtained for different $k$, and for a given choice of $k$ all factors are equally important -- they do not rank \cite{Stein-OBrien2018}.
+
+The problem of determining the optimal $k$ for a given expression matrix is addressed  by Kairov \etal\ \cite{Kairov2017}, based on optimizing the \emph{stability} of the components over multiple algorithm initializations.
 
 \subsection{Research goals}
-This work is concerned with analysing the patters gene expression in HGSOC by dimensionality reduction techniques of matrix factorization, and relating these patterns to clinical and genomic features.
-
-The following research questions were set out early in the project. 
+This work is concerned with analysing the patterns gene expression in HGSOC by dimensionality reduction techniques of matrix factorization, and relating these patterns to clinical and genomic features.
+
+The following research questions were set out early in the project. \emph{TODO: tweak these goals!}
 \begin{enumerate}
@@
 Our overall approach is to use \emph{unsupervised} machine learning methods to represent gene expression data in a small number of features, then to study whether these features correlate with clinical and biological variables.
-This is appropriate in our case, since of the two datasets available (described below), one (TCGA) has relatively large $n$ but little available metadata, while the other (AOCS) is much smaller but has more useful metadata of genomic features.
-This motivates an approach of unsupervised learning on the TCGA dataset.
+This is appropriate in our case, since of the two datasets available (described below), one (TCGA) has relatively large n but little available metadata, while the other (AOCS) is much smaller but has more useful metadata of genomic features.
+This motivates an approach of unsupervised learning on the TCGA dataset which is then transferred to the AOCS dataset.
 
@@
 
+Expression data for the AOCS and TCGA datasets was received for this project in a spreadsheet format, having been derived from the RNA-Seq data as described in \cite{Ewing2020}, including  normalization by variance stabilizing transformation.
+
 \subsubsection{Method outline}
-An overview of the methodology adopted in this work is shown in figure \ref{fig-method-overview}.
+An overview of the methodology adopted in this work is shown in figure \ref{fig-method-overview}, consisting of:
 
@@
 \end{figure}
-In outline our method is follows:
+
 \begin{enumerate}
@@
 \item
-Clustered heatmaps to visualise relationships between metasamples, patients and metadata.
+Scatter plots and heat maps to investigate the relationships between metasamples and genomic features available for the AOCS dataset.
 \end{enumerate}
 
-\afterpage{\clearpage}
+\clearpage
 
@@
 \item Python 3.6
-\item PyCharm IDE
+\item PyCharm Integrated Development Environment
 \item Numpy for high performance matrix manipulation
 \item Pandas for data frame handing
-\item Matplotlib for general plotting
-\item Seaborn for heatmaps
+\item Matplotlib and Seaborn for general plotting
 \item Scikit-learn for matrix factorization and k-means clustering
@@
 
-\subsection{Data normalization and quality control}
-
-Expression data for the AOCS and TCGA datasets was received for this project in a spreadsheet format, having been derived from the RNA-Seq data as described in \cite{Ewing2020}, including  normalization by variance stabilizing transformation.
-
 \subsection{Gene set intersection}
@@
 One of the aims of this work is to compare the efficacy of three methods of dimensionality reduction -- NMF, ICA and PCA -- as explained in the introduction.   
-The three methods have different properties and notational conventions, but to simplify the discussion here we adopt the notation $X \approx W H$ for all three methods, where $X$ is the (genes, patients) expression matrix  $W$ is the (genes, factors) metagenes matrix and $H$ is the (factors, patients) metasamples matrix.
+%The three methods have different properties and notational conventions, but to simplify the discussion here we adopt the notation $X \approx W H$ for all three methods, where $X$ is the (genes, patients) expression matrix  $W$ is the (genes, factors) metagenes matrix and $H$ is the (factors, patients) metasamples matrix.
 
@@
 
+\emph{TODO: probably best to state the chosen hyper parameters here}
+
 \subsection{Metagene selection by cluster coherence}
@@
 
-A complication arises in the computation of ICA and PCA facorizations, in that essentially the same factor can arise as $w$ or $-w$ vectors -- i.e.
-$180^\circ$ rotated.
+A complication arises in the computation of ICA and PCA facorizations, in that essentially the same factor can arise as $w$ or $-w$ -- i.e.
+$180^\circ$ rotated vectors.
 These would appear as separate cluster in repeated sampling, and confound attempts to collect and aggregate.
@@
 
-In order to obtain median estimates of the $k$ metagenes from the $50 k$ which were generated, k-means clustering was performed in the PCA reduced (20 dimension) space, delivering $k$ sets of points.
+In order to obtain median estimates of the $k$ metagenes from the $50 k$ which were generated, k-means clustering was performed in the PCA reduced ($r=20$ dimension) space, delivering $k$ sets of points.
 These were referenced back to the original 19,566 dimensioned metagenes and the per-dimension median calculated.
@@
 \href{https://en.wikipedia.org/wiki/Silhouette_(clustering)}{Wikipedia: Silouette (clustering)}.
+\emph{TODO: make this a reference}
 The score is in the range -1 to +1, with 0 implying random scatter and 1 implying all points of a cluster perfectly overlay.
 
-The t-SNE plots and silhouette scores were assessed visually (see figures \ref{fig-first-cluster-result} to \ref{fig-last-cluster-result}) to decide, for each of the three factorizers, the highest rank with good cluster coherence, those median metagenes to take forward to gene enrichment and survival analysis.
+The t-SNE plots and silhouette scores were assessed visually to decide, for each of the three factorizers, the highest rank with good cluster coherence, those median metagenes to take forward to gene enrichment and survival and genomic feature correlation analysis.
 Note that the t-SNE plots are used only for visualisation; they are not involved in the computation of median metagenes or silhouette score.
 
-Initial investigation on the N=80 AOCS dataset showed very poor cluster coherence, even for k=2.   Thus it was decided to compute and select metagenes only from the N=374 TCGA dataset.   
-The selection rationale is set out in the results section, but is convenient to state here that the following ranks were selected:
+Initial investigation on the N=80 AOCS dataset showed very poor cluster coherence, even for k=2 (this will be seen in the presented results).
+Thus it was decided to compute and select metagenes only from the N=374 TCGA dataset.
+The selection rationale is set out in the results section, but it is convenient to state here that the following ranks were selected:
 $K_{\mbox{\tiny NMF}} = 3, K_{\mbox{\tiny ICA}} = 5, K_{\mbox{\tiny PCA}} =3$.  Thus, there were $3 + 5 + 3 = 11$ metagenes taken forward for follow-on analysis.
@@
 \end{center}
-\caption{Pseudocode for generating median metagenes and assessing there stability to random algorithm initialization and sampling error.}
+\caption{Pseudocode for generating median metagenes and assessing their stability to random algorithm initialization and sampling error.}
 \label{fig-clustering-psuedocode}
 \end{figure}
+
+\subsection{Determining metagene similarity by Jaccard index}
+
+It is pertinent to ask ``how similar are the 11 components which emerge from the foregoing factorization and cluster analysis?''.
+We expect those coming from a single factorization method to be distinct by construction.
+But perhaps ICA and PCA identified similar components.
+Similarity with NMF derived components is possible but less likely due to the positivity constraint.
+A possible approach to measuring similarity might be to treat the components as vectors and determine the angle between them, via the scalar or ``dot'' product.
+But in such a high dimension space (19,566) any pair of vectors will be very close to $90^\circ$.
+
+An alternative to consider the intersection of the gene sets which each metagene highlights.
+In the gene enrichment analysis which follows, candidate genes will be identified for each meatagene.
+A standard measure of set similarity is the Jaccard index, (or similarity), defined on a pair of sets as:
+
+\[
+   \mbox{Jaccard}(A, B) = \frac{A \cap B}{A \cup B}.
+\]
+The measure lies between 0 (no similarity) and 1 (identical), and is symmetric.  Jaccard similarity was thus calculated for all pairs of the 11 genes sets and visualised as a heatmap.
 
@@
 Multiple hypothesis significance testing used the false discovery rate (FDR) method of Bengamini and Hochberg, the FDR threshold being set to 0.01.
-This is more stringent than the 0.05 value which is typically used, and was chosen since multiple metagenes are being analysed, implying multiple hypothesis testing over and above that which is accounted for by FDR filtering within a single gene enrichment analysis run.
+This is more stringent than the 0.05 value which is commonly used, and was chosen since multiple metagenes are being analysed, implying multiple hypothesis testing over and above that which is accounted for by FDR filtering within a single gene enrichment analysis run.
 
@@
 \caption{Example of a small section of the Gene Ontology (GO), focussed GO:0042773 (in purple), showing parents and children of the term.   Generated by GOATOOLS.}
-\label{fig:eg_go_graphic}
+\label{fig-eg_go_graphic}
 \end{center}
@@
 This list can be inspected directly for insights, but that misses the point of the GO, which is to organise the terms hierarchically by `is a' relationships.
-Thus, the enriched terms were rendered graphically to show them in the context of their parent terms.
+Thus, the enriched terms were rendered graphically to show them in the context of their parent terms; an example is shown in figure \ref{fig-eg_go_graphic}
 
@@
 
-To carry out patient level analysis -- survival analysis, heatmaps and boxplot -- as described in the following sections, it is necessary to compile tables (Panda DataFrames) which bring together per-patient metasamples relating to each of the selected metagenes, with associated metadata -- such as cellularity and survival information.
+To carry out patient level analysis -- survival analysis, heatmaps and boxplot -- as described in the following sections, it is necessary to compile tables (Pandas DataFrames) which bring together per-patient metasamples relating to each of the selected metagenes, with associated metadata -- such as cellularity and survival information.
 The metagenes always derive from the TCGA dataset (because of its larger size as explained in section \ref{sec-metagene-selection}), but we wish to apply these metagenes to the expression data and associated metadata of either the TCGA or AOCS datasets.
 The mathematics of transferring metagenes across datasets has been described above.
-Care is required to ensure that expression matrices aligned correctly with metagenes with respect to their genes (rows), and with the metadata with respect to patient identifiers in the columns.
+Care is required to ensure that expression matrices are aligned correctly with metagenes with respect to their genes (rows), and with the metadata with respect to patient identifiers in the columns.
 Note that the same transferring approach is taken when metasamples are required for TCGA, even though we could in that case obtain the $H$ matrix directly from the factorization.
@@
 Application to TCGA (for OS) thus represents an in-sample test, while application to AOCS (for OS and PFS) is a more exacting out-of-sample test.
-There are thus three analyses to consider: 1) TCGA$\rightarrow$TCGA (OS), 2) TCGA$\rightarrow$OACS (OS) and 3) TCGA$\rightarrow$OACS (PFS).
+There are thus three analyses to consider: 1) TCGA$\rightarrow$TCGA (OS), 2) TCGA$\rightarrow$AOCS (OS) and 3) TCGA$\rightarrow$AOCS (PFS).
+\emph{TODO: describe these three versions of analysis in words}.
 
@@
 Metasample values were binariased to 0, 1 by thresholding at the median value.
-Kaplan-Meier plots were made in respect of derived metasample for each of the three analyses -- making $11 \times 3$ plots each with two survival curves and showing with 95\% confidence intervals.
+Kaplan-Meier plots were made in respect of derived metasample for each of the three analyses -- making $11 \times 3$ plots each with two survival curves with 95\% confidence intervals.
 A hazard ratio (HR)
-\footnote{Do I need to explain the definition and meaning of HR?}
 was calculated for each plot by fitting Cox's proportional hazards model with p-value relating to the hypothesis that HR is significantly different to 1.0.
 
-\subsection{Metasample heatmap analysis}
-
-To further investigate the relationship between metasamples and to available patient metadata, clustered heatmaps were generated using the \href{https://seaborn.pydata.org/generated/seaborn.clustermap.html}{Seaborn {\tt clustermap()}} function from the reconciled metasamples and metadata (section \ref{sec-reconciling-metasamples-metadata}).
-Heatmaps were generated for TCGA and AOCS datasets (\emph{Should say something about stratifying the metadata, when I figure it out}).
-Heatmap clustering by correlation distance was applied.
-
-\subsection{Codebase}
-
-The described method was implemented in Python.
+\subsection{Investigating correlation between metasamples and genomic features}
+
+For the AOCS dataset (only) pre-determined per-patient high-level genomic features were available as follows:
+\begin{description}
+\item[WGD]: Whole genome doubling -- a marker of genome instability.  This is a binary feature.
+\item[Cellularity]: proportion of cells belonging to the tumour (as opposed to surrounding normal tissue).
+\item[HR Detect]: A predictor of homologous repair deficiency based on established mutational features \cite{Ewing2020}.
+\item[Mutational Load]: A measure of the total number of mutations present in the tumour genome.
+\item[CNV Load]: Copy-number variation, i.e. deviation from the normal diploid cell compliment.
+\item[SV Load]: Structural variation, a measure of the degree of chromosomal rearrangements such as translocations and inversions.
+\end{description}
+
+Scatter plots were generated between each metasample and each genomic features -- a grid of $11 \times 6 = 66$ plots in all.
+These were based on the reconciled metasamples and metadata as described in section \ref{sec-reconciling-metasamples-metadata}.
+Pearson's correlation coefficients ($r$) and associated p-values (i.e.
+the probability that $r$ does not differ from zero) were calculated.
+However, since WGD is a binary feature, the more appropriate Point-Biserial correlation was used in that case.
+A p-value significance threshold of 0.01 was chosen, although with 66 hypotheses being tested, this does risk false discovery.
+
+As an alternative visualisation of the relationship between metasamples and genomic features, and to see the relationship between metasamples themselves, a clustered heatmaps was generated using the \href{https://seaborn.pydata.org/generated/seaborn.clustermap.html}{Seaborn {\tt clustermap()}} function based on the same underlying data as above.
+
+\subsection{Codebase and plotting conventions}
+
+The described methodology was implemented in Python.
 %except for the KEGG gene enrichment analysis which was implemented in R.
@@
 Good software engineering practice has been followed, with an object-oriented design, frequent commits and unit testing.
-Unit tests are based on tiny expression matrices, just 100 genes by 10 patients, thus the whole test suite of over 60 tests executes in around 30 seconds.
-These tests are particularly valuable when re-factoring code by providing a rapid means of detecting coding errors.
-The total codebase is approximately 2,800 lines of Python.
+Unit tests are based on tiny expression matrices, just 100 genes by 10 patients, thus the whole test suite of over 60 tests executes in around 40 seconds and can be run frequently.
+These tests are particularly valuable when re-factoring code.
+The total codebase is approximately 2,900 lines of Python.
+
 All plots are generated in vector graphics pdf format to ensure smooth scaling to any resolution.
+A consistent colour scheme of \textcolor{blue}{blue}, \textcolor{orange}{orange} and \textcolor{olive}{green} was used for plots relating to NMF, ICA and PCA respectively.
+Metagenes are consistently referred to by, for example, ``NMF-2-of-3" -- meaning the 2nd component of the rank $k=3$ NMF factorization.
 
 %%%%%%%%%%% RESULTS %%%%%%%%%%%%%
-\afterpage{\clearpage}
+\clearpage
 
@@
 
-Results of applying the above described methodology are explained and presented here.
-Deeper interpretation and discussion is deferred to the following section.
-Throughout, consistent colours of \textcolor{blue}{blue}, \textcolor{orange}{orange} and \textcolor{green}{green} are used for results relating to NMF, ICA and PCA respectively.
-Specific metagenes are referred to by, for example, ``NMF-2-of-3" -- meaning the 2nd component of the rank $k=3$ NMF factorization.
-
 % \subsection{Investigation of optimal parameters}
-\subsection{Cluster coherence analysis results}
+\subsection{Consideration of sampling error is crucial to finding robust metagene signals}
 
@@
 Figure  \ref{fig-TCGA-ica-fixed-vs-bootstrap} makes the same comparison for the lager N=374 TCGA dataset, in which it can be seen that the impact of sampling error is not so severe.
-For brevity only results for ICA are shown, but a very similar effect is observable for PCA and NMF.
-
-For this reason, and in line with our overall strategy, it is appropriate to perform all metagene extraction on the larger TCGA dataset, in the expectation that the obtained metagenes will be more robust and likely to better generalise to other datasets.
-
-\def\S{0.4}
+For brevity only results for ICA are shown here; similar result for PCA and NMF can be found in the appendix, figures
+
+For this reason it is appropriate to perform all metagene extraction on the larger TCGA dataset, in the expectation that the obtained metagenes will be more robust and likely to better generalise to other datasets.
+
+\def\S{0.45}
 \begin{figure}[ht]
@@
 
-The critical decision of the choosing the factorization rank for each method -- $K_{\mbox{\tiny NMF}}$, $K_{\mbox{\tiny ICA}}$ and $K_{\mbox{\tiny PCA}}$ -- is based on figure \ref{fig-TCGA-all-bootstrap}.
-On this bases, considering both the t-SNE plots and silhouette scores, the following choices were made:
+In choosing the factorization rank for each method, both the t-SNE plots and silhouette scrore graphs in  figure \ref{fig-TCGA-all-bootstrap} were carefully considered.  
+Firstly, $k>2$ is desirable to have sufficient information to work with.
+NMF cluster coherence seems to deteriorate after $k=3$.
+Looking at the graph of silhouette scores (figure \ref{fig-TCGA-all-bootstrap}, bottom), ICA appears to have a sweet spot at $k=5$.
+For PCA, $k = 3$ or $4$ both seem reasonable.
+The following choices were made:
 $K_{\mbox{\tiny NMF}}  =  3,
 K_{\mbox{\tiny ICA}}  =  5,
-K_{\mbox{\tiny PCA}}  =  3$.
-
-The rationale for these choices is that, firstly $k>2$ is desirable to have sufficient information to work with.
-NMF cluster coherence seems to deteriorate after $k=3$.
-Looking at the graph of silhouette scores (figure \ref{fig-TCGA-all-bootstrap}, bottom), ICA appears to have a sweet spot at $k=5$.
-For PCA, $k = 3$ or $4$ both seem reasonable; $k=3$ was selected.
+K_{\mbox{\tiny PCA}}  =  3$.  These choices predicate all the results which follow.
 
@@
 It cannot be the $180^\circ$ rotation issue discussed earlier, since this does not apply to the all +ve components.
+\emph{TODO: hypothesis?}
 
@@
 
-\afterpage{\clearpage}
-
-\subsection{Gene Enrichment (against GO) Results}
-
-Gene enrichment analysis against the GO results in, for each of the 11 metagenes, a table of GO terms with a list of the candidate genes which have an annotated association to that term.   
-To take full advantage of the hierarchical nature of the GO, results are presented as lineage maps in figures \ref{fig-NMF-go-lineage} to \ref{fig-PCA-1-3-genesiso-lineage}.  
+\subsection{Gene sets identified by metagenes of different methods intersect in some cases}
+
+The Jaccard heatmap of figure \ref{fig-component-jaccard-similarities} shows that some pairs of metagenes identify overlapping sets of genes, in particular the pairs (ICA-1-of-5, PCA-2-of-3: J=0.61) and (ICA-5-of-5, PCA-3-of-3: J=0.49).
+As expected however, there is very little similarity between components \emph{within} a factorization method.
+
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[scale=0.5]{\plotspath{TCGA_OV_VST/GeneEnrichment/component_jaccard_heatmap.pdf}}
+\end{center}
+\caption{Heatmap of Jaccard similarities between the candidate genes identified by the 11 components.}
+\label{fig-component-jaccard-similarities}
+\end{figure}
+
+\clearpage
+
+\subsection{Metagenes highlight genes which are enriched for particular biological processes}
+
+Gene enrichment analysis against the GO results in, for each of the 11 metagenes, a table of GO terms with a list of the candidate genes which have an annotated association to that term; this table can be found in the appendix, table \ref{tab-go-enrichment-results}.
+To take full advantage of the hierarchical nature of the GO, results are presented as lineage maps in figures \ref{fig-NMF-go-lineage} to \ref{fig-PCA-1-3-go-lineage}.  
 The contributing candidate genes for all significantly enriched terms are shown beneath each figure.  
-(Ideally, the specific genes would be listed alongside each term, but this is difficult to achieve graphically).   
 High-level terms at depth less than 3 were removed, since these are generic (e.g. "regulation of biological process") and so uninteresting.   
@@
 For this component it was necessary to limit the graphic to the 12 terms having the largest number of associated candidate genes.
-The full table of terms for this component is available in appendix \ref{tab-ica-4-of-5-enrichment-table}.
 
 The four components NMF-3-of-3, ICA-1-of-5, ICA-2-of-5 and PCA-2-of-3 yielded no significant enriched terms (for FDR $\leq$ 0.01).
-
-\begin{figure}[ht]
-\begin{center}
-\includegraphics[scale=0.5]{\plotspath{TCGA_OV_VST/GeneEnrichment/component_jaccard_heatmap.pdf}}
-\end{center}
-\caption{Heatmap of Jaccard similarities between the candidate genes identified by the 11 components.}
-\label{fig-component-jaccard-similarities}
-\end{figure}
-
-To what extent do the 11 components overlap in the candidate genes they propose (independent of enrichment analysis)?
-Figure \ref{fig-component-jaccard-similarities} shows this by a heatmap of Jaccard similarities.
-Jaccard similarity (or index) is defined on a pair of sets, in this case sets of genes, as:
-\[
-   \mbox{Jaccard}(A, B) = \frac{A \cap B}{A \cup B}
-\]
-lying between 0 (no similarity) and 1 (identical).
-
-From this heatmap we can see that within a factorization method the components have almost no similarity -- they represent largely independent projections, as one would expect.
-Between methods, however, there is considerable overlap in some cases.
-In particular PCA 2\_of\_3 vs ICA 1\_of\_5 (J=0.61), PCA 3\_of\_3 vs ICA 5\_of\_5 (J=0.49) and NMF 1\_of\_3 vs ICA 3\_of\_5 (J=0.39).
 
@@
 
-\afterpage{\clearpage}
-
-\subsection{Survival analysis results}
-
-Kaplan-Meier overall survival (OS) plots stratified by each metasample component are shown for the TCGA dataset in figure \ref{fig-kaplan-meier-os-TCGA-TCGA} and for the AOCS dataset in figure \ref{fig-kaplan-meier-os-TCGA-OACS}.
-Plots for progression free survival (PFS) on OACS are shown in figure \ref{fig-kaplan-meier-pfs-TCGA-OACS}.
-All plots show 95\% confidence intervals and hazard ratio (HR) with associated p-value.
-
-All of these results are summarised in figure \ref{fig-summary-survival-plots}, which brings together the three sets of results -- TCGA (OS), OACS (OS) and OACS (PFS) --  with respect to the 11 metasample components.
-$\log_2{\mbox{HR}}$ is used in order that the \emph{sense} of the survival impact can be readily appreciated.
-If a metasample component has a robust correlation with survival, then we expect the p-value for all three sets of results to show significance \emph{and} for the effects to have the same sense -- be in the same direction.
-None of the 11 components pass this test.
-
-The component closest to correlation with survival would seem to be PCA 1-of-3, showing a reasonably consistent hazard ratio of around 1.3 ($2^{0.4}$) across the three experiments.
-This has $p < 0.05$ in the larger TCGA dataset, but not in the smaller AOCS dataset.
-
-\begin{figure}[ht]
-\begin{center}
-\includegraphics[scale=0.5]{\plotspath{SurvivalAnalysis/multiple_kaplan_meier_os_TCGA_TCGA.pdf}} 
-\caption{Kaplan-Meier plots for each metasample component, stratified at the median value, for TCGA $\rightarrow$ TCGA for overall survival (OS) case.
-Hazard ratio and p-value is shown for each case.
-The final plot (bottom right) is unstratified overall survival}
-\label{fig-kaplan-meier-os-TCGA-TCGA}
-\end{center}
-\end{figure}
-
-\begin{figure}[ht]
-\begin{center}
-\includegraphics[scale=0.5]{\plotspath{SurvivalAnalysis/multiple_kaplan_meier_os_TCGA_AOCS.pdf}} 
-\caption{Kaplan-Meier plots for TCGA $\rightarrow$ TCGA for overall survival (OS).}
-\label{fig-kaplan-meier-os-TCGA-OACS}
-\end{center}
-\end{figure}
-
-\begin{figure}[ht]
-\begin{center}
-\includegraphics[scale=0.5]{\plotspath{SurvivalAnalysis/multiple_kaplan_meier_pfs_TCGA_AOCS.pdf}} 
-\caption{Kaplan-Meier plots for TCGA $\rightarrow$ TCGA for progression free survival (PFS).  }
-\label{fig-kaplan-meier-pfs-TCGA-OACS}
-\end{center}
-\end{figure}
-
-\begin{figure}[ht]
-\begin{center}
-\includegraphics[scale=0.5]{\plotspath{/SurvivalAnalysis/survival_summary_barplots_NMF.pdf}} \\
-\includegraphics[scale=0.5]{\plotspath{/SurvivalAnalysis/survival_summary_barplots_ICA.pdf}} \\
-\includegraphics[scale=0.5]{\plotspath{/SurvivalAnalysis/survival_summary_barplots_PCA.pdf}}
-\caption{Visual summary of survival analysis as applied to TCGA(OS), AOCS(OS) and AOCS(PFS).
-Plots are divided by factorization method.
-Bar heights show $\log_2(\mbox{HR})$ with p-value also shown.
-For a component to have robust correlation with survival we expect all three of the analysis bars to pass a significance threshold {\em and} for the $\log{\mbox{HR}}$ to to be {\em in the same sense\/}.
-None achieve this.}
-\label{fig-summary-survival-plots}
-\end{center}
-\end{figure}
-
-\afterpage{\clearpage}
-
-\subsection{Metasample heatmap clustering results}
-As with survival analysis, metasamples were derived from the metagenes factorized from the TCGA dataset, then applied to both TCGA and AOCS.
-The resulting heatmaps are shown in figures \ref{fig-heatmap-TCGA-TCGA} and \ref{fig-heatmap-TCGA-OACS} respectively.
-Metadata was available for the AOCS dataset, and this is presented in colour coded columns on the left of figure \ref{fig-heatmap-TCGA-OACS}.
-The meanings of each column are as follows:
-\begin{description}
-\item[WGD]: Whole genome doubling -- a marker of genome instability.
-\item[Cellularity]: proportion of cells belonging to the tumour (as opposed to surrounding normal tissue).
-\item[HR Detect]: A predictor of homologous repair deficiency based on established mutational features \cite{Ewing2020}.
-\item[Mutational Load]: A measure of the total number of mutations present in the tumour genome.
-\item[CNV Load]: Copy-number variation, i.e. deviation from the normal diploid cell compliment.
-% \item[Donor survival time]: (should probably remove).
-\end{description}
-
-The dendrogram clustering of metasamples (columns, top of figures) gives insights into the similarity of each metasample component, while the clustering of patients (rows, left of figures) offers insights into sub-populations of patients and how (in the case of AOCS only) these relate to metadata variables such as cellularity and mutational load.
-These topics will be taken up in the following discussion section.
-
-\begin{figure}
-\begin{center}
-\includegraphics[scale=0.7]{\plotspath{HeatMaps/clustered_heatmap_TCGA_TCGA.pdf}} 
-\caption{Heatmap of the metasamples matrix for the TCGA dataset, computed from metagenes factorized also TCGA.
-Clustering dendrograms (based on a correlation distance) are shown between the metasamples (columns) and patients (rows).}
-\label{fig-heatmap-TCGA-TCGA}
-\end{center}
-\end{figure}
-
-\begin{figure}[ht]
-\begin{center}
-\includegraphics[scale=0.7]{\plotspath{HeatMaps/clustered_heatmap_TCGA_AOCS.pdf}} 
-\caption{Heatmap of the metasamples matrix for the AOCS dataset, computed from metagenes factorized from the TCGA dataset.
-Several patient metadata variables are shown in columns to the left.
-(ToDo: key for metadata colouring).}
-\label{fig-heatmap-TCGA-OACS}
-\end{center}
-\end{figure}
-
-\clearpage
-
-\subsection{Metasample correlation with genomic features}
-
-\begin{figure}[ht]
-\begin{center}
-\includegraphics[scale=0.6, draft=false]{\plotspath{BoxAndScatters/wgd_boxplot_NMF_TCGA_AOCS.pdf}}
-\includegraphics[scale=0.6, draft=false]{\plotspath{BoxAndScatters/wgd_boxplot_ICA_TCGA_AOCS.pdf}}
-\includegraphics[scale=0.6, draft=false]{\plotspath{BoxAndScatters/wgd_boxplot_PCA_TCGA_AOCS.pdf}}
-\caption{Boxplots showing relationship between metasamples (derived from TCGA metagenes, applied to AOCS data) and whole genome doubling (WGD) status.
-Metasamples from NMF, ICA and PCA are shown separately.
-To allow convenient visualisation, each metasample is normalised to zero mean unit variance, hence between metasample variation is removed.}
-\end{center}
-\label{fig-wgd-boxplots}
-\end{figure}
-
-\begin{figure}[ht]
-\begin{center}
-\includegraphics[scale=0.54, draft=false]{\plotspath{BoxAndScatters/genomic_feature_scatters_TCGA_AOCS.pdf}} 
-\caption{Grid of scatter plots to visualise the correlation between metasamples and genomic features.
-In the case of WGD which has binary values (0, 1), jitter is applied for visual effect only.
-Above each plot is shown a correlation coefficient $r$ and associated p-value, highlighted with '***' where $p < 0.01$.
-In the case of WGD , the Point-Biserial correlation is used; for all others Pearson's r is used.}
-\label{fig-genomic-scatters}
-\end{center}
-\end{figure}
-
-\clearpage
-\section{Discussion}
-
-One valuable lesson from this work is the importance of accounting for sampling error, by bootstrapping,  when considering the stability of factorizations and thence the selection of factorization rank.
-Sevaral authors comment on the variability of NMF and ICA factorization due to algorithm initialisation \cite{Kairov2017,Sompairac2019,WayGregory2019}, but most neglect sampling error which we have shown to be far more significant.
-Cantini \etal\ \cite{Cantini2019a} mention bootstrapping, in reference to the BIODICA package (for ICA factorization), but it is unclear whether this is applied to the input data to model sampling error.
-In the initial stages of the current work, BIODICA was used, and was found to recommend ranks of 14, 19 or 30.
-Yet we have seen that such high ranks on the small n=80 dataset leads to highly unstable factorizations when bootstrap modelling of sampling error is included.
-
-These considerations lead to the selection of much lower ranks (fewer metagenes) than would otherwise be the case.
-
-Considering the biological significance of each of the 11 extracted metagenes:
+Considering the biological significance of each of the 11 extracted metagenes, with reference to the GO enrichment results:
 \begin{description}
@@
 %
-\item[NMF 3-of-3]: no significant biological enrichment found.
-%
-\item[ICA 1-of-5 and ICA 2-of-5]: no significant biological enrichment found.
+\item[NMF 3-of-3, ICA 1-of-5 and ICA 2-of-5]: no significant biological enrichment found.
 %
@@
 It is known that tumours have a need for increased blood supply, and that expression level of factors promoting angigenesis are associated with aggresiveness of tumour growth \cite{Nishida2006}.
-Two distinct subtypes of ovarian cancer have been identified, defined on the expression of angiogenesis related genes.
-Further, these subtypes have been found to inform clinical outcome \cite{Glass2015}.
+We have already seen (section \ref{sec-subtype-classification}, \cite{Glass2015}) there is evidence for distinct subtypes of OC distinguished by angiogenesis related genes, and that these subtypes have been found to inform clinical outcome \cite{Glass2015}.
 Thus, this component might contain useful prognostic value.
@@
 We have noted above that both NMF 1-of-3 and ICA 3-of-5 may be related to tissue type heterogeneity.
-We have also seen from metasample heatmap analysis that these two metagenes are associated with closely correlated metasamples in both the TCGA and OACS datasets, and also to have a moderately high Jaccard similarity of 0.39.
+We have also seen from metasample heatmap analysis that these two metagenes are associated with closely correlated metasamples (in AOCS), and also to have a moderately high Jaccard similarity of 0.39.
 Further, we observed that these components correlate negatively with cellularity.
@@
 
-\subsection{Research questions revisited}
-
-TODO (sorry!).
+\clearpage
+
+\subsection{Association of unsupervised gene expression patterns with patient survival is inconclusive}
+
+Kaplan-Meier overall survival (OS) plots stratified by each metasample component are shown for the TCGA dataset in figure \ref{fig-kaplan-meier-os-TCGA-TCGA} and for the AOCS dataset in figure \ref{fig-kaplan-meier-os-TCGA-AOCS}.
+Plots for progression free survival (PFS) on AOCS can be found in the appendix, figure \ref{fig-kaplan-meier-pfs-TCGA-AOCS}.
+All plots show 95\% confidence intervals and hazard ratio (HR) with associated p-value.
+
+All of these results are summarised in figure \ref{fig-summary-survival-plots}, which brings together the three sets of results -- TCGA (OS), AOCS (OS) and AOCS (PFS) --  with respect to the 11 metasample components.
+$\log_2{\mbox{HR}}$ is used in order that the \emph{sense} of the survival impact can be readily appreciated.
+If a metasample component has a robust correlation with survival, then we expect the p-value for all three sets of results to show significance \emph{and} for the effects to have the same sense -- be in the same direction.
+None of the 11 components pass this test.
+
+The component with the largest observed effect on survival is PCA 1-of-3, showing a reasonably consistent hazard ratio of around 1.3 ($2^{0.4}$) across the three experiments.
+This has $p < 0.05$ in the larger TCGA dataset, but not in the smaller AOCS dataset.
+
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[scale=0.5]{\plotspath{SurvivalAnalysis/multiple_kaplan_meier_os_TCGA_TCGA.pdf}} 
+\caption{Kaplan-Meier plots for each metasample component, stratified at the median value, for TCGA $\rightarrow$ TCGA for overall survival (OS) case.
+Hazard ratio and p-value is shown for each case.
+The final plot (bottom right) is unstratified overall survival}
+\label{fig-kaplan-meier-os-TCGA-TCGA}
+\end{center}
+\end{figure}
+
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[scale=0.5]{\plotspath{SurvivalAnalysis/multiple_kaplan_meier_os_TCGA_AOCS.pdf}} 
+\caption{Kaplan-Meier plots for TCGA $\rightarrow$ TCGA for overall survival (OS).}
+\label{fig-kaplan-meier-os-TCGA-AOCS}
+\end{center}
+\end{figure}
+
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[scale=0.5]{\plotspath{/SurvivalAnalysis/survival_summary_barplots_NMF.pdf}} \\
+\includegraphics[scale=0.5]{\plotspath{/SurvivalAnalysis/survival_summary_barplots_ICA.pdf}} \\
+\includegraphics[scale=0.5]{\plotspath{/SurvivalAnalysis/survival_summary_barplots_PCA.pdf}}
+\caption{Visual summary of survival analysis as applied to TCGA(OS), AOCS(OS) and AOCS(PFS).  Plots are divided by factorization method.  Bar heights show $\log_2(\mbox{HR})$ with p-value also shown.}
+\label{fig-summary-survival-plots}
+\end{center}
+\end{figure}
+
+\clearpage
+
+%The dendrogram clustering of metasamples (columns, top of figures) gives insights into the similarity of each metasample component, while the clustering of patients (rows, left of figures) offers insights into sub-populations of patients and how (in the case of AOCS only) these relate to metadata variables such as cellularity and mutational load.
+%These topics will be taken up in the following discussion section.
+
+\subsection{Metasample correlation with genomic features}
+
+From the grid of scatter plots shown in figure \ref{fig-genomic-scatters} it can be seen there are some significant correlations.  
+The sense (sign) of the correlation is arbitrary -- particularly for ICA and PCA -- recall normalisation w.r.t $180^\circ$ was applied.
+
+The significant correlations are summarised in table \ref{tab-genomic-scatter-summary}.  
+WGD and Cellularity show correlation with at least one component from all three methods and indeed three components correlating with WGD also correlate with Cellularity and in the same sense.  
+This might suggest simply that WGD and Cellularity are themselves correlated, but this is not so: $r=-0.14, p=0.201$ for Point-Biserial correlation (data not shown).  % Latent factor??
+HR Detect correlates with two NMF and one ICA component, not intersecting with those above, suggesting distinct biological processes are involved.
+
+The heatmap dendrogram (figure \ref{fig-heatmap-TCGA-AOCS}) shows close correlation between metasamples pairs (NMF-1-of-3, ICA-5-of-5), (ICA-1-of-5, PCA-2-of-3) and (NMF-1-of-3, ICA-3-of-5).
+
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[scale=0.5,draft=false]{\plotspath{BoxAndScatters/genomic_feature_scatters_TCGA_AOCS.pdf}} 
+\caption{Grid of scatter plots to visualise the correlation between metasamples and genomic features.
+In the case of WGD which has binary values (0, 1), jitter is applied for visual effect only.
+Above each plot is shown a correlation coefficient $r$ and associated p-value, highlighted with '***' where $p < 0.01$.
+In the case of WGD , the Point-Biserial correlation is used; for all others Pearson's $r$ is used.}
+\label{fig-genomic-scatters}
+\end{center}
+\end{figure}
+
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[scale=0.7, draft=false]{\plotspath{HeatMaps/clustered_heatmap_TCGA_AOCS.pdf}} 
+\caption{Heatmap of the metasamples matrix for the AOCS dataset, computed from metagenes factorized from the TCGA dataset.   Several patient metadata variables are shown in columns to the left.}
+\label{fig-heatmap-TCGA-AOCS}
+\end{center}
+\end{figure}
+
+\begin{table}[ht]
+\begin{center}
+\begin{tabular}{|l|l|}
+\hline
+Genomic feature & Significant correlations \\
+\hline 
+WGD & NMF-1-of-3, ICA-5-of-5, PCA-3-of-3 \\
+Cellularity & NMF-1-of-3, ICA-3-of-5, ICA-5-of-5, PCA-3-of-3 \\
+HR Detect & NMF-2-of-3, NMF-3-of-3, ICA-2-of-5 \\
+Mutational load & None \\
+CNV load & None \\
+SV load & NMF-2-of-3 \\
+\hline
+\end{tabular}
+\end{center}
+\caption{Summary of metasample components showing a significant correlation with each genomic feature.}
+\label{tab-genomic-scatter-summary}
+\end{table}
+
+\clearpage
+
+\section{Discussion}
+
+One valuable lesson from this work is the importance of accounting for sampling error, by bootstrapping,  when considering the stability of factorizations and thence the selection of factorization rank.
+Sevaral authors comment on the variability of NMF and ICA factorization due to algorithm initialisation \cite{Kairov2017,Sompairac2019,WayGregory2019}, but most neglect sampling error which we have shown to be far more significant.
+Cantini \etal\ \cite{Cantini2019a} mention bootstrapping, in reference to the BIODICA package (for ICA factorization), but it is unclear whether this is applied to the input data to model sampling error.
+In the initial stages of the current work, BIODICA was used, and was found to recommend ranks of 14, 19 or 30.
+Yet we have seen that such high ranks on the small n=80 dataset leads to highly unstable factorizations when bootstrap modelling of sampling error is included.
+
+These considerations lead to the selection of much lower ranks (fewer metagenes) than would otherwise be the case.
+
+\emph{TODO: much more discussion!}
 
@@
 \item
-{\bf Contrasting gene expression patters between cancers}.
+{\bf Contrasting gene expression patterns between cancers}.
 The current work set out to find patters of gene expression in HGSOC.
@@
 \subsection{Additional figures and plots}
+
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[scale=\S]{\plotspath{AOCS_Protein/FactorClustering/multiple_single_factors_scatter_AOCS_NMF_2_3_5_10_fixed.pdf}} \\
+\includegraphics[scale=\S]{\plotspath{AOCS_Protein/FactorClustering/multiple_single_factors_scatter_AOCS_NMF_2_3_5_10_bootstrap.pdf}}\\
+\caption{Clustering of metagenes from NMF facorizations on the AOCS dataset, comparing fixed and bootstrap sampling}
+\label{fig-AOCS-nmf-fixed-vs-bootstrap}
+\end{center}
+\end{figure}
+
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[scale=\S]{\plotspath{AOCS_Protein/FactorClustering/multiple_single_factors_scatter_AOCS_PCA_2_3_5_10_fixed.pdf}} \\
+\includegraphics[scale=\S]{\plotspath{AOCS_Protein/FactorClustering/multiple_single_factors_scatter_AOCS_PCA_2_3_5_10_bootstrap.pdf}}\\
+\caption{Clustering of metagenes from PCA facorizations on the AOCS dataset, comparing fixed and bootstrap sampling}
+\label{fig-AOCS-pca-fixed-vs-bootstrap}
+\end{center}
+\end{figure}
+
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[scale=0.5]{\plotspath{SurvivalAnalysis/multiple_kaplan_meier_pfs_TCGA_AOCS.pdf}} 
+\caption{Kaplan-Meier plots for TCGA $\rightarrow$ TCGA for progression free survival (PFS).  }
+\label{fig-kaplan-meier-pfs-TCGA-AOCS}
+\end{center}
+\end{figure}
+
 TODO - I'll likely move the AOCS survivial plots here, and perhaps some of the GO lineages.
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[scale=0.6]{\plotspath{BoxAndScatters/wgd_boxplot_NMF_TCGA_AOCS.pdf}}
+\includegraphics[scale=0.6]{\plotspath{BoxAndScatters/wgd_boxplot_ICA_TCGA_AOCS.pdf}}
+\includegraphics[scale=0.6]{\plotspath{BoxAndScatters/wgd_boxplot_PCA_TCGA_AOCS.pdf}}
+\caption{Boxplots showing relationship between metasamples (derived from TCGA metagenes, applied to AOCS data) and whole genome doubling (WGD) status.
+Metasamples from NMF, ICA and PCA are shown separately.
+To allow convenient visualisation, each metasample is normalised to zero mean unit variance, hence between metasample variation is removed.}
+\end{center}
+\label{fig-wgd-boxplots}
+\end{figure}
+
 \subsection{Gene enrichment raw results}
-\label{tab-ica-4-of-5-enrichment-table}
+\label{tab-go-enrichment-results}
 TODO

### 2020-08-11 MSc_Dissertation_B156476.tex
Added batch effect results, figs moved to appendix, FloatBarrier, split up gene enrichment figs

@@
 
-\documentclass[tikz, 12pt,a4paper,oneside,fleqn]{article}
+\documentclass[draft, tikz, 12pt,a4paper,oneside,fleqn]{article}
 
@@
 \usepackage{setspace}
-\usepackage{afterpage}
+\usepackage{placeins}
 
@@
 
-\begin{figure}[ht]
+\begin{figure}[htb!]
 \begin{center}
@@
 
-\begin{figure}[ht]
+\begin{figure}[htb!]
 \def\M{2.2cm}
@@
 
-\clearpage
+\FloatBarrier
 
@@
 \def\S{0.45}
-\begin{figure}[ht]
+\begin{figure}[htb!]
 \begin{center}
@@
 Gene enrichment analysis against the GO results in, for each of the 11 metagenes, a table of GO terms with a list of the candidate genes which have an annotated association to that term; this table can be found in the appendix, table \ref{tab-go-enrichment-results}.
-To take full advantage of the hierarchical nature of the GO, results are presented as lineage maps in figures \ref{fig-NMF-go-lineage} to \ref{fig-PCA-1-3-go-lineage}.  
+To take full advantage of the hierarchical nature of the GO, results are presented as lineage maps in figures \ref{fig-NMF_1-go-lineage} to \ref{fig-PCA-3-3-go-lineage}.  
 The contributing candidate genes for all significantly enriched terms are shown beneath each figure.  
@@
 
-\begin{figure}[ht]
-\begin{center}
-\includegraphics[scale=0.14]{\plotspath{TCGA_OV_VST/GeneEnrichment/go_lineage_1_of_NMF_3.png}} \\
+\begin{figure}[htb!]
+\begin{center}
+\includegraphics[scale=0.1]{\plotspath{TCGA_OV_VST/GeneEnrichment/go_lineage_1_of_NMF_3.png}} \\
 \begin{spacing}{0.6}
@@
 {\small Component NMF-1-of-3} \\
-\vspace{1cm}
-\includegraphics[scale=0.14]{\plotspath{TCGA_OV_VST/GeneEnrichment/go_lineage_2_of_NMF_3.png}} \\
+\end{center}
+\caption{Lineage maps of enriched Gene Ongology (GO) terms for components NMF-1-of-3  In these diagrams, enriched terms are coloured purple, while there ancestors in the ontology are yellow.  }
+\label{fig-NMF_1-go-lineage}
+\end{figure}
+
+\begin{figure}[htb!]
+\begin{center}
+\includegraphics[scale=0.1]{\plotspath{TCGA_OV_VST/GeneEnrichment/go_lineage_2_of_NMF_3.png}} \\
 \begin{spacing}{0.6}
@@
 {\small Component NMF 2-of-3} \\
-\caption{Lineage maps of enriched Gene Ongology (GO) terms for components NMF-1-of-3 and NMF-2-of-3.
-In these diagrams, enriched terms are coloured purple, while there ancestors in the ontology are yellow.
-(NMF-3-of-3 produced no significant enrichment results)}
-\label{fig-NMF-go-lineage}
-\end{center}
-\end{figure}
-
-\begin{figure}[ht]
+\caption{Lineage maps of enriched GO terms for components NMF-2-of-3.   (NMF-3-of-3 produced no significant enrichment results)}
+\label{fig-NMF_2-go-lineage}
+\end{center}
+\end{figure}
+
+\begin{figure}[htb!]
 \begin{center}
@@
 {\small Component ICA-3-of-5} \\
-\vspace{1cm}
+\end{center}
+\caption{Lineage maps of enriched GO terms for components ICA-3-of-5}
+\label{fig-ICA-3-5-go-lineage}
+\end{figure}
+
+\begin{figure}[htb!]
+\begin{center}
 \includegraphics[scale=0.07]{\plotspath{TCGA_OV_VST/GeneEnrichment/go_lineage_5_of_ICA_5.png}} \\
@@
 {\small Component ICA-5-of-5} \\
-\caption{Lineage maps of enriched Gene Ongology (GO) terms for components ICA-3-of-5 and ICA-5-of-5}
-\label{fig-ICA-3-5-go-lineage}
-\end{center}
-\end{figure}
-
-\begin{sidewaysfigure}[ht]
+\caption{Lineage maps of enriched GO terms for components ICA-5-of-5}
+\label{fig-ICA-5-5-go-lineage}
+\end{center}
+\end{figure}
+
+\begin{sidewaysfigure}[htb!]
 \begin{center}
@@
 \vspace{0.5cm}
-\caption{Lineage maps of enriched Gene Ongology (GO) terms for component ICA-4-of-5. For this component it was necessary to additionally filter the enriched terms -- see text.}
+\caption{Lineage maps of enriched GO terms for component ICA-4-of-5. For this component it was necessary to additionally filter the enriched terms -- see text.}
 \label{fig-ICA-4-go-lineage}
@@
 
-\begin{figure}[ht]
-\begin{center}
-\includegraphics[width=0.5 \textwidth]{\plotspath{TCGA_OV_VST/GeneEnrichment/go_lineage_1_of_PCA_3.png}} \\
+\begin{figure}[htb!]
+\begin{center}
+\includegraphics[width=0.4 \textwidth]{\plotspath{TCGA_OV_VST/GeneEnrichment/go_lineage_1_of_PCA_3.png}} \\
 \begin{spacing}{0.6}
@@
 {\small Component PCA-1-of-3} \\
-\vspace{1cm}
+\end{center}
+\caption{Lineage maps of enriched GO terms for component PCA-1-of-3  }
+\label{fig-PCA-1-3-go-lineage}
+\end{figure}
+
+\begin{figure}[htb!]
+\begin{center}
 \includegraphics[width= \textwidth]{\plotspath{TCGA_OV_VST/GeneEnrichment/go_lineage_3_of_PCA_3.png}} \\
@@
 {\small Component PCA-3-of-3} \\
-\caption{Lineage maps of enriched Gene Ongology (GO) terms for components 1 and 3 from the rank=5 PCA factorization. (Component PCA-2-of-3 produced no significant enrichment results)  }
-\label{fig-PCA-1-3-go-lineage}
-\end{center}
+\end{center}
+\caption{Lineage maps of enriched GO terms for component PCA-3-of-3. (Component PCA-2-of-3 produced no significant enrichment results)  }
+\label{fig-PCA-3-3-go-lineage}
+
 \end{figure}
@@
 
-\begin{figure}[ht]
-\begin{center}
-\includegraphics[scale=0.5]{\plotspath{SurvivalAnalysis/multiple_kaplan_meier_os_TCGA_TCGA.pdf}} 
-\caption{Kaplan-Meier plots for each metasample component, stratified at the median value, for TCGA $\rightarrow$ TCGA for overall survival (OS) case.
-Hazard ratio and p-value is shown for each case.
-The final plot (bottom right) is unstratified overall survival}
-\label{fig-kaplan-meier-os-TCGA-TCGA}
-\end{center}
-\end{figure}
-
-\begin{figure}[ht]
+\begin{figure}[htb!]
 \begin{center}
@@
 
-\begin{table}[ht]
+\begin{table}[htb!]
 \begin{center}
@@
 
+\FloatBarrier
+
+\section{Matrix factorization discovers and effectively removes batch effects}
+
+Here are presented results of applying the three MF methods to the combined AOCS and TCGA datasets.
+
+\begin{figure}[htb!]
+\begin{center}
+\includegraphics[scale=0.7, draft=false]{\plotspath{HeatMaps/clustered_heatmap_BOTH_BOTH.pdf}} 
+\caption{Heatmap of the metasamples matrix for the combined AOCS + TCGA dataset.  The 'Which' field indicates the source dataset; purple: AOCS, yellow:TCGA.}
+\label{fig-heatmap-BOTH}
+\end{center}
+\end{figure}
+
+\begin{figure}[htb!]
+\begin{center}
+\includegraphics[scale=0.6]{\plotspath{BoxAndScatters/wgd_boxplot_NMF_BOTH_BOTH.pdf}}
+\includegraphics[scale=0.6]{\plotspath{BoxAndScatters/wgd_boxplot_ICA_BOTH_BOTH.pdf}}
+\includegraphics[scale=0.6]{\plotspath{BoxAndScatters/wgd_boxplot_PCA_BOTH_BOTH.pdf}}
+\caption{Boxplots of metasamples from the combined AOCS + TCGA datasets, showing relationship between metasamples dataset origin; Which=0 or 1 for AOCS or TCGA respectively.}
+\end{center}
+\label{fig-wgd-boxplots-BOTH}
+\end{figure}
+
+From the clustered heatmap of figure \ref{fig-heatmap-BOTH} it is clear that the five components clustered to the left are highly sensitive to the batch effect.
+
 \clearpage
@@
 
+\small
 \bibliography{OvarianCancer}
@@
 
-\begin{figure}[ht]
+\begin{figure}[htb!]
+\begin{center}
+\includegraphics[scale=0.5]{\plotspath{SurvivalAnalysis/multiple_kaplan_meier_os_TCGA_TCGA.pdf}} 
+\caption{Kaplan-Meier plots for each metasample component, stratified at the median value, for TCGA $\rightarrow$ TCGA for overall survival (OS) case.
+Hazard ratio and p-value is shown for each case.
+The final plot (bottom right) is unstratified overall survival}
+\label{fig-kaplan-meier-os-TCGA-TCGA}
+\end{center}
+\end{figure}
+
+\begin{figure}[htb!]
 \begin{center}
@@
 
-TODO - I'll likely move the AOCS survivial plots here, and perhaps some of the GO lineages.
-\begin{figure}[ht]
+\begin{figure}[htb!]
 \begin{center}
@@
 
+\begin{figure}[htb!]
+\begin{center}
+\includegraphics[scale=\S]{\plotspath{BOTH_AOCS_TCGA/FactorClustering/multiple_single_factors_scatter_BOTH_NMF_2_3_5_10_bootstrap.pdf}} \\
+\includegraphics[scale=\S]{\plotspath{BOTH_AOCS_TCGA/FactorClustering/multiple_single_factors_scatter_BOTH_ICA_2_3_5_10_bootstrap.pdf}} \\
+\includegraphics[scale=\S]{\plotspath{BOTH_AOCS_TCGA/FactorClustering/multiple_single_factors_scatter_BOTH_PCA_2_3_5_10_bootstrap.pdf}} \\
+%\includegraphics[scale=0.55]{\plotspath{/BOTH_AOCS_TCGA/FactorClustering/silhouette_plots_TCGA_2_3_4_5_6_7_8_9_10_bootstrap.pdf}} 
+\caption{Metagene clustering for all three methods applied to the combined N=80+374 AOCS + TCGA dataset with bootstrap sampling, over a range of factorization ranks.
+The silhouette scores are also plotted (bottom).
+}
+\label{fig-BOTH-all-bootstrap}
+\end{center}
+\end{figure}
+
+\begin{figure}[htb!]
+\begin{center}
+\includegraphics[scale=0.5]{\plotspath{SurvivalAnalysis/multiple_kaplan_meier_os_BOTH_BOTH.pdf}} 
+\caption{Kaplan-Meier plots for each metasample component, stratified at the median value, for the combined AOCS + TCGA dataset}
+\label{fig-kaplan-meier-os-BOTH}
+\end{center}
+\end{figure}
+
+\begin{figure}[htb!]
+\begin{center}
+\includegraphics[scale=0.5,draft=false]{\plotspath{BoxAndScatters/genomic_feature_scatters_BOTH_AOCS.pdf}} 
+\caption{Grid of scatter plots to visualise the correlation between metasamples derived from the metagenes of the combined AOCS + TCGA datasets, and genomic features of the AOCS.}
+\label{fig-genomic-scatters-BOTH-AOCS}
+\end{center}
+\end{figure}
+
 \subsection{Gene enrichment raw results}

### 2020-08-11 MSc_Dissertation_B156476.tex
Added integrative tables for main results and for combined ('BOTH') experiments; discussion of integrative results; other miprovement; some figures removed

@@
 
+\setlength{\parskip}{0em}
 \tableofcontents
-
+\setlength{\parskip}{1em}
 \newpage
@@
 Scatter plots and heat maps to investigate the relationships between metasamples and genomic features available for the AOCS dataset.
+\item
+Investigation of batch effects by repeating of metagene extraction and heatmaps on the \emph{combined} AOCS and TCGA datasets.
 \end{enumerate}
@@
 \begin{itemize}
+\itemsep -0.5em 
 \item Python 3.6
@@
 \end{itemize}
-
 A full list of software and libraries used, with versions, is given in appendix \ref{sec-software-versions}.
 
-\subsection{Gene set intersection}
+\subsection{Syncrhonisation of gene sets accross AOCS and TCGA}
 \label{sec:gene-set-intersection}
@@
 
-\subsection{Metagene selection by cluster coherence}
+\subsection{Metagene extraction by cluster coherence}
 \label{sec-metagene-selection}
@@
 \subsection{Investigating correlation between metasamples and genomic features}
+\label{sec-genomic-features}
 
@@
 
+\subsection{Investigating the impact of batch effects by analysis of a combined dataset}
+
+It has been claimed (e.g.
+\cite{Stein-OBrien2018}) that matrix factorization methods are an effective means of identifying and removing batch effects.
+To verify this, the TCGA and AOCS datasets (after gene set intersection), were horizontally concatenated producing an $n=80+274=354$ patient dataset with a substantial batch artefact.
+Survival data is available for both datasets, so this was preserved as metadata associated with the combined datasets.
+The previously described analyses of metagene extraction, metasample computation and clustered heatmap and survival analysis were repeated on this dataset.
+Ideally, the factorization rank selected for each method should be reconsidered, since the larger dataset could justify extracting more components.
+For simplicity this was not done however, the previously selected $k = 3,5,3$ for NMF, ICA and PCA respectively being maintained.
+Note however that we should not expect the extracted metagenes and derived metasamples to be equivalent.
+Genomic feature correlation w.r.t these newly extracted components was performed using the AOCS datasets.
+Theses results were inspected; is there evidence of some metasamples aligning with technical batch effects while others align with biological processes?
+
 \subsection{Codebase and plotting conventions}
@@
 Good software engineering practice has been followed, with an object-oriented design, frequent commits and unit testing.
-Unit tests are based on tiny expression matrices, just 100 genes by 10 patients, thus the whole test suite of over 60 tests executes in around 40 seconds and can be run frequently.
-These tests are particularly valuable when re-factoring code.
+A base class is used to present a uniform interface for the three factorization methods.
+Intermediate results are cached with lazy computation for efficient working without ordering dependency.
+Unit tests are based on tiny expression matrices, just 100 genes by 10 patients, thus the whole test suite of over 65 tests executes in around 45 seconds and so can be run frequently.
 The total codebase is approximately 2,900 lines of Python.
@@
 All plots are generated in vector graphics pdf format to ensure smooth scaling to any resolution.
-A consistent colour scheme of \textcolor{blue}{blue}, \textcolor{orange}{orange} and \textcolor{olive}{green} was used for plots relating to NMF, ICA and PCA respectively.
+A consistent colour scheme of \textcolor{blue}{blue}, \textcolor{orange}{orange} and \textcolor{olive}{green} is used for plots relating to NMF, ICA and PCA respectively.
 Metagenes are consistently referred to by, for example, ``NMF-2-of-3" -- meaning the 2nd component of the rank $k=3$ NMF factorization.
@@
 
-For this reason it is appropriate to perform all metagene extraction on the larger TCGA dataset, in the expectation that the obtained metagenes will be more robust and likely to better generalise to other datasets.
+For this reason it is was decided to perform all metagene extraction on the larger TCGA dataset, in the expectation that the obtained metagenes will be more robust and likely to better generalise to other datasets.
 
@@
 
-\subsection{Gene sets identified by metagenes of different methods intersect in some cases}
+\subsection{Jaccard similarity shows metagenes identify intersecting gene sets in some cases}
 
@@
 %
-\item[ICA 5-of-5]: processes relating to regulation of immune response are enriched in this component, featuring genes from the major histocompatibility complex group, in particular the HLA- genes, allowing the immune system to recognise self from non-self.
+\item[ICA 5-of-5]: processes relating to regulation of immune response are enriched in this component, featuring genes from the major histocompatibility complex (MHC) group, in particular the HLA- genes, allowing the immune system to recognise self from non-self.
 GO terms relating to the endoplasmic reticulum (ER) are also highlighted.
@@
 
-We have noted above that both NMF 1-of-3 and ICA 3-of-5 may be related to tissue type heterogeneity.
-We have also seen from metasample heatmap analysis that these two metagenes are associated with closely correlated metasamples (in AOCS), and also to have a moderately high Jaccard similarity of 0.39.
-Further, we observed that these components correlate negatively with cellularity.
-This is as one would expect: low cellularity implies a mixture of tumour and non-tumour cell types, resulting in variation of ECM composition.
-
-Interestingly, the above summary of the gene enrichment results tentatively suggests component PCA 1-of-3 is most likely to have clinical prognostic value, based on a cursory look at the literature.
-This perhaps supports an earlier observation that, while not conclusive, component PCA-1-of-3 comes closest to demonstrating correlation with patient survival (section \ref{sec-survival-analysis}, figure \ref{fig-summary-survival-plots}).
-
-How should we interpret those four components which demonstrate no significant enrichment with biological meaning?
-It is notable that all three methods have at least one such component.
-Among these four, PCA 1-of-3 and ICA 1-of-5 have high (the highest) Jaccard similarity of 0.61.
-For all other pairs among these four Jaccard similarity is negligible.
-Perhaps these two components are picking up on similar technical variation between the samples?
-
 \FloatBarrier
@@
 
-\FloatBarrier
-
 %The dendrogram clustering of metasamples (columns, top of figures) gives insights into the similarity of each metasample component, while the clustering of patients (rows, left of figures) offers insights into sub-populations of patients and how (in the case of AOCS only) these relate to metadata variables such as cellularity and mutational load.
@@
 
-\subsection{Metasample correlation with genomic features}
-
-From the grid of scatter plots shown in figure \ref{fig-genomic-scatters} it can be seen there are some significant correlations.  
-The sense (sign) of the correlation is arbitrary -- particularly for ICA and PCA -- recall normalisation w.r.t $180^\circ$ was applied.
-
-The significant correlations are summarised in table \ref{tab-genomic-scatter-summary}.  
-WGD and Cellularity show correlation with at least one component from all three methods and indeed three components correlating with WGD also correlate with Cellularity and in the same sense.  
-This might suggest simply that WGD and Cellularity are themselves correlated, but this is not so: $r=-0.14, p=0.201$ for Point-Biserial correlation (data not shown).  % Latent factor??
-HR Detect correlates with two NMF and one ICA component, not intersecting with those above, suggesting distinct biological processes are involved.
-
-The heatmap dendrogram (figure \ref{fig-heatmap-TCGA-AOCS}) shows close correlation between metasamples pairs (NMF-1-of-3, ICA-5-of-5), (ICA-1-of-5, PCA-2-of-3) and (NMF-1-of-3, ICA-3-of-5).
+\subsection{Metasamples correlate with genomic features}
+
+From the grid of scatter plots shown in figure \ref{fig-genomic-scatters} it can be seen there are some significant correlations at play.  These are summarised and analysed in the next section.
 
@@
 
-\begin{figure}[htb!]
-\begin{center}
-\includegraphics[scale=0.7, draft=false]{\plotspath{HeatMaps/clustered_heatmap_TCGA_AOCS.pdf}} 
-\caption{Heatmap of the metasamples matrix for the AOCS dataset, computed from metagenes factorized from the TCGA dataset.   Several patient metadata variables are shown in columns to the left.}
-\label{fig-heatmap-TCGA-AOCS}
-\end{center}
-\end{figure}
+\FloatBarrier
+
+\subsection{Integrative analysis of results}
+
+Table \ref{tab-integrative-analysis} summarises the key results of all the forgoing analysis.
 
@@
 \begin{center}
-\begin{tabular}{|l|l|}
+\begingroup
+\scriptsize
+\renewcommand{\arraystretch}{1.4} % Default value: 1
+\begin{tabular}{|l||c|c|c|c|c|c|c|l|l|}
 \hline
-Genomic feature & Significant correlations \\
+Metagene & S'val & WGD & Cell. & HRD & Mut.L. & CN.L. & SV.L. & Enriched GO terms & Similar to\\
 \hline
-WGD & NMF-1-of-3, ICA-5-of-5, PCA-3-of-3 \\
-Cellularity & NMF-1-of-3, ICA-3-of-5, ICA-5-of-5, PCA-3-of-3 \\
-HR Detect & NMF-2-of-3, NMF-3-of-3, ICA-2-of-5 \\
-Mutational load & None \\
-CNV load & None \\
-SV load & NMF-2-of-3 \\
+%     & S & W & C & H % M & C & S & Enriched % Sim.
+NMF-1 & . & Y-& Y-& . & . & . & . & Extra-cellular matrix (ECM) & ICA-3, PCA-1 \\
+NMF-2 & . & . & . & Y+& . & . & Y+& Ribo. subunit & - \\
+NMF-3 & - & - & - & Y-& . & . & . & . & . \\
+\hline
+ICA-1 & . & . & . & . & . & . & . & . & PCA-2 \\
+ICA-2 & . & . & . & . & . & . & . & . & . \\
+ICA-3 & . & . & Y-& . & . & . & Y-& ECM, multicellular org. & NMF-1, PCA-1 \\
+ICA-4 & . & . & . & . & . & . & . & Ribo subunit, Mitoch. NADH dehydr. & - \\
+ICA-5 & . & Y+& Y+& . & . & . & . & Immune response, MHC & PCA-3 \\
+\hline
+PCA-1 & Y+& . & Y-& . & . & . & . & ECM, Reg. angiogenesis & NMF-1, ICA-3 \\
+PCA-2 & . & . & . & . & . & . & . & . & ICA-1 \\
+PCA-3 & . & Y-& Y-& . & . & . & . & Immune response, MHC, chemotaxis & ICA-5 \\
 \hline
 \end{tabular}
-\end{center}
-\caption{Summary of metasample components showing a significant correlation with each genomic feature.}
-\label{tab-genomic-scatter-summary}
+\endgroup
+\end{center}
+\caption{For each metagenes a 'Y' indicates that a significant correlation was observed with: Survival, WGD, HRD, Mutational load,  CNV load and SV load.
+The next column gives a terse summary of the gene enrichment analysis results.
+The final column lists the metagenes which are apparently similar -- having a Jaccard similarity of $\geq 0.35$.}
+\label{tab-integrative-analysis}
 \end{table}
-
-\FloatBarrier
-
-\section{Matrix factorization discovers and effectively removes batch effects}
-
-Here are presented results of applying the three MF methods to the combined AOCS and TCGA datasets.
+From this summary table we can make some interesting observations:
+\begin{description}
+\item[ECM relationship to Cellularity?]
+We see that NMF-1, ICA-3 and PCA-1 each have an ECM association and we have noted this could be related to tissue type heterogeneity.
+These three components (one from each of the methods) are also observed to be correlated with Cellularity.
+The sense (+/-1) of the correlation is meaningful for NMF (not so for ICA or PCA) and we see that ECM and thence tissue heterogeneity is negatively correlated with cellularity.
+This seems reasonable, in that a low Cellularity implies that a greater proportion of surrounding non-tumour tissue is included in the sample, so increasing heterogeneity.
+That these three components turn out to be similar by the Jaccard index is as would be expected given their functional similarity.
+\item[Angiogenesis link with survival?]
+The only component showing (albeit weak) correlation with patient survival is PCA-1.
+Interestingly, this same component is also found to be linked with Go terms relating to the regulation of angiogenesis, which in turn is related to aggressiveness of tumour growth and outcomes.
+Thus, these observations are telling a consistent story.
+\item[Normal cell activity?]
+Component ICA-4 highlights the greatest number of biological processes -- indeed the GO lineage map \ref{fig-ICA-4-go-lineage} needed to be pruned to fit on a page.
+And yet this component shows no correlation with any disease related factors, neither survival nor genomic features.
+It may be that this component is simply picking up on the normal metabolic activity of the cells,
+\item[Technical batch effects?]
+There are four components (NMF-3, ICA-2, ICA-4 and PCA-2) showing no enrichment of GO terms.
+These components also show almost no correlation with disease related factors (the exception being NMF-3 linking with HRD).
+With that exception aside, we could conclude that these components have little biological meaning, and are instead picking up technical effects.
+I do not know how the TCGA dataset was assembled; perhaps there are multiple batches within the set?
+\item[WGD and Cellularity have curious common influences.]
+The three components NMF-1, ICA-5 and PCA-3 (one from each method) correlate with both WGD and Cellularity, and in the same sense. 
+This might suggest simply that WGD and Cellularity are themselves positively correlated, but this is not so: $r=-0.14, p=0.201$ for Point-Biserial correlation (data not shown).
+Neither is there any close similarity between the three components by Jaccard, although ICA-5 and PCA-3 are both linked to immune response and MHC.
+Is there some curious non-linear interaction?
+\item[Mutational load CNV load are not captured by these components.]  This is somewhat surprising, perhaps the link would emerge with more extracted components.
+\end{description}
+
+\subsection{Matrix factorization discovers and effectively removes batch effects}
+
+Here are presented results of applying the three MF methods to the \emph{combined} AOCS and TCGA datasets.
 
@@
 From the clustered heatmap of figure \ref{fig-heatmap-BOTH} it is clear that the five components clustered to the left are highly sensitive to the batch effect.
+
+\begin{table}[htb!]
+\begin{center}
+\begingroup
+\scriptsize
+\renewcommand{\arraystretch}{1.4} % Default value: 1
+\begin{tabular}{|l||c|c|c|c|c|c|c|c|l|l|}
+\hline
+Metagene & Which & WGD & Cell. & HRD & Mut.L. & CN.L. & SV.L. & Enrichment? \\
+\hline
+%     & w & W & C & H % M & C & S & Enrichment?
+NMF-1 & Y & . & . & . & . & . & . &  Y \\
+NMF-2 & Y & Y & Y & . & . &   & . &  Y \\
+NMF-3 & . & Y & Y & . & . & . & . &  Y \\
+\hline
+ICA-1 & Y & . & . & . & . & . & . &  . \\
+ICA-2 & Y & . & . & . & . & . & . &  . \\
+ICA-3 & . & . & . & . & . &   & . &  Y \\
+ICA-4 & . & Y & Y & . & . & . & . &  Y \\
+ICA-5 & . & . & . & . & . & . & . &  Y \\
+\hline
+PCA-1 & . & Y & Y & . & . & . & . &  Y \\
+PCA-2 & Y & . & . & . & . & . & . &  . \\
+PCA-3 & Y & . & . & . & . & . & . &  . \\
+\hline
+\end{tabular}
+\endgroup
+\end{center}
+\caption{In a similar format as table \ref{tab-integrative-analysis}, this summarises results of in respect of the \emph{combined AOCS and TCGA} datasets.
+The key point here is that a known batch effect has been introduced, reflected in the 'Which' column.
+Note that since these metagenes have been re-extracted from different data, they are distinct from those analysed in the previous section}
+\label{tab-integrative-analysis-BOTH}
+\end{table}
 
@@
 This should result in many (most?) metagene elements reducing to zero, and might offer a more robust means of selecting candidate genes to feed into gene enrichment analysis -- replacing the current arbitrary 3 SD from the mean rule.
-\item
-{\bf Discovering batch effects}.
-It has been claimed (e.g.
-\cite{Stein-OBrien2018}) that matrix factorization methods are an effective means of identifying and removing batch effects.
-This could be explored by horizontally concatenating the TCGA and AOCS datasets (after gene set intersection), producing an $n=80+274=354$ dataset with a substantial batch artefact.
-Factorizations of this combined dataset should be more robust because of the larger $n$.
-Does each of NMF, ICA and PCA naturally produce a component which correlates with batch?
 \item
@@
 \begin{center}
-\includegraphics[scale=0.6]{\plotspath{BoxAndScatters/wgd_boxplot_NMF_TCGA_AOCS.pdf}}
-\includegraphics[scale=0.6]{\plotspath{BoxAndScatters/wgd_boxplot_ICA_TCGA_AOCS.pdf}}
-\includegraphics[scale=0.6]{\plotspath{BoxAndScatters/wgd_boxplot_PCA_TCGA_AOCS.pdf}}
-\caption{Boxplots showing relationship between metasamples (derived from TCGA metagenes, applied to AOCS data) and whole genome doubling (WGD) status.
-Metasamples from NMF, ICA and PCA are shown separately.
-To allow convenient visualisation, each metasample is normalised to zero mean unit variance, hence between metasample variation is removed.}
-\end{center}
-\label{fig-wgd-boxplots}
+\includegraphics[scale=0.7]{\plotspath{HeatMaps/clustered_heatmap_TCGA_AOCS.pdf}} 
+\caption{Heatmap of the metasamples matrix for the AOCS dataset, computed from metagenes factorized from the TCGA dataset.   Several patient metadata variables are shown in columns to the left.}
+\label{fig-heatmap-TCGA-AOCS}
+\end{center}
 \end{figure}
@@
 \begin{center}
-\includegraphics[scale=0.5]{\plotspath{SurvivalAnalysis/multiple_kaplan_meier_os_BOTH_BOTH.pdf}} 
-\caption{Kaplan-Meier plots for each metasample component, stratified at the median value, for the combined AOCS + TCGA dataset}
-\label{fig-kaplan-meier-os-BOTH}
-\end{center}
-\end{figure}
-
-\begin{figure}[htb!]
-\begin{center}
 \includegraphics[scale=0.5,draft=false]{\plotspath{BoxAndScatters/genomic_feature_scatters_BOTH_AOCS.pdf}} 

### 2020-08-12 MSc_Dissertation_B156476.tex
Added section on MF comparative research, and completed research gaps and questions

@@
 
-\documentclass[tikz, 12pt,a4paper,oneside,fleqn]{article}
+\documentclass[draft,tikz, 12pt,a4paper,oneside,fleqn]{article}
 
@@
 \bf
-Disentangling patterns of gene expression in high grade serous ovarian cancer
+Disentangling patterns of gene expression in high grade serous ovarian cancer by matrix factorization
 \vspace{0.2in}
@@
 \label{sec-subtype-classification}
-
-A relatively early (2003) expression microarray based work demonstrated clustering into lymphocyte related genes (IGH3, IGKC, IGLJ3), extracellular matrix/stromal related (COL11A1, COL3A1, MMP2, SPARC, RBBP1) and six other clusters \cite{Schaner2003}.
 
@@
 The main difference in the existing approaches consists in what is considered as source signal matrix in the decomposition" \cite{Cantini2019}.
-According to Cantini \etal\, references \cite{Au-Yeung2014}, \cite{Kairov2017}, \cite{Kong2008} and \cite{Lee2003} optimise metagenes, while references \cite{Meng2016} and \cite{Barillot2013} optimises metasamples.
+According to \cite{Cantini2019}, references \cite{Au-Yeung2014,Kairov2017,Kong2008,Lee2003},  optimise metagenes, while references \cite{Meng2016,Barillot2013} optimises metasamples.
 
@@
 
-\subsection{Research goals}
-This work is concerned with analysing the patterns gene expression in HGSOC by dimensionality reduction techniques of matrix factorization, and relating these patterns to clinical and genomic features.
-
-The following research questions were set out early in the project. \emph{TODO: tweak these goals!}
+\subsubsection{Comparative studies of dimensionality reduction methods for gene expression analysis}
+
+Recent work by Way \etal\ \cite{WayGregory2019} was aimed directly at answering the question which dimensionality reduction method was best suited to gene expression data.
+They considered PCA, ICA, NMF and also non-linear neural network methods -- directional autoencoders  and variational autoencoders .
+Briefly, these latter methods work by training a neural network to reconstruct an input signal while being forced through a ``bottleneck'' in dimensionality.
+Their conclusion was that no single method was optimal, but best performance -- judged by capturing the most biological pathway-associated features -- was achieved by taking the extracted latent dimensions of all the methods, an approach they call the ``BioBombe''.
+
+Meng \etal\ \cite{Meng2016} extensively review dimensionality reduction methods applied to multi-omic datasets.
+PCA, ICA and NMF are discussed, but the main focus is on less well known algorithms applicable to multiple datasets, including multiple coinertia analysis and consensus PCA.
+The essential purpose of these methods is to transform diverse datasets -- transcriptomics, genomics and metabolomics  -- into a common and manageable representation space.
+
+The question of reproducibility across datasets of matrix factorization methods is considered by Cantini \etal.
+\cite{Cantini2019}.
+They experiment with several MF methods applied to a total of 26 cancer transcriptomic datasets (including ovarian).
+They conclude the best approach to be ICA, stabilized by the use of Reciprocally Best Hit graphs; in essence requiring reciprocal identification of metagenes between at least two datasets.
+
+\subsection{Research gaps and questions}
+Considering the foregoing review of the literature, there has clearly been much work by large teams of researchers directed at understanding the gene expression patterns in HGSOC, as one would expect given its clinical importance.
+Dimensionality reduction is a huge and active field -- the bread-and-butter of machine learning research. 
+As we have seen, conclusions both on the underlying biology and the optimal methodology are far from consistent, however.
+Some approaches -- such as the BioBombe reviewed above -- are complex with results likely to be difficult to reproduce.
+
+Robustness of research results is of course fundamental.
+MF methods offer a means of \emph{unsupervised} learning on transcriptomic datasets to identify key genes active in a disease process.
+Would two teams working with identical data and using similar MF methods obtain similar metagenes and so arrive at the same biological conclusions?
+The \emph{stability} of MF methods is well considered by many authors, although this has typically been focussed on stability of the underlying algorithm -- NMF and ICA are stochastic in nature.
+Much less considered is stability in the presence of \emph{sampling error} which inevitably exists particularly in small datasets.
+
+These and wider considerations suggest the following research questions.
+
 \begin{enumerate}
-\item Which are the influential genes in HGSOC according to gene expression data?   Do these confirm published results?
-\item Which of PCA, ICA and NMF is best suited to this analysis?
-\item For identified genes, what is the underlying biology?
-\item Can we discriminate subtypes of HGSOC from expression data?
-\item Is there value in incorporating somatic and germline genomic data?
+\item Which are the influential genes, or groups of genes, active in HGSOC according to gene expression data?   
+\item How do these patterns of gene expression relate to underlying biological processes.  Do these confirm published results?
+\item Which MF method is best suited to this analysis and how should they be applied to achieve robust results in the presence of limited dataset size?
+\item Do MF methods yield factors which are predictive of patient survival?
+\item How do the patterns of gene expression uncovered relate to genome level features, such as those marking genome instability?
 \end{enumerate}
@@
 
-\subsubsection{Method outline}
+\subsubsection{Methods}
 An overview of the methodology adopted in this work is shown in figure \ref{fig-method-overview}, consisting of:
@@
 Scatter plots and heat maps to investigate the relationships between metasamples and genomic features available for the AOCS dataset.
-\item
-Investigation of batch effects by repeating of metagene extraction and heatmaps on the \emph{combined} AOCS and TCGA datasets.
+
 \end{enumerate}
+
+These steps are detailed in the following sections.
 
@@
 
-\subsection{Metagene extraction by cluster coherence}
+\subsection{Metagene extraction by cluster coherence under bootstrap resampling}
 \label{sec-metagene-selection}
@@
        for j in 1..50:
-          X = bootstrap sample N from N
+          X = bootstrap sample N from N with replacement
           W, H = factorizer(X, rank=k, seed=j)
@@
 \]
-The measure lies between 0 (no similarity) and 1 (identical), and is symmetric.  Jaccard similarity was thus calculated for all pairs of the 11 genes sets and visualised as a heatmap.
+The measure lies between 0 (no similarity) and 1 (identical), and is symmetric.  Jaccard similarity was thus calculated for all pairs of the 11 genes sets and visualised as a square heatmap.
 
@@
 \begin{center}
-\includegraphics[scale=0.7, draft=false]{\plotspath{HeatMaps/clustered_heatmap_BOTH_BOTH.pdf}} 
+\includegraphics[scale=0.5
+, draft=false]{\plotspath{HeatMaps/clustered_heatmap_BOTH_BOTH.pdf}} 
 \caption{Heatmap of the metasamples matrix for the combined AOCS + TCGA dataset.  The 'Which' field indicates the source dataset; purple: AOCS, yellow:TCGA.}
@@
 \begin{center}
-\includegraphics[scale=0.6]{\plotspath{BoxAndScatters/wgd_boxplot_NMF_BOTH_BOTH.pdf}}
-\includegraphics[scale=0.6]{\plotspath{BoxAndScatters/wgd_boxplot_ICA_BOTH_BOTH.pdf}}
-\includegraphics[scale=0.6]{\plotspath{BoxAndScatters/wgd_boxplot_PCA_BOTH_BOTH.pdf}}
+\includegraphics[scale=0.4]{\plotspath{BoxAndScatters/wgd_boxplot_NMF_BOTH_BOTH.pdf}}
+\includegraphics[scale=0.4]{\plotspath{BoxAndScatters/wgd_boxplot_ICA_BOTH_BOTH.pdf}}
+\includegraphics[scale=0.4]{\plotspath{BoxAndScatters/wgd_boxplot_PCA_BOTH_BOTH.pdf}}
 \caption{Boxplots of metasamples from the combined AOCS + TCGA datasets, showing relationship between metasamples dataset origin; Which=0 or 1 for AOCS or TCGA respectively.}
@@
 In this way, a larger number of components could be obtained without use of high factorization ranks which we have seen to be unstable.
+In fact this is similar to the approach of \cite{WayGregory2019} mentioned in section \ref{sec-comparative-dimensionality-studies}.
 \item
@@
 TODO
-\subsection{Software libraries and versions}
-TODO
+
 \label{sec-software-versions}

### 2020-08-12 MSc_Dissertation_B156476.tex
Discussion, corrected PCA-1 / NMF-1 mixup, acknowldgments, better references.  98% done

@@
 Finding a clear, robust correlation between gene expression patterns and survival is not straight forward.  
-One substantial meta-analysis applied 16 previously published gene sets to two studies from GEO (199 samples) which included survival information, and found no significant predictive power \cite{Fekete2012}.
+One substantial meta-analysis applied 16 previously published gene sets to two studies from the Gene Expression Omnibus (GEO) (199 samples) which included survival information, and found no significant predictive power \cite{Fekete2012}.
 Although, within the same study, the authors did identify predictive genes, in particular SNGC, MAPT, ESR2 and PGR.
@@
 \subsubsection{Comparative studies of dimensionality reduction methods for gene expression analysis}
+\label{sec-comparative-dimensionality-studies}
 
@@
 \itemsep -0.5em 
-\item Python 3.6
-\item PyCharm Integrated Development Environment
-\item Numpy for high performance matrix manipulation
-\item Pandas for data frame handing
-\item Matplotlib and Seaborn for general plotting
-\item Scikit-learn for matrix factorization and k-means clustering
-\item GOATOOLS package for gene enrichment analysis against GO
-\item Lifelines package for survival analysis.
+\item DELL XPS 13, 16GB RAM, running Linux Mint version v19.
+\item Python v3.6.8
+\item PyCharm Integrated Development Environment, v2020.1.3
+\item Numpy for high performance matrix manipulation, v1.18.15
+\item Pandas for data frame handing, v1.0.5
+\item Matplotlib and Seaborn for general plotting, v3.2.2 and v0.10.1
+\item SciPy for least-square solutions and statistical tests, v1.5.0
+\item Scikit-learn for matrix factorization and k-means clustering, v0.23.1
+\item GOATOOLS package for gene enrichment analysis against GO, v1.0.6
+\item Lifelines package for survival analysis, v0.24.16.
+\item Github for source control.
 \end{itemize}
-A full list of software and libraries used, with versions, is given in appendix \ref{sec-software-versions}.
 
@@
 It has been claimed (e.g.
-\cite{Stein-OBrien2018}) that matrix factorization methods are an effective means of identifying and removing batch effects.
+\cite{Stein-OBrien2018,Renard2016}) that matrix factorization methods are an effective means of identifying and removing batch effects.
 To verify this, the TCGA and AOCS datasets (after gene set intersection), were horizontally concatenated producing an $n=80+274=354$ patient dataset with a substantial batch artefact.
@@
 \begin{center}
-\includegraphics[scale=0.1]{\plotspath{TCGA_OV_VST/GeneEnrichment/go_lineage_1_of_NMF_3.png}} \\
+\includegraphics[scale=0.09]{\plotspath{TCGA_OV_VST/GeneEnrichment/go_lineage_1_of_NMF_3.png}} \\
 \begin{spacing}{0.6}
@@
 %
-\item[NMF 1-of-3]: this is related to the cellular structures and processes of the extra-cellular matrix (ECM), that is the proteins such as collagens which mediate the three-dimensional organisation of cells in a tissues.
-ECM molecular composition will vary substantially between tissue types, but is also known to play a part in many disease processes \cite{Theocharis2019}.
-Thus, this metagene may simply reflect heterogeneity of tissues in the biopsied sample, or may have some deeper disease related significance.
+\item[NMF 1-of-3]: there is mention of ECM related terms, as in ICA 3-of-5 and NMF 1-of-3 below.
+However, there are also terms relating to the regulation of angiogenesis.
+It is known that tumours have a need for increased blood supply, and that expression level of factors promoting angigenesis are associated with aggresiveness of tumour growth \cite{Nishida2006}.
+We have already seen (section \ref{sec-subtype-classification}, \cite{Glass2015}) there is evidence for distinct subtypes of OC distinguished by angiogenesis related genes, and that these subtypes have been found to inform clinical outcome \cite{Glass2015}.
+Thus, this component might contain useful prognostic value.
 %
@@
 \item[ICA 3-of-5]: this component relates to processes of multicellular / extracellular organisation, and the ECM.  It is therefore similar to NMF 1-of-3 above.
-\item[ICA 4-of-5]: this component is seen to relate to the ribosomal subunit (as NMF 2-of-3), additionally membrane proteins in the respiratory complex, mitochondrial and the NADH dehydrogenase complex.
+\item[ICA 4-of-5]: many processes are highlighted by the component.
+These include processes relating to the ribosomal subunit (as NMF 2-of-3), additionally membrane proteins in the respiratory complex, mitochondrial and the NADH dehydrogenase complex.
 Processes of organonitrogen biosynthesis are also highlighted.
@@
 %
-\item[PCA 1-of-3]: there is mention of ECM related terms, as for ICA 3-of-5 and NMF 1-of-3.
-However, there are also terms relating to the regulation of angiogenesis.
-It is known that tumours have a need for increased blood supply, and that expression level of factors promoting angigenesis are associated with aggresiveness of tumour growth \cite{Nishida2006}.
-We have already seen (section \ref{sec-subtype-classification}, \cite{Glass2015}) there is evidence for distinct subtypes of OC distinguished by angiogenesis related genes, and that these subtypes have been found to inform clinical outcome \cite{Glass2015}.
-Thus, this component might contain useful prognostic value.
+\item[PCA 1-of-3]: this is related to the cellular structures and processes of the extra-cellular matrix (ECM), that is the proteins such as collagens which mediate the three-dimensional organisation of cells in a tissues.
+ECM molecular composition will vary substantially between tissue types, but is also known to play a part in many disease processes \cite{Theocharis2019}.
+Thus, this metagene may simply reflect heterogeneity of tissues in the biopsied sample, or may have some deeper disease related significance.
 %
@@
 \subsection{Integrative analysis of results}
+\label{sec-integrative-analysis}
 
@@
 %     & S & W & C & H % M & C & S & Enriched % Sim.
-NMF-1 & . & Y-& Y-& . & . & . & . & Extra-cellular matrix (ECM) & ICA-3, PCA-1 \\
-NMF-2 & . & . & . & Y+& . & . & Y+& Ribo. subunit & - \\
+NMF-1 & . & Y-& Y-& . & . & . & . & ECM, Reg. angiogenesis & ICA-3, PCA-1 \\
+NMF-2 & . & . & . & Y+& . & . & Y+& Ribo. subunit, RNA-binding & - \\
 NMF-3 & - & - & - & Y-& . & . & . & . & . \\
@@
 \hline
-PCA-1 & Y+& . & Y-& . & . & . & . & ECM, Reg. angiogenesis & NMF-1, ICA-3 \\
-PCA-2 & . & . & . & . & . & . & . & . & ICA-1 \\
+PCA-1 & ?+& . & Y-& . & . & . & . & Extra-cellular matrix (ECM)& NMF-1, ICA-3 \\
+PCA-2 & ?+& . & . & . & . & . & . & . & ICA-1 \\
 PCA-3 & . & Y-& Y-& . & . & . & . & Immune response, MHC, chemotaxis & ICA-5 \\
@@
 That these three components turn out to be similar by the Jaccard index is as would be expected given their functional similarity.
-\item[Angiogenesis link with survival?]
-The only component showing (albeit weak) correlation with patient survival is PCA-1.
-Interestingly, this same component is also found to be linked with Go terms relating to the regulation of angiogenesis, which in turn is related to aggressiveness of tumour growth and outcomes.
-Thus, these observations are telling a consistent story.
-\item[Normal cell activity?]
+\item[No consistent link with survival.]
+The only components showing even correlation with patient survival PCA-1 and PCA-2.
+The first of these are associated with ECM terms, which may simply be due to Cellularity.
+The other term showes no biological enrichment.]
 Component ICA-4 highlights the greatest number of biological processes -- indeed the GO lineage map \ref{fig-ICA-4-go-lineage} needed to be pruned to fit on a page.
 And yet this component shows no correlation with any disease related factors, neither survival nor genomic features.
-It may be that this component is simply picking up on the normal metabolic activity of the cells,
+It may be that this component is simply picking up on the normal metabolic activity of the cells.
+\item[Is HRD linked with ribosomal processing?].
+HRD is correlated with NMF-2 and NMF-3.
+The first of these is associated with the ribonucleoprotein complex while the second shows no biological enrichment.
 \item[Technical batch effects?]
@@
 \hline
-Metagene & Which & WGD & Cell. & HRD & Mut.L. & CN.L. & SV.L. & Enrichment? \\
+Metagene & Batch & WGD & Cell. & HRD & Mut.L. & CN.L. & SV.L. & Enrichment? \\
 \hline
-%     & w & W & C & H % M & C & S & Enrichment?
+%     & B & W & C & H % M & C & S & Enrichment?
 NMF-1 & Y & . & . & . & . & . & . &  Y \\
@@
 
-One valuable lesson from this work is the importance of accounting for sampling error, by bootstrapping,  when considering the stability of factorizations and thence the selection of factorization rank.
+The high level findings of this work are discussed below, under biological and methodological headings.
+
+\subsection{Biological findings}
+
+The integrative analysis set out in section \ref{sec-integrative-analysis} suggests that  Cellularity is correlated to the processes of the the ECM, reflecting heterogeneity of tissue type.
+Of course Cellularity is not really a biological characteristic -- it is a technical artefact of the tumour biopsy process, and so we might conclude that signatures relating to ECM might also be considered as technical artefacts.
+We found that whole genome doubling (WGD) and Cellularity correlate together with three components even though they are not themselves directly correlated.
+
+We found a weak evidence of a link between homologous repair deficiency (HRD) and RNA-binding.
+Being the site of protein synthesis, ribosomes will be associated with most of what goes on in a the cell.
+One paper suggests a link : ``a comprehensive review of the literature [] reveals a role for conventional DNA repair proteins in ribosome biogenesis, and conversely, ribosome biogenesis proteins in DNA repair" \cite{Ogawa2018}.
+Seek and ye shall find!
+
+Expression signature correlations with mutational load and copy-number variation (CNV) load have not been detected.
+These markers of genome instability are known to be very noisy, so if a correlation with expression does exist it would likely be a weak signal, below the detection power of the small (N=80) AOCS dataset.
+
+We found no convincing evidence of a metasample correlation with patient survival.
+
+\subsection{Methodological findings}
+
+One valuable lesson from this work is the importance of accounting for sampling error  when considering the stability of factorizations and thence the selection of factorization rank.
 Sevaral authors comment on the variability of NMF and ICA factorization due to algorithm initialisation \cite{Kairov2017,Sompairac2019,WayGregory2019}, but most neglect sampling error which we have shown to be far more significant.
 Cantini \etal\ \cite{Cantini2019a} mention bootstrapping, in reference to the BIODICA package (for ICA factorization), but it is unclear whether this is applied to the input data to model sampling error.
-In the initial stages of the current work, BIODICA was used, and was found to recommend ranks of 14, 19 or 30.
+In the initial stages of this project, experiments were conducted with BIODICA on the AOCS dataset.
+The package recommend ranks of 14, 19 or 30.
 Yet we have seen that such high ranks on the small n=80 dataset leads to highly unstable factorizations when bootstrap modelling of sampling error is included.
-
-These considerations lead to the selection of much lower ranks (fewer metagenes) than would otherwise be the case.
-
-\emph{TODO: much more discussion!}
+Suspicion of this result lead to me roll my own implementation based on bootstrap resampling, resulting in much lower ranks (fewer metagenes) being selected than would otherwise be the case if only algorithmic randomness were considered.
+
+We have been able to confirm the observations of \cite{Zhou2018a,Renard2016,Meng2016} that detection and removal of technical  batch effects naturally emerge from MF methods.
+That is, the technical and biological variation largely (not completely) separate into different components.
+This is a useful property, particularly as the growing availability of transcriptomics datasets from resources such as the TCGA and GEO allows for the simultaneous analysis of dozens of distinct datasets with likely different technical provenance, for which hand crafted normalisation would be infeasible.
 
@@
 That said, there will likely be many more sophisticated approached in the literature, given the clinical importance of the topic.
+A direct supervised approach -- find the axis of variation which best separates the two cancers -- may be more appropriate.
 \item
 {\bf Systematic comparison with published gene expression patterns}.
-In analysing the meaning of each metagene above, some tentative links were made with the research literature.
-But this was anecdotal and frankly not particularly scientific; searching on PubMed for GO terms associated with HGSOC and looking at one or two hits!
+In analysing the meaning of each metagene above, some tentative links were made with the research literature, by informal searching.
 A more systematic and ideally automated approach is required.
@@
 TODO
+
+\section*{Acknowledgements}
+I am most grateful to Dr.
+Ailith Ewing for proposing this project and making available datasets.
+Ailith provided excellent, inciteful supervision throughout and gave helpful feedback on an early draft of this dissertation.
+
+I'd like to thank my employer, Canon Medical Research Ltd., for allowing me the flexibility and study leave to undertake this MSc.
+
+Thank you to my family for excusing me from cooking and washing up duties during the more stressful phases of MSc study!
 

### 2020-08-12 MSc_Dissertation_B156476.tex
Pipeline execution time, correction to psuedocode, abstract

@@
 \begin{abstract}
-Here is the abstract
+Gene expression analysis -- transcriptomics -- of a set biopsied tumour tissue can identify genes whose expression is linked to the pathological process.
+In contrast to supervised contrastive approaches (e.g.
+comparing normal with tumour samples to identify over/under expressed genes), unsupervised approaches work with datasets of mixed and unknown tumour status.
+Matrix factorization (MF) methods have become a popular method of reducing and ``disentangling'' the salient components of variation in such datasets, known as metagenes.
+From these metagenes, sets of genes most associated with component of variation can be identified, and when combined with gene enrichment analysis against Gene Ontology (GO) linked annotations, can provide insights unto the underlying biological processes in play.
+
+In this work the biological target is high-grade serous ovarian cancer (HGSOC) -- the most common and aggressive type of ovarian cancer.
+Two gene expression datasets are analysed; the Austrailian Ovarian Cancer Study dataset (n=80) has patient survival and genomic metadata, while the larger dataset from The Cancer Genome Atlas (n=374) has only survival metadata.
+This motivates an approach of unsupervised metagenes discovered by MF techniques in the larger n=374 dataset then transferring that information to the smaller n=80 dataset to test for their power to stratify patient survival by Cox's proportional hazards analysis, and for correlation with genomic features.
+
+Three MF methods are applied and compared: non-negative matrix factorization (NMF), independent component analysis (ICA) and principal component analysis (PCA).
+While these are all linear factorization methods, they have very different properties which are explained.
+A key problem in applying MF is deciding on the number of components to extract -- the factorization rank.
+Too high a rank with result in metagenes which are unstable and fail to generalise to other datasets.
+From a review of the literature, it seems that the role of sampling error in the instability is an under appreciated issue, which this work addresses by use of bootstrap resampling.
+The similarity of the metagenes across the three methods is assessed via the Jaccard index.
+
+An integrative analysis of the GO enrichment results, survival analysis, genomic feature correlations and metagene similarity allows some tentative biological findings to be suggested and corroboration sought in the literature.
+The main methodological findings are that MF is effective in identifying and removing technical batch effects, but to beware of instability due to sampling error when applying MF to small datasets.
+The question of which of the three methods -- NMF, ICA or PCA -- is most effective for these purposes is considered but not clear conclusion can be drawn.
+
 \end{abstract}
@@
        save median_metagenes to file by factorizer and k
-       for each factorizer:
-          plot score vs k
+    plot score vs for factorizer k
 
@@
 Unit tests are based on tiny expression matrices, just 100 genes by 10 patients, thus the whole test suite of over 65 tests executes in around 45 seconds and so can be run frequently.
+The fully automated analysis pipeline takes approximately three hours from scratch, the majority of that time being taken for the repeated factorizations over the range of considered ranks.
 The total codebase is approximately 2,900 lines of Python.

### 2020-08-13 MSc_Dissertation_B156476.tex
Tweaked abstract, GEA, substantial addition to Discussion, going through research questions

@@
 \begin{abstract}
-Gene expression analysis -- transcriptomics -- of a set biopsied tumour tissue can identify genes whose expression is linked to the pathological process.
+Gene expression analysis -- transcriptomics -- of a set biopsied tumour tissues can identify genes whose expression is linked to the pathological process.
 In contrast to supervised contrastive approaches (e.g.
@@
 Matrix factorization (MF) methods have become a popular method of reducing and ``disentangling'' the salient components of variation in such datasets, known as metagenes.
-From these metagenes, sets of genes most associated with component of variation can be identified, and when combined with gene enrichment analysis against Gene Ontology (GO) linked annotations, can provide insights unto the underlying biological processes in play.
+From these metagenes, sets of genes most associated with component of variation can be identified, and when subjected to gene enrichment analysis against the Gene Ontology (GO) can provide insights into the underlying biological processes in play.
 
@@
 Three MF methods are applied and compared: non-negative matrix factorization (NMF), independent component analysis (ICA) and principal component analysis (PCA).
-While these are all linear factorization methods, they have very different properties which are explained.
+While these are all linear factorization methods, they have very different properties.
 A key problem in applying MF is deciding on the number of components to extract -- the factorization rank.
-Too high a rank with result in metagenes which are unstable and fail to generalise to other datasets.
-From a review of the literature, it seems that the role of sampling error in the instability is an under appreciated issue, which this work addresses by use of bootstrap resampling.
-The similarity of the metagenes across the three methods is assessed via the Jaccard index.
+Too high a rank will result in metagenes which are unstable and fail to generalise to other datasets.
+From a review of the literature, it seems that the role of sampling error in the instability is an under appreciated issue, which this work addresses by  bootstrap resampling then assessing  coherence through k-means clustering and the Silhouette score.
+The similarity of the metagenes found by the three methods is assessed via the Jaccard index.
 
@@
 The main methodological findings are that MF is effective in identifying and removing technical batch effects, but to beware of instability due to sampling error when applying MF to small datasets.
-The question of which of the three methods -- NMF, ICA or PCA -- is most effective for these purposes is considered but not clear conclusion can be drawn.
+The question of which of the three methods -- NMF, ICA or PCA -- is most effective for these purposes is considered but no clear conclusion can be drawn.
 
@@
 The platinum based drug Cisplatin is a key treatment in ovarian cancer, yet tumours often develop resistance, possibly related to the host's immune response.  
-Understanding and predicting such resistance is therefore of clinical importance and is addressed by Maringer \etal\ \cite{Mairinger2019} through expression analysis using the NanoString platform with a panel of 770 immune related genes.
+Understanding and predicting such resistance is therefore of clinical importance and is addressed by Mairinger \etal\ \cite{Mairinger2019} through expression analysis using the NanoString platform with a panel of 770 immune related genes.
 They find the following genes to be significantly related to platinum resistance: KLRC1, TCF7, CD274, HSD11B1, COLEC12, PDGFC, FCF1, BMI1, TNFRSF9, ATG10, EWSR1.
@@
 CD38 has a role in the breakdown of nicotinamide adenine dinucleotide ($\mbox{NAD}^+$), and is also a cell marker of lymphocytes.
-CD38 is over expressed in myloma cells, thence its importance as a prognostic marker and a potential drug treatment target.  
+CD38 is over expressed in myeloma cells, thence its importance as a prognostic marker and a potential drug treatment target.  
 It is worth mentioning that the analysis in this paper made use of the Gene Expression Profiling Interactive Analysis (GEPIA) database, which conveniently supports survival analysis with respect of any given gene, against a range of TCGA gene expression databases.
@@
 In the case of gene expression analysis, the dimensionality is of the order 20,000, i.e.
-the number of transcripts (mRMA transcribed genes) being considered.
+the number of transcripts (mRNA transcribed genes) being considered.
 However, it is known that sets of genes act in concert, their expression thus being correlated, so reducing the true degrees of freedom in the data.
@@
 PCA is a well known technique for dimensionality reduction, based on eigenvalue decomposition.
-It is arguably the ideal, efficient solution for data which is multivariate Gaussian distributed -- or can assumed to be so.
+It is arguably the ideal, efficient solution for data which is multivariate Gaussian distributed -- or can be assumed to be so.
 PCA results in components which are ordered by the proportion of total variance they explain, and are mutually orthogonal.
@@
 Recent work by Way \etal\ \cite{WayGregory2019} was aimed directly at answering the question which dimensionality reduction method was best suited to gene expression data.
-They considered PCA, ICA, NMF and also non-linear neural network methods -- directional autoencoders  and variational autoencoders .
+They considered PCA, ICA, NMF and also non-linear neural network methods -- directional autoencoders (DAE)  and variational autoencoders (VAE).
 Briefly, these latter methods work by training a neural network to reconstruct an input signal while being forced through a ``bottleneck'' in dimensionality.
@@
 PCA, ICA and NMF are discussed, but the main focus is on less well known algorithms applicable to multiple datasets, including multiple coinertia analysis and consensus PCA.
-The essential purpose of these methods is to transform diverse datasets -- transcriptomics, genomics and metabolomics  -- into a common and manageable representation space.
+The essential purpose of these methods is to transform diverse datasets -- transcriptomics, genomics and metabolomics  -- into a crommon and manageable representation space.
 
@@
 \begin{enumerate}
-\item Which are the influential genes, or groups of genes, active in HGSOC according to gene expression data?   
-\item How do these patterns of gene expression relate to underlying biological processes.  Do these confirm published results?
-\item Which MF method is best suited to this analysis and how should they be applied to achieve robust results in the presence of limited dataset size?
+\item What underlying biological processes are at play in HGSOC as seen through the patterns of gene expression?  Do these confirm published results?
+\item How do the patterns of gene expression uncovered relate to genome level features, such as those marking genome instability?
 \item Do MF methods yield factors which are predictive of patient survival?
-\item How do the patterns of gene expression uncovered relate to genome level features, such as those marking genome instability?
+\item How should MF methods be applied to achieve robust results on datasets of limited size?
+\item Do MF detect and remove technical batch effects?
+\item Which MF method is best suited to transcripomics analysis?
 \end{enumerate}
@@
 \item The Cancer Genome Atlas (TCGA) derived, n=274, g=19,601, with metadata on survival
-\item Austrailian Ovarian Cancer Study (AOCS) \cite{Patch2015}, n=80, g=19,730, with metadata on survival, cellularity and additional genomic features.
+\item Australian Ovarian Cancer Study (AOCS) \cite{Patch2015}, n=80, g=19,730, with metadata on survival, cellularity and additional genomic features.
 Several of the studies reviewed earlier also use data from the Australian Ovarian Cancer Study (AOCS): \cite{Patch2015,Ewing2020,Cuello2018,Au-Yeung2014}.
@@
 \item
-Investigation of biological significance of determined factors by gene enrichment analysis against the Gene Ontology (GO).
+Investigation of biological significance of determined factors by gene enrichment analysis (GEA) against the Gene Ontology (GO).
 \item
@@
 \item Scikit-learn for matrix factorization and k-means clustering, v0.23.1
-\item GOATOOLS package for gene enrichment analysis against GO, v1.0.6
+\item GOATOOLS package for GEA against GO, v1.0.6
 \item Lifelines package for survival analysis, v0.24.16.
@@
 An alternative to consider the intersection of the gene sets which each metagene highlights.
-In the gene enrichment analysis which follows, candidate genes will be identified for each meatagene.
+In the GEA which follows, candidate genes will be identified for each meatagene.
 A standard measure of set similarity is the Jaccard index, (or similarity), defined on a pair of sets as:
@@
 
-\subsection{Gene enrichment analysis}
-
+\subsection{Gene enrichment analysis (GEA)}
+\label{sec-gea}
 The metagenes extracted by the above described factorization and clustering process provide valuable information into which genes vary in expression in the study samples; thus in our case, which genes are influential in HGSOC.
-In order to gain insights into what biological processes are involved, gene enrichment analysis against the Gene Ontology (GO) was carried out for each of the 11 metagenes individually.
-
-The essence of gene enrichment analysis is to compare a candidate set of genes with many functional gene sets, asking the question ``are there significantly more (or less, i.e.
+In order to gain insights into what biological processes are involved, GEA against the Gene Ontology (GO) was carried out for each of the 11 metagenes individually.
+
+The essence of GEA is to compare a candidate set of genes with many functional gene sets, asking the question ``are there significantly more (or less, i.e.
 depletion) intersecting genes than would be expected for the same number of genes being drawn at random (without replacement) from the total population of genes in the study.
@@
 Multiple hypothesis significance testing used the false discovery rate (FDR) method of Bengamini and Hochberg, the FDR threshold being set to 0.01.
-This is more stringent than the 0.05 value which is commonly used, and was chosen since multiple metagenes are being analysed, implying multiple hypothesis testing over and above that which is accounted for by FDR filtering within a single gene enrichment analysis run.
+This is more stringent than the 0.05 value which is commonly used, and was chosen since multiple metagenes are being analysed, implying multiple hypothesis testing over and above that which is accounted for by FDR filtering within a single GEA run.
 
@@
 The described methodology was implemented in Python.
-%except for the KEGG gene enrichment analysis which was implemented in R.
+%except for the KEGG GEA which was implemented in R.
 All code is available in a github repository: \url{https://github.com/ipoole/HgsocTromics}.
@@
 
-Gene enrichment analysis against the GO results in, for each of the 11 metagenes, a table of GO terms with a list of the candidate genes which have an annotated association to that term; this table can be found in the appendix, table \ref{tab-go-enrichment-results}.
+GEA against the GO results in, for each of the 11 metagenes, a table of GO terms with a list of the candidate genes which have an annotated association to that term; this table can be found in the appendix, table \ref{tab-go-enrichment-results}.
 To take full advantage of the hierarchical nature of the GO, results are presented as lineage maps in figures \ref{fig-NMF_1-go-lineage} to \ref{fig-PCA-3-3-go-lineage}.  
@@
 \begin{center}
-\includegraphics[scale=0.5,draft=false]{\plotspath{BoxAndScatters/genomic_feature_scatters_TCGA_AOCS.pdf}} 
+\includegraphics[scale=0.5,draft]{\plotspath{BoxAndScatters/genomic_feature_scatters_TCGA_AOCS.pdf}} 
 \caption{Grid of scatter plots to visualise the correlation between metasamples and genomic features.
@@
 \caption{For each metagenes a 'Y' indicates that a significant correlation was observed with: Survival, WGD, HRD, Mutational load,  CNV load and SV load.
-The next column gives a terse summary of the gene enrichment analysis results.
+The next column gives a terse summary of the GEA results.
 The final column lists the metagenes which are apparently similar -- having a Jaccard similarity of $\geq 0.35$.}
@@
 \begin{center}
-\includegraphics[scale=0.5
-, draft=false]{\plotspath{HeatMaps/clustered_heatmap_BOTH_BOTH.pdf}} 
+\includegraphics[scale=0.5]{\plotspath{HeatMaps/clustered_heatmap_BOTH_BOTH.pdf}} 
 \caption{Heatmap of the metasamples matrix for the combined AOCS + TCGA dataset.  The 'Which' field indicates the source dataset; purple: AOCS, yellow:TCGA.}
@@
 
-The high level findings of this work are discussed below, under biological and methodological headings.
-
-\subsection{Biological findings}
+Addressing each of our research questions in turn:
+
+\subsubsection*{What underlying biological processes are at play in HGSOC as seen through the patterns of gene expression? }
+
+Biological processes identified by GEA have been set out in section \ref{sec-gea}.
+Processes relating to the ECM and extracellular organisation may reflect cell type heterogeneity -- perhaps within the tumour its self, or due to sampling of adjacent normal tissue.
+Processes of negative regulation of angiogenesis could reflect tumour growth and therefore be of prognostic value.
+
+These observations are based on GEA on the GO only.
+Possibly more insights could have been gained by GEA against KEGG\footnote{\href{https://www.genome.jp/kegg/pathway.html}{Kyoto Encyclopedia of Genes and Genomes}} Pathways; this was intended but had to be dropped due to shortage of time.
+
+\subsubsection*{How do the patterns of gene expression uncovered relate to genome level features, such as those marking genome instability?}
 
 The integrative analysis set out in section \ref{sec-integrative-analysis} suggests that  Cellularity is correlated to the processes of the the ECM, reflecting heterogeneity of tissue type.
-Of course Cellularity is not really a biological characteristic -- it is a technical artefact of the tumour biopsy process, and so we might conclude that signatures relating to ECM might also be considered as technical artefacts.
-We found that whole genome doubling (WGD) and Cellularity correlate together with three components even though they are not themselves directly correlated.
-
-We found a weak evidence of a link between homologous repair deficiency (HRD) and RNA-binding.
+Of course, Cellularity is not really a biological characteristic -- it is a technical artefact of the tumour biopsy process, and so we might conclude that signatures relating to ECM might also be considered as technical artefacts.
+We found that whole genome doubling (WGD) and Cellularity correlate together with three components even though they are not themselves directly correlated -- something which warrants further investigation.
+
+Weak evidence of a link between homologous repair deficiency (HRD) and RNA-binding was foud.
 Being the site of protein synthesis, ribosomes will be associated with most of what goes on in a the cell.
-One paper suggests a link : ``a comprehensive review of the literature [] reveals a role for conventional DNA repair proteins in ribosome biogenesis, and conversely, ribosome biogenesis proteins in DNA repair" \cite{Ogawa2018}.
+One paper suggests a link : ``a comprehensive review of the literature reveals a role for conventional DNA repair proteins in ribosome biogenesis, and conversely, ribosome biogenesis proteins in DNA repair" \cite{Ogawa2018}.
 Seek and ye shall find!
 
-Expression signature correlations with mutational load and copy-number variation (CNV) load have not been detected.
+Expression signature correlations with mutational load and  CNV load have not been detected.
 These markers of genome instability are known to be very noisy, so if a correlation with expression does exist it would likely be a weak signal, below the detection power of the small (N=80) AOCS dataset.
 
-We found no convincing evidence of a metasample correlation with patient survival.
-
-\subsection{Methodological findings}
+\subsubsection*{Do MF methods yield factors which are predictive of patient survival?}
+
+We found no convincing evidence of this, for an individual factor (metasample).
+Some components were predictive in one or other of the two datasets, but not both, which is unconvincing.
+The factor with GEA showed indicated regulation of angiogenesis (NMF-1), and thence might have prognostic value, performed poorly in survival analysis.
+It is possible that a more sophisticated multiple regression model would be successful, but this would risk over-fitting given the limited dataset size.
+
+\subsubsection*{How should MF methods be applied to achieve robust results on datasets of limited size?}
 
@@
 
+\subsubsection*{Do MF detect and remove technical batch effects?}
+
+Yes.
 We have been able to confirm the observations of \cite{Zhou2018a,Renard2016,Meng2016} that detection and removal of technical  batch effects naturally emerge from MF methods.
@@
 This is a useful property, particularly as the growing availability of transcriptomics datasets from resources such as the TCGA and GEO allows for the simultaneous analysis of dozens of distinct datasets with likely different technical provenance, for which hand crafted normalisation would be infeasible.
+
+\noindent
+\emph{Which MF method is best suited to transcripomics analysis?}
+
+The original focus of this project was ICA and NMF since these are know to be able to disentangle variation in data with complex structure, not multivariate normally distributed.
+PCA was added as a baseline dimensionality reduction method, just for comparison.
+What would we have been missed if only PCA had been used?
+Reviewing the integrative analysis of results in table \ref{tab-integrative-analysis} we see that all three methods picked up on WGD and cellularity.
+PCA included the two components having the largest observed effect on overall survival.
+NMF and ICA both showed sensitivity to SV load.
+NMF further added sensitivity to HRD and identified genes relating to angiogenesis with possible pathological significance.
+
+So it's a mixed picture.
+Of course, had more components been selected from each method then likely more would have been found, but at the expense of robustness and thence false discovery, as we have discussed.
+It seems we must concur with the conclusion of Way \etal \cite{WayGregory2019} that no one method excels, each brings individual value.
+At least these three methods are simple and linear (unlike the VAE for example), and so can be used together easily.
 
@@
 For example, NMF has parameters to encourage sparsity through L1 regularization.
-This should result in many (most?) metagene elements reducing to zero, and might offer a more robust means of selecting candidate genes to feed into gene enrichment analysis -- replacing the current arbitrary 3 SD from the mean rule.
+This should result in many (most?) metagene elements reducing to zero, and might offer a more robust means of selecting candidate genes to feed into GEA -- replacing the current arbitrary 3 SD from the mean rule.
 \item
@@
 \begin{center}
-\includegraphics[scale=0.5,draft=false]{\plotspath{BoxAndScatters/genomic_feature_scatters_BOTH_AOCS.pdf}} 
+\includegraphics[scale=0.5]{\plotspath{BoxAndScatters/genomic_feature_scatters_BOTH_AOCS.pdf}} 
 \caption{Grid of scatter plots to visualise the correlation between metasamples derived from the metagenes of the combined AOCS + TCGA datasets, and genomic features of the AOCS.}

### 2020-08-13 MSc_Dissertation_B156476.tex
Added CD38 results

@@
 Some components were predictive in one or other of the two datasets, but not both, which is unconvincing.
-The factor with GEA showed indicated regulation of angiogenesis (NMF-1), and thence might have prognostic value, performed poorly in survival analysis.
-It is possible that a more sophisticated multiple regression model would be successful, but this would risk over-fitting given the limited dataset size.
+The factor which GEA indicated regulation of angiogenesis (NMF-1), and thence might have prognostic value, performed poorly in survival analysis.
+
+Is there a signal for survival to be found in these data?
+We have noted that according to \cite{Zhu2020}, the single gene CD38 is highly predictive of survival in EOC.
+Does this hold for the TCGA and AOCS datasets we are working with?
+This is easy to check, running survival analysis stratified by a hand-crafted metagene, setting the CD38 element to 1, all others to 0.
+This was applied as in section \ref{sec-survival-analysis} to TCGA (OS),  AOCS (OS) and AOCS (PFS) -- see figure \ref{fig-cd38 survival}.
+The effects are in a consistent direction across the three tests, with high CD38 expression predicting favourable survival.
+This is highly significant in the TCGA dataset and $p < 0.05 $ in the smaller AOCS.
+So there is a signal, but it did not emerge unbidden from matrix factorization.
+(CD38 is in fact highlighted by component ICA-5, see figure \ref{fig-ICA-5-5-go-lineage}, but must be diluted by other genes).
+
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.7,draft=false]{\plotspath{SurvivalAnalysis/gene_specific_CD38_survival.pdf}}
+\end{center}
+\caption{Survival analysis plots stratified by above/below median expression of CD38 only.}
+\label{fig-cd38 survival}
+\end{figure}
 
@@
 
-\noindent
-\emph{Which MF method is best suited to transcripomics analysis?}
+\subsubsection*{Which MF method is best suited to transcripomics analysis?}
 
@@
 
+\FloatBarrier
+
 \section{Conclusions}
@@
 I am most grateful to Dr.
-Ailith Ewing for proposing this project and making available datasets.
-Ailith provided excellent, inciteful supervision throughout and gave helpful feedback on an early draft of this dissertation.
-
+Ailith Ewing for proposing this project and making available the datasets.
+Ailith provided excellent, insightful supervision throughout and gave valuable feedback on an early draft of this dissertation.
 I'd like to thank my employer, Canon Medical Research Ltd., for allowing me the flexibility and study leave to undertake this MSc.
-
 Thank you to my family for excusing me from cooking and washing up duties during the more stressful phases of MSc study!
@@
 
+\subsection{Additional plots relating to the batch effect investigation}
 \begin{figure}[htb!]
@@
 \includegraphics[scale=\S]{\plotspath{BOTH_AOCS_TCGA/FactorClustering/multiple_single_factors_scatter_BOTH_PCA_2_3_5_10_bootstrap.pdf}} \\
-%\includegraphics[scale=0.55]{\plotspath{/BOTH_AOCS_TCGA/FactorClustering/silhouette_plots_TCGA_2_3_4_5_6_7_8_9_10_bootstrap.pdf}} 
+\includegraphics[scale=0.55]{\plotspath{/BOTH_AOCS_TCGA/FactorClustering/silhouette_plots_BOTH_2_3_4_5_6_7_8_9_10_bootstrap.pdf}} 
 \caption{Metagene clustering for all three methods applied to the combined N=80+374 AOCS + TCGA dataset with bootstrap sampling, over a range of factorization ranks.

### 2020-08-13 MSc_Dissertation_B156476.tex
Substantial edits from a full read through, many improvements.  .pdf included in commit for safety.   Prety much done :-)

@@
 
-\documentclass[draft,tikz, 12pt,a4paper,oneside,fleqn]{article}
+\documentclass[tikz, 12pt,a4paper,oneside,fleqn]{article}
 
@@
 \begin{abstract}
-Gene expression analysis -- transcriptomics -- of a set biopsied tumour tissues can identify genes whose expression is linked to the pathological process.
+Gene expression analysis -- transcriptomics -- of a set biopsied tumour tissues can identify genes whose expression is linked to the pathological processes of the tumour.
 In contrast to supervised contrastive approaches (e.g.
@@
 Matrix factorization (MF) methods have become a popular method of reducing and ``disentangling'' the salient components of variation in such datasets, known as metagenes.
-From these metagenes, sets of genes most associated with component of variation can be identified, and when subjected to gene enrichment analysis against the Gene Ontology (GO) can provide insights into the underlying biological processes in play.
+From these metagenes the most influential genes can be identified, and when subjected to gene enrichment analysis against the Gene Ontology (GO) can provide insights into the underlying biological processes in play.
 
 In this work the biological target is high-grade serous ovarian cancer (HGSOC) -- the most common and aggressive type of ovarian cancer.
-Two gene expression datasets are analysed; the Austrailian Ovarian Cancer Study dataset (n=80) has patient survival and genomic metadata, while the larger dataset from The Cancer Genome Atlas (n=374) has only survival metadata.
-This motivates an approach of unsupervised metagenes discovered by MF techniques in the larger n=374 dataset then transferring that information to the smaller n=80 dataset to test for their power to stratify patient survival by Cox's proportional hazards analysis, and for correlation with genomic features.
+Two gene expression datasets are analysed; the Austrailian Ovarian Cancer Study dataset (N=80) has patient survival and genomic metadata, while the larger dataset from The Cancer Genome Atlas (N=374) has only survival metadata.
+This motivates an approach of unsupervised metagenes discovered by MF techniques in the larger N=374 dataset then transferring that information to the smaller N=80 dataset to test for their power to stratify patient survival by Cox's proportional hazards analysis, and for correlation with genomic features.
 
@@
 \item[\emph{High grade serous OC (HGSOC)}] affects the epithelium of the ovaries and fallopian tubes.
-It is the most common type of OC, accounting for 70\% of cases, and most serious of the OC subtypes and the focus of this dissertation.
+It is the most common type of OC, accounting for 70\% of cases, and most serious of the OC subtypes.
+HGSOC is the focus of this dissertation.
 \end{description}
@@
 HGSOC typically occurs in older patients and is diagnosed at a late stage.   Histology of HGSOC is similar to the low grade subtype, but the cancer develops along distinct molecular pathways. 
-
-Treatment options for HGSOC depend on tumour stage and include surgery (tumour debulking) or platinum-based chemotherapy (e.g.
+Treatment options depend on tumour stage and include surgery (tumour debulking) or platinum-based chemotherapy (e.g.
 Cisplatin).
@@
 They leave open the question whether DNA repair by PARP plays a role.
-The relationship between BRCA1/2 and DNA repair -- specifically homologous repair deficiency (HRD) -- is picked up by Ewing \etal \cite{Ewing2020}.
+The relationship between BRCA1/2 and DNA repair -- specifically homologous repair deficiency (HRD) -- is picked up by Ewing \etal\ \cite{Ewing2020}.
 They study the complex interplay of single nucleotide variants and large structural variants as they impact BRCA1/2 disruption and thence HRD.
@@
 ``All Array": used expression array data from 1650 patients across 14 studies evaluating 9 supervised learning algorithms by bootstrap, selecting an AdaBoost -like method.
-The ``TCGA" team used 434 patients from The Cancer Genome Atlas (TCGA\footnote{I'm always amused by the genetics pun in the naming of this resource!}) evaluating 5 algorithms by cross-validation selecting a random forest.
-"All Array" had the advantage of more data but needed attention to batch effects.
-A final classifier took a consensus of the two methods based on a minimal gene set (order 40 of genes), validated by a leave-one-out (patient level) approach.
+The ``TCGA" team used 434 patients from The Cancer Genome Atlas (TCGA\footnote{I'm always amused by the genetics pun in the naming of this resource!}) evaluating 5 algorithms by cross-validation, selecting a random forest.
+``All Array" had the advantage of more data but needed careful attention to batch effects.
+A final classifier took a consensus of the two methods based on a minimal gene set (around 40), validated by a leave-one-out (patient level) approach.
 Survival analysis was carried out stratified by predicted subtype.
@@
 Single-sample gene set enrichment analysis (ssGSEA) is used to obtain scores for previously reported expression signatures: differentiated, immunoreative, mesenchymal and proliferative.  
-An outcome prediction model based on these signatures gave good survival stratification (likelihood ratio 10.8 between above/below median groups) on a validation dataset) which improved to 15.7 when augmented with BRCA1/2 mutation status and other post-operative clinical factors.
+An outcome prediction model based on these signatures gave good survival stratification,, with likelihood ratio 10.8 between above/below median groups on a validation dataset which improved to 15.7 when augmented with BRCA1/2 mutation status and other post-operative clinical factors.
 
 An explicit supervised learning approach can be taken, as in Berchuck \etal, where a hybrid decision tree plus linear discriminant model demonstrated survival stratification in both cross-validation and external validation experiments, based on microarray expression data.
-They found expression of CSFT3, ABCD3, MAL and APMCF1 to be the most influential; in fact the decision tree was almost entirely driven by the first two of these
+They found expression of CSFT3, ABCD3, MAL and APMCF1 to be the most influential; in fact the decision tree was almost entirely driven by the first two of these.
 
 Mairinger \etal\ screened 770 immune related genes to identify 11 differentially expressed genes associated with response to platinum treatment.  
-They find that expression of HS11B1, DNBT1, CKLF, NUP107. CCL18, LY96, ATG7, SLAMF7, CXCL9 is associated with better survival, while IKBKG and SDHA associate with with poor survival.
+They find that expression of HS11B1, DNBT1, CKLF, NUP107, CCL18, LY96, ATG7, SLAMF7, CXCL9 is associated with better survival, while IKBKG and SDHA associate with with poor survival.
 
@@
 CD38 is over expressed in myeloma cells, thence its importance as a prognostic marker and a potential drug treatment target.  
-It is worth mentioning that the analysis in this paper made use of the Gene Expression Profiling Interactive Analysis (GEPIA) database, which conveniently supports survival analysis with respect of any given gene, against a range of TCGA gene expression databases.
+It is worth mentioning that the analysis in this paper made use of the Gene Expression Profiling Interactive Analysis (GEPIA) database, which conveniently supports survival analysis with respect of any given gene, against a range of TCGA hosted gene expression databases.
 
@@
 They used NMF based clustering to demonstrate that expression of genes linked with obesity and lipid metabolism impart poorer progression free survival, independent of known HGSOC driver gene expression.
-The link is complex however, there been many confounding non-molecular based reasons for a link between high BMI and HGSOC survival, including later diagnosis and compromised operability.
+The link is complex however, there being many confounding non-molecular based reasons for a link between high BMI and HGSOC survival, including later diagnosis and compromised operability.
 Some studies (e.g.
@@
 A particular value of NMF is that it allows for input samples to be directly assigned to one of $k$ clusters.   
-The optimization function on which NMF is based can be tuned to achieve a \emph{sparse} factorization result, that is one in which only a small number of elements in the achieved factorization are non-zero, a property which further simplifies interpretation.
-\cite{Brunet2004}.
+The optimization function on which NMF is based can be tuned to achieve a \emph{sparse} factorization result, that is one in which only a small number of elements in the achieved factorization are non-zero, a property which further simplifies interpretation \cite{Brunet2004}.
 %
@@
 Where the data is non-Gaussian however, ICA results in components which separate out independent sources of variation.
-See \cite{Sompairac2019} for a full explanation, from which figure \ref{fig-pca-ica-nmp-explanation} is taken.
+
 \subsubsection{Principal component analysis (PCA)}
@@
 
-The contrasting results of the three methods is illustrated in figure \ref{fig-pca-ica-nmp-explanation}. \emph{TODO: add explanation of this diagram}
-
-\begin{figure}[htb!]
-\begin{center}
-\includegraphics[scale=0.4]{Sompairac_ICA_NMF_PCA.png}
-\end{center}
-\caption{Illustrating the differing results of component extraction by PCA, ICA and NMF.  From \cite{Sompairac2019}}
-\label{fig-pca-ica-nmp-explanation}
-\end{figure}
-
-Conventionally, transcriptomics expression arrays are oriented with genes (or transcripts) as rows, and samples (e.g. patients) in columns.   
+%\begin{figure}[htb!]
+%\begin{center}
+%\includegraphics[scale=0.4]{Sompairac_ICA_NMF_PCA.png}
+%\end{center}
+%\caption{Illustrating the differing results of component extraction by PCA, ICA and NMF.  From \cite{Sompairac2019}}
+%\label{fig-pca-ica-nmp-explanation}
+%\end{figure}
+
+Conventionally, expression arrays are oriented with genes (or transcripts) in rows, and samples (e.g. patients) in columns.   
 A typical expression array might have in the order of tens of thousands of rows and hundreds of columns.  
-MF methods reduces this large $M \times N$ (genes $\times$ samples) matrix into two smaller matrices.
-In the general terminology of Stein-O'Brien \etal\ these are the $M \times K$ \emph{pattern matrix} and the $K \times N$ \emph{amplitude matrix}.
-$K$ (or $k$ as we will use) is typically small of the order 10, and refers to the number of extracted \emph{factors} (or components); it is the rank of the factorization.
+MF methods reduces this large $m \times n$ (genes $\times$ samples) matrix into two smaller matrices.
+In the general terminology of Stein-O'Brien \etal\ \cite{Stein-OBrien2018} these are the $m \times k$ \emph{pattern matrix} and the $k \times n$ \emph{amplitude matrix}.
+$k$ is typically small of the order 10, and refers to the number of extracted \emph{factors} (or components); it is the rank of the factorization.
 Since we are focussing on transriptomics, rather than pattern and amplitude we will use the terms \emph{metagenes} and \emph{metasamples} respectively, which are in common useage.
@@
 NMF generally uses $W$ and $H$ respectively while ICA generally uses $S$ and $A$.
-PCA tends to be differently formulated, but $W$ and $T$ sometimes used.
+PCA tends to be differently formulated, but $W$ and $T$ are often used.
 The original matrix to be factorized is variously named $X$ or $V$.
@@
 \begin{description}
-\item[Expression matrix:] $X$ $(M \times N)$
-\item[Metagene matrix:] $W$ $(M \times K)$
-\item[Metasample matrix:] $H$ $(K \times N)$
+\item[Expression matrix:] $X$ $(m \times n)$
+\item[Metagene matrix:] $W$ $(m \times k)$
+\item[Metasample matrix:] $H$ $(k \times n)$
 \end{description}
-where $N$ is the number of patients (or samples), $M$ is the number of genes and $K$ is the factorization rank, i.e. the number of components or factors.  Thus the factorization is written
+where $N$ is the number of patients (or samples), $m$ is the number of genes and $K$ is the factorization rank, i.e. the number of components or factors.  Thus the factorization is written
 \begin{equation}
@@
 \begin{equation}
-X_{M,N} \approx W_{M,K} H_{K,N}
+X_{m,n} \approx W_{m,k} H_{k,n}
 \end{equation}
@@
 \subfloat[Optimising independence of Meta\underline{genes}]{
-\begin{tikzpicture}[node distance = 0.25cm]
+\begin{tikzpicture}[node distance = 0.30cm, scale=1.3, every node/.style={scale=1.3}]
 	\node (X) [draw, fill=green, minimum height=\M, minimum width=\N] {\large X};
@@
 
-	\node [below=\fudge of X] {\tiny{$N$ Samples}};
-	\node [left=\fudge of X] {\rotatebox{90}{\tiny{$M$ Genes}}};
+	\node [below=\fudge of X] {\tiny{$n$ Samples}};
+	\node [left=\fudge of X] {\rotatebox{90}{\tiny{$n$ Genes}}};
 	\node [below=of X, text width=2cm, align=center] {\small Expression matrix};
 
-	\node [below=\fudge of S] {\tiny{$K$ Metagenes}};
-	\node [left=\fudge of S] {\rotatebox{90}{\tiny{$M$ Genes}}};
+	\node [below=\fudge of S] {\tiny{$k$ Metagenes}};
+	\node [left=\fudge of S] {\rotatebox{90}{\tiny{$m$ Genes}}};
 	\node [below=of S, text width=2cm, align=center] {\small Metagenes};
 
-	\node [below=\fudge of A] {\tiny{$N$ Samples}};
-	\node [left=\fudge of A] {\tiny{$K$}};
+	\node [below=\fudge of A] {\tiny{$n$ Samples}};
+	\node [left=\fudge of A] {\tiny{$k$}};
 	\node [below=of A, text width=2.5cm, align=center] {\small Metasamples};
@@
 \subfloat[Optimising independence of Meta\underline{samples}]{
-\begin{tikzpicture}[node distance = 0.25cm]
+\begin{tikzpicture}[node distance = 0.30cm, scale=1.3, every node/.style={scale=1.3}]
 	\node (X) [draw, fill=green, minimum height=\N, minimum width=\M] {\large X};
@@
 
-	\node [below=\fudge of X] {\tiny{$M$ Genes}};
-	\node [left=\fudge of X] {\rotatebox{90}{\tiny{$N$ Samples}}};
+	\node [below=\fudge of X] {\tiny{$m$ Genes}};
+	\node [left=\fudge of X] {\rotatebox{90}{\tiny{$n$ Samples}}};
 	\node [below=of X, text width=2cm, align=center] {\small Expression matrix};
 
-	\node [below=\fudge of S] {\tiny{$K$ Metasamples}};
-	\node [left=\fudge of S] {\rotatebox{90}{\tiny{$N$ Samples}}};
+	\node [below=\fudge of S] {\tiny{$k$ Metasamples}};
+	\node [left=\fudge of S] {\rotatebox{90}{\tiny{$nN$ Samples}}};
 	\node [below=of S, text width=2.5cm, align=center] {\small Metasamples};
 
-	\node [below=\fudge of A] {\tiny{$M$ Genes}};
-	\node [left=\fudge of A] {\tiny{$K$}};
+	\node [below=\fudge of A] {\tiny{$m$ Genes}};
+	\node [left=\fudge of A] {\tiny{$k$}};
 	\node [below=of A, text width=2cm, align=center] {\small Metagenes};
@@
 It is perfectly possible to apply these algorithms to gene expression analysis with exchanged and transposed meanings of $W$ and $H$, and this is illustrated in figure \ref{fig-matrix-factorization} (b).
-In this case it is the properties of the \emph{metasamples} which will be optimised.
+In this case it is the properties of the \emph{metasamples} which are optimised.
 Thus, the two forms are different in substance, not simply in notational convention.
@@
 In this regard, PCA differs from ICA and NMF.
-In PCA it is reasonable to extract all factors, setting $k=N$, the factors being ranked by the associated eigenvalue which also ranks the proportion of variance explained.
+In PCA it is reasonable to extract all factors, setting $k=n$, the factors being ranked by the associated eigenvalue which also ranks the proportion of variance explained.
 However, that approach is not valid for ICA and NMF, since differing sets of factors will be obtained for different $k$, and for a given choice of $k$ all factors are equally important -- they do not rank \cite{Stein-OBrien2018}.
@@
 
-\subsubsection{Comparative studies of dimensionality reduction methods for gene expression analysis}
+\subsubsection{Comparisons dimensionality reduction methods for gene expression analysis}
 \label{sec-comparative-dimensionality-studies}
 
-Recent work by Way \etal\ \cite{WayGregory2019} was aimed directly at answering the question which dimensionality reduction method was best suited to gene expression data.
+Recent work by Way \etal\ \cite{WayGregory2019} was aimed directly at answering the question of which dimensionality reduction method was best suited to gene expression data.
 They considered PCA, ICA, NMF and also non-linear neural network methods -- directional autoencoders (DAE)  and variational autoencoders (VAE).
-Briefly, these latter methods work by training a neural network to reconstruct an input signal while being forced through a ``bottleneck'' in dimensionality.
+Briefly, these latter methods work by training a neural network to reconstruct an input signal while being forced through a network ``bottleneck''.
 Their conclusion was that no single method was optimal, but best performance -- judged by capturing the most biological pathway-associated features -- was achieved by taking the extracted latent dimensions of all the methods, an approach they call the ``BioBombe''.
@@
 Would two teams working with identical data and using similar MF methods obtain similar metagenes and so arrive at the same biological conclusions?
-The \emph{stability} of MF methods is well considered by many authors, although this has typically been focussed on stability of the underlying algorithm -- NMF and ICA are stochastic in nature.
+The \emph{stability} of MF methods is well considered by many authors, although this has typically been focussed on stability of the underlying algorithm -- NMF and ICA are stochastic in nature, sensitive to initial conditions.
 Much less considered is stability in the presence of \emph{sampling error} which inevitably exists particularly in small datasets.
@@
 This is appropriate in our case, since of the two datasets available (described below), one (TCGA) has relatively large n but little available metadata, while the other (AOCS) is much smaller but has more useful metadata of genomic features.
-This motivates an approach of unsupervised learning on the TCGA dataset which is then transferred to the AOCS dataset.
+This motivates an approach of unsupervised learning on the TCGA dataset which is then transferred to the AOCS dataset for evaluation.
 
@@
 \begin{enumerate}
-\item The Cancer Genome Atlas (TCGA) derived, n=274, g=19,601, with metadata on survival
-\item Australian Ovarian Cancer Study (AOCS) \cite{Patch2015}, n=80, g=19,730, with metadata on survival, cellularity and additional genomic features.
+\item The Cancer Genome Atlas (TCGA) derived, n=274 patiens, m=19,601 genes, with metadata on survival
+\item Australian Ovarian Cancer Study (AOCS) \cite{Patch2015}, n=80, m=19,730, with metadata on survival, cellularity and additional genomic features.
 Several of the studies reviewed earlier also use data from the Australian Ovarian Cancer Study (AOCS): \cite{Patch2015,Ewing2020,Cuello2018,Au-Yeung2014}.
 \end{enumerate}
-TODO: \emph{more explanation of the provenance of these datasets}.
-
-Expression data for the AOCS and TCGA datasets was received for this project in a spreadsheet format, having been derived from the RNA-Seq data as described in \cite{Ewing2020}, including  normalization by variance stabilizing transformation.
+
+Expression data for the AOCS and TCGA datasets was received for this project in spreadsheet format, having been derived from the RNA-Seq data with normalization by variance stabilizing transformation applied.
+The provenance and processing of these datasets is described in \cite{Ewing2020}.
 
@@
 The diagramatic convention here is that at each stage, both datasets -- TCGA and AOCS -- are processed separately by the given algorithm.
-The exception is Gene Set Intersection, which involves interaction between the datasets.}
+The exception is Gene Set Intersection, which involves interaction between the datasets.
+(The investigation of batch effects on the combined dataset is not shown in this diagram to avoid over complexity).
+}
 \label{fig-method-overview}
@@
 \item
-Investigation of biological significance of determined factors by gene enrichment analysis (GEA) against the Gene Ontology (GO).
+Investigation of biological significance of extracted metagenes by gene enrichment analysis (GEA) against the Gene Ontology (GO).
 \item
@@
 \item 
-Survival analysis on TCGA and AOCS to investigate relationship between metasamples and patient survival.
+Survival analysis on TCGA and AOCS to investigate relationships between metasamples and patient survival.
 \item
 Scatter plots and heat maps to investigate the relationships between metasamples and genomic features available for the AOCS dataset.
+\item
+Investigation of batch effects by repeating the above analysis pipeline on the \emph{combined} AOCS / TCGA dataset (n=80+273=353).
 
@@
 
-These steps are detailed in the following sections.
+These steps are detailed in the following main sections.
 
@@
 One of the aims of this work is to compare the efficacy of three methods of dimensionality reduction -- NMF, ICA and PCA -- as explained in the introduction.     
-%The three methods have different properties and notational conventions, but to simplify the discussion here we adopt the notation $X \approx W H$ for all three methods, where $X$ is the (genes, patients) expression matrix  $W$ is the (genes, factors) metagenes matrix and $H$ is the (factors, patients) metasamples matrix.
-
 Some of the key algorithm hyper parameters of each method were investigated and tuned with respect to reconstruction accuracy, specifically the root-mean square (RMS) difference between $X$ and $W H$.
@@
 \item[NMF:] Parameters {\tt max\_iter} (algorithm iterations) and {\tt tol} (tolerance of convergence) were optimised for good accuracy and acceptable execution time.   
+{\tt max\_iter} = 5000 and {\tt tol} = 0.01 was used.
 Other parameters of interest are {\tt alpha} (multipluer for regulation term) and {\tt l1\_ratio} (multiplier for L1 regularization, when {\tt alpha} $> 0$).
@@
 \item[ICA:]  Parameters {\tt max\_iter} and {\tt tol} were investigated as for NMF above. 
+{\tt max\_iter} = 5000 and {\tt tol} = ($10^{-5}$) was used.
 Additionally, options for the entropy functional which forms the basis of the optimiziation were investigated.
@@
 However, {\tt sklearn.decomposition.PCA} uses a more efficient 'randomized' algorithm when the given matrix is larger than 500 in both dimensions.
-Thus in our use case PCA is seen to have stochastic behaviour.
+Thus in our use case PCA is seen to have (slightly) stochastic behaviour.
 \end{description}
-
-\emph{TODO: probably best to state the chosen hyper parameters here}
 
@@
 A second and more fundamental source of variation of relevance to all three factorization methods is \emph{sampling error}.
-Our factorizations are based on a small ($N=80$ or $N=374$) sample of patients drawn from the population of HGSOC patients; our particular datasets are just two examples of many different `draws' which could have been made from that population.
+Our factorizations are based on a small ($N=80$ or $N=374$) sample of patients drawn from the population of HGSOC patients; our particular datasets are just two examples of many different ``draws" which could have been made from that population.
 
@@
 For this work 50 repeats were performed, being a compromise between achieving an adequate simulation without overly burdensome computation time.
-The overall process is summarised as pseudocode in figure \ref{fig-clustering-psuedocode}.
+
+The overall process is described below and summarised as pseudocode in figure \ref{fig-clustering-psuedocode}.
 
@@
 $180^\circ$ rotated vectors.
-These would appear as separate cluster in repeated sampling, and confound attempts to collect and aggregate.
+These would appear as separate clusters in repeated sampling, and confound attempts to collect and aggregate.
 The solution adopted here is to normalise each factor by requiring that the most extreme element -- i.e.
@@
 If sampling and algorithm initialisation error are modest then we expect to see $k$ tight clusters of points.
-To avoid lengthy computation in the t-SNE clustering, the metagenes were first reduced to $r=20$ dimensions by PCA; brief experiments showed that this reduction had negligible effect on the t-SNE visualisation for $r \geq 10$.
+To avoid lengthy computation in the t-SNE clustering, the metagenes were first reduced to $r=20$ dimensions by PCA; brief experiments showed that this reduction had negligible effect on the t-SNE visualisation providing $r \geq 10$.
 
@@
 
-The k-means clustering further allowed for a quantitative assessment of cluster coherence for the particular choice of factorizer and rank, via the \emph{silhouette score}.
-In brief, this is a measure of how close (by Euclidean distance) each point in a cluster is to other points in the cluster versus points \emph{not} in the same cluster.
-See
-\href{https://en.wikipedia.org/wiki/Silhouette_(clustering)}{Wikipedia: Silouette (clustering)}.
-\emph{TODO: make this a reference}
+The k-means clustering further allowed for a quantitative assessment of cluster coherence for the particular choice of factorizer and rank, via the \emph{silhouette score} (see \cite{Wikipedia}).
+In brief, this is a measure of how close (by Euclidean distance) each point in a cluster is to other points in the cluster versus points in \emph{other} clusters.
 The score is in the range -1 to +1, with 0 implying random scatter and 1 implying all points of a cluster perfectly overlay.
 
-The t-SNE plots and silhouette scores were assessed visually to decide, for each of the three factorizers, the highest rank with good cluster coherence, those median metagenes to take forward to gene enrichment and survival and genomic feature correlation analysis.
+The t-SNE plots and silhouette scores were assessed visually to decide, for each of the three factorizers, the highest rank with good cluster coherence.
+The median points of those clusters then formed the metagenes to take forward to gene enrichment, survival and genomic feature correlation analysis.
 Note that the t-SNE plots are used only for visualisation; they are not involved in the computation of median metagenes or silhouette score.
 
-Initial investigation on the N=80 AOCS dataset showed very poor cluster coherence, even for k=2 (this will be seen in the presented results).
-Thus it was decided to compute and select metagenes only from the N=374 TCGA dataset.
+Initial investigation on the N=80 AOCS dataset showed very poor cluster coherence, even for k=2.   Thus, it was decided to compute and select metagenes only from the N=374 TCGA dataset.   
 The selection rationale is set out in the results section, but it is convenient to state here that the following ranks were selected:
 $K_{\mbox{\tiny NMF}} = 3, K_{\mbox{\tiny ICA}} = 5, K_{\mbox{\tiny PCA}} =3$.  Thus, there were $3 + 5 + 3 = 11$ metagenes taken forward for follow-on analysis.
+
+A point to be aware of is that because PCA metagenes are treated the same as NMF and ICA metagenes, being extracted by the same clustering approach, the \emph{ordering} of PCA components is not necessarily maintained.
+Thus, component 'PCA-1' is not necessarily the first PCA component  in the usual sense -- i.e.
+the component capturing greatest variance.
+This could be fixed but would require special case code.
 
@@
 
-It is pertinent to ask ``how similar are the 11 components which emerge from the foregoing factorization and cluster analysis?''.
+It is pertinent to ask ``how similar are the 11 metagenes which emerge from the foregoing factorization and cluster analysis?''.
 We expect those coming from a single factorization method to be distinct by construction.
@@
 
-An alternative to consider the intersection of the gene sets which each metagene highlights.
+An alternative is to consider the intersection of the gene sets which each metagene highlights.
 In the GEA which follows, candidate genes will be identified for each meatagene.
@@
 This is more stringent than the 0.05 value which is commonly used, and was chosen since multiple metagenes are being analysed, implying multiple hypothesis testing over and above that which is accounted for by FDR filtering within a single GEA run.
+(Arguably, $0.05 / 11 \approx 0.005$ should have been used).
 
@@
 
-The result of this analysis, per metagene, is a list of enriched (or depleted, however in our case no depleted terms were found) GO terms, each with an associated list of involved genes and a FDR significance level.
+The result of this analysis, per metagene, is a list of enriched (or depleted, however no depleted terms were found) GO terms, each with an associated list of involved genes and a FDR significance level.
 This list can be inspected directly for insights, but that misses the point of the GO, which is to organise the terms hierarchically by `is a' relationships.
@@
 
-\subsection{Transfer of learned metagenes to novel a dataset}
+\subsection{Transfer of learned metagenes to a novel dataset}
 \label{sec-transfer-to-novel}
@@
 
-It is fundamental to our approach that \emph{metagenes} determined on the basis of one cohort of data can be used to generate \emph{metasamples} for a different cohort.
+It is fundamental to our approach that \emph{metagenes} determined on the basis of one dataset of data can be used to generate \emph{metasamples} for a different dataset.
 In the exposition below we refer to these as the \emph{training} and \emph{validation} cohorts respectively.
-
 Rank $k$ matrix factorization on the training dataset $\trainset$ (e.g. TCGA) results in:
@@
 \end{equation}
-where $X_\trainset$ is the expression matrix of shape $(g_\trainset, n_\trainset)$, with $g_\trainset$ the number genes and $n_\trainset$ the number of patients.
-$W_\trainset$ is the metagene matrix of shape $(g_\trainset, k)$ and $H_t$ is the metasamples matrix of shape $(k, n_\trainset)$.
-
-We wish to apply the factorization learned on $\trainset$ to a novel dataset $\validset$ (specifically the AOCS dataset) of shape $(g_\validset, n_\validset)$. 
+where $X_\trainset$ is the expression matrix of shape $(m_\trainset, n_\trainset)$, with $m_\trainset$ the number genes and $n_\trainset$ the number of patients.
+$W_\trainset$ is the metagene matrix of shape $(m_\trainset, k)$ and $H_\trainset$ is the metasamples matrix of shape $(k, n_\trainset)$.
+
+We wish to apply the factorization learned on $\trainset$ to a novel dataset $\validset$ (specifically the AOCS dataset) of shape $(m_\validset, n_\validset)$. 
 Importantly, $n_\validset = 1$ reflects the application of these methods to a single patient in a clinical setting.
@@
 However, if patients in datasets $\trainset$ and $\validset$ are drawn from the same population (ovarian cancer patients), then $X_\trainset$ and $X_\validset$ can be expected to have similar distributions w.r.t to their columns, and thus $W_\trainset$ and $W_\validset$ can be expected to be equivalent within sampling error.
-This only makes sense if the two expression matrices $X_\trainset$ and $X_\validset$ are defined over the \emph{same set of genes}, so that $g_\trainset = g_\validset = g$, in which case $W_\trainset$ and $W_\validset$ have the same shape of $(g, k)$.
+This only makes sense if the two expression matrices $X_\trainset$ and $X_\validset$ are defined over the \emph{same set of genes}, so that $m_\trainset = m_\validset = m$, in which case $W_\trainset$ and $W_\validset$ have the same shape of $(m, k)$.
 
@@
 The Python library function {\tt scipy.optimize.nnls} is used.
-
-For ICA and PCA we use ordinary least square regression, formulated a above but dropping the $H\geq 0$ constraint, using {\tt scipy.linalg.lstsq}.
+For ICA and PCA we use ordinary least square regression, formulated as above but dropping the $H\geq 0$ constraint, using {\tt scipy.linalg.lstsq}.
 
@@
 
-Since in this work we wish to evaluate the efficacy of the three factorization methods -- NMF, ICA and PCA -- we in fact take forward three different $H$ matrices, each having ranks as for the associated metagenes.
+Since in this work we wish to evaluate the efficacy of the three factorization methods -- NMF, ICA and PCA -- we in fact take forward three different $H$ matrices, each having ranks as for the associated metagenes -- 3, 5 and 3 respectively.
 
@@
 
-To carry out patient level analysis -- survival analysis, heatmaps and boxplot -- as described in the following sections, it is necessary to compile tables (Pandas DataFrames) which bring together per-patient metasamples relating to each of the selected metagenes, with associated metadata -- such as cellularity and survival information.
-The metagenes always derive from the TCGA dataset (because of its larger size as explained in section \ref{sec-metagene-selection}), but we wish to apply these metagenes to the expression data and associated metadata of either the TCGA or AOCS datasets.
+To carry out patient level analysis -- survival analysis, heatmaps and genome correlation -- as described in the following sections, it is necessary to compile tables (Pandas DataFrames) which bring together per-patient metasamples relating to each of the selected metagenes, with associated metadata -- such as cellularity and survival information.
+The metagenes always derive from the TCGA dataset (because of its larger size as explained in section \ref{sec-metagene-selection}), but we wish to apply these metagenes as metasamples to the expression data and associated metadata of either the TCGA or AOCS datasets.
 The mathematics of transferring metagenes across datasets has been described above.
-Care is required to ensure that expression matrices are aligned correctly with metagenes with respect to their genes (rows), and with the metadata with respect to patient identifiers in the columns.
+Care is required to ensure that expression matrices are aligned correctly with metagenes with respect to their genes, and that metadata is aligned correctly with metasamples with respect to patient identifiers.
 Note that the same transferring approach is taken when metasamples are required for TCGA, even though we could in that case obtain the $H$ matrix directly from the factorization.
@@
 Application to TCGA (for OS) thus represents an in-sample test, while application to AOCS (for OS and PFS) is a more exacting out-of-sample test.
-There are thus three analyses to consider: 1) TCGA$\rightarrow$TCGA (OS), 2) TCGA$\rightarrow$AOCS (OS) and 3) TCGA$\rightarrow$AOCS (PFS).
-\emph{TODO: describe these three versions of analysis in words}.
+There are thus three analyses to consider:
+1) TCGA$\rightarrow$TCGA (OS) meaning TCGA derived metagenes are tested against OS on the TCGA dataset, 2) TCGA$\rightarrow$AOCS (OS) meaning TCGA derived metagenes are tested against OS on the AOCS dataset, and 3) TCGA$\rightarrow$AOCS (PFS)...
+as 2) but for PFS.
 
@@
 
-For the AOCS dataset (only) pre-determined per-patient high-level genomic features were available as follows:
+For the AOCS dataset (only) per-patient high-level genomic features were available as follows:
 \begin{description}
@@
 the probability that $r$ does not differ from zero) were calculated.
-However, since WGD is a binary feature, the more appropriate Point-Biserial correlation was used in that case.
+However, since WGD is a binary feature, the more appropriate Point-Biserial correlation (see \cite{Wikipediab}) was used in that case.
 A p-value significance threshold of 0.01 was chosen, although with 66 hypotheses being tested, this does risk false discovery.
@@
 \cite{Stein-OBrien2018,Renard2016}) that matrix factorization methods are an effective means of identifying and removing batch effects.
-To verify this, the TCGA and AOCS datasets (after gene set intersection), were horizontally concatenated producing an $n=80+274=354$ patient dataset with a substantial batch artefact.
-Survival data is available for both datasets, so this was preserved as metadata associated with the combined datasets.
-The previously described analyses of metagene extraction, metasample computation and clustered heatmap and survival analysis were repeated on this dataset.
+To verify this, the TCGA and AOCS datasets (after gene set intersection), were horizontally concatenated producing an $N=80+274=354$ patient dataset with a known batch artefact.
+All of previously described analyses were repeated for this combined dataset (however, survival analysis is not shown).
 Ideally, the factorization rank selected for each method should be reconsidered, since the larger dataset could justify extracting more components.
@@
 Genomic feature correlation w.r.t these newly extracted components was performed using the AOCS datasets.
-Theses results were inspected; is there evidence of some metasamples aligning with technical batch effects while others align with biological processes?
+These results were inspected for evidence of some metasamples aligning with technical batch effects while others align with biological processes.
 
@@
 A base class is used to present a uniform interface for the three factorization methods.
-Intermediate results are cached with lazy computation for efficient working without ordering dependency.
+Intermediate results are cached to file with lazy computation for efficient working without imposing evaluation order dependency.
 Unit tests are based on tiny expression matrices, just 100 genes by 10 patients, thus the whole test suite of over 65 tests executes in around 45 seconds and so can be run frequently.
@@
 A consistent colour scheme of \textcolor{blue}{blue}, \textcolor{orange}{orange} and \textcolor{olive}{green} is used for plots relating to NMF, ICA and PCA respectively.
-Metagenes are consistently referred to by, for example, ``NMF-2-of-3" -- meaning the 2nd component of the rank $k=3$ NMF factorization.
+In figures metagenes / metasamples are consistently referred to by, for example, ``NMF-2-of-3" -- meaning the 2nd component of the rank $k=3$ NMF factorization.
+In text this is shortened to ``NMF-2'' where the rank is understood.
 
@@
 
-From figure \ref{fig-AOCS-ica-fixed-vs-bootstrap} it can be seen that with sampling error excluded, clusters appear reasonably coherent, but when sampling error is modelled by bootstrap sampling then  the factorizations become much less stable.
+From figure \ref{fig-AOCS-ica-fixed-vs-bootstrap} it can be seen that with sampling error excluded (top), clusters appear reasonably coherent, but when sampling error is modelled by bootstrap sampling then  the factorizations become much less stable.
 This demonstrates that \emph{sampling} error is far greater than errors due to algorithm initialisation when the dataset is small (N=80), making metagene extraction unreliable.
 Figure  \ref{fig-TCGA-ica-fixed-vs-bootstrap} makes the same comparison for the lager N=374 TCGA dataset, in which it can be seen that the impact of sampling error is not so severe.
-For brevity only results for ICA are shown here; similar result for PCA and NMF can be found in the appendix, figures
-
-For this reason it is was decided to perform all metagene extraction on the larger TCGA dataset, in the expectation that the obtained metagenes will be more robust and likely to better generalise to other datasets.
+For brevity only results for ICA are shown here; similar result for NMF and PCA can be found in the appendix, figures \ref{fig-AOCS-nmf-fixed-vs-bootstrap} and \ref{fig-AOCS-pca-fixed-vs-bootstrap}.
+
+It is for this reason it is was decided to perform all metagene extraction on the larger TCGA dataset, in the expectation that the obtained metagenes will be more robust and likely to better generalise to other datasets.
 
@@
 For PCA, $k = 3$ or $4$ both seem reasonable.
-The following choices were made:
+As already noted earlier, the following choices were made:
 $K_{\mbox{\tiny NMF}}  =  3,
 K_{\mbox{\tiny ICA}}  =  5,
-K_{\mbox{\tiny PCA}}  =  3$.  These choices predicate all the results which follow.
+K_{\mbox{\tiny PCA}}  =  3$.
 
@@
 The artefact is difficult to explain.
-It cannot be the $180^\circ$ rotation issue discussed earlier, since this does not apply to the all +ve components.
-\emph{TODO: hypothesis?}
+It cannot be the $180^\circ$ rotation issue discussed earlier, since this does not apply to the all +ve components; I have no explanation.
 
@@
 
-The Jaccard heatmap of figure \ref{fig-component-jaccard-similarities} shows that some pairs of metagenes identify overlapping sets of genes, in particular the pairs (ICA-1-of-5, PCA-2-of-3: J=0.61) and (ICA-5-of-5, PCA-3-of-3: J=0.49).
+The Jaccard heatmap of figure \ref{fig-component-jaccard-similarities} shows that some pairs of metagenes identify overlapping sets of genes, in particular the pairs (ICA-1, PCA-2: J=0.61) and (ICA-5, PCA-3: J=0.49).
 As expected however, there is very little similarity between components \emph{within} a factorization method.
@@
 The contributing candidate genes for all significantly enriched terms are shown beneath each figure.  
-High-level terms at depth less than 3 were removed, since these are generic (e.g. "regulation of biological process") and so uninteresting.   
-One component -- ICA-4-of-5 -- was problematic in that 49 GO terms (with depth $\geq 3$) were identified as significant.  
+High-level terms at depth less than 3 were removed, since these are generic (e.g. ``regulation of biological process") and so uninteresting.   
+One component -- ICA-4 -- was problematic in that 49 GO terms (with depth $\geq 3$) were identified as significant.  
 For this component it was necessary to limit the graphic to the 12 terms having the largest number of associated candidate genes.
 
-The four components NMF-3-of-3, ICA-1-of-5, ICA-2-of-5 and PCA-2-of-3 yielded no significant enriched terms (for FDR $\leq$ 0.01).
+The four components NMF-3, ICA-1, ICA-2 and PCA-2 yielded no significant enriched terms (for FDR $\leq$ 0.01).
 
@@
 \vspace{0.5cm}
-{\small Component NMF-1-of-3} \\
-\end{center}
-\caption{Lineage maps of enriched Gene Ongology (GO) terms for components NMF-1-of-3  In these diagrams, enriched terms are coloured purple, while there ancestors in the ontology are yellow.  }
+{\small Component NMF-1} \\
+\end{center}
+\caption{Lineage maps of enriched Gene Ongology (GO) terms for components NMF-1  In these diagrams, enriched terms are coloured purple, while there ancestors in the ontology are yellow.  }
 \label{fig-NMF_1-go-lineage}
@@
 \vspace{0.5cm}
-{\small Component NMF 2-of-3} \\
-\caption{Lineage maps of enriched GO terms for components NMF-2-of-3.   (NMF-3-of-3 produced no significant enrichment results)}
+{\small Component NMF 2} \\
+\caption{Lineage maps of enriched GO terms for components NMF-2.   (NMF-3 produced no significant enrichment results)}
 \label{fig-NMF_2-go-lineage}
@@
 \vspace{0.5cm}
-{\small Component ICA-3-of-5} \\
-\end{center}
-\caption{Lineage maps of enriched GO terms for components ICA-3-of-5}
+{\small Component ICA-3} \\
+\end{center}
+\caption{Lineage maps of enriched GO terms for components ICA-3}
 \label{fig-ICA-3-5-go-lineage}
@@
 \vspace{0.5cm}
-{\small Component ICA-5-of-5} \\
-\caption{Lineage maps of enriched GO terms for components ICA-5-of-5}
+{\small Component ICA-5} \\
+\caption{Lineage maps of enriched GO terms for components ICA-5}
 \label{fig-ICA-5-5-go-lineage}
@@
 \vspace{0.5cm}
-\caption{Lineage maps of enriched GO terms for component ICA-4-of-5. For this component it was necessary to additionally filter the enriched terms -- see text.}
+\caption{Lineage maps of enriched GO terms for component ICA-4. For this component it was necessary to additionally filter the enriched terms -- see text.}
 \label{fig-ICA-4-go-lineage}
@@
 \vspace{0.5cm}
-{\small Component PCA-1-of-3} \\
-\end{center}
-\caption{Lineage maps of enriched GO terms for component PCA-1-of-3  }
+{\small Component PCA-1} \\
+\end{center}
+\caption{Lineage maps of enriched GO terms for component PCA-1  }
 \label{fig-PCA-1-3-go-lineage}
@@
 \vspace{0.5cm}
-{\small Component PCA-3-of-3} \\
-\end{center}
-\caption{Lineage maps of enriched GO terms for component PCA-3-of-3. (Component PCA-2-of-3 produced no significant enrichment results)  }
+{\small Component PCA-3} \\
+\end{center}
+\caption{Lineage maps of enriched GO terms for component PCA-3. (Component PCA-2 produced no significant enrichment results)  }
 \label{fig-PCA-3-3-go-lineage}
@@
 %
-\item[NMF 1-of-3]: there is mention of ECM related terms, as in ICA 3-of-5 and NMF 1-of-3 below.
+\item[NMF-1]: there is mention of extra-cellular matrix (ECM) related terms, as in ICA-3 and NMF-1 below.
 However, there are also terms relating to the regulation of angiogenesis.
@@
 %
-\item[NMF 2-of-3]: here we see enrichment of genes relating to the ribosomal subunit and the processes of RNA binding, implying perhaps a link with assembly of the ribosomal RNA-protein complex.
+\item[NMF-2]: here we see enrichment of genes relating to the ribosomal subunit and the processes of RNA binding, implying perhaps a link with assembly of the ribosomal RNA-protein complex.
 Ribosomes are known to have a role in carcinogenesis, by dysregulation the RNA $\rightarrow$ protein translation, or mutations in ribosomal subunits impacting on cellular metabolism \cite{Sulima2017}.
 %
-\item[NMF 3-of-3, ICA 1-of-5 and ICA 2-of-5]: no significant biological enrichment found.
+\item[NMF-3, ICA-1 and ICA-2]: no significant biological enrichment found.
 %
-\item[ICA 3-of-5]: this component relates to processes of multicellular / extracellular organisation, and the ECM.  It is therefore similar to NMF 1-of-3 above.
-\item[ICA 4-of-5]: many processes are highlighted by the component.
-These include processes relating to the ribosomal subunit (as NMF 2-of-3), additionally membrane proteins in the respiratory complex, mitochondrial and the NADH dehydrogenase complex.
+\item[ICA-3]: this component relates to processes of multicellular / extracellular organisation, and the ECM.  It is therefore similar to NMF-1 above.
+\item[ICA-4]: many processes are highlighted by the component.
+These include processes relating to the ribosomal subunit (as NMF-2), additionally membrane proteins in the respiratory complex, mitochondrial and the NADH dehydrogenase complex.
 Processes of organonitrogen biosynthesis are also highlighted.
 %
-\item[ICA 5-of-5]: processes relating to regulation of immune response are enriched in this component, featuring genes from the major histocompatibility complex (MHC) group, in particular the HLA- genes, allowing the immune system to recognise self from non-self.
+\item[ICA-5]: processes relating to regulation of immune response are enriched in this component, featuring genes from the major histocompatibility complex (MHC) group, in particular the HLA- genes, allowing the immune system to recognise self from non-self.
 GO terms relating to the endoplasmic reticulum (ER) are also highlighted.
-It may be that this component is mainly sensitive to the immunohistochemical signature of patients, and therefore not of clinical interest.
+It may be that this component is mainly sensitive to the immunohistochemical signature of patients, and therefore not of clinical interest \footnote{This turns out to be an incorrect conclusion: note presence of gene CD38 and see the Discussion!}.
 %
-\item[PCA 1-of-3]: this is related to the cellular structures and processes of the extra-cellular matrix (ECM), that is the proteins such as collagens which mediate the three-dimensional organisation of cells in a tissues.
+\item[PCA-1]: this is related to the cellular structures and processes of the ECM, that is the proteins such as collagens which mediate the three-dimensional organisation of cells in a tissues.
 ECM molecular composition will vary substantially between tissue types, but is also known to play a part in many disease processes \cite{Theocharis2019}.
@@
 %
-\item[PCA 2-of-3]: no significant biological enrichment found.
+\item[PCA-2]: no significant biological enrichment found.
 %
-\item[PCA 3-of-3]: this component has some similarity with ICA 5-of-5, in that it refers to regulation of immune processes (HLA- genes) and ER membrane.
+\item[PCA-3]: this component has some similarity with ICA-5, in that it refers to regulation of immune processes (HLA- genes) and ER membrane.
 However, chemotaxis (cell movement) is also highlighted.
@@
 
-Kaplan-Meier overall survival (OS) plots stratified by each metasample component are shown for the TCGA dataset in figure \ref{fig-kaplan-meier-os-TCGA-TCGA} and for the AOCS dataset in figure \ref{fig-kaplan-meier-os-TCGA-AOCS}.
-Plots for progression free survival (PFS) on AOCS can be found in the appendix, figure \ref{fig-kaplan-meier-pfs-TCGA-AOCS}.
+Kaplan-Meier overall survival (OS) plots stratified by each metasample component are shown for the for the AOCS dataset in figure \ref{fig-kaplan-meier-os-TCGA-AOCS} and for the TCGA dataset in figure \ref{fig-kaplan-meier-os-TCGA-TCGA} (appendix).
+Plots for progression free survival (PFS) on AOCS are also in the appendix, figure \ref{fig-kaplan-meier-pfs-TCGA-AOCS}.
 All plots show 95\% confidence intervals and hazard ratio (HR) with associated p-value.
@@
 
-The component with the largest observed effect on survival is PCA 1-of-3, showing a reasonably consistent hazard ratio of around 1.3 ($2^{0.4}$) across the three experiments.
+The component with the largest observed effect on survival is PCA-1, showing a reasonably consistent hazard ratio of around 1.3 ($2^{0.4}$) across the three experiments.
 This has $p < 0.05$ in the larger TCGA dataset, but not in the smaller AOCS dataset.
@@
 
-%The dendrogram clustering of metasamples (columns, top of figures) gives insights into the similarity of each metasample component, while the clustering of patients (rows, left of figures) offers insights into sub-populations of patients and how (in the case of AOCS only) these relate to metadata variables such as cellularity and mutational load.
-%These topics will be taken up in the following discussion section.
-
-\subsection{Metasamples correlate with genomic features}
+\subsection{Metasamples correlate with some genomic features}
 
@@
 \end{center}
-\caption{For each metagenes a 'Y' indicates that a significant correlation was observed with: Survival, WGD, HRD, Mutational load,  CNV load and SV load.
+\caption{For each metagenes a 'Y' indicates that a significant correlation was observed with: Survival, WGD, HRD, Mutational load,  CNV load and SV load;  '?' indicates a questionable link; +ve / -ve correlation is indicated.
 The next column gives a terse summary of the GEA results.
@@
 These three components (one from each of the methods) are also observed to be correlated with Cellularity.
-The sense (+/-1) of the correlation is meaningful for NMF (not so for ICA or PCA) and we see that ECM and thence tissue heterogeneity is negatively correlated with cellularity.
+The sense (+/-) of the correlation is meaningful for NMF (not so for ICA or PCA) and we see that ECM and thence tissue heterogeneity is negatively correlated with cellularity.
 This seems reasonable, in that a low Cellularity implies that a greater proportion of surrounding non-tumour tissue is included in the sample, so increasing heterogeneity.
@@
 \item[No consistent link with survival.]
-The only components showing even correlation with patient survival PCA-1 and PCA-2.
+The only components showing even questionable correlation with patient survival are  PCA-1 and PCA-2.
 The first of these are associated with ECM terms, which may simply be due to Cellularity.
-The other term showes no biological enrichment.]
+The second shows no biological enrichment.
 Component ICA-4 highlights the greatest number of biological processes -- indeed the GO lineage map \ref{fig-ICA-4-go-lineage} needed to be pruned to fit on a page.
@@
 It may be that this component is simply picking up on the normal metabolic activity of the cells.
-\item[Is HRD linked with ribosomal processing?].
-HRD is correlated with NMF-2 and NMF-3.
+\item[Is HRD linked with ribosomal processing?]  HRD is correlated with NMF-2 and NMF-3.
 The first of these is associated with the ribonucleoprotein complex while the second shows no biological enrichment.
@@
 
-Here are presented results of applying the three MF methods to the \emph{combined} AOCS and TCGA datasets.
-
 \begin{figure}[htb!]
@@
 \includegraphics[scale=0.5]{\plotspath{HeatMaps/clustered_heatmap_BOTH_BOTH.pdf}} 
+\end{center}
 \caption{Heatmap of the metasamples matrix for the combined AOCS + TCGA dataset.  The 'Which' field indicates the source dataset; purple: AOCS, yellow:TCGA.}
 \label{fig-heatmap-BOTH}
-\end{center}
 \end{figure}
@@
 \includegraphics[scale=0.4]{\plotspath{BoxAndScatters/wgd_boxplot_PCA_BOTH_BOTH.pdf}}
+
+\end{center}
 \caption{Boxplots of metasamples from the combined AOCS + TCGA datasets, showing relationship between metasamples dataset origin; Which=0 or 1 for AOCS or TCGA respectively.}
-\end{center}
-\label{fig-wgd-boxplots-BOTH}
-\end{figure}
-
-From the clustered heatmap of figure \ref{fig-heatmap-BOTH} it is clear that the five components clustered to the left are highly sensitive to the batch effect.
+\label{fig-boxplots-BOTH}
+\end{figure}
+
+Here are presented results of applying the three MF methods to the \emph{combined} AOCS and TCGA datasets. 
+From the clustered heatmap of figure \ref{fig-heatmap-BOTH} it is clear that the five components clustered to the left are highly sensitive to the batch effect, while in the other components batch is hardly discernable.
+This is confirmed by the boxplots in figure \ref{fig-boxplots-BOTH}.
+
+Table \ref{tab-integrative-analysis-BOTH} summarises observation on this combined dataset.
+It can be seen that, with a few exceptions, a component is sensitive \emph{either} to Batch \emph{or} to some biological process, but seldom both.
 
@@
 \begingroup
-\scriptsize
 \renewcommand{\arraystretch}{1.4} % Default value: 1
@@
 \caption{In a similar format as table \ref{tab-integrative-analysis}, this summarises results of in respect of the \emph{combined AOCS and TCGA} datasets.
-The key point here is that a known batch effect has been introduced, reflected in the 'Which' column.
+The key point here is that a known batch effect has been introduced, reflected in the 'Batch' column.
+The `Enrichment' column simply indicates whether or not biological enrichment was found (the details have not been analysed).
 Note that since these metagenes have been re-extracted from different data, they are distinct from those analysed in the previous section}
@@
 These observations are based on GEA on the GO only.
-Possibly more insights could have been gained by GEA against KEGG\footnote{\href{https://www.genome.jp/kegg/pathway.html}{Kyoto Encyclopedia of Genes and Genomes}} Pathways; this was intended but had to be dropped due to shortage of time.
+Possibly more insights could have been gained by GEA against KEGG\footnote{\href{https://www.genome.jp/kegg/pathway.html}{Kyoto Encyclopedia of Genes and Genomes}} Pathways.
+(This was begun but had to be dropped due to shortage of time).
 
@@
 
-Weak evidence of a link between homologous repair deficiency (HRD) and RNA-binding was foud.
+Weak evidence of a link between homologous repair deficiency (HRD) and RNA-binding was found.
 Being the site of protein synthesis, ribosomes will be associated with most of what goes on in a the cell.
@@
 Some components were predictive in one or other of the two datasets, but not both, which is unconvincing.
-The factor which GEA indicated regulation of angiogenesis (NMF-1), and thence might have prognostic value, performed poorly in survival analysis.
+The factor for which GEA indicated regulation of angiogenesis (NMF-1), and thence might have prognostic value, performed poorly in survival analysis.
+So overall a negative result.
 
@@
 The package recommend ranks of 14, 19 or 30.
-Yet we have seen that such high ranks on the small n=80 dataset leads to highly unstable factorizations when bootstrap modelling of sampling error is included.
-Suspicion of this result lead to me roll my own implementation based on bootstrap resampling, resulting in much lower ranks (fewer metagenes) being selected than would otherwise be the case if only algorithmic randomness were considered.
+Yet we have seen that such high ranks on the small N=80 dataset leads to highly unstable factorizations when bootstrap modelling of sampling error is included.
+Suspicion of this result lead to me roll my own implementation based on bootstrap resampling as described, resulting in much lower ranks (fewer metagenes) being selected than would otherwise be the case if only algorithmic randomness were considered.
 
@@
 That is, the technical and biological variation largely (not completely) separate into different components.
-This is a useful property, particularly as the growing availability of transcriptomics datasets from resources such as the TCGA and GEO allows for the simultaneous analysis of dozens of distinct datasets with likely different technical provenance, for which hand crafted normalisation would be infeasible.
+This is a useful property, particularly as the growing availability of transcriptomics datasets from resources such as TCGA and GEO allows for the simultaneous analysis of dozens of distinct datasets with likely different technical provenance, for which hand crafted normalisation would be infeasible.
 
@@
 
-The original focus of this project was ICA and NMF since these are know to be able to disentangle variation in data with complex structure, not multivariate normally distributed.
+The original technical focus of this dissertation was ICA and NMF since these are know to be able to disentangle variation in data with complex structure, not multivariate normally distributed.
 PCA was added as a baseline dimensionality reduction method, just for comparison.
@@
 The Jaccard similarity heatmap could be used to verify this, and reject components which showed low cross-dataset similarity.
+This proposal chimes to some extent with the use of Reciprocally Best Hit graphs in \cite{Cantini2019} reviewed in section \ref{sec-comparative-dimensionality-studies}.
 \item
 {\bf Contrasting gene expression patterns between cancers}.
-The current work set out to find patters of gene expression in HGSOC.
+The current work set out to find patterns of gene expression in HGSOC.
 Yet it is hard to say the highlighted patterns are specific to HGSOC, or generic to all cancers.

### 2020-08-13 MSc_Dissertation_B156476.tex
Organised appendices

@@
 \usepackage{placeins}
+\usepackage{longtable}
 
@@
 
-GEA against the GO results in, for each of the 11 metagenes, a table of GO terms with a list of the candidate genes which have an annotated association to that term; this table can be found in the appendix, table \ref{tab-go-enrichment-results}.
+GEA against the GO results in, for each of the 11 metagenes, a table of GO terms with a list of the candidate genes which have an annotated association to that term.
+See appendix \ref{tab-go-enrichment-results} to access the full table.
 To take full advantage of the hierarchical nature of the GO, results are presented as lineage maps in figures \ref{fig-NMF_1-go-lineage} to \ref{fig-PCA-3-3-go-lineage}.  
@@
 \begin{center}
-\includegraphics[scale=0.5,draft]{\plotspath{BoxAndScatters/genomic_feature_scatters_TCGA_AOCS.pdf}} 
+\includegraphics[scale=0.4]{\plotspath{BoxAndScatters/genomic_feature_scatters_TCGA_AOCS.pdf}} 
 \caption{Grid of scatter plots to visualise the correlation between metasamples and genomic features.
@@
 \bibliographystyle{ieeetr}
-
+\normalsize
+
+\newpage
 \section{Appendices}
-\subsection{Additional figures and plots}
-
+\subsection{Metagene clustering -- additional figures}
+\def\S{0.35}
 \begin{figure}[htb!]
@@
 
-\begin{figure}[htb!]
-\begin{center}
-\includegraphics[scale=0.5]{\plotspath{SurvivalAnalysis/multiple_kaplan_meier_os_TCGA_TCGA.pdf}} 
+\FloatBarrier
+\subsection{Survival analysis -- additional figures}
+
+\begin{figure}[htb!]
+\begin{center}
+\includegraphics[scale=0.4]{\plotspath{SurvivalAnalysis/multiple_kaplan_meier_os_TCGA_TCGA.pdf}} 
 \caption{Kaplan-Meier plots for each metasample component, stratified at the median value, for TCGA $\rightarrow$ TCGA for overall survival (OS) case.
@@
 \begin{center}
-\includegraphics[scale=0.5]{\plotspath{SurvivalAnalysis/multiple_kaplan_meier_pfs_TCGA_AOCS.pdf}} 
+\includegraphics[scale=0.4]{\plotspath{SurvivalAnalysis/multiple_kaplan_meier_pfs_TCGA_AOCS.pdf}} 
 \caption{Kaplan-Meier plots for TCGA $\rightarrow$ TCGA for progression free survival (PFS).  }
@@
 \begin{center}
-\includegraphics[scale=0.7]{\plotspath{HeatMaps/clustered_heatmap_TCGA_AOCS.pdf}} 
+\includegraphics[scale=0.5]{\plotspath{HeatMaps/clustered_heatmap_TCGA_AOCS.pdf}} 
 \caption{Heatmap of the metasamples matrix for the AOCS dataset, computed from metagenes factorized from the TCGA dataset.   Several patient metadata variables are shown in columns to the left.}
@@
 
-\subsection{Additional plots relating to the batch effect investigation}
+\FloatBarrier
+
+\subsection{Batch effect investigation -- Additional figures}
 \begin{figure}[htb!]
@@
 \begin{center}
-\includegraphics[scale=0.5]{\plotspath{BoxAndScatters/genomic_feature_scatters_BOTH_AOCS.pdf}} 
+\includegraphics[scale=0.4]{\plotspath{BoxAndScatters/genomic_feature_scatters_BOTH_AOCS.pdf}} 
 \caption{Grid of scatter plots to visualise the correlation between metasamples derived from the metagenes of the combined AOCS + TCGA datasets, and genomic features of the AOCS.}
@@
 
-\subsection{Gene enrichment raw results}
+\FloatBarrier
+
+\subsection{Access to gene enrichment analysis table}
 \label{tab-go-enrichment-results}
-TODO
-
-\label{sec-software-versions}
+
+The table of GEA results is too large to conveniently include here.
+If required it can be obtained from my github repository at \url{https://github.com/ipoole/HgsocTromics/blob/master/Results/combined_gea_fdr.tsv}.
 

### 2020-08-14 MSc_Dissertation_B156476.tex
Improved integrative analysis table, tweaks to discussion, added conclusions.  Done?

@@
 
-%\usepackage{titlesec}
-%\titleformat{\subsubsection}
-%   {\itshape\normalsize}{\thesubsubsection}{1em}{}
+\usepackage{titlesec}
+\titleformat{\subsubsection}
+   {\itshape\normalsize}{\thesubsubsection}{1em}{}
 
@@
 \subsection{Patterns of gene expression in HGSOC}
-
+\label{sec-review-of-gene-expression}
 Identifying the patterns -- or ``signatures'' -- of gene expression has been an active research area, for the purposes of understanding the carcinogenesis pathway, subtype classification, predicting survival prognosis, predicting treatment response and understanding the causative mechanism of risk factors; there is much cross-over between these areas, however.
@@
 \item How should MF methods be applied to achieve robust results on datasets of limited size?
-\item Do MF detect and remove technical batch effects?
+\item Does MF detect and remove technical batch effects?
 \item Which MF method is best suited to transcripomics analysis?
@@
 These include processes relating to the ribosomal subunit (as NMF-2), additionally membrane proteins in the respiratory complex, mitochondrial and the NADH dehydrogenase complex.
-Processes of organonitrogen biosynthesis are also highlighted.
+Alsp processes of organonitrogen biosynthesis are also highlighted, purine nucleotide biosynthesis, RNA binding, proton membrane transport.
 %
@@
 \begin{center}
-\includegraphics[scale=0.4]{\plotspath{BoxAndScatters/genomic_feature_scatters_TCGA_AOCS.pdf}} 
+\includegraphics[scale=0.45]{\plotspath{BoxAndScatters/genomic_feature_scatters_TCGA_AOCS.pdf}} 
 \caption{Grid of scatter plots to visualise the correlation between metasamples and genomic features.
@@
 \renewcommand{\arraystretch}{1.4} % Default value: 1
-\begin{tabular}{|l||c|c|c|c|c|c|c|l|l|}
+\begin{tabular}{|l||c|c|c|c|c|c|c|p{4.5cm}|l|}
 \hline
@@
 NMF-1 & . & Y-& Y-& . & . & . & . & ECM, Reg. angiogenesis & ICA-3, PCA-1 \\
-NMF-2 & . & . & . & Y+& . & . & Y+& Ribo. subunit, RNA-binding & - \\
-NMF-3 & - & - & - & Y-& . & . & . & . & . \\
+NMF-2 & . & . & . & Y+& . & . & Y+& Ribo. subunit, RNA-binding & . \\
+NMF-3 & . & . & . & Y-& . & . & . & . & . \\
 \hline
@@
 ICA-2 & . & . & . & . & . & . & . & . & . \\
-ICA-3 & . & . & Y-& . & . & . & Y-& ECM, multicellular org. & NMF-1, PCA-1 \\
-ICA-4 & . & . & . & . & . & . & . & Ribo subunit, Mitoch. NADH dehydr. & - \\
-ICA-5 & . & Y+& Y+& . & . & . & . & Immune response, MHC & PCA-3 \\
+ICA-3 & . & . & Y-& . & . & . & Y-& ECM, multicellular/extracellular organisation & NMF-1, PCA-1 \\
+ICA-4 & . & . & . & . & . & . & . & Ribo. subunit, Mitochondria, NADH dehydrogenase, purine nucleotide synthesis, RNA binding, Proton membrane transport & . \\
+ICA-5 & . & Y+& Y+& . & . & . & . & ER, Immune response, MHC & PCA-3 \\
 \hline
-PCA-1 & ?+& . & Y-& . & . & . & . & Extra-cellular matrix (ECM)& NMF-1, ICA-3 \\
+PCA-1 & ?+& . & Y-& . & . & . & . & Extra-cellular matrix / organisation, & NMF-1, ICA-3 \\
 PCA-2 & ?+& . & . & . & . & . & . & . & ICA-1 \\
-PCA-3 & . & Y-& Y-& . & . & . & . & Immune response, MHC, chemotaxis & ICA-5 \\
+PCA-3 & . & Y-& Y-& . & . & . & . & ER, Immune response, MHC, chemotaxis & ICA-5 \\
 \hline
@@
 \begin{center}
-\includegraphics[scale=0.5]{\plotspath{HeatMaps/clustered_heatmap_BOTH_BOTH.pdf}} 
+\includegraphics[scale=0.5,draft=false]{\plotspath{HeatMaps/clustered_heatmap_BOTH_BOTH.pdf}} 
 \end{center}
@@
 Here are presented results of applying the three MF methods to the \emph{combined} AOCS and TCGA datasets. 
-From the clustered heatmap of figure \ref{fig-heatmap-BOTH} it is clear that the five components clustered to the left are highly sensitive to the batch effect, while in the other components batch is hardly discernable.
+From the clustered heatmap of figure \ref{fig-heatmap-BOTH} it is clear that the five components clustered to the left are highly sensitive to (100\% predictive of) the batch, while in the other components batch is hardly discernable.
 This is confirmed by the boxplots in figure \ref{fig-boxplots-BOTH}.
-
-Table \ref{tab-integrative-analysis-BOTH} summarises observation on this combined dataset.
+Within the larger TCGA dataset there is clear clustering into two groups with a roughly 5:6 ratio split.
+Does this perhaps indicate that the TCGA set is collected from two separate sources?
+Is the difference technical or are there substantially different cohorts of patients?
+
+Table \ref{tab-integrative-analysis-BOTH} summarises observations on this combined dataset.
 It can be seen that, with a few exceptions, a component is sensitive \emph{either} to Batch \emph{or} to some biological process, but seldom both.
+This is a useful property.
 
@@
 \begingroup
+\small
 \renewcommand{\arraystretch}{1.4} % Default value: 1
@@
 
+A note of caution: these results are based on testing $6 \times 11 = 66$ hypotheses at a significance level of $< 0.01$, so there will likely be at least one false discovery.
+A significance level of 0.001 would have been more appropriate.
+
 Expression signature correlations with mutational load and  CNV load have not been detected.
-These markers of genome instability are known to be very noisy, so if a correlation with expression does exist it would likely be a weak signal, below the detection power of the small (N=80) AOCS dataset.
 
@@
 
+Several other specific genes have been noted in section \ref{sec-review-of-gene-expression}; it would be interesting to test these also.
+
 \subsubsection*{How should MF methods be applied to achieve robust results on datasets of limited size?}
@@
 
-\subsubsection*{Do MF detect and remove technical batch effects?}
+\subsubsection*{Does MF detect and remove technical batch effects?}
 
@@
 In fact this is similar to the approach of \cite{WayGregory2019} mentioned in section \ref{sec-comparative-dimensionality-studies}.
+It will be important to take a principled approach to the setting of the FDR theshold when applying GEA, to account for the increased number of hypothesis being tested.
 \item
@@
 
-\FloatBarrier
+\clearpage
 
 \section{Conclusions}
-TODO
-
-\section*{Acknowledgements}
+
+This work has investigated the application of unsupervised matrix factorization (MF) methods  -- specifically non-negative matrix factorization (NMF), independent component analysis (ICA) and principal component analysis (PCA) -- to gene expression analysis in the context of high grade serous ovarian cancer (HGSOC).
+These methods aim to find metagenes which capture biologically or technically significant variation in the gene expression signal.
+Metagenes were extracted from a N=374 patient dataset.
+Key findings are:
+\begin{itemize}
+\item
+Biological processes relating to  extracellular matrix (ECM), ribosomal subunits, RNA binding, angiogenesis and immune response have been identified and may have significance to HGSOC.
+\item
+Metagenes have been found with possible correlation to genome level features, such as whole chomosome doubling, homologous repair deficiency, structural variant load.
+However, these findings have marginal statistical significance given the small size (N=80) of the available dataset.
+\item
+A metagene predictive of patient survival did \emph{not} emerge from these unsupervised methods.
+\item
+However, expression of a single gene, CD38, identified from the literature, was confirmed to have substantial impact on patient survival.
+\item
+When applying MF methods, particularly to small datasets, it is important to consider the impact of sampling error when deciding the number of metagenes to extract.
+Bootstrap resampling and cluster analysis is an effective approach.
+\item
+MF methods have been found effective in separating technical variation (batch effect) from biological variation.
+\item
+NMF, ICA and PCA each find mostly different metagenes. No clear evidence has been found to support one over the others.
+
+\end{itemize}
+
+\clearpage
+\subsection*{Acknowledgements}
 I am most grateful to Dr.
@@
 Ailith provided excellent, insightful supervision throughout and gave valuable feedback on an early draft of this dissertation.
-I'd like to thank my employer, Canon Medical Research Ltd., for allowing me the flexibility and study leave to undertake this MSc.
-Thank you to my family for excusing me from cooking and washing up duties during the more stressful phases of MSc study!
+I'd like to thank my employer, Canon Medical Research Europe Ltd., for allowing me the flexibility and study leave to undertake this MSc.
+Thank you to my family for excusing me from cooking and washing up duties during the more stressful phases of MSc study and dissertation writing!
 

### 2020-08-14 MSc_Dissertation_B156476.tex
Minor revisions after reiewing Ailith's comments.  DONE!

@@
     filecolor=blue,      
-    urlcolor=darkblue,
+    urlcolor=darkblue
 }
@@
 Two gene expression datasets are analysed; the Austrailian Ovarian Cancer Study dataset (N=80) has patient survival and genomic metadata, while the larger dataset from The Cancer Genome Atlas (N=374) has only survival metadata.
-This motivates an approach of unsupervised metagenes discovered by MF techniques in the larger N=374 dataset then transferring that information to the smaller N=80 dataset to test for their power to stratify patient survival by Cox's proportional hazards analysis, and for correlation with genomic features.
+This motivates an approach of unsupervised metagenes discovered by MF techniques in the larger dataset then transferring that information to the smaller dataset to test for their power to stratify patient survival by Cox's proportional hazards analysis, and for correlation with genomic features.
 
@@
 Too high a rank will result in metagenes which are unstable and fail to generalise to other datasets.
-From a review of the literature, it seems that the role of sampling error in the instability is an under appreciated issue, which this work addresses by  bootstrap resampling then assessing  coherence through k-means clustering and the Silhouette score.
+From a review of the literature, it seems that the role of sampling error in the instability is an under appreciated issue, which this work addresses by  bootstrap resampling then assessing  stability through k-means clustering and the Silhouette score.
 The similarity of the metagenes found by the three methods is assessed via the Jaccard index.
@@
 
-\subsubsection{Carcinogenesis pathways including homologous repair deficiency}
+\subsubsection{Carcinogenesis pathways including homologous repair deficiency (HRD)}
 
@@
 They leave open the question whether DNA repair by PARP plays a role.
-The relationship between BRCA1/2 and DNA repair -- specifically homologous repair deficiency (HRD) -- is picked up by Ewing \etal\ \cite{Ewing2020}.
+The relationship between BRCA1/2 and DNA repair -- specifically homologous repair deficiency -- is picked up by Ewing \etal\ \cite{Ewing2020}.
 They study the complex interplay of single nucleotide variants and large structural variants as they impact BRCA1/2 disruption and thence HRD.
 This work suggests that when BRCA1/2 loss is detected in HGSOC patients, there may be a clinical role for PARP inhibitors, since this would prevent error-prone non-homologous repair by PARP, and so selectively kill BRCA1/2 disrupted cells.
+
+A gene expression signature of HRD has been reported \cite{Lu2014} and shown to be predictive of overall and progression free survival in OC patients.
+The signature was derived by a supervised approach, from tumour samples known to be with or without HRD.
+Examining the paper's supplementary material, the signature is based on 116 genes, with +1 /-1 weights.
+Notable positively weighted genes include ABCF1, ACTL6B and ARID5A; negatively weighted include ZNF780A, WDR36 and TRPM7.
 
@@
 Our overall approach is to use \emph{unsupervised} machine learning methods to represent gene expression data in a small number of features, then to study whether these features correlate with clinical and biological variables.
+Unsupervised methods have the advantage of being applicable to large datasets for which metadata (e.g.
+clinical information) is not available.
+Results can then be transferred to smaller datasets which do have metadata.
 This is appropriate in our case, since of the two datasets available (described below), one (TCGA) has relatively large n but little available metadata, while the other (AOCS) is much smaller but has more useful metadata of genomic features.
@@
 \subsubsection{Datasets}
-Two gene expression datasets were assailable for this work (n refers to number of patients, g refers to number of genes).
+Two gene expression datasets were made available for this work (N refers to number of patients, m refers to number of protein coding genes).
 \begin{enumerate}
-\item The Cancer Genome Atlas (TCGA) derived, n=274 patiens, m=19,601 genes, with metadata on survival
-\item Australian Ovarian Cancer Study (AOCS) \cite{Patch2015}, n=80, m=19,730, with metadata on survival, cellularity and additional genomic features.
+\item The Cancer Genome Atlas (TCGA) derived, N=374 patiens, m=19,601 genes, with metadata on survival
+\item Australian Ovarian Cancer Study (AOCS) \cite{Patch2015}, N=80, m=19,730, with metadata on survival, cellularity and additional genomic features.
 Several of the studies reviewed earlier also use data from the Australian Ovarian Cancer Study (AOCS): \cite{Patch2015,Ewing2020,Cuello2018,Au-Yeung2014}.
@@
 Expression data for the AOCS and TCGA datasets was received for this project in spreadsheet format, having been derived from the RNA-Seq data with normalization by variance stabilizing transformation applied.
+The RNA-Seq data is publicaly available, downloadable from  the \href{https://dcc.icgc.org/pcawg}{Pan-cancer Analysis of Whole Genomes (PCAWG) data portal} for AOCS, and from the \href{https://portal.gdc.cancer.gov/}{Genomic Data Commonas (GDC) data portal} for TCGA.
 The provenance and processing of these datasets is described in \cite{Ewing2020}.
@@
 \item
-Computation by least squares optimisation of metasamples associated with with each TCGA derived metagene against TCGA and AOCS datasets.
+Computation by least squares optimisation of metasamples associated with each TCGA derived metagene against TCGA and AOCS datasets.
 \item
@@
 \item
-Investigation of batch effects by repeating the above analysis pipeline on the \emph{combined} AOCS / TCGA dataset (n=80+273=353).
+Investigation of batch effects by repeating the above analysis pipeline on the \emph{combined} AOCS / TCGA dataset (N=80+273=353).
 
@@
 {\tt max\_iter} = 5000 and {\tt tol} = 0.01 was used.
-Other parameters of interest are {\tt alpha} (multipluer for regulation term) and {\tt l1\_ratio} (multiplier for L1 regularization, when {\tt alpha} $> 0$).
+Other parameters of interest are {\tt alpha} (multiplier for regulation term) and {\tt l1\_ratio} (multiplier for L1 regularization, when {\tt alpha} $> 0$).
 L1 regularization  favours elements being precisely zero, whereas L2 regularization will encourage them to be small.
@@
 \item[ICA:]  Parameters {\tt max\_iter} and {\tt tol} were investigated as for NMF above. 
-{\tt max\_iter} = 5000 and {\tt tol} = ($10^{-5}$) was used.
-Additionally, options for the entropy functional which forms the basis of the optimiziation were investigated.
+{\tt max\_iter} = 5000 and {\tt tol} = $10^{-5}$ was used.
+Additionally, options for the entropy function which forms the basis of the optimiziation were investigated.
 \item[PCA:] This is in principle a deterministic algorithm based on eigenvector decomposition.
@@
 
-A complication arises in the computation of ICA and PCA facorizations, in that essentially the same factor can arise as $w$ or $-w$ -- i.e.
+A complication arises in the computation of ICA and PCA factorizations, in that essentially the same factor can arise as $w$ or $-w$ -- i.e.
 $180^\circ$ rotated vectors.
@@
 \label{sec-gea}
-The metagenes extracted by the above described factorization and clustering process provide valuable information into which genes vary in expression in the study samples; thus in our case, which genes are influential in HGSOC.
+The metagenes extracted by the above described factorization and clustering process provide valuable information into which genes vary in expression in the study samples; thus in our case, which genes are influential in the dominant expression signal in HGSOC.
 In order to gain insights into what biological processes are involved, GEA against the Gene Ontology (GO) was carried out for each of the 11 metagenes individually.
@@
 
-It is fundamental to our approach that \emph{metagenes} determined on the basis of one dataset of data can be used to generate \emph{metasamples} for a different dataset.
+It is fundamental to our approach that \emph{metagenes} determined on the basis of one (large but metadata poor) dataset can be used to generate \emph{metasamples} for a different (small but metadata rich) dataset.
 In the exposition below we refer to these as the \emph{training} and \emph{validation} cohorts respectively.
@@
 
-Since in this work we wish to evaluate the efficacy of the three factorization methods -- NMF, ICA and PCA -- we in fact take forward three different $H$ matrices, each having ranks as for the associated metagenes -- 3, 5 and 3 respectively.
+Since in this work the efficacy of three factorization methods are being studied (NMF, ICA and PCA), three different $H$ matrices are taken forward, each having ranks as for the associated metagenes (3, 5 and 3 respectively).
 
@@
 
-To carry out patient level analysis -- survival analysis, heatmaps and genome correlation -- as described in the following sections, it is necessary to compile tables (Pandas DataFrames) which bring together per-patient metasamples relating to each of the selected metagenes, with associated metadata -- such as cellularity and survival information.
+To carry out patient level analysis -- survival analysis, heatmaps and genome feature correlation -- as described in the following sections, it is necessary to compile tables (Pandas DataFrames) which bring together per-patient metasamples relating to each of the selected metagenes, with associated metadata -- such as cellularity and survival information.
 The metagenes always derive from the TCGA dataset (because of its larger size as explained in section \ref{sec-metagene-selection}), but we wish to apply these metagenes as metasamples to the expression data and associated metadata of either the TCGA or AOCS datasets.
@@
 There are thus three analyses to consider:
-1) TCGA$\rightarrow$TCGA (OS) meaning TCGA derived metagenes are tested against OS on the TCGA dataset, 2) TCGA$\rightarrow$AOCS (OS) meaning TCGA derived metagenes are tested against OS on the AOCS dataset, and 3) TCGA$\rightarrow$AOCS (PFS)...
+1) TCGA$\rightarrow$TCGA (OS) meaning TCGA derived metagenes are tested for OS on the TCGA dataset, 2) TCGA$\rightarrow$AOCS (OS) meaning TCGA derived metagenes are tested for OS on the AOCS dataset, and 3) TCGA$\rightarrow$AOCS (PFS)...
 as 2) but for PFS.
@@
 %Metasamples ($H$ matrices) were calculated from the metagenes and expression matrix as explained in section \ref{sec-transfer-to-novel} below.  
-Metasample values were binariased to 0, 1 by thresholding at the median value.
+Metasample values were binarized to 0, 1 by thresholding at the median value.
 Kaplan-Meier plots were made in respect of derived metasample for each of the three analyses -- making $11 \times 3$ plots each with two survival curves with 95\% confidence intervals.
 A hazard ratio (HR)
-was calculated for each plot by fitting Cox's proportional hazards model with p-value relating to the hypothesis that HR is significantly different to 1.0.
+was calculated for each analysis by fitting Cox's proportional hazards model with p-value relating to the hypothesis that HR is significantly different to 1.0.
 
@@
 \cite{Stein-OBrien2018,Renard2016}) that matrix factorization methods are an effective means of identifying and removing batch effects.
-To verify this, the TCGA and AOCS datasets (after gene set intersection), were horizontally concatenated producing an $N=80+274=354$ patient dataset with a known batch artefact.
-All of previously described analyses were repeated for this combined dataset (however, survival analysis is not shown).
+To verify this, the TCGA and AOCS datasets (after gene set intersection), were horizontally concatenated producing an $N=80+374=454$ patient dataset with a known batch artefact.
+All of the previously described analyses were repeated for this combined dataset (however, survival analysis is not shown).
 Ideally, the factorization rank selected for each method should be reconsidered, since the larger dataset could justify extracting more components.
@@
 Unit tests are based on tiny expression matrices, just 100 genes by 10 patients, thus the whole test suite of over 65 tests executes in around 45 seconds and so can be run frequently.
-The fully automated analysis pipeline takes approximately three hours from scratch, the majority of that time being taken for the repeated factorizations over the range of considered ranks.
+The fully automated (except for rank selection) analysis pipeline takes approximately three hours from scratch, the majority of that time being taken for the repeated factorizations over the range of considered ranks.
 The total codebase is approximately 2,900 lines of Python.
@@
 This demonstrates that \emph{sampling} error is far greater than errors due to algorithm initialisation when the dataset is small (N=80), making metagene extraction unreliable.
-Figure  \ref{fig-TCGA-ica-fixed-vs-bootstrap} makes the same comparison for the lager N=374 TCGA dataset, in which it can be seen that the impact of sampling error is not so severe.
+Figure  \ref{fig-TCGA-ica-fixed-vs-bootstrap} makes the same comparison for the larger N=374 TCGA dataset, in which it can be seen that the impact of sampling error is not so severe.
 For brevity only results for ICA are shown here; similar result for NMF and PCA can be found in the appendix, figures \ref{fig-AOCS-nmf-fixed-vs-bootstrap} and \ref{fig-AOCS-pca-fixed-vs-bootstrap}.
 
-It is for this reason it is was decided to perform all metagene extraction on the larger TCGA dataset, in the expectation that the obtained metagenes will be more robust and likely to better generalise to other datasets.
+For this reason it is was decided to perform all metagene extraction on the larger TCGA dataset, in the expectation that the obtained metagenes will be more robust and likely to better generalise to other datasets.
 
@@
 
+\FloatBarrier
+
 \subsection{Jaccard similarity shows metagenes identify intersecting gene sets in some cases}
@@
 
-\subsection{Metagenes highlight genes which are enriched for particular biological processes}
+\subsection{Metagenes highlight genes which are enriched for biological processes}
 
@@
 \begin{center}
-\includegraphics[width= \textwidth]{\plotspath{TCGA_OV_VST/GeneEnrichment/go_lineage_3_of_PCA_3.png}} \\
+\includegraphics[width= 0.7\textwidth]{\plotspath{TCGA_OV_VST/GeneEnrichment/go_lineage_3_of_PCA_3.png}} \\
 \begin{spacing}{0.6}
@@
 \includegraphics[scale=0.5]{\plotspath{SurvivalAnalysis/multiple_kaplan_meier_os_TCGA_AOCS.pdf}} 
-\caption{Kaplan-Meier plots for TCGA $\rightarrow$ TCGA for overall survival (OS).}
+\caption{Kaplan-Meier plots for TCGA $\rightarrow$ AOCS for overall survival (OS).}
 \label{fig-kaplan-meier-os-TCGA-AOCS}
@@
 PCA-1 & ?+& . & Y-& . & . & . & . & Extra-cellular matrix / organisation, & NMF-1, ICA-3 \\
-PCA-2 & ?+& . & . & . & . & . & . & . & ICA-1 \\
+PCA-2 & . & . & . & . & . & . & . & . & ICA-1 \\
 PCA-3 & . & Y-& Y-& . & . & . & . & ER, Immune response, MHC, chemotaxis & ICA-5 \\
@@
 \item[ECM relationship to Cellularity?]
-We see that NMF-1, ICA-3 and PCA-1 each have an ECM association and we have noted this could be related to tissue type heterogeneity.
+NMF-1, ICA-3 and PCA-1 each have an ECM association and we have noted this could be related to tissue type heterogeneity.
 These three components (one from each of the methods) are also observed to be correlated with Cellularity.
@@
 \item[No consistent link with survival.]
-The only components showing even questionable correlation with patient survival are  PCA-1 and PCA-2.
-The first of these are associated with ECM terms, which may simply be due to Cellularity.
-The second shows no biological enrichment.
+The only component showing even questionable correlation with patient survival is  PCA-1.  This is associated with ECM terms, which may simply be due to Cellularity.  
 Component ICA-4 highlights the greatest number of biological processes -- indeed the GO lineage map \ref{fig-ICA-4-go-lineage} needed to be pruned to fit on a page.
@@
 \item
-Metagenes have been found with possible correlation to genome level features, such as whole chomosome doubling, homologous repair deficiency, structural variant load.
+Metagenes have been found with possible correlation to genome level features of whole chomosome doubling, homologous repair deficiency and structural variant load.
 However, these findings have marginal statistical significance given the small size (N=80) of the available dataset.

### 2020-08-14 MSc_Dissertation_B156476.tex
Grrr, needed to adjust margins which required some squeezing of figures an text.  Adding .pdf.  Hopefully no really done!

@@
 
-\usepackage[margin=1.5cm]{geometry}
+%\usepackage[margin=1.5cm]{geometry}
+\usepackage
+[
+        left=3cm,
+        right=2cm,
+        top=2cm,
+        bottom=3cm,
+        % use vmargin=2cm to make vertical margins equal to 2cm.
+        % us  hmargin=3cm to make horizontal margins equal to 3cm.
+        % use margin=3cm to make all margins  equal to 3cm.
+]{geometry}
 \usepackage[utf8]{inputenc}
@@
 \begin{center}
-\huge
+\Large
 \vspace{0.3in}
@@
 Gene expression analysis -- transcriptomics -- of a set biopsied tumour tissues can identify genes whose expression is linked to the pathological processes of the tumour.
-In contrast to supervised contrastive approaches (e.g.
-comparing normal with tumour samples to identify over/under expressed genes), unsupervised approaches work with datasets of mixed and unknown tumour status.
-Matrix factorization (MF) methods have become a popular method of reducing and ``disentangling'' the salient components of variation in such datasets, known as metagenes.
+In contrast to supervised contrastive approaches unsupervised approaches work with datasets of mixed and unknown tumour status.
+Matrix factorization (MF) methods have become a popular means of reducing and ``disentangling'' the salient components of variation in such datasets, known as metagenes.
 From these metagenes the most influential genes can be identified, and when subjected to gene enrichment analysis against the Gene Ontology (GO) can provide insights into the underlying biological processes in play.
@@
 Two gene expression datasets are analysed; the Austrailian Ovarian Cancer Study dataset (N=80) has patient survival and genomic metadata, while the larger dataset from The Cancer Genome Atlas (N=374) has only survival metadata.
-This motivates an approach of unsupervised metagenes discovered by MF techniques in the larger dataset then transferring that information to the smaller dataset to test for their power to stratify patient survival by Cox's proportional hazards analysis, and for correlation with genomic features.
+This motivates an approach of unsupervised metagenes discovered by MF techniques in the larger dataset then transferring that information to the smaller dataset to test for their power to stratify patient survival and for correlation with genomic features.
 
@@
 While these are all linear factorization methods, they have very different properties.
-A key problem in applying MF is deciding on the number of components to extract -- the factorization rank.
-Too high a rank will result in metagenes which are unstable and fail to generalise to other datasets.
-From a review of the literature, it seems that the role of sampling error in the instability is an under appreciated issue, which this work addresses by  bootstrap resampling then assessing  stability through k-means clustering and the Silhouette score.
-The similarity of the metagenes found by the three methods is assessed via the Jaccard index.
+A key problem is deciding on the number of components to extract -- the factorization rank.
+Too high a rank will result in metagenes which are unstable and so fail to generalise.
+The role of sampling error in the instability seems under appreciated in the literature, which this work addresses by bootstrap resampling then assessing  stability through k-means clustering and the Silhouette score.
+Similarity of metagenes found by the three methods is assessed via the Jaccard index.
 
@@
 \subfloat[Optimising independence of Meta\underline{genes}]{
-\begin{tikzpicture}[node distance = 0.30cm, scale=1.3, every node/.style={scale=1.3}]
+\begin{tikzpicture}[node distance = 0.30cm, scale=1.1, every node/.style={scale=1.1}]
 	\node (X) [draw, fill=green, minimum height=\M, minimum width=\N] {\large X};
@@
 \subfloat[Optimising independence of Meta\underline{samples}]{
-\begin{tikzpicture}[node distance = 0.30cm, scale=1.3, every node/.style={scale=1.3}]
+\begin{tikzpicture}[node distance = 0.30cm, scale=1.1, every node/.style={scale=1.1}]
 	\node (X) [draw, fill=green, minimum height=\N, minimum width=\M] {\large X};
@@
 \begin{center}
-\includegraphics[scale=0.7]{methodlology_overview.pdf}
+\includegraphics[scale=0.6]{methodlology_overview.pdf}
 \end{center}
@@
 For this work 50 repeats were performed, being a compromise between achieving an adequate simulation without overly burdensome computation time.
-
-The overall process is described below and summarised as pseudocode in figure \ref{fig-clustering-psuedocode}.
+The process is described below and summarised as pseudocode in figure \ref{fig-clustering-psuedocode}.
 
@@
 where ${\lvert \mathbf{\cdot} \rvert}_2$ indicates Euclidean distance or L2-norm.
-The Python library function {\tt scipy.optimize.nnls} is used.
+The function {\tt scipy.optimize.nnls} is used.
 For ICA and PCA we use ordinary least square regression, formulated as above but dropping the $H\geq 0$ constraint, using {\tt scipy.linalg.lstsq}.
@@
 \begin{description}
-\item[WGD]: Whole genome doubling -- a marker of genome instability.  This is a binary feature.
+\item[WGD]: Whole genome doubling -- a marker of genome instability, a binary feature.
 \item[Cellularity]: proportion of cells belonging to the tumour (as opposed to surrounding normal tissue).
@@
 
-\subsection{Investigating the impact of batch effects by analysis of a combined dataset}
+\subsection{Investigating batch effects by analysis of a combined dataset}
 
@@
 Intermediate results are cached to file with lazy computation for efficient working without imposing evaluation order dependency.
-Unit tests are based on tiny expression matrices, just 100 genes by 10 patients, thus the whole test suite of over 65 tests executes in around 45 seconds and so can be run frequently.
+Unit tests are based on tiny expression matrices, just 100 genes by 10 patients, thus the whole test suite of over 65 tests executes in around 45 seconds.
 The fully automated (except for rank selection) analysis pipeline takes approximately three hours from scratch, the majority of that time being taken for the repeated factorizations over the range of considered ranks.
@@
 
-\def\S{0.45}
+\def\S{0.4}
 \begin{figure}[htb!]
@@
 \begin{center}
-\includegraphics[scale=0.5]{\plotspath{TCGA_OV_VST/GeneEnrichment/component_jaccard_heatmap.pdf}}
+\includegraphics[scale=0.4]{\plotspath{TCGA_OV_VST/GeneEnrichment/component_jaccard_heatmap.pdf}}
 \end{center}
@@
 To take full advantage of the hierarchical nature of the GO, results are presented as lineage maps in figures \ref{fig-NMF_1-go-lineage} to \ref{fig-PCA-3-3-go-lineage}.
+(The large graphic for ICA-4 is in the appendix, figure \ref{fig-ICA-4-go-lineage}).
 The contributing candidate genes for all significantly enriched terms are shown beneath each figure.  
@@
 \begin{center}
-\includegraphics[scale=0.09]{\plotspath{TCGA_OV_VST/GeneEnrichment/go_lineage_1_of_NMF_3.png}} \\
+\includegraphics[scale=0.08]{\plotspath{TCGA_OV_VST/GeneEnrichment/go_lineage_1_of_NMF_3.png}} \\
 \begin{spacing}{0.6}
@@
 \begin{center}
-\includegraphics[scale=0.1]{\plotspath{TCGA_OV_VST/GeneEnrichment/go_lineage_3_of_ICA_5.png}} \\
+\includegraphics[scale=0.09]{\plotspath{TCGA_OV_VST/GeneEnrichment/go_lineage_3_of_ICA_5.png}} \\
 \begin{spacing}{0.6}
@@
 \begin{center}
-\includegraphics[scale=0.07]{\plotspath{TCGA_OV_VST/GeneEnrichment/go_lineage_5_of_ICA_5.png}} \\
+\includegraphics[scale=0.06]{\plotspath{TCGA_OV_VST/GeneEnrichment/go_lineage_5_of_ICA_5.png}} \\
 \begin{spacing}{0.6}
@@
 \end{figure}
-
-\begin{sidewaysfigure}[htb!]
-\begin{center}
-\includegraphics[width=\textwidth]{\plotspath{TCGA_OV_VST/GeneEnrichment/go_lineage_4_of_ICA_5.png}} \\
-\begin{spacing}{0.6}
-{\small Genes: \tiny \input{/home/ipoole/Documents/gitrepos/HgsocTromics/Plots/TCGA_OV_VST/GeneEnrichment/go_gene_set_4_of_ICA_5.txt}}
-\end{spacing}
-\vspace{0.5cm}
-\caption{Lineage maps of enriched GO terms for component ICA-4. For this component it was necessary to additionally filter the enriched terms -- see text.}
-\label{fig-ICA-4-go-lineage}
-\end{center}
-\end{sidewaysfigure}
 
@@
 \begin{center}
-\includegraphics[scale=0.5]{\plotspath{SurvivalAnalysis/multiple_kaplan_meier_os_TCGA_AOCS.pdf}} 
+\includegraphics[scale=0.4]{\plotspath{SurvivalAnalysis/multiple_kaplan_meier_os_TCGA_AOCS.pdf}} 
 \caption{Kaplan-Meier plots for TCGA $\rightarrow$ AOCS for overall survival (OS).}
@@
 \begin{center}
-\includegraphics[scale=0.5]{\plotspath{/SurvivalAnalysis/survival_summary_barplots_NMF.pdf}} \\
-\includegraphics[scale=0.5]{\plotspath{/SurvivalAnalysis/survival_summary_barplots_ICA.pdf}} \\
+\includegraphics[scale=0.4]{\plotspath{/SurvivalAnalysis/survival_summary_barplots_NMF.pdf}} \\
+\includegraphics[scale=0.4]{\plotspath{/SurvivalAnalysis/survival_summary_barplots_ICA.pdf}} \\
 \includegraphics[scale=0.5]{\plotspath{/SurvivalAnalysis/survival_summary_barplots_PCA.pdf}}
@@
 \begin{center}
-\includegraphics[scale=0.5,draft=false]{\plotspath{HeatMaps/clustered_heatmap_BOTH_BOTH.pdf}} 
+\includegraphics[scale=0.4,draft=false]{\plotspath{HeatMaps/clustered_heatmap_BOTH_BOTH.pdf}} 
 \end{center}
@@
 \begin{center}
-\includegraphics[scale=0.4]{\plotspath{BoxAndScatters/wgd_boxplot_NMF_BOTH_BOTH.pdf}}
-\includegraphics[scale=0.4]{\plotspath{BoxAndScatters/wgd_boxplot_ICA_BOTH_BOTH.pdf}}
-\includegraphics[scale=0.4]{\plotspath{BoxAndScatters/wgd_boxplot_PCA_BOTH_BOTH.pdf}}
+\includegraphics[scale=0.36]{\plotspath{BoxAndScatters/wgd_boxplot_NMF_BOTH_BOTH.pdf}}
+\includegraphics[scale=0.36]{\plotspath{BoxAndScatters/wgd_boxplot_ICA_BOTH_BOTH.pdf}}
+\includegraphics[scale=0.36]{\plotspath{BoxAndScatters/wgd_boxplot_PCA_BOTH_BOTH.pdf}}
 
@@
 \begingroup
-\small
+\scriptsize
 \renewcommand{\arraystretch}{1.4} % Default value: 1
@@
 
-\FloatBarrier
+\clearpage
 
@@
 
-Addressing each of our research questions in turn:
+Addressing each research questions in turn:
 
@@
 \begin{center}
-\includegraphics[scale=0.7,draft=false]{\plotspath{SurvivalAnalysis/gene_specific_CD38_survival.pdf}}
+\includegraphics[scale=0.5,draft=false]{\plotspath{SurvivalAnalysis/gene_specific_CD38_survival.pdf}}
 \end{center}
@@
 \item
-{\bf Contrasting gene expression patterns between cancers}.
-The current work set out to find patterns of gene expression in HGSOC.
-Yet it is hard to say the highlighted patterns are specific to HGSOC, or generic to all cancers.
-One way of teasing out cancer specific patterns might be to take a similar approach to the previous item, but instead of looking for metagenes which are consistent between datasets, find those which are distinct.
-That said, there will likely be many more sophisticated approached in the literature, given the clinical importance of the topic.
-A direct supervised approach -- find the axis of variation which best separates the two cancers -- may be more appropriate.
-\item
 {\bf Systematic comparison with published gene expression patterns}.
@@
 \subsection{Metagene clustering -- additional figures}
-\def\S{0.35}
+\def\S{0.3}
 \begin{figure}[htb!]
@@
 \end{figure}
+
+\subsection{GEA -- additional plot}
+
+\begin{sidewaysfigure}[htb!]
+\begin{center}
+\includegraphics[width=\textwidth]{\plotspath{TCGA_OV_VST/GeneEnrichment/go_lineage_4_of_ICA_5.png}} \\
+\begin{spacing}{0.6}
+{\small Genes: \tiny \input{/home/ipoole/Documents/gitrepos/HgsocTromics/Plots/TCGA_OV_VST/GeneEnrichment/go_gene_set_4_of_ICA_5.txt}}
+\end{spacing}
+\vspace{0.5cm}
+\caption{Lineage maps of enriched GO terms for component ICA-4. For this component it was necessary to additionally filter the enriched terms -- see text.}
+\label{fig-ICA-4-go-lineage}
+\end{center}
+\end{sidewaysfigure}
 

### 2020-08-17 MSc_Dissertation_B156476.tex
Post submission: fixed typo in abstract; added my name.  Will now make repository public

@@
 \vspace{0.2in}
-Student Exam Number: \bf{B156476}
+Ian Poole \\
+Student Examination Number: \bf{B156476}
 \vspace{0.2in}
@@
 Gene expression analysis -- transcriptomics -- of a set biopsied tumour tissues can identify genes whose expression is linked to the pathological processes of the tumour.
-In contrast to supervised contrastive approaches unsupervised approaches work with datasets of mixed and unknown tumour status.
+In contrast to supervised contrastive approaches unsupervised techniques work with datasets of mixed and unknown tumour status.
 Matrix factorization (MF) methods have become a popular means of reducing and ``disentangling'' the salient components of variation in such datasets, known as metagenes.
@@
 In this work the biological target is high-grade serous ovarian cancer (HGSOC) -- the most common and aggressive type of ovarian cancer.
-Two gene expression datasets are analysed; the Austrailian Ovarian Cancer Study dataset (N=80) has patient survival and genomic metadata, while the larger dataset from The Cancer Genome Atlas (N=374) has only survival metadata.
+Two gene expression datasets are analysed; the Australian Ovarian Cancer Study dataset (N=80) has patient survival and genomic metadata, while the larger dataset from The Cancer Genome Atlas (N=374) has only survival metadata.
 This motivates an approach of unsupervised metagenes discovered by MF techniques in the larger dataset then transferring that information to the smaller dataset to test for their power to stratify patient survival and for correlation with genomic features.

### 2021-01-01 Methods_Paper.tex
Basic edits to paper rather than dissertation style

--- 2020-08-17 MSc_Dissertation_B156476.tex
@@
 
-\documentclass[tikz, 12pt,a4paper,oneside,fleqn]{article}
-
-%\usepackage[margin=1.5cm]{geometry}
-\usepackage
-[
-        left=3cm,
-        right=2cm,
-        top=2cm,
-        bottom=3cm,
-        % use vmargin=2cm to make vertical margins equal to 2cm.
-        % us  hmargin=3cm to make horizontal margins equal to 3cm.
-        % use margin=3cm to make all margins  equal to 3cm.
-]{geometry}
+\documentclass[tikz, 11pt,a4paper,oneside,fleqn]{article}
+
+\usepackage[margin=2.5cm]{geometry}
+%\usepackage
+%[
+%        left=3cm,
+%        right=2cm,
+%        top=2cm,
+%        bottom=3cm,
+%        % use vmargin=2cm to make vertical margins equal to 2cm.
+%        % us  hmargin=3cm to make horizontal margins equal to 3cm.
+%        % use margin=3cm to make all margins  equal to 3cm.
+%]{geometry}
 \usepackage[utf8]{inputenc}
@@
 
-\linespread{1.5}
-% This sets Arial font (bit ugly!)
-\usepackage{helvet}
-\renewcommand{\familydefault}{\sfdefault}
-\setlength{\parindent}{2em}
-\setlength{\parskip}{1em}
-\renewcommand{\baselinestretch}{1.5}
-
-\usepackage{titlesec}
-\titleformat{\subsubsection}
-   {\itshape\normalsize}{\thesubsubsection}{1em}{}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% SUBMISSION PAGE %%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\thispagestyle{empty}
-
-\title{Disentangling patterns of gene expression in high grade serous ovarian cancer by matrix factorization}
-\author{}
+%%%%%%%%%%%%%%%%%%%%%%%%%%% Start of paper for real %%%%%%%%%%%%%%%%%%%
+
+\title{MF Methods paper}
+\author{Ian Poole and Ailith Ewing}
+\date{January 2021}
 
@@
 
-%%%%%% Assesment Title Page
-\begin{center}
-\begin{center}
-\includegraphics[scale=0.3]{images/UoE_SBO_logo.png}
-\end{center}
-
-\vspace{0.3in}
-
-\begin{mdframed}
-\begin{center}
-\Large
-\vspace{0.3in}
-\bf
-Disentangling patterns of gene expression in high grade serous ovarian cancer by matrix factorization
-\vspace{0.2in}
-\end{center}
-\vspace{0.2in}
-\end{mdframed}
-
-\vspace{0.3in}
-
-\begin{mdframed}
-\begin{center}
-\large
-\vspace{0.2in}
-Ian Poole \\
-Student Examination Number: \bf{B156476}
-\vspace{0.2in}
-\end{center}
-\end{mdframed}
-
-\vspace{0.3in}
-
-\begin{mdframed}
-\begin{center}
-\large
-\vspace{0.2in}
-In partial fulfilment of the requirement for the Degree of
-Master of Science in Systems and Synthetic Biology at the
-University of Edinburgh,
-2019 / 2020
-\vspace{0.2in}
-\end{center}
-\end{mdframed}
-
-\vspace{0.3in}
-
-\begin{mdframed}
-\begin{center}
-\large
-\vspace{0.2in}
-
-Disssertation Supervisor:  Dr. Ailith Ewing
-\vspace{0.2in}
-\end{center}
-\end{mdframed}
-\end{center}
-
-\newpage
-\setcounter{page}{1}
-%\pagestyle{headings}
-
-\setlength{\parskip}{0em}
-\tableofcontents
-\setlength{\parskip}{1em}
-\newpage
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%% Start of report for real %%%%%%%%%%%%%%%%%%%
+\maketitle
 
@@
 \end{abstract}
+
+\maketitle
 \newpage
@@
 	\node [below=\fudge of X] {\tiny{$n$ Samples}};
-	\node [left=\fudge of X] {\rotatebox{90}{\tiny{$n$ Genes}}};
+	\node [left=\fudge of X] {\rotatebox{90}{\tiny{$m$ Genes}}};
 	\node [below=of X, text width=2cm, align=center] {\small Expression matrix};
@@
 	\node [below=\fudge of S] {\tiny{$k$ Metasamples}};
-	\node [left=\fudge of S] {\rotatebox{90}{\tiny{$nN$ Samples}}};
+	\node [left=\fudge of S] {\rotatebox{90}{\tiny{$n$ Samples}}};
 	\node [below=of S, text width=2.5cm, align=center] {\small Metasamples};

### 2021-01-24 Methods_Paper.tex
starting to slim it down...

@@
 
-\documentclass[tikz, 11pt,a4paper,oneside,fleqn]{article}
+\documentclass[tikz, 11pt,a4paper,oneside,fleqn, draft]{article}
 
 \usepackage[margin=2.5cm]{geometry}
-%\usepackage
-%[
-%        left=3cm,
-%        right=2cm,
-%        top=2cm,
-%        bottom=3cm,
-%        % use vmargin=2cm to make vertical margins equal to 2cm.
-%        % us  hmargin=3cm to make horizontal margins equal to 3cm.
-%        % use margin=3cm to make all margins  equal to 3cm.
-%]{geometry}
+
 \usepackage[utf8]{inputenc}
@@
 
-\title{MF Methods paper}
+\title{Accounting for sampling error in when applying matrix factorization to gene expression data}
+
 \author{Ian Poole and Ailith Ewing}
@@
 
+\tableofcontents
+
+\newpage
+
 \section{Introduction}
 
-\subsection{Epithelial Ovarian Cancer}
+\subsection{- - Epithelial Ovarian Cancer}
 \label{sec-epithelial-ovarian-cancer}
@@
 
-\subsection{Patterns of gene expression in HGSOC}
+\subsection{- - Patterns of gene expression in HGSOC}
 \label{sec-review-of-gene-expression}
@@
 
-\subsubsection{Carcinogenesis pathways including homologous repair deficiency (HRD)}
-
-BRCA1 inactivation is known to cause chromosomal instability in many cancers.
-Pradhan \etal\ \cite{Pradhan2010} investigated the role of BRCA1 in HGSOC by copy number and expression analysis, finding surprisingly that inactivation has no relationship with gross genomic alteration.
-They leave open the question whether DNA repair by PARP plays a role.
-The relationship between BRCA1/2 and DNA repair -- specifically homologous repair deficiency -- is picked up by Ewing \etal\ \cite{Ewing2020}.
-They study the complex interplay of single nucleotide variants and large structural variants as they impact BRCA1/2 disruption and thence HRD.
-This work suggests that when BRCA1/2 loss is detected in HGSOC patients, there may be a clinical role for PARP inhibitors, since this would prevent error-prone non-homologous repair by PARP, and so selectively kill BRCA1/2 disrupted cells.
-
-A gene expression signature of HRD has been reported \cite{Lu2014} and shown to be predictive of overall and progression free survival in OC patients.
-The signature was derived by a supervised approach, from tumour samples known to be with or without HRD.
-Examining the paper's supplementary material, the signature is based on 116 genes, with +1 /-1 weights.
-Notable positively weighted genes include ABCF1, ACTL6B and ARID5A; negatively weighted include ZNF780A, WDR36 and TRPM7.
-
-Studying the combined landscape of chromosomal rearrangements, driver mutations (particularly TP53, BRCA1/2), methylation and gene expression profiles has lead to a better understanding of the molecular events involved in the progression HGSOC and the development of chemoresistance \cite{Patch2015}.
-The sequence of events is complex, however, and not easily summarised here.
-A similar integrative / multiplatform approach is taken in \cite{He2017a} and \cite{Hoadley2014}.
-
-\subsubsection{Subtype classification}
-\label{sec-subtype-classification}
-
-A relatively early (2003) expression microarray based work demonstrated clustering into lymphocyte related genes (IGH3, IGKC, IGLJ3), extracellular matrix/stromal related (COL11A1, COL3A1, MMP2, SPARC, RBBP1) and six other clusters \cite{Schaner2003}.
-
-Wang \etal\ \cite{Wang2017c} use NMF clustering (see section \ref{sec-nmf}) to identify five subtypes of HGSOC informative of outcomes: 1:mesenchymal, 2:immunoreactive, 3:proliferative, 4:differentiated and 5:anti-mesenchymal.
-Subtypes 2 and 5 are found to be associated with better survival.
-
-Disruption of the PI3K-AKT signalling pathway is known to be significant in several cancers, due to its role in regulating apoptosis.  
-Espinosa \etal\ \cite{Espinosa2011} used unsupervised clustering techniques to investigate expression of 22 genes of this pathway.  
-They found that HGSOC cases formed two separate clusters; the cluster with high expression of CASP3, XIAP, NFKB1, FAS and GSK3B was linked with better outcomes.
-
-An integrative network approach has identified two distinct subtypes of ovarian cancer, distinguished by differences in expression of genes related to transcription factors which influence angiogenesis \cite{Glass2015}.
-The authors suggest that the regulatory networks highlighted by their analysis can lead to targetting with specific drugs.
-
-A recent pre-publication develops a pipeline for sub-type prediction with the potential to be used clinically \cite{Talhouk2020}.
-The NanoString gene expression platform is used.  Their strategy was to focus on small set of 513 genes (vs $\sim$21,000 protein coding genes) known from the literature to have relevance to subtyping.
-Two approaches were followed by different teams.
-``All Array": used expression array data from 1650 patients across 14 studies evaluating 9 supervised learning algorithms by bootstrap, selecting an AdaBoost -like method.
-The ``TCGA" team used 434 patients from The Cancer Genome Atlas (TCGA\footnote{I'm always amused by the genetics pun in the naming of this resource!}) evaluating 5 algorithms by cross-validation, selecting a random forest.
-``All Array" had the advantage of more data but needed careful attention to batch effects.
-A final classifier took a consensus of the two methods based on a minimal gene set (around 40), validated by a leave-one-out (patient level) approach.
-Survival analysis was carried out stratified by predicted subtype.
-
-\subsubsection{Treatment response and survival}
-
-As we have seen in section \ref{sec-epithelial-ovarian-cancer}, EOC subtypes are closely linked with survival, never-the-less much research is directed at finding direct links with expression.
-
-The prognostic value of expression signatures is demonstrated by the ``Classification of Ovarian Cancer" (CLOVAR) system \cite{Verhaak2013}.  
-Single-sample gene set enrichment analysis (ssGSEA) is used to obtain scores for previously reported expression signatures: differentiated, immunoreative, mesenchymal and proliferative.  
-An outcome prediction model based on these signatures gave good survival stratification,, with likelihood ratio 10.8 between above/below median groups on a validation dataset which improved to 15.7 when augmented with BRCA1/2 mutation status and other post-operative clinical factors.
-
-An explicit supervised learning approach can be taken, as in Berchuck \etal, where a hybrid decision tree plus linear discriminant model demonstrated survival stratification in both cross-validation and external validation experiments, based on microarray expression data.
-They found expression of CSFT3, ABCD3, MAL and APMCF1 to be the most influential; in fact the decision tree was almost entirely driven by the first two of these.
-
-Mairinger \etal\ screened 770 immune related genes to identify 11 differentially expressed genes associated with response to platinum treatment.  
-They find that expression of HS11B1, DNBT1, CKLF, NUP107, CCL18, LY96, ATG7, SLAMF7, CXCL9 is associated with better survival, while IKBKG and SDHA associate with with poor survival.
-
-The platinum based drug Cisplatin is a key treatment in ovarian cancer, yet tumours often develop resistance, possibly related to the host's immune response.  
-Understanding and predicting such resistance is therefore of clinical importance and is addressed by Mairinger \etal\ \cite{Mairinger2019} through expression analysis using the NanoString platform with a panel of 770 immune related genes.
-They find the following genes to be significantly related to platinum resistance: KLRC1, TCF7, CD274, HSD11B1, COLEC12, PDGFC, FCF1, BMI1, TNFRSF9, ATG10, EWSR1.
-
-Finding a clear, robust correlation between gene expression patterns and survival is not straight forward.  
-One substantial meta-analysis applied 16 previously published gene sets to two studies from the Gene Expression Omnibus (GEO) (199 samples) which included survival information, and found no significant predictive power \cite{Fekete2012}.
-Although, within the same study, the authors did identify predictive genes, in particular SNGC, MAPT, ESR2 and PGR.
-
-Expression of a single gene -- CD38, which codes for a transmembrane glycoprotein -- has recently been shown to have survival prognostic value on its own \cite{Zhu2020}.
-CD38 has a role in the breakdown of nicotinamide adenine dinucleotide ($\mbox{NAD}^+$), and is also a cell marker of lymphocytes.
-CD38 is over expressed in myeloma cells, thence its importance as a prognostic marker and a potential drug treatment target.  
-It is worth mentioning that the analysis in this paper made use of the Gene Expression Profiling Interactive Analysis (GEPIA) database, which conveniently supports survival analysis with respect of any given gene, against a range of TCGA hosted gene expression databases.
-
-\subsubsection{Association with other risk factors}
-
-Obesity has been shown to negatively impact prognosis in ovarian cancer, possibly through difficulties in matching chemotherapy dose with patient BMI  \cite{Au-Yeung2014}.
-A more direct molecular link is proposed by Cuello \etal\ \cite{Cuello2018}.
-They used NMF based clustering to demonstrate that expression of genes linked with obesity and lipid metabolism impart poorer progression free survival, independent of known HGSOC driver gene expression.
-The link is complex however, there being many confounding non-molecular based reasons for a link between high BMI and HGSOC survival, including later diagnosis and compromised operability.
-Some studies (e.g.
-\cite{Hew2013}) conclude no overall link at all.
-
-%Stratification of HGSOC by immunohistochemistry is addresses Kamble \etal.
-\cite{Kamble2019}.
-They identify a panel of six biomarkers -- TCF21, E-cadherin, PARP1, Slug and AnnexinA2.
-These markers allow stratification into categories including Mesenchymal-to-epithelial transition, Homologous recombination repair and Epithelial-to-mesenchymal transition.
-This, they argue, allows selection of class-specific inhibitor drugs such as Olaparib and Rucaparib.
-
-\subsection{Matrix factorization for dimensionality reduction}
+\subsection{+ Matrix factorization for dimensionality reduction}
 \label{sec-matrix-factorization-intro}
@@
 The number of target dimensions (or components, factors, metagenes -- these terms are used somewhat synonymously) is referred to as the rank, $k$.
-\subsubsection{Non-negative matrix factorization (NMF)}
+\subsubsection{+ Non-negative matrix factorization (NMF)}
 \label{sec-nmf}
@@
 %
-\subsubsection{Independent component analysis (ICA)} 
+\subsubsection{+ Independent component analysis (ICA)} 
 The goal of ICA is to represent the given matrix with components which are statistically independent of each other.
@@
 
-\subsubsection{Principal component analysis (PCA)}
+\subsubsection{+ Principal component analysis (PCA)}
 PCA is a well known technique for dimensionality reduction, based on eigenvalue decomposition.
@@
 
-\subsubsection{Determining the optimum number of factors}
+\subsubsection{- - Determining the optimum number of factors}
 
@@
 
-\subsubsection{Comparisons dimensionality reduction methods for gene expression analysis}
+\subsubsection{- - Comparisons dimensionality reduction methods for gene expression analysis}
 \label{sec-comparative-dimensionality-studies}
@@
 
-\subsection{Research gaps and questions}
+\subsection{- - Research gaps and questions}
 Considering the foregoing review of the literature, there has clearly been much work by large teams of researchers directed at understanding the gene expression patterns in HGSOC, as one would expect given its clinical importance.
@@
 
-\section{Methodology}
-
-\subsection{Outline}
+\section{- - Methodology}
+
+\subsection{- - Outline}
 
@@
 
-\subsubsection{Datasets}
+\subsubsection{+ Datasets}
 Two gene expression datasets were made available for this work (N refers to number of patients, m refers to number of protein coding genes).
@@
 
-\subsubsection{Methods}
+\subsubsection{- - Methods}
 An overview of the methodology adopted in this work is shown in figure \ref{fig-method-overview}, consisting of:
@@
 
-\subsubsection{Tools}
+\subsubsection{- - Tools}
 \begin{itemize}
@@
 
-\subsection{Syncrhonisation of gene sets accross AOCS and TCGA}
-\label{sec:gene-set-intersection}
-
-In order to allow factorizations found in the TCGA dataset to be applied to the AOCS dataset it is necessary (or at least convenient) to synchronize the set of genes over which the expression matrices are defined.
-As provided to this project, the TCGA and AOCS datasets cover 19,610 and 19,730 protein coding genes respectively, 19,566 of which are common to both (according to ENSG encodings).
-Thus, both datasets were pruned to the 19,566 intersection set and ordered consistently.
-
-\subsection{Matrix factorization computation}
+\subsection{- - Matrix factorization computation}
 One of the aims of this work is to compare the efficacy of three methods of dimensionality reduction -- NMF, ICA and PCA -- as explained in the introduction.     
@@
 
-\subsection{Metagene extraction by cluster coherence under bootstrap resampling}
+\subsection{+ Metagene extraction by cluster coherence under bootstrap resampling}
 \label{sec-metagene-selection}
@@
 
-\subsection{Determining metagene similarity by Jaccard index}
-
-It is pertinent to ask ``how similar are the 11 metagenes which emerge from the foregoing factorization and cluster analysis?''.
-We expect those coming from a single factorization method to be distinct by construction.
-But perhaps ICA and PCA identified similar components.
-Similarity with NMF derived components is possible but less likely due to the positivity constraint.
-A possible approach to measuring similarity might be to treat the components as vectors and determine the angle between them, via the scalar or ``dot'' product.
-But in such a high dimension space (19,566) any pair of vectors will be very close to $90^\circ$.
-
-An alternative is to consider the intersection of the gene sets which each metagene highlights.
-In the GEA which follows, candidate genes will be identified for each meatagene.
-A standard measure of set similarity is the Jaccard index, (or similarity), defined on a pair of sets as:
-
-\[
-   \mbox{Jaccard}(A, B) = \frac{A \cap B}{A \cup B}.
-\]
-The measure lies between 0 (no similarity) and 1 (identical), and is symmetric.  Jaccard similarity was thus calculated for all pairs of the 11 genes sets and visualised as a square heatmap.
-
-\subsection{Gene enrichment analysis (GEA)}
+\subsection{- - Gene enrichment analysis (GEA)}
 \label{sec-gea}
@@
 
-\subsection{Transfer of learned metagenes to a novel dataset}
+\subsection{- - Transfer of learned metagenes to a novel dataset}
 \label{sec-transfer-to-novel}
@@
 
-\subsection{Reconciling computed metasamples and metadata}
-\label{sec-reconciling-metasamples-metadata}
-
-To carry out patient level analysis -- survival analysis, heatmaps and genome feature correlation -- as described in the following sections, it is necessary to compile tables (Pandas DataFrames) which bring together per-patient metasamples relating to each of the selected metagenes, with associated metadata -- such as cellularity and survival information.
-The metagenes always derive from the TCGA dataset (because of its larger size as explained in section \ref{sec-metagene-selection}), but we wish to apply these metagenes as metasamples to the expression data and associated metadata of either the TCGA or AOCS datasets.
-The mathematics of transferring metagenes across datasets has been described above.
-Care is required to ensure that expression matrices are aligned correctly with metagenes with respect to their genes, and that metadata is aligned correctly with metasamples with respect to patient identifiers.
-Note that the same transferring approach is taken when metasamples are required for TCGA, even though we could in that case obtain the $H$ matrix directly from the factorization.
-
-\subsection{Survival analysis}
-\label{sec-survival-analysis}
-
-Survival analysis was performed to investigate whether the metasamples derived from the selected metagenes correlate with patient survival.
-Overall  survival (OS) data is available for both the TCGA and AOCS datsets.
-Additionally, progression free survival (PFS) data is available for AOCS.
-As described above,  metagenes were obtained by factorization on the TCGA dataset only.
-Application to TCGA (for OS) thus represents an in-sample test, while application to AOCS (for OS and PFS) is a more exacting out-of-sample test.
-There are thus three analyses to consider:
-1) TCGA$\rightarrow$TCGA (OS) meaning TCGA derived metagenes are tested for OS on the TCGA dataset, 2) TCGA$\rightarrow$AOCS (OS) meaning TCGA derived metagenes are tested for OS on the AOCS dataset, and 3) TCGA$\rightarrow$AOCS (PFS)...
-as 2) but for PFS.
-
-Analysis was performed using the Python {\tt Lifelines} package \cite{Davidson-Pilon2020}.  
-%Metasamples ($H$ matrices) were calculated from the metagenes and expression matrix as explained in section \ref{sec-transfer-to-novel} below.  
-Metasample values were binarized to 0, 1 by thresholding at the median value.
-Kaplan-Meier plots were made in respect of derived metasample for each of the three analyses -- making $11 \times 3$ plots each with two survival curves with 95\% confidence intervals.
-A hazard ratio (HR)
-was calculated for each analysis by fitting Cox's proportional hazards model with p-value relating to the hypothesis that HR is significantly different to 1.0.
-
-\subsection{Investigating correlation between metasamples and genomic features}
+\subsection{- - Investigating correlation between metasamples and genomic features}
 \label{sec-genomic-features}
@@
 
-\subsection{Investigating batch effects by analysis of a combined dataset}
+\subsection{- - Investigating batch effects by analysis of a combined dataset}
 
@@
 
-\subsection{Codebase and plotting conventions}
-
-The described methodology was implemented in Python.
-%except for the KEGG GEA which was implemented in R.
-All code is available in a github repository: \url{https://github.com/ipoole/HgsocTromics}.
-Good software engineering practice has been followed, with an object-oriented design, frequent commits and unit testing.
-A base class is used to present a uniform interface for the three factorization methods.
-Intermediate results are cached to file with lazy computation for efficient working without imposing evaluation order dependency.
-Unit tests are based on tiny expression matrices, just 100 genes by 10 patients, thus the whole test suite of over 65 tests executes in around 45 seconds.
-The fully automated (except for rank selection) analysis pipeline takes approximately three hours from scratch, the majority of that time being taken for the repeated factorizations over the range of considered ranks.
-The total codebase is approximately 2,900 lines of Python.
-
-All plots are generated in vector graphics pdf format to ensure smooth scaling to any resolution.
-A consistent colour scheme of \textcolor{blue}{blue}, \textcolor{orange}{orange} and \textcolor{olive}{green} is used for plots relating to NMF, ICA and PCA respectively.
-In figures metagenes / metasamples are consistently referred to by, for example, ``NMF-2-of-3" -- meaning the 2nd component of the rank $k=3$ NMF factorization.
-In text this is shortened to ``NMF-2'' where the rank is understood.
-
 %%%%%%%%%%% RESULTS %%%%%%%%%%%%%
@@
 
-\section{Results}
+\section{- - Results}
 
 % \subsection{Investigation of optimal parameters}
-\subsection{Consideration of sampling error is crucial to finding robust metagene signals}
+\subsection{+ Consideration of sampling error is crucial to finding robust metagene signals}
 
@@
 
-\subsection{Jaccard similarity shows metagenes identify intersecting gene sets in some cases}
-
-The Jaccard heatmap of figure \ref{fig-component-jaccard-similarities} shows that some pairs of metagenes identify overlapping sets of genes, in particular the pairs (ICA-1, PCA-2: J=0.61) and (ICA-5, PCA-3: J=0.49).
-As expected however, there is very little similarity between components \emph{within} a factorization method.
-
-\begin{figure}[htb!]
-\begin{center}
-\includegraphics[scale=0.4]{\plotspath{TCGA_OV_VST/GeneEnrichment/component_jaccard_heatmap.pdf}}
-\end{center}
-\caption{Heatmap of Jaccard similarities between the candidate genes identified by the 11 components.}
-\label{fig-component-jaccard-similarities}
-\end{figure}
-
-\FloatBarrier
-
-\subsection{Metagenes highlight genes which are enriched for biological processes}
+\subsection{- - Metagenes highlight genes which are enriched for biological processes}
 
@@
 
-\subsection{Association of unsupervised gene expression patterns with patient survival is inconclusive}
-
-Kaplan-Meier overall survival (OS) plots stratified by each metasample component are shown for the for the AOCS dataset in figure \ref{fig-kaplan-meier-os-TCGA-AOCS} and for the TCGA dataset in figure \ref{fig-kaplan-meier-os-TCGA-TCGA} (appendix).
-Plots for progression free survival (PFS) on AOCS are also in the appendix, figure \ref{fig-kaplan-meier-pfs-TCGA-AOCS}.
-All plots show 95\% confidence intervals and hazard ratio (HR) with associated p-value.
-
-All of these results are summarised in figure \ref{fig-summary-survival-plots}, which brings together the three sets of results -- TCGA (OS), AOCS (OS) and AOCS (PFS) --  with respect to the 11 metasample components.
-$\log_2{\mbox{HR}}$ is used in order that the \emph{sense} of the survival impact can be readily appreciated.
-If a metasample component has a robust correlation with survival, then we expect the p-value for all three sets of results to show significance \emph{and} for the effects to have the same sense -- be in the same direction.
-None of the 11 components pass this test.
-
-The component with the largest observed effect on survival is PCA-1, showing a reasonably consistent hazard ratio of around 1.3 ($2^{0.4}$) across the three experiments.
-This has $p < 0.05$ in the larger TCGA dataset, but not in the smaller AOCS dataset.
-
-\begin{figure}[htb!]
-\begin{center}
-\includegraphics[scale=0.4]{\plotspath{SurvivalAnalysis/multiple_kaplan_meier_os_TCGA_AOCS.pdf}} 
-\caption{Kaplan-Meier plots for TCGA $\rightarrow$ AOCS for overall survival (OS).}
-\label{fig-kaplan-meier-os-TCGA-AOCS}
-\end{center}
-\end{figure}
-
-\begin{figure}[htb!]
-\begin{center}
-\includegraphics[scale=0.4]{\plotspath{/SurvivalAnalysis/survival_summary_barplots_NMF.pdf}} \\
-\includegraphics[scale=0.4]{\plotspath{/SurvivalAnalysis/survival_summary_barplots_ICA.pdf}} \\
-\includegraphics[scale=0.5]{\plotspath{/SurvivalAnalysis/survival_summary_barplots_PCA.pdf}}
-\caption{Visual summary of survival analysis as applied to TCGA(OS), AOCS(OS) and AOCS(PFS).  Plots are divided by factorization method.  Bar heights show $\log_2(\mbox{HR})$ with p-value also shown.}
-\label{fig-summary-survival-plots}
-\end{center}
-\end{figure}
-
-\subsection{Metasamples correlate with some genomic features}
+\subsection{- - Metasamples correlate with some genomic features}
 
@@
 
-\subsection{Integrative analysis of results}
+\subsection{+ Integrative analysis of results}
 \label{sec-integrative-analysis}
@@
 
-\subsection{Matrix factorization discovers and effectively removes batch effects}
+\subsection{- - Matrix factorization discovers and effectively removes batch effects}
 
@@
 
-\section{Discussion}
+\section{- - Discussion}
 
@@
 
-\subsubsection*{What underlying biological processes are at play in HGSOC as seen through the patterns of gene expression? }
-
-Biological processes identified by GEA have been set out in section \ref{sec-gea}.
-Processes relating to the ECM and extracellular organisation may reflect cell type heterogeneity -- perhaps within the tumour its self, or due to sampling of adjacent normal tissue.
-Processes of negative regulation of angiogenesis could reflect tumour growth and therefore be of prognostic value.
-
-These observations are based on GEA on the GO only.
-Possibly more insights could have been gained by GEA against KEGG\footnote{\href{https://www.genome.jp/kegg/pathway.html}{Kyoto Encyclopedia of Genes and Genomes}} Pathways.
-(This was begun but had to be dropped due to shortage of time).
-
-\subsubsection*{How do the patterns of gene expression uncovered relate to genome level features, such as those marking genome instability?}
-
-The integrative analysis set out in section \ref{sec-integrative-analysis} suggests that  Cellularity is correlated to the processes of the the ECM, reflecting heterogeneity of tissue type.
-Of course, Cellularity is not really a biological characteristic -- it is a technical artefact of the tumour biopsy process, and so we might conclude that signatures relating to ECM might also be considered as technical artefacts.
-We found that whole genome doubling (WGD) and Cellularity correlate together with three components even though they are not themselves directly correlated -- something which warrants further investigation.
-
-Weak evidence of a link between homologous repair deficiency (HRD) and RNA-binding was found.
-Being the site of protein synthesis, ribosomes will be associated with most of what goes on in a the cell.
-One paper suggests a link : ``a comprehensive review of the literature reveals a role for conventional DNA repair proteins in ribosome biogenesis, and conversely, ribosome biogenesis proteins in DNA repair" \cite{Ogawa2018}.
-Seek and ye shall find!
-
-A note of caution: these results are based on testing $6 \times 11 = 66$ hypotheses at a significance level of $< 0.01$, so there will likely be at least one false discovery.
-A significance level of 0.001 would have been more appropriate.
-
-Expression signature correlations with mutational load and  CNV load have not been detected.
-
-\subsubsection*{Do MF methods yield factors which are predictive of patient survival?}
-
-We found no convincing evidence of this, for an individual factor (metasample).
-Some components were predictive in one or other of the two datasets, but not both, which is unconvincing.
-The factor for which GEA indicated regulation of angiogenesis (NMF-1), and thence might have prognostic value, performed poorly in survival analysis.
-So overall a negative result.
-
-Is there a signal for survival to be found in these data?
-We have noted that according to \cite{Zhu2020}, the single gene CD38 is highly predictive of survival in EOC.
-Does this hold for the TCGA and AOCS datasets we are working with?
-This is easy to check, running survival analysis stratified by a hand-crafted metagene, setting the CD38 element to 1, all others to 0.
-This was applied as in section \ref{sec-survival-analysis} to TCGA (OS),  AOCS (OS) and AOCS (PFS) -- see figure \ref{fig-cd38 survival}.
-The effects are in a consistent direction across the three tests, with high CD38 expression predicting favourable survival.
-This is highly significant in the TCGA dataset and $p < 0.05 $ in the smaller AOCS.
-So there is a signal, but it did not emerge unbidden from matrix factorization.
-(CD38 is in fact highlighted by component ICA-5, see figure \ref{fig-ICA-5-5-go-lineage}, but must be diluted by other genes).
-
-\begin{figure}
-\begin{center}
-\includegraphics[scale=0.5,draft=false]{\plotspath{SurvivalAnalysis/gene_specific_CD38_survival.pdf}}
-\end{center}
-\caption{Survival analysis plots stratified by above/below median expression of CD38 only.}
-\label{fig-cd38 survival}
-\end{figure}
-
-Several other specific genes have been noted in section \ref{sec-review-of-gene-expression}; it would be interesting to test these also.
-
-\subsubsection*{How should MF methods be applied to achieve robust results on datasets of limited size?}
+\subsubsection*{- - How should MF methods be applied to achieve robust results on datasets of limited size?}
 
@@
 
-\subsubsection*{Does MF detect and remove technical batch effects?}
+\subsubsection*{+ Does MF detect and remove technical batch effects?}
 
@@
 
-\subsubsection*{Which MF method is best suited to transcripomics analysis?}
-
-The original technical focus of this dissertation was ICA and NMF since these are know to be able to disentangle variation in data with complex structure, not multivariate normally distributed.
-PCA was added as a baseline dimensionality reduction method, just for comparison.
-What would we have been missed if only PCA had been used?
-Reviewing the integrative analysis of results in table \ref{tab-integrative-analysis} we see that all three methods picked up on WGD and cellularity.
-PCA included the two components having the largest observed effect on overall survival.
-NMF and ICA both showed sensitivity to SV load.
-NMF further added sensitivity to HRD and identified genes relating to angiogenesis with possible pathological significance.
-
-So it's a mixed picture.
-Of course, had more components been selected from each method then likely more would have been found, but at the expense of robustness and thence false discovery, as we have discussed.
-It seems we must concur with the conclusion of Way \etal \cite{WayGregory2019} that no one method excels, each brings individual value.
-At least these three methods are simple and linear (unlike the VAE for example), and so can be used together easily.
-
-\subsection{Further work}
-
-This work has thrown up several possible lines of further study:
-\begin{enumerate}
-\item
-{\bf Algorithm hyper-parameters}.
-NMF and ICA algorithms have a number of hyper-parameters, only a few of which were explored in this work.
-For example, NMF has parameters to encourage sparsity through L1 regularization.
-This should result in many (most?) metagene elements reducing to zero, and might offer a more robust means of selecting candidate genes to feed into GEA -- replacing the current arbitrary 3 SD from the mean rule.
-\item
-{\bf Selecting components from several factorization ranks}.
-In the current work, a single rank was selected for each method -- 3, 5 an 3 for NMF, ICA and PCA respectively.
-Unlike PCA, components extracted by NMF and ICA  do not `nest' with rank; that is adding rank in general yields a new set of components.
-Thus one might, for NMF and ICA, perform factorizations at $k=2,3,4,5$, say, yielding $2 \times 15 = 30$ potential components in total.
-Jaccard similarity could be used to identify that subset of components which had the least overlap in detected genes.
-In this way, a larger number of components could be obtained without use of high factorization ranks which we have seen to be unstable.
-In fact this is similar to the approach of \cite{WayGregory2019} mentioned in section \ref{sec-comparative-dimensionality-studies}.
-It will be important to take a principled approach to the setting of the FDR theshold when applying GEA, to account for the increased number of hypothesis being tested.
-\item
-{\bf Cross-dataset factorization stability}.
-We went to considerable effort -- through bootstrap sampling and cluster analysis -- to select factorization ranks which we hoped would be robust and generalise well to other datasets.
-A way to confirm this robustness would be to perform the factorization / clustering pipeline on \emph{both} datasets separately, yielding two sets of metagenes.
-In a perfect world, these metagenes would pair up identically (in-so-far as the two datasets had identical technical characteristics, drawn from identical population of patients).
-The Jaccard similarity heatmap could be used to verify this, and reject components which showed low cross-dataset similarity.
-This proposal chimes to some extent with the use of Reciprocally Best Hit graphs in \cite{Cantini2019} reviewed in section \ref{sec-comparative-dimensionality-studies}.
-\item
-{\bf Systematic comparison with published gene expression patterns}.
-In analysing the meaning of each metagene above, some tentative links were made with the research literature, by informal searching.
-A more systematic and ideally automated approach is required.
-The \href{https://www.ncbi.nlm.nih.gov/geoprofiles}{Geo Profiles database} at NCBI, or the \href{https://www.ebi.ac.uk/gxa/home}{Expression Atlas} at EMBL-BI might be possible starting points.
-
-\end{enumerate}
-
 \clearpage
 
-\section{Conclusions}
+\section{- - Conclusions}
 
@@
 
-\clearpage
-\subsection*{Acknowledgements}
-I am most grateful to Dr.
-Ailith Ewing for proposing this project and making available the datasets.
-Ailith provided excellent, insightful supervision throughout and gave valuable feedback on an early draft of this dissertation.
-I'd like to thank my employer, Canon Medical Research Europe Ltd., for allowing me the flexibility and study leave to undertake this MSc.
-Thank you to my family for excusing me from cooking and washing up duties during the more stressful phases of MSc study and dissertation writing!
-
-\small
 \bibliography{OvarianCancer}
@@
 
-\newpage
-\section{Appendices}
-\subsection{Metagene clustering -- additional figures}
-\def\S{0.3}
-\begin{figure}[htb!]
-\begin{center}
-\includegraphics[scale=\S]{\plotspath{AOCS_Protein/FactorClustering/multiple_single_factors_scatter_AOCS_NMF_2_3_5_10_fixed.pdf}} \\
-\includegraphics[scale=\S]{\plotspath{AOCS_Protein/FactorClustering/multiple_single_factors_scatter_AOCS_NMF_2_3_5_10_bootstrap.pdf}}\\
-\caption{Clustering of metagenes from NMF facorizations on the AOCS dataset, comparing fixed and bootstrap sampling}
-\label{fig-AOCS-nmf-fixed-vs-bootstrap}
-\end{center}
-\end{figure}
-
-\begin{figure}[htb!]
-\begin{center}
-\includegraphics[scale=\S]{\plotspath{AOCS_Protein/FactorClustering/multiple_single_factors_scatter_AOCS_PCA_2_3_5_10_fixed.pdf}} \\
-\includegraphics[scale=\S]{\plotspath{AOCS_Protein/FactorClustering/multiple_single_factors_scatter_AOCS_PCA_2_3_5_10_bootstrap.pdf}}\\
-\caption{Clustering of metagenes from PCA facorizations on the AOCS dataset, comparing fixed and bootstrap sampling}
-\label{fig-AOCS-pca-fixed-vs-bootstrap}
-\end{center}
-\end{figure}
-
-\subsection{GEA -- additional plot}
-
-\begin{sidewaysfigure}[htb!]
-\begin{center}
-\includegraphics[width=\textwidth]{\plotspath{TCGA_OV_VST/GeneEnrichment/go_lineage_4_of_ICA_5.png}} \\
-\begin{spacing}{0.6}
-{\small Genes: \tiny \input{/home/ipoole/Documents/gitrepos/HgsocTromics/Plots/TCGA_OV_VST/GeneEnrichment/go_gene_set_4_of_ICA_5.txt}}
-\end{spacing}
-\vspace{0.5cm}
-\caption{Lineage maps of enriched GO terms for component ICA-4. For this component it was necessary to additionally filter the enriched terms -- see text.}
-\label{fig-ICA-4-go-lineage}
-\end{center}
-\end{sidewaysfigure}
-
-\FloatBarrier
-\subsection{Survival analysis -- additional figures}
-
-\begin{figure}[htb!]
-\begin{center}
-\includegraphics[scale=0.4]{\plotspath{SurvivalAnalysis/multiple_kaplan_meier_os_TCGA_TCGA.pdf}} 
-\caption{Kaplan-Meier plots for each metasample component, stratified at the median value, for TCGA $\rightarrow$ TCGA for overall survival (OS) case.
-Hazard ratio and p-value is shown for each case.
-The final plot (bottom right) is unstratified overall survival}
-\label{fig-kaplan-meier-os-TCGA-TCGA}
-\end{center}
-\end{figure}
-
-\begin{figure}[htb!]
-\begin{center}
-\includegraphics[scale=0.4]{\plotspath{SurvivalAnalysis/multiple_kaplan_meier_pfs_TCGA_AOCS.pdf}} 
-\caption{Kaplan-Meier plots for TCGA $\rightarrow$ TCGA for progression free survival (PFS).  }
-\label{fig-kaplan-meier-pfs-TCGA-AOCS}
-\end{center}
-\end{figure}
-
-\begin{figure}[htb!]
-\begin{center}
-\includegraphics[scale=0.5]{\plotspath{HeatMaps/clustered_heatmap_TCGA_AOCS.pdf}} 
-\caption{Heatmap of the metasamples matrix for the AOCS dataset, computed from metagenes factorized from the TCGA dataset.   Several patient metadata variables are shown in columns to the left.}
-\label{fig-heatmap-TCGA-AOCS}
-\end{center}
-\end{figure}
-
-\FloatBarrier
-
-\subsection{Batch effect investigation -- Additional figures}
-\begin{figure}[htb!]
-\begin{center}
-\includegraphics[scale=\S]{\plotspath{BOTH_AOCS_TCGA/FactorClustering/multiple_single_factors_scatter_BOTH_NMF_2_3_5_10_bootstrap.pdf}} \\
-\includegraphics[scale=\S]{\plotspath{BOTH_AOCS_TCGA/FactorClustering/multiple_single_factors_scatter_BOTH_ICA_2_3_5_10_bootstrap.pdf}} \\
-\includegraphics[scale=\S]{\plotspath{BOTH_AOCS_TCGA/FactorClustering/multiple_single_factors_scatter_BOTH_PCA_2_3_5_10_bootstrap.pdf}} \\
-\includegraphics[scale=0.55]{\plotspath{/BOTH_AOCS_TCGA/FactorClustering/silhouette_plots_BOTH_2_3_4_5_6_7_8_9_10_bootstrap.pdf}} 
-\caption{Metagene clustering for all three methods applied to the combined N=80+374 AOCS + TCGA dataset with bootstrap sampling, over a range of factorization ranks.
-The silhouette scores are also plotted (bottom).
-}
-\label{fig-BOTH-all-bootstrap}
-\end{center}
-\end{figure}
-
-\begin{figure}[htb!]
-\begin{center}
-\includegraphics[scale=0.4]{\plotspath{BoxAndScatters/genomic_feature_scatters_BOTH_AOCS.pdf}} 
-\caption{Grid of scatter plots to visualise the correlation between metasamples derived from the metagenes of the combined AOCS + TCGA datasets, and genomic features of the AOCS.}
-\label{fig-genomic-scatters-BOTH-AOCS}
-\end{center}
-\end{figure}
-
-\FloatBarrier
-
-\subsection{Access to gene enrichment analysis table}
-\label{tab-go-enrichment-results}
-
-The table of GEA results is too large to conveniently include here.
-If required it can be obtained from my github repository at \url{https://github.com/ipoole/HgsocTromics/blob/master/Results/combined_gea_fdr.tsv}.
-
 \end{document}

### 2021-01-24 Methods_Paper.tex
Copied images, inputs and .bib to directory so self-contained * All required images copied from HgsocTromics/Plots/... to ./images * All required .tex inputs copied to ./inputs * Methods_Paper.tex edited to use local copies above * OvarianCancer.bib copied directly (was a symbolic link)

@@
 
-\documentclass[tikz, 11pt,a4paper,oneside,fleqn, draft]{article}
+\documentclass[tikz, 11pt,a4paper,oneside,fleqn]{article}
 
@@
 
-\graphicspath{{/home/ipoole/Documents/gitrepos/MScDissertation/images/}}  % trailin '/' is required!
-\newcommand{\plotspath}[1]{{/home/ipoole/Documents/gitrepos/HgsocTromics/Plots/#1}}
+\graphicspath{{/home/ipoole/Documents/gitrepos/MScDissertation/images/}}  % trailing '/' is required!
 
@@
 \begin{center}
-\includegraphics[scale=\S]{\plotspath{AOCS_Protein/FactorClustering/multiple_single_factors_scatter_AOCS_ICA_2_3_5_10_fixed.pdf}} \\
-\includegraphics[scale=\S]{\plotspath{AOCS_Protein/FactorClustering/multiple_single_factors_scatter_AOCS_ICA_2_3_5_10_bootstrap.pdf}}\\
+\includegraphics[scale=\S]{multiple_single_factors_scatter_AOCS_ICA_2_3_5_10_fixed.pdf} \\
+\includegraphics[scale=\S]{multiple_single_factors_scatter_AOCS_ICA_2_3_5_10_bootstrap.pdf}\\
 \caption{Clustering of metagenes from ICA facorizations on the N=80 AOCS dataset, comparing \emph{fixed} sampling (top row) with \emph{bootstrap} sampling (bottom row) for a selection of factorization ranks.
@@
 \begin{center}
-\includegraphics[scale=\S]{\plotspath{TCGA_OV_VST/FactorClustering/multiple_single_factors_scatter_TCGA_ICA_2_3_5_10_fixed.pdf}} 
-\includegraphics[scale=\S]{\plotspath{TCGA_OV_VST/FactorClustering/multiple_single_factors_scatter_TCGA_ICA_2_3_5_10_bootstrap.pdf}}
+\includegraphics[scale=\S]{multiple_single_factors_scatter_TCGA_ICA_2_3_5_10_fixed.pdf} 
+\includegraphics[scale=\S]{multiple_single_factors_scatter_TCGA_ICA_2_3_5_10_bootstrap.pdf}
 \caption{Clustering of metagenes from ICA facorizations on the N=374 TCGA dataset, again comparing fixed and bootstrap sampling.}
@@
 \begin{center}
-\includegraphics[scale=\S]{\plotspath{TCGA_OV_VST/FactorClustering/multiple_single_factors_scatter_TCGA_NMF_2_3_5_10_bootstrap.pdf}} \\
-\includegraphics[scale=\S]{\plotspath{TCGA_OV_VST/FactorClustering/multiple_single_factors_scatter_TCGA_ICA_2_3_5_10_bootstrap.pdf}} \\
-\includegraphics[scale=\S]{\plotspath{TCGA_OV_VST/FactorClustering/multiple_single_factors_scatter_TCGA_PCA_2_3_5_10_bootstrap.pdf}} \\
-\includegraphics[scale=0.55]{\plotspath{/TCGA_OV_VST/FactorClustering/silhouette_plots_TCGA_2_3_4_5_6_7_8_9_10_bootstrap.pdf}} 
+\includegraphics[scale=\S]{multiple_single_factors_scatter_TCGA_NMF_2_3_5_10_bootstrap.pdf} \\
+\includegraphics[scale=\S]{multiple_single_factors_scatter_TCGA_ICA_2_3_5_10_bootstrap.pdf} \\
+\includegraphics[scale=\S]{multiple_single_factors_scatter_TCGA_PCA_2_3_5_10_bootstrap.pdf} \\
+\includegraphics[scale=0.55]{silhouette_plots_TCGA_2_3_4_5_6_7_8_9_10_bootstrap.pdf} 
 \caption{Metagene clustering for all three methods applied to the N=374 TCGA dataset with bootstrap sampling, over a range of factorization ranks.
@@
 \begin{center}
-\includegraphics[scale=0.08]{\plotspath{TCGA_OV_VST/GeneEnrichment/go_lineage_1_of_NMF_3.png}} \\
+\includegraphics[scale=0.08]{go_lineage_1_of_NMF_3.png} \\
 \begin{spacing}{0.6}
-{\small Genes: \tiny \input{/home/ipoole/Documents/gitrepos/HgsocTromics/Plots/TCGA_OV_VST/GeneEnrichment/go_gene_set_1_of_NMF_3.txt}}
+{\small Genes: \tiny \input{inputs/TCGA_go_gene_set_1_of_NMF_3.txt}}
 \end{spacing}
@@
 \begin{center}
-\includegraphics[scale=0.1]{\plotspath{TCGA_OV_VST/GeneEnrichment/go_lineage_2_of_NMF_3.png}} \\
+\includegraphics[scale=0.1]{go_lineage_2_of_NMF_3.png} \\
 \begin{spacing}{0.6}
-{\small Genes: \tiny \input{/home/ipoole/Documents/gitrepos/HgsocTromics/Plots/TCGA_OV_VST/GeneEnrichment/go_gene_set_2_of_NMF_3.txt}}
+{\small Genes: \tiny \input{inputs/TCGA_go_gene_set_2_of_NMF_3.txt}}
 \end{spacing}
@@
 \begin{center}
-\includegraphics[scale=0.09]{\plotspath{TCGA_OV_VST/GeneEnrichment/go_lineage_3_of_ICA_5.png}} \\
+\includegraphics[scale=0.09]{go_lineage_3_of_ICA_5.png} \\
 \begin{spacing}{0.6}
-{\small Genes: \tiny \input{/home/ipoole/Documents/gitrepos/HgsocTromics/Plots/TCGA_OV_VST/GeneEnrichment/go_gene_set_3_of_ICA_5.txt}}
+{\small Genes: \tiny \input{inputs/TCGA_go_gene_set_3_of_ICA_5.txt}}
 \end{spacing}
@@
 \begin{center}
-\includegraphics[scale=0.06]{\plotspath{TCGA_OV_VST/GeneEnrichment/go_lineage_5_of_ICA_5.png}} \\
+\includegraphics[scale=0.06]{go_lineage_5_of_ICA_5.png} \\
 \begin{spacing}{0.6}
-{\small Genes: \tiny \input{/home/ipoole/Documents/gitrepos/HgsocTromics/Plots/TCGA_OV_VST/GeneEnrichment/go_gene_set_5_of_ICA_5.txt}}
+{\small Genes: \tiny \input{inputs/TCGA_go_gene_set_5_of_ICA_5.txt}}
 \end{spacing}
@@
 \begin{center}
-\includegraphics[width=0.4 \textwidth]{\plotspath{TCGA_OV_VST/GeneEnrichment/go_lineage_1_of_PCA_3.png}} \\
+\includegraphics[width=0.4 \textwidth]{go_lineage_1_of_PCA_3.png} \\
 \begin{spacing}{0.6}
-{\small Genes: \tiny \input{/home/ipoole/Documents/gitrepos/HgsocTromics/Plots/TCGA_OV_VST/GeneEnrichment/go_gene_set_1_of_PCA_3.txt}}
+{\small Genes: \tiny \input{inputs/TCGA_go_gene_set_1_of_PCA_3.txt}}
 \end{spacing}
@@
 \begin{center}
-\includegraphics[width= 0.7\textwidth]{\plotspath{TCGA_OV_VST/GeneEnrichment/go_lineage_3_of_PCA_3.png}} \\
+\includegraphics[width= 0.7\textwidth]{go_lineage_3_of_PCA_3.png} \\
 \begin{spacing}{0.6}
-{\small Genes: \tiny \input{/home/ipoole/Documents/gitrepos/HgsocTromics/Plots/TCGA_OV_VST/GeneEnrichment/go_gene_set_3_of_PCA_3.txt}}
+{\small Genes: \tiny \input{inputs/TCGA_go_gene_set_3_of_PCA_3.txt}}
 \end{spacing}
@@
 \begin{center}
-\includegraphics[scale=0.45]{\plotspath{BoxAndScatters/genomic_feature_scatters_TCGA_AOCS.pdf}} 
+\includegraphics[scale=0.45]{genomic_feature_scatters_TCGA_AOCS.pdf}
 \caption{Grid of scatter plots to visualise the correlation between metasamples and genomic features.
@@
 \begin{center}
-\includegraphics[scale=0.4,draft=false]{\plotspath{HeatMaps/clustered_heatmap_BOTH_BOTH.pdf}} 
+\includegraphics[scale=0.4,draft=false]{clustered_heatmap_BOTH_BOTH.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics[scale=0.36]{\plotspath{BoxAndScatters/wgd_boxplot_NMF_BOTH_BOTH.pdf}}
-\includegraphics[scale=0.36]{\plotspath{BoxAndScatters/wgd_boxplot_ICA_BOTH_BOTH.pdf}}
-\includegraphics[scale=0.36]{\plotspath{BoxAndScatters/wgd_boxplot_PCA_BOTH_BOTH.pdf}}
-
+\includegraphics[scale=0.36]{wgd_boxplot_NMF_BOTH_BOTH.pdf}
+\includegraphics[scale=0.36]{wgd_boxplot_ICA_BOTH_BOTH.pdf}
+\includegraphics[scale=0.36]{wgd_boxplot_PCA_BOTH_BOTH.pdf}
 \end{center}
